### 2017-08-03 paper.tex
copy from isrr

\documentclass[usletter, 10pt, conference]{ieeeconf}      % for icra

\usepackage{graphicx}
\usepackage{booktabs}
%\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{subfig}
\usepackage[ruled,lined]{algorithm2e}
\usepackage{multirow}
%\usepackage[bookmarks=false]{hyperref}
%\usepackage{setspace}
\pdfminorversion=4

%\IEEEoverridecommandlockouts                              % This command is only
%\overrideIEEEmargins

\newcommand{\red}[1]{\textcolor{red}{#1}}
%\newcommand{\tylde}{\raisebox{-0.85ex}{\textasciitilde}} %does not work for vojta
\newcommand{\tylde}{$\sim$}

%\hypersetup{colorlinks=true,linkcolor=black,citecolor=black}

\SetKwInput{KwData}{Global params.}

\title{    Hey we need a title}

\author{ Vojt\v ech Von\' asek$^{1}$,
    Barbora Kozl\'\i kov\'a$^{2}$,
    Adam Jur\v{c}\'\i k$^{2}$,
    Martin Saska$^{1}$
\thanks{$^{1}$ Faculty of Electrical Engineering,  Czech Technical University in Prague, 
Technick\'a 2, 166 27 Prague, Czech Republic,
{\tt vonasek@labe.felk.cvut.cz}}
\thanks{$^{2}$        
Faculty of Informatics,   Masaryk University, 
Botanick\'a 68a, 602~00 Brno, Czech Republic,\\
The presented work has been supported by the Czech Science Foundation (GA{\v C}R) under research project No. 17-07690S.
Access to computing and storage facilities owned by parties and projects contributing to the National Grid Infrastructure MetaCentrum, provided under the programme "Projects of Large Infrastructure for Research, Development, and Innovations" (LM2010005), is greatly appreciated.}
}

%D. Bedn\'{a}\v{r}:
%Faculty of Science, Masaryk University, Kamenice 5, 625 00 Brno, Czech Republic\\

%\makeindex

\def\qrand{q_{rand}}
\def\qstart{q_{start}}
\def\qinit{\qstart}
\def\qgoal{q_{goal}}
\def\qnear{q_{near}}
\def\qnew{q_{new}}
\def\T{\mathcal{T}}

\def\C{\mathcal{C}}
\def\CF{\mathcal{C}_{free}}
\def\CFD{{\mathcal{C}^s_{free}}}
\def\dt{d_{tunnel}}
\def\da{d_{atom}}
\def\R{\mathbb{R}}
\def\QI{Q_{init}}
\def\RI{R_{init}}

\def\rv{R_{tunnel}}

\def\Imax{I_{max}} %max number of iterations of RRT-based planners

\def\dist{\mathrm{dist}}
\def\dists{\mathrm{dist}_{\mathrm{s}}}

\SetKw{return}{return}

\def\smin{s_{min}}
\def\smax{s_{max}}
\def\sdelta{s_{\Delta}}

\def\probe{r_{\mathrm{probe}}}
\def\Sprobe{S_{\mathrm{probe}}}

\def\gprobe{r_{\mathrm{out}}}
\def\Sgprobe{S_{\mathrm{out}}}

\def\CG{\mathcal{C}_{goal}}
\def\SB{\mathbf{S}_{blocking}}
\def\SS{\mathbf{S}}

%spacing for algorithm environment. 1.0 mean normal spacing
\def\gb{p_{tunnel}}

\def\L{L}
\def\LA{L_1}
\def\LB{L_2}

% ==================================================================================

\begin{document}

\maketitle
%\thispagestyle{empty}
%\pagestyle{empty}

%\frontmatter          % for the preliminaries

\begin{abstract}
%Understanding of interactions between enzymes and ligands is very important in enzymology, protein engineering or drug design. 
Molecular tunnels connect catalytic active sites with the solvent and they play an important role in catalytic and thermodynamic properties.
The geometrical and physico-chemical properties of tunnels need to be determined in order to understand the transport of a ligand from and into the protein active site. 
Current methods describing the ligand passing through the tunnel are based either solely on the geometry of the tunnel or on molecular dynamics (MD) simulations.
The tunnels are usually computed using Voronoi diagrams and therefore they can show possible movements only for a single atom.
Ligand movements inside protein can be computed using MD simulations, but they are too slow for a daily routine.
The tunnels computed for a single atom can however be used as a hint for ligand transport from or to the active site.
The task of ligand transportation through a tunnel considering its shape and also conformation changes can be formulated
 as a motion planning problem.
We propose to evaluate the tunnels based on ligand trajectories computed by motion planning, namely using
Rapidly Exploring Random Tree (RRT).
We present modifications to RRT in order to compute the trajectories around a given tunnel and also to consider the flexibility of the ligand.
Besides, methods for visualization of the trajectories and their properties are presented.
The goal of the paper is to show that motion planning can provide additional insight about tunnel properties, being
less computationally demanding than the MD simulations.
The proposed approach can be used to improve virtual screening results which are usually based only on determination of ligand binding in the active site but no information about ligand transport through the tunnel is provided.
\end{abstract}

\section{Introduction \& Motivation}

\begin{figure}[b]
\centering
{\footnotesize
\renewcommand{\arraystretch}{0.1}
\renewcommand{\tabcolsep}{0pt}
\begin{tabular}{ccccc}
\includegraphics[width=0.15\textwidth]{fig/motiv1} &
\includegraphics[width=0.17\textwidth]{fig/motiv2lab} &
\includegraphics[width=0.16\textwidth]{fig/motiv3} &
\hbox{
\vbox{
\hbox{\includegraphics[width=0.25\textwidth]{fig/ta-1} }
\hbox{\includegraphics[width=0.25\textwidth]{fig/ta-433}}
} 
}
\\
Protein 1CQW & Active site & Detected tunnels & Example of a  \\
             &            & (orange)          & trajectory
\end{tabular}
}
\caption{\label{fig::motiv}
    Tunnels in haloalkane dehalogenase with a possible trajectory of 1-chlorpropan ligand.
}
\end{figure}

Understanding of interactions between proteins and other small molecules is crucial in many research fields, including drug design and protein engineering. 
These interactions are highly influenced by the ability to transport the small molecule, called ligand, to the protein active site.
The active site can be considered as a deeply buried inner cavity with the ability to interact with the incoming ligand.
To get the ligand to the active site, there has to be a transportation path connecting the protein outer environment with the active site.
These paths, called tunnels, have to be wide enough and the physico-chemical properties of the surrounding amino acids have to be compatible with the ligand (Fig.~\ref{fig::motiv}).

Traditionally, tunnels are detected using Voronoi diagrams~\cite{yaffe2008,caver3} assuming
%The widely used tools for tunnel detection are based on Voronoi diagrams~\cite{yaffe2008,caver3}.  %or a 3D grid~\cite{sehnal2013mole,petrek2006caver}.
 a spherical ligand (probe) and they are represented as a sequence of spheres.
Biochemists decide if a tunnel can be used to transport a given ligand mainly based on the tunnel length and bottleneck (i.e., the radius of the smallest sphere that forms the tunnel).
This is used for example in protein engineering, where the task is to change selected properties of a protein, e.g., its stability under different outer conditions~\cite{Koudelakova2013} or its activity of the protein towards other molecules~\cite{Pavlova2009}.
%The design of suitable ligands causing the desired changes can be speeded up by detecting and analyzing the tunnels leading to the active sites.
%This can be achieved by detecting and studying so called tunnels in proteins which can serve as the transportation paths for the 
%ligand from the outside environment to the active site or vice versa.

Whereas the tunnel computation is already a well established research field, the simulation of ligand transportation through the detected tunnels is rather new.
Ligands are typically of non-spherical shape, therefore it is difficult to estimate their traversability through tunnels computed for a spherical probe.
The decision based only on spherical tunnels requires previous expertise in the domain and yet it may be imprecise.
To provide biochemists  a better insight into the behavior of non-spherical ligands, it is necessary to compute the trajectories of the ligand considering its shape and possibly also its conformation changes.

We propose to analyze the traversability of the ligands in the tunnels using motion planning.
The principle of Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} planner is used and further modified.
The main idea of the proposed planner is to generate the random samples around a virtual sphere moving through the tunnel, which
guides the search in the configuration space.
This allows us to focus the sampling of the configuration space around the tunnel.
The flexibility of the ligand is modeled using a predefined set of typical conformations.
To enable ligand movements in narrow tunnels, a scaled-down version of the ligand is used, similarly, e.g., to~\cite{cortes2005path}.
It helps to keep also potential solutions which do not fit to the geometric restrains but can be still feasible because of the physico-chemical properties.
%This strategy helps to overcome the limitations of the conformational discretation introduced by the rotamer approximation.

The proposed method can be seen as an extension for the traditionally used tunnel detection tools.
%Besides trajectory computation, visualization of the results in important. %we also propose methods for their visualization.
Our motivation is to help the biochemists to perform so called virtual screening, where they test the traversability of a ligand through a given tunnel.
In order to predict the success of ligand traversability, the virtual screening performs hundreds of thousands of tests and checks if the ligand passed through the tunnel.
The advantage of the proposed approach is that we can search the ligand path inside a specific tunnel which substantially decreases the computational time and resources required for the virtual screening.
%We further propose visualization techniques for presenting the results to biochemists.
%Therefore, methods for visualization of the results  are discussed.

\section{Related Work}

The analysis of protein structure aiming to reveal the tunnels has been supported by different computational software tools which take the geometry of the protein as an input and explore the inner void space (e.g., CAVER 1.0~\cite{petrek2006caver} or MOLE~\cite{Petrek20071357}). 
Early methods for tunnel detection utilized a discretized 3D grid, where each cell is considered as occupied or free depending
on the presence of atoms of the protein.
Tunnels can be then searched using standard graph-search methods, like Dijkstra's algorithm.
Besides, the grid can be used to identify other relevant properties like 
pockets, cavities, or channels~\cite{sehnal2013mole,petrek2006caver}.
%One of the first grid-based approaches to the detection of tunnels in protein is the CAVER 1.0 algorithm by Pet\v{r}ek et al.~\cite{citeulike:6257975}.
The obvious disadvantage of the grid-based methods is the high memory demand and their dependency on the grid resolution.
%Due to the high memory consumption, these methods are not suitable for tunnel detection in dynamic proteins and therefore they
%are used primarily for analysis of static molecules or individual snapshots of molecular dynamics.

Currently the most widely used approach to tunnel detection is based on ordinary Voronoi diagrams (VD) or Weighted Voronoi Diagrams (WVD). %~\cite{caver3,yaffe2008}.
The ordinary VD is computed on points representing centers of all atoms, without considering the radii of atoms.
%This may lead to detection of tunnels with incorrect bottlenecks. % i.e., incorrect radius of a smallest probe that can traverse the tunnels.
To consider atoms with different radii, the weights of individual points are determined by the van der Walls radii of atoms in WVD.
An alternative solution is to compute a non-weighted VD on an extended point set, where 
each atom is approximated by several spheres with a small radius~\cite{yaffe2008,caver3}.
VD-based methods are memory less demanding and also faster than the grid-based methods.
%The extension of VD-based methods to dynamic molecules requires to construct VD in the frames being analyzed and 
%finding correspondences between them.
%The existing approaches often use hierarchical clustering to match Voronoi vertices and edges from different frames, which is computationally demanding~\cite{lindow2012dynamic,caverDetails}.

The tunnels detected by the above mentioned approaches are evaluated using basic characteristics like bottleneck, length, curvature, and a list of surrounding
residues, which are later used to estimate the interaction possibilities.
The biggest disadvantage of both grid-based and VD-based methods is that the shape of the ligand is not taken into account during the tunnel detection and it is therefore not easy to estimate if (and how) a ligand might traverse the tunnels.
%The protein tunnels computed for spherical probes contain valuable information about the protein void space.
%Based on the biochemical properties, the chemists can decide if a given tunnel is relevant and can be possibly used by a ligand to exit the protein.
%The biochemically relevant tunnels can be then analyzed separately.

In order to determine if a ligand can pass the tunnel, it is necessary to compute a trajectory considering the shape of the ligand.
This can be formulated as a motion planning problem in a high-dimensional configuration space.
The configuration space $\C$ is formed by all possible configurations of the ligand in the tunnel, i.e., considering its rotation, translation, and possibly also other degrees of freedom responsible for the conformation changes.
The dimension of the configuration space is given by the degrees of freedom (DOF) of the ligand, i.e., 6D for a rigid ligand and 6D + $n$ for a flexible
ligand with $n$ DOFs.
Sampling-based motion planning methods can be used to search this high-dimensional configuration space~\cite{Lav06}.
The idea of sampling-based motion planning is to randomly sample $\C$ and classify the samples as free or non-free using collision detection.
The free samples are stored in a roadmap (a graph structure), in which a path can be searched using standard graph-search methods.

Sampling-based motion planning methods are suitable for computing the trajectories of the ligand as they 
can cope with many-DOF robots (objects) of arbitrary shapes.
The flexibility of ligands (or even proteins) can be modeled using a multi-link kinematic chain, where torsional angles
can change~\cite{songPFpath}.
This is useful, e.g., in the protein folding studies~\cite{al2012motion,gipson2012computational,amato2002using,raveh2009rapid} or analysis of loop motions~\cite{cortes2004geometric}.
%This is useful, e.g., in the protein folding studies~\cite{al2012motion,gipson2012computational,amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro} or analysis of loop motions~\cite{cortes2004geometric}.
%Sampling-based methods have also been used for 
% tunnel detection~\cite{vonasek2016application,vonasek2017tunnel} and
% exit pathway computation~\cite{cortes2010simulating,guieysse2008structure}.
%none of the existing work focuses on the trajectory generation in tunnels.
%However, not all conformations are feasible, which has to be checked using potential energy, which is time consuming.

Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} is a single-query sampling-based motion planning method that 
incrementally builds a configuration tree $\T$ rooted at the initial configuration $\qinit$.
In each iteration of RRT, a random configuration $\qrand \in \C$ is generated and its nearest node $\qnear \in \T$ in the tree is found.
A new configuration $\qnew$ is constructed on the line connecting $\qnear$ and $\qrand$ in the distance $\varepsilon$ from $\qnear$.
If $\qnew$ is collision-free, it is added to the tree.
The algorithm terminates if the tree approaches the goal configuration close enough.
%RRT is a suitable candidate for computing the trajectories of flexible ligands due to its ability to cope with many-DOF systems.

%rrt for molecular: \cite{al2012motion}
%survey \cite{gipson2012computational}
%In our previous works, we proposed to employ RRT also for the purp
%Solutions for tunnel detection using sampling-based methods however have not been discussed yet.
%Sampling-based planners have been used in various applications in robotics~\cite{elbanhawi2014sampling} and also  %,latombe1999motion} and also
% to study proteins, e.g. in  
%loop motions~\cite{cortes2004geometric},
%protein folding~\cite{amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},
%protein folding~\cite{raveh2009rapid,novinskaya2015improving},
%or protein folding combined with ligand diffusion~\cite{cortes2010simulating}.

%Two main issues need to be addressed: the increased dimension of the configuration space and the narrow passage problem.
%A well known issue of sampling-based planners is the narrow passage problem~\cite{hannaWIS}.

Motion planning for flexible ligands in protein tunnels brings two main issues: 
the ligand flexibility increases the dimension of the configuration space, and the necessity to plan in protein tunnels
leads to the narrow passage problem.
A narrow passage is a region in the configuration space whose removal changes the connectivity of the free space~\cite{hannaWIS}.
Narrow passages have smaller volume than other regions and it is therefore difficult to sample them dense enough using uniform distribution, that is used in the basic sampling-based planners.
The presence of narrow passages in the configuration space, especially if they contain part of the solution, requires many iterations
in order to put enough samples there and it consequently increases the planning time.
Typical tunnels have bottlenecks smaller than $1.0$~\AA, so they can already be considered as narrow passages for ligands with more than two atoms.
%The protein tunnels have the properties of the narrow passages, as they are very narrow
To cope with the narrow passages, they have to be sampled more dense.
For the family of RRT planners, it is useful to change the distribution of random samples according to the growth of the tree~\cite{vonasek2009rrt,vonasekphd,denny2016dynamic}.
%about their 3D position, which is represented by the tunnel centerline.
%The centerline can be used to guide the growth of the RRT tree through the configuration space~\cite{vonasek2009rrt,denny2016dynamic}.

%The additional degrees of freedom required to model the flexibility increase the dimension of the configuration space.
The kinematic chain representation used to model the ligand flexibility can generate all possible conformations but it also increases the dimension of the configuration space.
Not all of conformations are however feasible and it is therefore necessary to verify the feasibility of a given conformation based on energy, which is time consuming.
%RRT-based planners are sensitive to the employed metric, that should consider also the flexibility-related DOFs.
An alternative solution is to employ a library of known conformations.
The conformation changes of protein amino acids and ligand are represented by so called rotamers, stored in different rotamer libraries (e.g., Dunbrack library~\cite{dunbrack}).
%Rotamers differ in the angles between their atoms.
Possible rotamer conformations correspond to energetically and geometrically favorable positions.
%This solution is common also in different types of calculations, e.g.~\cite{kellogg}.

A RRT-based method for computing exit pathways for a small flexible molecule was presented in~\cite{cortes2010simulating}.
To cope with many-DOF ligands, the RRT-ML variant~\cite{cortes2007mlrrt} was employed.
RRT-ML expands the tree primarily using those DOF that are essential for achieving motion of the ligand (i.e., rotation
and translation) and it employs the other DOFs (i.e., that are responsible for conformation changes) if they hinder the growth of the tree.
The approach~\cite{cortes2010simulating} may however suffer from the narrow passage problem, as it aims to find some exit pathway for a given ligand, which requires to search the whole configuration space of the ligand/protein complex.
Although there are usually multiple pathways from a given active site, not all of them are traversable by a given ligand.
Many iterations are needed to find a solution, which increases the computational time.

In this paper, we propose to analyze the traversability separately for each tunnel.
This can bring several advantages for the biochemists.
The tunnels in a protein being studied can be computed using standard tunnel detection tools for spherical probes and assessed
according to their length, curvature, bottleneck, and biochemical properties (e.g., partial charge).
These characteristics are already used in many research studies.
Selected tunnels can further be analyzed with a given ligand to determine the traversability.
According to the found trajectories, researches may decided of the given ligand can reach the active site, which helps to 
organize subsequent MD simulations or other experiments.

%It is not necessary to first run time consuming MD simulations with both protein and ligand being studied.
%but it is suitable to detected tunnels in the protein only.

The proposed work employs the RRT planning principle with three extensions: 
a) the configuration space is not sampled uniformly, but uses a guided sampling along the tunnel centerline,
b) the radii of the ligand atoms can be decreased, and 
c) ligand flexibility is modeled using a library of predefined conformations.
The first two extensions are designed to cope with the narrow passage problem.
By reducing the scale of the atomic radii, the ligand can move more without collisions, which widens the narrow passages and increases
the probability of placing samples into them.
The scaling-down technique have been used e.g. for motion planning of deformable objects~\cite{gayle2005path,alterovitz2008motion}, and
they are also used in the most related MoMa-LigPath tool~\cite{cortes2005path}.
By using the predefined set of conformations, the number of dimensions of the configuration space does not depend on the DOF of the ligand.
Library of conformations are used also in different types of calculations, e.g.~\cite{kellogg}. 
%Using a library of predefined conformations has been used also in other tasks, e.g.~\cite{kellogg}.

%is used also in other related tools like like MoMa-LigPath~\cite{cortes2005path}
%Such techniques have been used for motion planning of deformable objects, 
%e.g., in~\cite{frank2008efficient,bayazit2001ligand,alterovitz2008motion,lamiraux2001flexible,kavraki1998towards,gayle2005path}, 
%e.g., in~\cite{frank2008efficient,bayazit2001ligand,alterovitz2008motion,lamiraux2001flexible,gayle2005path}, 
%and for motion planning of rigid objects among 
%flexible obstacles~\cite{rodriguez2006planning,frank2008efficient,phillips2014representation}.

%and the third modification
%is introduced in order to decrease the dimension of the configuration space.

%Ligands are small molecules interacting with the protein. 
%Due to the interaction, the relative positions of the ligand's atoms can change.
% in cortes2005path:
%In this paper two kinds of large-amplitude motion are treated: protein loop conformational changes (involving pro-
%tein backbone flexibility) and ligand trajectories to deep active sites in proteins (involving ligand and protein side-chain flex-
%ibility). First studies performed using our two-stage approach (geometric search followed by energy refinements) show that,
%    compared to classical molecular modeling methods, quite   similar results can be obtained with a performance gain of
% several orders of magnitude. Furthermore, our results also  indicate that the geometric stage can provide highly valuable information to biologists.
%This technique is similar to generation of samples near the medial exist of the environment
%medial axis~\cite{wilmarthMAPRM,foskey01hybrid,guibas1999probabilistic,hoff2000interactive,yang2004adapting,amatoOBRRT}.
%or by guiding the growth of the tree around a general path in the workspace~\cite{vonasek2009rrt,denny2014marrt}.

\section{Traversability of Tunnels}

\subsection{Preliminaries}

Proteins and ligands are represented by the hard sphere model, where the radius of each sphere (atom) is given by its van der Waals radius.
The flexibility of a ligand is modeled using the set $\L$ of its conformations.
The conformations $L$ are used from a library (e.g.,~\cite{dunbrack}), or they can be prepared considering the potential energy.

%or downloaded from a library~\cite{dunbrack}.
%We assume that a conformation $l \in \L$ can be switched to all other conformations in $\L$.

%Another approach is to model the ligand flexibility using a predefined set of feasible conformations $\L$
%To decrease the computational burden, a set of feasible conformations $\L$ is prepared, e.g. using Rosseta tool.
%The conformations can be prepared using Molecular Dynamics software such as Rosetta.
%For each conformation $c\in \L$ an energy potential is also calculated.
%Let $\L_i$ denote the set of conformations that 
%The transition matrix $M$ defines the possible changes of conformations, i.e., $M(i,j)=1$ if the conformation $c_i \in C$
%can be switched to the confomration $c_j \in C$.

A protein tunnel is described by a sequence of collision-free spheres 
$T=( (c_1, r_1),\ldots,(c_n,r_n) )$, where $n$ denotes the number of spheres,
$c_i \in \R^3$ is their 3D position and $r_i > 0$ denotes the maximum collision-free radius of a sphere centered at $c_i$. 
The tunnels can be found by tunnel detection tools like CAVER 3.0~\cite{caver3}.

%from wiki:
%Multiple static structures experimentally determined for the same protein in different conformations are often used to emulate receptor flexibility.[19] Alternatively rotamer libraries of amino acid side chains that surround the binding cavity may be searched to generate alternate but energetically reasonable protein conformations.[20][21]
% 19 = \cite{totrov2008flexible}
% 20 = \cite{hartmann2009docking}
% 21 = \cite{taylor2003fds}

%Conformations of the ligand may be generated in the absence of the receptor and subsequently docked[13] 
%or conformations may be generated on-the-fly in the presence of the receptor binding cavity,[14] 
%or with full rotational flexibility of every dihedral angle using fragment based docking.[15] 
%Force field energy evaluation are most often used to select energetically reasonable conformations,[16] but 
%knowledge-based methods have also been used.[17]
%13 = https://link.springer.com/article/10.1007/BF00123666

%The proteins are dense structures and the tunnels are typically narrow.
%Depending on the protein, the tunnel bottlenecks can be even less than $1$~\AA, which is too narrow for ligands with more than 2 atoms.
To enable motion of ligands in the narrow tunnels, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
A discrete set of scales is used, i.e., $s \in \{\smin, \smin+\sdelta, \ldots, \smax\}$, where 
$\smin$ is the minimal allowed scale, $\smax=1$ is the maximal allowed scale and $\sdelta$ is the minimal difference between two scales.

A configuration of the ligand $q=(x,y,z,r_x,r_y,r_z,l,s)$  is described
by the 3D position $(x,y,z)$ of the reference point of the ligand (e.g., its geometric center), rotation around $x$, $y$, and $z$ axes,
index of the conformation $l\in \L$ and the actual scale $s$.
All possible configurations form the configuration space $\C$. 
A configuration is collision-free if none of the ligand atoms scaled by $s$ and placed at the
position defined by $q$ collides with the protein atoms.
%The set of all feasible configurations at scale $s$ is denoted $\CFD \subseteq \C$.

\subsection{Computing Initial Configuration}

The analysis of tunnel traversability is based on computation of multiple trajectories of the ligand inside the tunnel, which
requires to generate a set $\QI$ of collision-free initial configurations.
In this paper we assume that the ligand has to travel from the beginning to the end of the tunnel.
As the tunnels are computed for a spherical probe, the ligand may not fit exactly to the first sphere of the tunnel.
The initial configurations have to be searched around the first sphere of the tunnel.
To find a new initial configuration, a random sample $q$ is generated around $c_i$ in the distance $\RI$ (translation and rotation
 parts of $q$ are generated randomly, the scale is set to $\smin$ and the conformation index is set randomly).
If the sample $q$ is collision-free, it can be considered as a new starting configuration, so $q$ is added to $\QI$.
Similarly, a single goal configuration $\qgoal$ is found around the end of the tunnel. 
For each starting configuration $\qinit \in \QI$,  $K$ trajectories are created.
Each trajectory is computed using a modified RRT, which is introduced in the following section.

\subsection{Computing Single Trajectory of the Ligand}

The task of the trajectory computation is to find a trajectory for a ligand in the given tunnel. 
% which can be solved using RRT~\cite{lavalleRRT}.
Here, the original RRT is extended to cope with the specific requirements needed for the traversability of ligands.
First, the trajectory has to be found around the given tunnel, but deflection from the tunnel centerline is allowed.
%The deviation of the trajectory from the tunnel centerline is influenced by the parameter $\rv$.
%We assume, that the distance of the ligand from the tunnel centerline is at most $\rv$.
Due to this requirement, sampling process of RRT has to be adapted in order to follow the tunnel, and to prevent
construction of trajectories in the rest of the protein.
%The other modifications aim to cope with various scales of the ligand and multiple conformations,

The main loop of the proposed method is described in Alg.~\ref{alg::main}.
In each iteration, a random sample $\qrand$ is generated and its nearest node $\qnear\in\T$ in the tree is found.
The nearest-neighbor search between $\qrand$ and the tree is performed using the weighted 6D Euclidean metric considering
both 3D rotation and 3D translation.

To guide the growth of the tree through the tunnel, a moving virtual goal is used~\cite{vonasek2009rrt}.
The virtual goal $v, 1\le v \le n$, is the index of a sphere of the tunnel.
The random samples $\qrand$ are generated around the sphere $c_v \in T$ with the probability $\gb$, and from the whole $\C$ otherwise.
%with probability $1-\gb$ from $\C$, otherwise they are generated around the sphere $c_v \in T$.
After the tree reaches the sphere $c_v$, i.e., the distance of the tree to $c_v$ is
less than a predefined threshold $\dt$, the virtual goal is moved to the successor of the last sphere in the tunnel
that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
Setting the virtual goal to this successor allows the tree to avoid such parts of the tunnels that are not traversable or reachable by the ligand.
This is necessary in the dense protein structures, where it is not always possible to exactly follow the tunnel.
The algorithm terminates after a predefined number of planning trials $\Imax$ or if the tree reaches
the last sphere in the tunnel, i.e., when $v = n$.

To generate the samples $\qrand$ around the virtual goal $v$, the translation part $(x,y,z)$ of $\qrand$ is generated
from $N(c_v,\Sigma)$, where $\Sigma$ is the diagonal matrix with diagonal entries equal to the parameter $\rv$, and the rotational
part of $\qrand$ is generated using techniques described in~\cite{kuffnerES}.
The other two parameters (ligand index $l$ and scale $s$) of $\qrand$ can be left zero, as these are not used in the employed
metric for the nearest-neighbor search.
The parameter $\rv$ influences the distribution of samples around the tunnel centerline. 
By setting $\rv$ to a small value, the planner attempts to find the trajectories inside the tunnel, while higher values
of $\rv$ cause  exploration of paths around the tunnel.
We propose to set this parameter to the average tunnel width.

\linesnumbered
\begin{algorithm}[h]
{\small
%\setstretch{0.88}
\caption{\label{alg::main}Main loop of the RRT planner}
\KwIn{
    tunnel $T=( (c_i, r_i) )$, $i=1,\ldots,n$, with spheres centers $c_i \in \R^3$ and radii $r_i$,
    initial configuration $\qinit$
%    goal configuration $\qgoal$,
}
\KwData{
   ligand conformations $\L$,
   scale limits $\smin, \smax$ and $\sdelta$
}
\KwOut{
    configuration tree $\T$\;
}
\hrule
$v = 1$; // index of the virtual goal\\
$iteration = 0$\;
\While{$iteration < \Imax$ {\bf and}  $v < n$}{
    \eIf{$rand() < \gb$}{
        $\qrand$ = random sample around actual virtual goal $c_v \in T$\;
    }{
        $\qrand$ = random sample from $\C$\;
    }
    $\qnear$ = nearest node in $\T$ towards $\qrand$\;
    expand($\qnear,\qrand$)\;
    \For{$i= n-1,n-2,\ldots,v+1,v$}{ \nllabel{alg::main:a}
        $d$ = nearest node in the tree towards sphere $c_i$\;
        \If{$d < \dt $}{
            $v = i+1$; // new virtual goal found\;
            {\bf  break}\;
        }
    } \nllabel{alg::main:b}
    $iteration = iteration+1$\;
}
\return $\T$\;
}
\end{algorithm}

The core of the proposed planner is the expansion procedure (Alg.~\ref{alg::expand}) which generates new collision-free nodes around $\qnear\in\T$.
For each ligand conformation $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with a maximal scale.
First, the maximal scale $\smax$ is tested and $m$ random samples are generated around $\qrand$ and tested for collision.
The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
If none of the tested samples is collision-free, the scale is reduced to $\smax-\sdelta$ and the search continues
until a collision-free sample is found or until the minimal reduced-scale $\smin$ is reached.
The random samples are generated similarly as in the case of $\qrand$ samples, only their translation
part is generated around $\qnear$.

The sampling-based methods are sensitive to the employed metric, especially if the objects are not symmetrical, which is the
case of the non-spherical ligands.
To consider the actual shape of the ligand (which is different in each conformation) and to support finding such configurations
that maximally approach $\qrand$, the distance between newly generated configurations and $\qrand$ is measured as the smallest
3D distance between an atom of the ligand placed at $q$ and the 3D position of $\qrand$ ($\da(q,\qrand)$ on line~\ref{alg::expand:a} in
Alg.~\ref{alg::expand}).
By computing the distance between the nearest atoms, the shape of the ligand is actually considered.
This metric supports retraction of the ligand towards $\qrand$.
%As the expansion routine attempts to find such sample that minimizes this distance to $\qrand$, it supports retraction of the ligand towards $\qrand$.

\begin{algorithm}[h]
{\small
%\setstretch{0.88}
%\setstretch{\straa}
\caption{\label{alg::expand}expand}
\KwIn{
   configuration $\qnear$ to be expanded,
   random configuration $\qrand$
}
\KwData{
   ligand conformations $\L$,
   scale limits $\smin, \smax$, and $\sdelta$,
   tree $\T$
}
%\KwOut{
%    set $S$ of collision-free configurations reachable from $\qnear$\;
%}
\hrule
\ForEach{$l \in \L$}{
    \ForEach{$s \in (\smax,\smax-\sdelta, \ldots, \smin+\sdelta, \smin)$}{
        $\qnew = \emptyset$; // empty configuration\\
        \For{$i = 1,\ldots,m$}{
            $q=\qnear$\;
            $q.position$ = random 3D position around $\qnear$\;
            $q.rotation$ = random 3D rotation\;
            $q.l = l$\; 
            $q.s = s$\;
            \If{isCollisionFree($q$)}{
                \If{$\qnew = \emptyset$ {\bf or} $\da(q, \qrand) < \da(\qnew,\qrand)$}{ \nllabel{alg::expand:a}
                    $\qnew = q$\;
                }
            } 
        }
        \If{$\qnew \ne \emptyset$} {
            $\T$.addNode($\qnew$)\;
            $\T$.addEdge($\qnear,\qnew$)\;
            {\bf break;} // go to next conformation
        }
    }
}
}
\end{algorithm}

The result of each planning trial is the tree $\T$ of collision-free configurations in which a path
between $\qinit$ (root of the tree) and $q'$ is found, where
 $q'$ is the nearest node in the tree towards $\qgoal$ (using 3D Euclidean metric). 
The path $P=(q_i), q_i \in \C$ is represented as a sequence of collision-free configurations.
The path is found in the tree even if the tree does not approach $\qgoal$ close enough.
Considering also these non-feasible solutions is necessary to evaluate difficult areas of the tunnels, e.g. bottlenecks.
The utilization of all computed paths for evaluation of tunnel difficulty is described in the next section.

\subsection{Traversability Characteristics}

For each initial configuration $\qinit \in \QI$, $K$ trajectories are computed, which results in the set of $K |\QI|$ trajectories.
All these trajectories are used to compute the following properties of the tunnel $T$.
A trajectory $P$ reaches the tunnel sphere $c_i \in T$, if the 3D Euclidean distance of the 
nearest configuration $q \in P$ towards $c_i$ is less than $r_i$ (radius of the sphere $c_i$).
Let $N(i)$ denote the number of trajectories that reached $i-$th sphere of the tunnel, $i=1,\ldots,n$.
Three basic characteristics of the tunnel are computed from the trajectories: accessibility, throughput, and the scaling factor.

The accessibility $A(i)=N(i)/N$ of the sphere $i$ is the probability of reaching the sphere $i$, where $N=K|\QI|$ is the total number of trajectories.
The accessibility shows how probable it is to pass the tunnel up to the sphere $i$.
Obviously, the most important is $A(n)$ of the last sphere of the tunnel, which can be considered as the overall difficulty of the tunnel.
The  ligand passage may however be strongly affected by the first bottleneck, so the parts of the tunnel located behind
the bottleneck have low accessibility.

%To better measure the probability of passing a given tunnel part, we should take into account only trajectories that even have chance
%to visit this part, e.g., the trajectories that reached the preceding tunnel part.
The throughput $T(i)$ is the ratio of trajectories that passed sphere $i$ (i.e., visited sphere $i+1$) and reached the sphere $i$,
i.e., $T(i) = N(i+1) / N(i)$. 
The throughput is not computed for the last sphere ($i=N$).
The throughput shows a local accessibility of tunnel parts and it can be used to detect places where most of the trajectories ends.

%The probability of the ligand that it reaches a tunnel's part.
%We employ the tunnel representation from CAVER 3.0 and we define a tunnel part as a sphere $P = (x, r)$ where $x \in \mathbb{R}^3$ is sphere center and $r \in \R$ is its radius.
%Then, for each trajectory $t$ we define its ending tunnel part $P_t^e$ as the part whose center is closest to the end position of $t$.
%Finally, the accessibility of $P$ is defined as a ratio of trajectories that passed $P$ or ended in $P$ over all trajectories.
%On the other hand, the probability does not show any details about the ligand's passage, e.g., how complicated it is to pass through some part of the tunnel, which trajectory is the most used one, etc.
%Moreover, it is not feasible to directly visualize and explore the resulting trajectories due to the visual clutter (see Fig.~\ref{fig:trajectories} left) caused by their high number and close proximity.

The proposed planner is allowed to scale down the radii of ligand spheres up to the allowed scale $\smin$. 
It can be expected that narrower parts of the tunnel are more often passed with a more scaled-down ligand than
the wider parts.
Ligand scale $L(i)$ at the tunnel sphere $i$ is the average scale of ligand that reaches the sphere $i$.
%The ligand scale is computed as the average value of scales of the nearest configurations of the trajectories towards tunnel sphere~$i$.

%--- The scale of the ligand when it passes a given tunnel's part.
%For the trajectory $t$, the ligand scale $s(P_t)$ in the tunnel part $P_t$ is defined as the scale of such sample of $t$ whose position is the closest one to the center of $P_t$.
%Then, the average ligand scale is defined as $s(P) = \frac{1}{|T|} \sum_{t \in T} s(P_t)$.
%Similarly, the minimum and maximum ligand scales can be defined.

\subsection{Visualization of Results}

%In order to fursupport the tunnel exploration process performed by biochemists.
The above defined characteristics can be shown even as a graph, or better, presented visually by mapping them using colors to the tunnel spheres.
 Alternatively, the surface representation can be used to visualize the tunnel.
In this case, a 3D point on the tunnel surface is colored according to the property value in its nearest tunnel part $i$,  i.e., such part $i$ whose center $c_i$ is the closest to the point among all tunnel parts.

The examples of the color mapping are shown in Fig.~\ref{fig:properties}.
The accessibility (Fig.~\ref{fig:properties}a) shows that more than the half of the tunnel is not accessible (red part of the tunnel).
The throughput shows (Fig.~\ref{fig:properties}b) that the only difficult is the part around the bottleneck (red color in Fig.~\ref{fig:properties}b), and the second half of the tunnel is also traversable.

\begin{figure}[h]
\centering
\begin{tabular}{ccc}
\includegraphics[width=0.3\textwidth]{fig/accessibility} &
\includegraphics[width=0.3\textwidth]{fig/throughput} &
\includegraphics[width=0.3\textwidth]{fig/ligand-scale} \\
  (a) & (b) & (c) \\                     
\end{tabular}
\caption{Tunnel properties mapped onto its surface. The tunnel begins at the top left corner.
(a) Accessibility --- green denotes the parts that were accessed by the ligand easily ($A(i) > 80$~\%), 
    while red denotes parts that were accessible  in $<10$~\% of cases.
(b) Throughput --- green denotes parts that are highly probable to be passed by the ligand ($T(i) > 80$~\%), 
    while red denotes those that were hard to pass ($T(i) < 20$~\%).
(c) Ligand scale --- the average scale of the ligand was \tylde 50\% in orange parts, \tylde 60\% in yellow parts, \tylde 80\% in yellow to green parts and \tylde 100\% in green parts.
\label{fig:properties}
}
\end{figure}

\def\cstart{\overline{c_{start}}}
\def\cend{\overline{c_{end}}}

Besides the color mapping, it is also necessary to show the computed trajectories.
%This is useful, for example, to find out in which part of the tunnel the trajectories detour.
Simple visualization of all trajectories could however be too slow for an interactive work.
Therefore, the trajectories are first clustered and then only the clusters are visualized.
Due to different lengths of the trajectories, they are first converted to a normalized form.
Let $\cstart$ represent the average starting position of all trajectories and let $d_{max}$ be the 3D Euclidean distance
of the most distant configuration from $\cstart$ among all trajectories.
A set of $M$ spheres centered at $\cstart$ are created with radii $r'_i=i {d_{max} \over M}$, where $i=0,\ldots, M-1$.
The trajectory $P=(q_1,\ldots,q_n)$ of length $n$ is represented by the normalized vector $v=(x_1,\ldots,x_M)$ of length $M$,
where $x_i$ is the 3D position of the nearest configuration $q \in P$ to the surface of the $i$-th sphere with the radius $r'_i$.
The distance between two normalized trajectories $v_i$ and $v_j$ is defined as
$d(v_i, v_j) = \frac{1}{M} \sum_{1 \leq k \leq N} |x_k^i x_k^j|$.
This distance is used in the UPGMA clustering technique~\cite{sokal1958statistical}.
The trajectories can be visualized using a representative of each cluster.
The number of trajectories in each cluster is represented by the width of the polyline (Fig.~\ref{fig:trajectories}).
%In this manner, we are able to convey the information about all different trajectories together (see Fig.~\ref{fig:trajectories} right).

\begin{figure}
\centering
\begin{tabular}{cc}
\includegraphics[width=0.4\textwidth]{fig/trajectories-all} &
\includegraphics[width=0.4\textwidth]{fig/trajectories-clustered-21} \\
(a) & (b) \\                       
\end{tabular}
\caption{Visualization of the trajectories. The tunnel begins in the top left corner.
(a) All trajectories ($\sim5000$) colored according to whether they reached the end of the tunnel (green) or not (red). 
(b) Visualization using clusters of trajectories. 
%Trajectories clustered according to their proximity colored by distinct colors.
%The different line widths represent the number of trajectories in a cluster.
\label{fig:trajectories}
}
\end{figure}

\section{Experimental Verification}

The tunnels in Haloalkane dehalogenase protein (PDB ID 1CQW) has been analyzed using the proposed approach.
Three tunnels were detected using CAVER 3.0~\cite{caver3} for the spherical probe of radius 0.9~\AA (Fig.~\ref{fig::tunnel}a).
The traversability was evaluated for 1-chlorpropan (denoted as $\LA$) with 11 atoms and 
% from David: m003 je "1-chlorpropan" a m004 "1-chlorbutan"
and 1-chlorbutan (denoted as $\LB$) with 14 atoms.
$\LA$ was represented by 12 conformations, and $\LB$ by a set of $114$ conformations.
Examples of three conformations of $\LA$ and $\LB$ are depicted in Fig.~\ref{fig::tunnel}b.
Further information about the experiments can be found at %{\url{http://mrs.felk.cvut.cz/isrr2017}}.

Both tested ligands have more than 10 atoms and therefore they cannot fit into the tunnel of with 0.9~\AA, so
the radii of ligands were scaled down. 
Four different scaling-down factors were used: $\smin=\{0.3,0.4,0.5,0.6\}$.
No trajectories were found for $\smin > 0.6$.
For each scale, 50 collision-free initial configurations were found in the $\RI=2$\AA\ radius around the first sphere of the tunnel.
For each ligand, each minimal scale $\smin$ and each initial configuration, 100 trajectories were computed using the proposed planner.
The parameters of the planner were: $\Imax=10,000$, $m=50$, $\gb=0.9$, $\dt=1.5$\AA, $\rv=2$~\AA.
For each initial configuration, the success rate is computed as the ratio of trajectories that reached the end of the tunnel (to the distance 
 $\rv=2$~\AA or less) over 100 trials.

The average runtimes, success rates and sizes of the built trees for the first tunnel are shown in Tab.~\ref{tab::rrt}.
The highest average success rate was achieved for the most scaled-down ligands ($\smin=0.3$) and it decreases with the increasing $\smin$. 
The difference between the minimal and maximal success rates indicates how the initial configurations of the ligand influence the ability to pass the tunnels.
For example, the most difficult initial configuration of $\LA$ and $\smin=0.3$ lead to success rate 78~\% (the easiest 
        initial configuration lead to success rate $100$~\%), but for the scale $\smin=0.4$, the worst success rate is
only $42$~\% and the best 96~\%.
The traversability of larger ligands ($\smin=0.4$ vs. $\smin=0.3$) is therefore more sensitive to the initial configuration.
This shows the importance of testing the traversability from multiple initial configurations.
The runtime is significantly higher for the $\LB$ ligand, which is caused by the larger number of conformations that need to be
examined in each expansion.
The average values of the runtimes, which are in the order of minutes, are negligible in the comparison of runtime of MD simulations.

%The tree can be expanded by up to $|\L|$ conformations in each iterations, which gives theoretical
%size of the tree $110\cdot 10^3$ for $\LA$ and $1140\cdot 10^3$ for $\LB$.
%The constructed tree are however significantly smaller

\def\tmpa{0.13\textwidth}

\begin{figure}
\centering
\begin{tabular}{cc}
%\includegraphics[width=0.38\textwidth]{fig/protein-tunnel-gray} &
\includegraphics[width=0.3\textwidth]{fig/t05proteintunnels} &
\includegraphics[width=0.45\textwidth]{fig/ligAll}  \\
(a) & (b)                        
\end{tabular}                       
\caption{\label{fig::tunnel}
    (a) The protein 1CQW visualized using the cartoon representation (gray) with 
        three tunnels (red, green, blue) detected by CAVER 3.0.
        The first tunnel is depicted red.
%        its highest ranked tunnel (green) which was detected using CAVER 3.0.
    (b) Examples of three conformations of $\LA$ (top) and $\LB$ (bottom)
}
\end{figure}

\begin{table}
\centering
\caption{\label{tab::rrt}
    Runtime and success ratio for ligand $\LA$ (left) and $\LB$ (right).
}
{
\small
\renewcommand{\tabcolsep}{1pt}
\input table.003.tex
\hskip 4pt   
\input table.004.tex
}
\end{table}

Trajectories for $\LA$ in all detected tunnels were classified as successful if they reached the last sphere of the tunnel
to distance $\dt=2$~\AA and they were considered unsuccessful otherwise.
The throughput computed from the trajectories shows that the tunnels are difficult not only around bottlenecks, but
also in other places.
The comparison of the classic bottleneck (i.e., measured by the radius of spherical probe) and throughput is depicted in Fig.~\ref{fig::tunnel2}a.
The trajectories for $\smin=0.4$ are depicted in Fig.~\ref{fig::tunnel2}b and for $\smin=0.5$ in Fig.~\ref{fig::tunnel2}c.
The successful trajectories for $\smin=0.4$ reveal that the end of the tunnel No. 3 can be approached by two different pathways (one in the tunnel and another one outside the tunnel).
The detail is depicted in Fig.~\ref{fig::detail}.
Despite the low bottleneck of this tunnel, the ligand may reach its end using the alternative pathway.

\begin{figure}
\centering
{
\renewcommand{\tabcolsep}{-5pt}
\begin{tabular}{ccc}
\includegraphics[width=0.38\textwidth]{fig/t05proteinBottle2T} & 
\includegraphics[width=0.38\textwidth]{fig/t04goodT} & 
\includegraphics[width=0.38\textwidth]{fig/t05goodT} \\
\includegraphics[width=0.38\textwidth]{fig/t05thpT} &
\includegraphics[width=0.38\textwidth]{fig/t04badT} &
\includegraphics[width=0.38\textwidth]{fig/t05badT} \\ 
(a) & (b) & (c)                       
\end{tabular}                       
\caption{\label{fig::tunnel2}
    (a) Classic bottleneck for spherical probe (top) and visualization of throughput (bottom) for the ligand $\LA$.
    (b) Successful (green, top) trajectories that reached end points of the tunnels,
        and unsuccessful ones (red, bottom) for $\smin=0.4$.
    (c) Successful and unsuccessful trajectories for $\smin=0.5$.
}
}
\end{figure}

\begin{figure}
\centering
{
\renewcommand{\tabcolsep}{0pt}
\begin{tabular}{ccc}
\includegraphics[width=0.325\textwidth]{fig/tunne4ac}& 
\includegraphics[width=0.325\textwidth]{fig/tunne4bc} & 
\includegraphics[width=0.325\textwidth]{fig/tunne4cc} \\
(a) & (b) & (c)                        
\end{tabular}
}
\caption{\label{fig::detail}
    Detailed view to alternative pathway around the third tunnel in 1CQW.
    The successful trajectories are in green (a) the unsuccessful in red (b).
    (c) Shows visualization of protein atoms around the tunnel.
}
\end{figure}

\section{Discussion}

%Trajectories computed for a ligand moving in a tunnel can reveal such properties that cannot be computed
%only based on the geometry of the tunnels.

The computed trajectories cannot be considered as `real', they should rather be considered as a hint for the biochemists.
One of the reason is that the trajectories are computed inside a static protein and static tunnels.
Real proteins are dynamic structures and consequently also the tunnels are dynamic: they move, merge or even disappear due to motions of protein atoms.
Contrary, usually static tunnels are used in protein engineering.
%multiple tunnels are detected separately in each frame of molecular dynamics and the tunnels are then aggregated to a single one.
Despite such a simplification, researches often used information about the static tunnels  to estimate the possibility of chemical reactions
at the active sites.

The second limitation is caused by the utilization of scaled-down ligands.
The protein atoms fill the internal void space therefore the tunnels identified inside proteins without ligand tend
to be very narrow.
To enable at least some motion of the ligands inside such narrow tunnels, the scaling-down of the ligand atoms is necessary.
It is used also in other related tool, e.g.~\cite{cortes2005path}.
%By allowing more aggressive scaling-down, more easier a trajectory can be found.
%The tunnels therefore should be always evaluated considering various minimal scales.

Due to these reasons, the computed trajectories can be considered either as too optimistic
(e.g. because they are computed on a wide tunnel that can be in fact closed due to molecular dynamics),
or too pessimistic (e.g. the trajectory is not found because of narrow passage around the tunnel, which can be 
opened due to the molecular dynamics).
Despite these limitations, testing the tunnel traversability using motion planning technique can provide chemists more information
than the simple bottleneck radius, which is used nowadays.
For example, possible detours from the tunnel can be identified, which indicates that even a tunnel with a small
bottleneck can be traversed (Fig.~\ref{fig::detail}).

%To better utilize the results of motion planning, it is necessary to observe behavior of ligands in many simulations with various proteins.
%To the best knowledge of the authors, this is the first work attempt to analyze traversability of ligands in protein tunnels.
%Most related works, e.g. those based on motion planning, are usually focused on protein folding, or finding arbitrary pathways
%from proteins~\cite{cortes2010simulating,guieysse2008structure}.

The proposed planner is very simple and it can be improved in many ways: by using a less aggressive scaling-down technique (in order
to obtain better trajectories) or by using only a subset of conformations and a better expansion (in order to speed it up).
Having a better motion planner (in the term of the speed or even success rate) however does not automatically mean that the resulting trajectories will help the biochemists more.
The presented work shows possible advantages of the analysis based on ligand trajectories.
By comparing the planned trajectories with trajectories observed in MD simulations, thresholds for the success rate or minimal tunnel throughput can be set, which will help biochemists to make decisions.

%The runtime of the planner is currently given mainly by the collision-detection.
%As the ligand is assumed to travel only along the tunnel centerline (in the distance defined by $\rv$),
%only atoms located in the vicinity of the tunnel can be considered for collision detection.
%Fast collision-detection can be achieved either using fast hierarchical data structures as OBB (implemented, e.g., in~\cite{ozcollide})
%    or using collision-detection system developed directly for proteins~\cite{ruiz2005biocd}.
%We assume that a conformation $l \in \L$ can be switched to all other conformations in $\L$.

%parmas:
%$\RI$
%$\rv$
%$\dt$
%The presented expansion procedure favors the samples with maximal scale (they are tested first), the trajectory contains samples with higher scales at the easy part of the tunnels (e.i., wide areas), and the smaller scales are used only if the ligand has to fit into a narrow space.

\section{Conclusion }
%In the future, we will focus on the following extensions proposed by the domain experts.
%Only ligand flexibility is included in the current algorithm but the flexibility of amino acid side-chains would greatly improve the prediction. 
%Moreover, incorporation of some scoring function for energy evaluation would introduce the characterization of not only geometrical but also physico-chemical properties and would further improve the prediction ability of our approach.

The protein tunnels are transporting pathways for ligands to the  active sites.
In this paper, we propose to analyze the traversability in the tunnels using motion planning, that can consider both shape
of the protein and ligand.
We proposed modifications to the RRT method to compute trajectories for a ligand represented by a library of common conformations.
The corresponding configuration space is sampled using the guided sampling, where a virtual sphere moves along the tunnel and attracts growth
of the tree towards it.
To enable trajectory computation in narrow tunnels, atomic radii of the ligand are scaled down.
Computed trajectories allows us to evaluate the traversability of the tunnels using
accessibility, throughput and also by visualizing typical pathways found by the planner.
These properties can help to better understand importance of the tunnels.

%which can be useful in the virtual screening

\bibliographystyle{plain}
\bibliography{paper}

\end{document}

### 2017-08-03 converttojpg.pl

@@ -1,0 +1,9 @@
+#!/usr/bin/env perl
+
+foreach my $f (glob "*.png") {
+    print "Converting $f .. ";
+    my $nf = $f;
+    $nf =~ s/png$/jpg/g;
+    print "convert $f $nf\n";
+    system "convert $f $nf\n";
+}

### 2017-08-23 paper.tex
intro

@@
 \newcommand{\red}[1]{\textcolor{red}{#1}}
-%\newcommand{\tylde}{\raisebox{-0.85ex}{\textasciitilde}} %does not work for vojta
 \newcommand{\tylde}{$\sim$}
@@
 
-\title{    Hey we need a title}
+\title{Motion planning for assessing accessibility of protein tunnels}
 
-\author{ Vojt\v ech Von\' asek$^{1}$,
+\author{Vojt\v ech Von\' asek$^{1}$,
     Barbora Kozl\'\i kov\'a$^{2}$,
@@
 \begin{abstract}
-%Understanding of interactions between enzymes and ligands is very important in enzymology, protein engineering or drug design. 
-Molecular tunnels connect catalytic active sites with the solvent and they play an important role in catalytic and thermodynamic properties.
-The geometrical and physico-chemical properties of tunnels need to be determined in order to understand the transport of a ligand from and into the protein active site. 
-Current methods describing the ligand passing through the tunnel are based either solely on the geometry of the tunnel or on molecular dynamics (MD) simulations.
-The tunnels are usually computed using Voronoi diagrams and therefore they can show possible movements only for a single atom.
-Ligand movements inside protein can be computed using MD simulations, but they are too slow for a daily routine.
-The tunnels computed for a single atom can however be used as a hint for ligand transport from or to the active site.
-The task of ligand transportation through a tunnel considering its shape and also conformation changes can be formulated
- as a motion planning problem.
-We propose to evaluate the tunnels based on ligand trajectories computed by motion planning, namely using
-Rapidly Exploring Random Tree (RRT).
-We present modifications to RRT in order to compute the trajectories around a given tunnel and also to consider the flexibility of the ligand.
-Besides, methods for visualization of the trajectories and their properties are presented.
-The goal of the paper is to show that motion planning can provide additional insight about tunnel properties, being
-less computationally demanding than the MD simulations.
-The proposed approach can be used to improve virtual screening results which are usually based only on determination of ligand binding in the active site but no information about ligand transport through the tunnel is provided.
+Chemical interactions of proteins and other molecules takes place in actives sites that are usually deeply burid inside the proteins.
+Motion planning can be used to analyze accessibility of tunnels by small molecules (ligands), which can help chemists to ..
+The ligands are not rigid structures, but they are rather flexible.
+In this paper, we present sampling-based motion planinng method to compute trajectories of the ligands inside know protein tunnels.
+The tunnels are represented by a set of spheres and they are used to focus sampling of the configuration space.
+To decrease the number of degrees of freedom, the ligand flexibility is modeled by a set of predefined conformations.
+We propose a new metric to measrue distances between the configurations which considers the shape of the ligands.
 \end{abstract}
@@
 \renewcommand{\tabcolsep}{0pt}
-\begin{tabular}{ccccc}
+\begin{tabular}{ccc}
 \includegraphics[width=0.15\textwidth]{fig/motiv1} &
 \includegraphics[width=0.17\textwidth]{fig/motiv2lab} &
-\includegraphics[width=0.16\textwidth]{fig/motiv3} &
-\hbox{
-\vbox{
-\hbox{\includegraphics[width=0.25\textwidth]{fig/ta-1} }
-\hbox{\includegraphics[width=0.25\textwidth]{fig/ta-433}}
-} 
-}
-\\
-Protein 1CQW & Active site & Detected tunnels & Example of a  \\
-             &            & (orange)          & trajectory
+\includegraphics[width=0.16\textwidth]{fig/motiv3}  \\
+%\hbox{
+%\vbox{
+%\hbox{\includegraphics[width=0.25\textwidth]{fig/ta-1} }
+%\hbox{\includegraphics[width=0.25\textwidth]{fig/ta-433}}
+%} 
+%}
+Protein 1CQW & Active site & Detected tunnels \\ %& Example of a  \\
+             &            & (orange)          %& trajectory
 \end{tabular}
@@
 Traditionally, tunnels are detected using Voronoi diagrams~\cite{yaffe2008,caver3} assuming 
-%The widely used tools for tunnel detection are based on Voronoi diagrams~\cite{yaffe2008,caver3}.  %or a 3D grid~\cite{sehnal2013mole,petrek2006caver}.
 a spherical ligand (probe) and they are represented as a sequence of spheres.
 Biochemists decide if a tunnel can be used to transport a given ligand mainly based on the tunnel length and bottleneck (i.e., the radius of the smallest sphere that forms the tunnel).
-This is used for example in protein engineering, where the task is to change selected properties of a protein, e.g., its stability under different outer conditions~\cite{Koudelakova2013} or its activity of the protein towards other molecules~\cite{Pavlova2009}.
+This is used for example in protein engineering, where the task is to change selected properties of a protein, e.g., its stability under different outer conditions~\cite{Koudelakova2013}.
+% or its activity of the protein towards other molecules~\cite{Pavlova2009}.
 %The design of suitable ligands causing the desired changes can be speeded up by detecting and analyzing the tunnels leading to the active sites.
@@
 %ligand from the outside environment to the active site or vice versa. 
-
 Whereas the tunnel computation is already a well established research field, the simulation of ligand transportation through the detected tunnels is rather new.
@@
 The decision based only on spherical tunnels requires previous expertise in the domain and yet it may be imprecise.
+
 To provide biochemists  a better insight into the behavior of non-spherical ligands, it is necessary to compute the trajectories of the ligand considering its shape and possibly also its conformation changes.
+This can be solved using motion planning, where the ligand is considered as a robot moving amongst protein's atoms.
+Sampling-based motion planning methods are suitable candidates for this task, as they can consider non-spherical shape of the ligand as well as the conformational changes.
 
-We propose to analyze the traversability of the ligands in the tunnels using motion planning.
-The principle of Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} planner is used and further modified.
-The main idea of the proposed planner is to generate the random samples around a virtual sphere moving through the tunnel, which
-guides the search in the configuration space.
-This allows us to focus the sampling of the configuration space around the tunnel.
+Proteins are dense structures which limits movements of the ligand insider the tunnels, which leads to the narrow passage problem.
+To cope with this problem, we propose to utilize the principle of guided sampling~\cite{vonasek2009rrt,denny2016dynamic}.
+The random samples are not generated from the whole configuration space, but mainly around the tunnel.
+To enable ligand movements in narrow tunnels, a scaled-down version of the ligand is used, similarly, e.g., to~\cite{cortes2005path}.
+%We propose to analyze the traversability of the ligands in the tunnels using motion planning.
 The flexibility of the ligand is modeled using a predefined set of typical conformations.
-To enable ligand movements in narrow tunnels, a scaled-down version of the ligand is used, similarly, e.g., to~\cite{cortes2005path}.
-It helps to keep also potential solutions which do not fit to the geometric restrains but can be still feasible because of the physico-chemical properties.
+%It helps to keep also potential solutions which do not fit to the geometric restrains but can be still feasible because of the physico-chemical properties.
 %This strategy helps to overcome the limitations of the conformational discretation introduced by the rotamer approximation.
 
-The proposed method can be seen as an extension for the traditionally used tunnel detection tools.
+%The proposed method can be seen as an extension for the traditionally used tunnel detection tools.
 %Besides trajectory computation, visualization of the results in important. %we also propose methods for their visualization.
-Our motivation is to help the biochemists to perform so called virtual screening, where they test the traversability of a ligand through a given tunnel.
-In order to predict the success of ligand traversability, the virtual screening performs hundreds of thousands of tests and checks if the ligand passed through the tunnel.
-The advantage of the proposed approach is that we can search the ligand path inside a specific tunnel which substantially decreases the computational time and resources required for the virtual screening.
+%Our motivation is to help the biochemists to perform so called virtual screening, where they test the traversability of a ligand through a given tunnel.
+%In order to predict the success of ligand traversability, the virtual screening performs hundreds of thousands of tests and checks if the ligand passed through the tunnel.
+%The advantage of the proposed approach is that we can search the ligand path inside a specific tunnel which substantially decreases the computational time and resources required for the virtual screening.
 %We further propose visualization techniques for presenting the results to biochemists.

### 2017-08-24 paper.tex
intro,sota,bib

@@
-\documentclass[usletter, 10pt, conference]{ieeeconf}      % for icra
+%\documentclass[a4paper, 10pt]{IEEEtran}      % for icra/RAL final
+\documentclass[letterpaper, 10 pt, conference]{ieeeconf} % initial submission to RAL/ICRA
 
@@
 The presented work has been supported by the Czech Science Foundation (GA{\v C}R) under research project No. 17-07690S.
-Access to computing and storage facilities owned by parties and projects contributing to the National Grid Infrastructure MetaCentrum, provided under the programme "Projects of Large Infrastructure for Research, Development, and Innovations" (LM2010005), is greatly appreciated.}
-}
+Access to computing and storage facilities owned by parties and projects contributing to the National Grid Infrastructure MetaCentrum, provided under the programme "Projects of Large Infrastructure for Research, Development, and Innovations" (LM2010005), is greatly appreciated.}}
 
@@
 %Faculty of Science, Masaryk University, Kamenice 5, 625 00 Brno, Czech Republic\\
-
-%\makeindex
 
@@
 \maketitle
-%\thispagestyle{empty}
-%\pagestyle{empty}
+\thispagestyle{empty}
+\pagestyle{empty}
 
@@
 \begin{abstract}
-Chemical interactions of proteins and other molecules takes place in actives sites that are usually deeply burid inside the proteins.
+Chemical interactions of proteins and other molecules take place in actives sites that are usually deeply burid inside the proteins.
 Motion planning can be used to analyze accessibility of tunnels by small molecules (ligands), which can help chemists to ..
@@
 
+The idea of sampling-based motion planning is to randomly sample the configuration space of the robot and connect the collision-free samples
+into a graph structure (roadmap).
+A path in the roadmap then corresponds to a motion in the workspace.
+The two most used sampling-based techniques are Probabilistic Roadmaps (PRM)~\cite{kavrakiForPP} and Rapidly Exploring Random Trees (RRT)~\cite{lavalleRRT}.
+Originally, both samples utilize uniform distribution to sample the configuration space.
+As the random samples are classify as free or non-free using collision detection, the sampling-based methods can found paths for roobts of various shapes. 
+Another important advantage of the sampling-based planners is that they can cope with robots with many degrees of freedom (DOF).
+The well known issues is the narrow passage problem and the sensitiviy to distance metric in the configuration space.
+
+A narrow passage is a small region in the configuration space whose removal changes connectivity of the space.
+As the volume of the narrow passage is significantly smaller than then volume of the whole configuration space, it is difficult to 
+sample it dense enough using uniform ditribution.
+Consequently, many iterations are needed in order to put enough samples into it, which increases the computation time.
+
+A proper distance metric in the configuration space is necessary in order to determine which samples should be connected together (in the case of PRM), or which node of the tree should be expanded (in the case of RRT).
+Generally, it is not easy to combine the rotational and translational components of the configuration space.
+Moreover, the selection of the metric is also influence by the shape of the objects (robots).
+
+Sampling-based motion planners has been intensively studied in robotics, but they have been applied also in other research fields.
+In bio-chemistry, the planners have been used to study
+loop motions~\cite{cortes2004geometric},
+%protein folding~\cite{amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},
+protein folding~\cite{raveh2009rapid,novinskaya2015improving},
+or for tunnel detection~\cite{vonasek2017tunnel}.
+
+%or protein folding combined with ligand diffusion~\cite{cortes2010simulating}.
+%rrt for molecular: \cite{al2012motion}
+%survey \cite{gipson2012computational}
+%Sampling-based planners have been used in various applications in robotics~\cite{elbanhawi2014sampling} and also  %,latombe1999motion} and also
+%Solutions for tunnel detection using sampling-based methods however have not been discussed yet.
+%Besides robotics, sampling-based planners have been used also in applications in bio-i
+
 Understanding of interactions between proteins and other small molecules is crucial in many research fields, including drug design and protein engineering. 
@@
 To provide biochemists  a better insight into the behavior of non-spherical ligands, it is necessary to compute the trajectories of the ligand considering its shape and possibly also its conformation changes.
-This can be solved using motion planning, where the ligand is considered as a robot moving amongst protein's atoms.
+This can be formulated as the path planning problem where the ligand is considered as a robot moving amongst protein's atoms.
 Sampling-based motion planning methods are suitable candidates for this task, as they can consider non-spherical shape of the ligand as well as the conformational changes.
@@
 The random samples are not generated from the whole configuration space, but mainly around the tunnel.
-To enable ligand movements in narrow tunnels, a scaled-down version of the ligand is used, similarly, e.g., to~\cite{cortes2005path}.
+The dilatation of the ligand free space can be achieved by shrinking the atom radii, similarly, e.g., to~\cite{cortes2005path}.
 %We propose to analyze the traversability of the ligands in the tunnels using motion planning.
@@
 %Therefore, methods for visualization of the results  are discussed.
+
+energetic constraints are translated into geometric ones by considering a steric model of the molecules and applying collision detection
+\cite{deAngulo2005biocd}
+
+First the biredge test is applied to identify regions aeround narrow passages, optimization-based retraction operatoin sleectively only at those regions
+
+selective rrt~\cite{lee2012srrrt}
 
@@
 %survey \cite{gipson2012computational}
-%In our previous works, we proposed to employ RRT also for the purp
 %Solutions for tunnel detection using sampling-based methods however have not been discussed yet.
@@
 
-%Two main issues need to be addressed: the increased dimension of the configuration space and the narrow passage problem.
-%A well known issue of sampling-based planners is the narrow passage problem~\cite{hannaWIS}.
+The well known issue of sampling-based motion planners is the narrow passage problem. 
+Many approaches have been proposed to improve performance in the presence of narrow passages.
+The performance of PRM-based methods can be improved by filtering out samples that are not located insider narrow passages.
+For example, in~\cite{hsuBridge}, a random sample is classified as..
+
+In the Gaussian sampling strategy~\cite{overmarsGauss}, the samples are generated more frequently around obstacles.
+A random configuration $q_1\in\C$ is generated uniformly and another random configuration
+$q_2\in\C$ is drawn around using Normal distribution $N(q_1,\sigma^2)$.
+If both $q_1$ and $q_2$ are free or both are non-free, they are not added to the roadmap. 
+If only one of these configurations is free, then the free one is added to the roadmap. 
+The disadvantage of this strategy is the necessity to choose suitable value of the parameter $\sigma^2$, which depends
+on the used map and the shape of the robot.
+The Bridge-Test sampling~\cite{hsuBridge} employs the Gaussian strategy to generate two samples in order to compute their midpoint.
+%The Gaussian strategy was improved in the Bridge-Test~\cite{hsuBridge}. % to generate samples inside a narrow passage.
+%Two random configurations $q,q'\in\C$ are generated in same way as in the Gaussian strategy.
+%The midpoint $p$ on a segment $q,q'$ is constructed.
+If the midpoint is collision-free and both end configurations are non-free, the midpoint lies in a narrow passage, and it is added to the roadmap.
+Both Gaussian and Bridge-Test strategies have to be combined with the uniform sampling in order to ensure sampling of free-regions~\cite{sun2005narrow}.
+Despite the simplicity of these two modifications, they were proven to significantly improve performance of PRM~\cite{hsuOnProb,geraertsRA,geraertsPRMA,wang10adaptive}.
+
+Another techniques to sample the narrow passages dense enough is to retract the random samples towards regions that are believed
+to be located in narrow passages.
+In the Retraction-based RRT~\cite{zhangRetraction}, the random sample $\qrand$ is connected to the tree if the line segment $\qnear,\qrand$ is collision-free, otherwise, the retraction-step is performed.
+The task of the retraction step is to find a contact configuration  that minimizes the distance to $\qrand$.
+It is necessary to use an apporpriated distance metric in the configuration space.
+Generally, it is hard to combine the translational and rotational components in the configuration space.
+
+In~\cite{zhangRetraction}, the DISP metric~\cite{zhang2008fast}.
+
+The contact configuration $q$ is found on the line and its neighborhood is searched to find another contact configuration $q'$ such that 
+$\varrho(q',\qrand) < \varrho(q,\qrand)$.
+The retraction is terminated after several steps or if no new contact configuration can be found.
+The configuration trees built by RRT--Retraction can therefore contain more nodes than is the number of planning iterations,
+as the tree can be expanded by upto n contact configurations in each iteration.
+
+\cite{zhangRetraction}
+
+\cite{amatoOBRRT}
+
+dilatation
+\cite{hsu06multilevel}
+
+\cite{saha2005finding}
+
+DISP metric 
+\cite{zhang2008fast}
 

### 2017-08-28 paper.tex
abstract

@@
 \begin{abstract}
+Sampling-based motion planners can compute trajectories for many-DOF (Degrees Of Freedom) robots (objects) of various shapes, which finds
+applicatoins even beyond robotics, e.g. for studing accesible pathways of proteins in computational bio-chemistry.
 Chemical interactions of proteins and other molecules take place in actives sites that are usually deeply burid inside the proteins.
+The active site are accesible throu one or more tunnels (pathways), which have been traditionally computed using Voronoi diagrams.
+The Voronoi-based tunnel computation is however not suitable for ligands of arbitrary shapes.
+In this paper, we propose to analyze the accessibility of the active sites using motion planning.
+The require to cope with the narrow passage problem, as the proteins are dense structures which limits possible movements of the ligands.
+To cope with the narrow passage issue, we propose to guide the search in the corresponding configuration space using the Voronoi-based tunnels.
+We propose to imitate the ligands flexibility by utilizing a library of known (typical) conformations, which significantly decreases 
+the dimension of the configuration space.
+We propose to measure distances between the configurations using a metric that considers also the shape of the ligand.
+Old:
 Motion planning can be used to analyze accessibility of tunnels by small molecules (ligands), which can help chemists to ..

### 2017-09-02 paper.tex
intro

--- 2017-08-24 paper.tex
@@
 \begin{abstract}
-Chemical interactions of proteins and other molecules take place in actives sites that are usually deeply burid inside the proteins.
-Motion planning can be used to analyze accessibility of tunnels by small molecules (ligands), which can help chemists to ..
-The ligands are not rigid structures, but they are rather flexible.
-In this paper, we present sampling-based motion planinng method to compute trajectories of the ligands inside know protein tunnels.
-The tunnels are represented by a set of spheres and they are used to focus sampling of the configuration space.
-To decrease the number of degrees of freedom, the ligand flexibility is modeled by a set of predefined conformations.
-We propose a new metric to measrue distances between the configurations which considers the shape of the ligands.
+Chemical interactions of proteins and other molecules (ligands) take place in actives sites that are usually deeply burid inside the proteins.
+The active site are accesible throu one or more tunnels (pathways), which have been traditionally computed using Voronoi diagrams.
+The Voronoi-based tunnel computation is however not suitable for ligands of arbitrary shapes neither for flexible
+ligands, which can changes their conformation.
+In this paper, we propose to analyze the accessibility of the active sites using sampling-based motion planning.
+Thing brings three challenges: how to find path in a dense protein structures, where the tunnels behave like narrow passages,
+how to model the conformation changes of the ligands, and how to measure distance in the related configuration space.
+To cope with the narrow passage problem, we generate the random samples areound a Voronoi-based tunnel and we shrink
+tha atomic radii of the ligand.
+The flexibility of the ligand is modeled using a fixed set of known conformations.
+The distance between the configurations is measured using the nearest distance between two atoms.
 \end{abstract}
@@
 Originally, both samples utilize uniform distribution to sample the configuration space.
-As the random samples are classify as free or non-free using collision detection, the sampling-based methods can found paths for roobts of various shapes. 
+As the random samples are classify as free or non-free using collision detection, the sampling-based methods can found paths for robots of various shapes. 
 Another important advantage of the sampling-based planners is that they can cope with robots with many degrees of freedom (DOF).
-The well known issues is the narrow passage problem and the sensitiviy to distance metric in the configuration space.
-
-A narrow passage is a small region in the configuration space whose removal changes connectivity of the space.
-As the volume of the narrow passage is significantly smaller than then volume of the whole configuration space, it is difficult to 
-sample it dense enough using uniform ditribution.
-Consequently, many iterations are needed in order to put enough samples into it, which increases the computation time.
-
-A proper distance metric in the configuration space is necessary in order to determine which samples should be connected together (in the case of PRM), or which node of the tree should be expanded (in the case of RRT).
-Generally, it is not easy to combine the rotational and translational components of the configuration space.
-Moreover, the selection of the metric is also influence by the shape of the objects (robots).
-
-Sampling-based motion planners has been intensively studied in robotics, but they have been applied also in other research fields.
-In bio-chemistry, the planners have been used to study
-loop motions~\cite{cortes2004geometric},
+Thanks to these features, sampling-based planners has been utilized in wide range of applications in robotics, but also in 
+other research domain, namely in bio-chemistry 
+%Sampling-based motion planners has been intensively studied in robotics, but they have been applied also in other research fields.
+%In bio-chemistry, the planners have been used to study
+to study loop motions~\cite{cortes2004geometric},
 %protein folding~\cite{amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},
@@
 %Solutions for tunnel detection using sampling-based methods however have not been discussed yet.
-%Besides robotics, sampling-based planners have been used also in applications in bio-i
-
-Understanding of interactions between proteins and other small molecules is crucial in many research fields, including drug design and protein engineering. 
-These interactions are highly influenced by the ability to transport the small molecule, called ligand, to the protein active site.
-The active site can be considered as a deeply buried inner cavity with the ability to interact with the incoming ligand.
+%Understanding of interactions between proteins and other small molecules is crucial in many research fields, including drug design and protein engineering. 
+The problem of tunnel detection is to find a tunnel (pathways) leading to a specific place (called active site) inside the protein.
+The knowledge about the existence of the tunnels and their properties (e.g., their bottlenect and chemical properties around the tunnel),
+is important to understand interactions between proteins and other molecules (ligands).
 To get the ligand to the active site, there has to be a transportation path connecting the protein outer environment with the active site.
@@
 
+%These interactions are highly influenced by the ability to transport the small molecule, called ligand, to the protein active site.
+%The active site can be considered as a deeply buried inner cavity with the ability to interact with the incoming ligand.
 Traditionally, tunnels are detected using Voronoi diagrams~\cite{yaffe2008,caver3} assuming 
 a spherical ligand (probe) and they are represented as a sequence of spheres.
-Biochemists decide if a tunnel can be used to transport a given ligand mainly based on the tunnel length and bottleneck (i.e., the radius of the smallest sphere that forms the tunnel).
-This is used for example in protein engineering, where the task is to change selected properties of a protein, e.g., its stability under different outer conditions~\cite{Koudelakova2013}.
+However, ligands typically have non-spherical shape and it is therefore difficult to estimate their traversability through
+the Voronoi-based tunnels.
+
+%Biochemists decide if a tunnel can be used to transport a given ligand mainly based on the tunnel length and bottleneck (i.e., the radius of the smallest sphere that forms the tunnel).
+%This is used for example in protein engineering, where the task is to change selected properties of a protein, e.g., its stability under different outer conditions~\cite{Koudelakova2013}.
 % or its activity of the protein towards other molecules~\cite{Pavlova2009}.
@@
 %ligand from the outside environment to the active site or vice versa. 
-Whereas the tunnel computation is already a well established research field, the simulation of ligand transportation through the detected tunnels is rather new.
-Ligands are typically of non-spherical shape, therefore it is difficult to estimate their traversability through tunnels computed for a spherical probe.
-The decision based only on spherical tunnels requires previous expertise in the domain and yet it may be imprecise.
+%Whereas the tunnel computation is already a well established research field, the simulation of ligand transportation through the detected tunnels is rather new.
+%Ligands are typically of non-spherical shape, therefore it is difficult to estimate their traversability through tunnels computed for a spherical probe.
+%The decision based only on spherical tunnels requires previous expertise in the domain and yet it may be imprecise.
 
@@
 This can be formulated as the path planning problem where the ligand is considered as a robot moving amongst protein's atoms.
-Sampling-based motion planning methods are suitable candidates for this task, as they can consider non-spherical shape of the ligand as well as the conformational changes.
+Sampling-based planners are suitable candidates to solve this problem due to their ability to cope with many-DOF problems and objects with arbitrary shapes.
 
-Proteins are dense structures which limits movements of the ligand insider the tunnels, which leads to the narrow passage problem.
-To cope with this problem, we propose to utilize the principle of guided sampling~\cite{vonasek2009rrt,denny2016dynamic}.
+The well known issues of the sampling-based planners are the narrow passage problem and the sensitiviy to distance metric in the configuration space.
+A narrow passage is a small region in the configuration space whose removal changes the connectivity of the space.
+Due to their low volumes, the probability of sampling the narrow passages is low.
+Consequently, many iterations are needed in order to put enough samples into the passages, which increases the computation time.
+A proper distance metric in the configuration space is necessary in order to determine which samples should be connected together (in the case of PRM), or which node of the tree should be expanded (in the case of RRT).
+Generally, it is not easy to combine the rotational and translational components of the configuration space.
+Moreover, the selection of the metric is also influence by the shape of the objects (robots).
+
+To cope with these issues, we propose to guide the sampling of the configuration space using the Voronoi-baed tunnels.
 The random samples are not generated from the whole configuration space, but mainly around the tunnel.
-The dilatation of the ligand free space can be achieved by shrinking the atom radii, similarly, e.g., to~\cite{cortes2005path}.
-%We propose to analyze the traversability of the ligands in the tunnels using motion planning.
-The flexibility of the ligand is modeled using a predefined set of typical conformations.
+To increse movements of the ligands in the narrow tunnels, the free-space is dilated by 
+shrinking the atom radii, similarly, e.g., to~\cite{cortes2005path,hsuMultiLevel}.
+The ligand can be considered as a multi-link kinematic chain and by chaning the torsion angles, the flexibility can be modeled~\cite{songPFpath,cortes2005path}.
+This however increases the dimension of the configuration space.
+Instead, we model the flexibility using a fixed set of known conformations.
+
+%Proteins are dense structures which limits movements of the ligand insider the tunnels, which leads to the narrow passage problem.
+%To cope with this problem, we propose to utilize the principle of guided sampling~\cite{vonasek2009rrt,denny2016dynamic}.
 %It helps to keep also potential solutions which do not fit to the geometric restrains but can be still feasible because of the physico-chemical properties.
@@
 
+Free-space dilation is an effective aproach for narrow passage sampling~\cite{hsuMultiLevel}. 
+It is necessary to define proper method for dilating and determin the amount of dilation.
+
 \section{Related Work}
 
-The analysis of protein structure aiming to reveal the tunnels has been supported by different computational software tools which take the geometry of the protein as an input and explore the inner void space (e.g., CAVER 1.0~\cite{petrek2006caver} or MOLE~\cite{Petrek20071357}).
+%The analysis of protein structure aiming to reveal the tunnels has been supported by different computational software tools which take the geometry of the protein as an input and explore the inner void space (e.g., CAVER 1.0~\cite{petrek2006caver} or MOLE~\cite{Petrek20071357}).
 Early methods for tunnel detection utilized a discretized 3D grid, where each cell is considered as occupied or free depending
-on the presence of atoms of the protein.
+on the presence of atoms of the protein~\cite{petrek2006caver,Petrek20071357}.
 Tunnels can be then searched using standard graph-search methods, like Dijkstra's algorithm.
@@
 %are used primarily for analysis of static molecules or individual snapshots of molecular dynamics.
-
 Currently the most widely used approach to tunnel detection is based on ordinary Voronoi diagrams (VD) or Weighted Voronoi Diagrams (WVD). %~\cite{caver3,yaffe2008}.

### 2017-09-02 paper.tex
sota

@@
 These paths, called tunnels, have to be wide enough and the physico-chemical properties of the surrounding amino acids have to be compatible with the ligand (Fig.~\ref{fig::motiv}).
-
 %These interactions are highly influenced by the ability to transport the small molecule, called ligand, to the protein active site.
 %The active site can be considered as a deeply buried inner cavity with the ability to interact with the incoming ligand.
-Traditionally, tunnels are detected using Voronoi diagrams~\cite{yaffe2008,caver3} assuming 
+The tunnels are usually detected using Voronoi diagrams~\cite{yaffe2008,caver3} assuming 
 a spherical ligand (probe) and they are represented as a sequence of spheres.
@@
 To increse movements of the ligands in the narrow tunnels, the free-space is dilated by 
-shrinking the atom radii, similarly, e.g., to~\cite{cortes2005path,hsuMultiLevel}.
+shrinking the atom radii, similarly, e.g., to~\cite{cortes2005path,hsu06multilevel}.
 The ligand can be considered as a multi-link kinematic chain and by chaning the torsion angles, the flexibility can be modeled~\cite{songPFpath,cortes2005path}.
@@
 
-Free-space dilation is an effective aproach for narrow passage sampling~\cite{hsuMultiLevel}. 
+Free-space dilation is an effective aproach for narrow passage sampling~\cite{hsu06multilevel}. 
 It is necessary to define proper method for dilating and determin the amount of dilation.
@@
 The biggest disadvantage of both grid-based and VD-based methods is that the shape of the ligand is not taken into account during the tunnel detection and it is therefore not easy to estimate if (and how) a ligand might traverse the tunnels.
+In order to determine if a ligand can pass the tunnel, it is necessary to compute a trajectory considering the shape of the ligand.
+This can be formulated as a motion planning problem in a high-dimensional configuration space and solved using sampling-based planners.
+
 %The protein tunnels computed for spherical probes contain valuable information about the protein void space.
@@
 
-In order to determine if a ligand can pass the tunnel, it is necessary to compute a trajectory considering the shape of the ligand.
-This can be formulated as a motion planning problem in a high-dimensional configuration space.
-The configuration space $\C$ is formed by all possible configurations of the ligand in the tunnel, i.e., considering its rotation, translation, and possibly also other degrees of freedom responsible for the conformation changes.
-The dimension of the configuration space is given by the degrees of freedom (DOF) of the ligand, i.e., 6D for a rigid ligand and 6D + $n$ for a flexible
-ligand with $n$ DOFs.
-Sampling-based motion planning methods can be used to search this high-dimensional configuration space~\cite{Lav06}.
-The idea of sampling-based motion planning is to randomly sample $\C$ and classify the samples as free or non-free using collision detection.
-The free samples are stored in a roadmap (a graph structure), in which a path can be searched using standard graph-search methods.
+%The configuration space $\C$ is formed by all possible configurations of the ligand in the tunnel, i.e., considering its rotation, translation, and possibly also other degrees of freedom responsible for the conformation changes.
+%The dimension of the configuration space is given by the degrees of freedom (DOF) of the ligand, i.e., 6D for a rigid ligand and 6D + $n$ for a flexible
+%ligand with $n$ DOFs.
+%Sampling-based motion planning methods can be used to search this high-dimensional configuration space~\cite{Lav06}.
+%The idea of sampling-based motion planning is to randomly sample $\C$ and classify the samples as free or non-free using collision detection.
+%The free samples are stored in a roadmap (a graph structure), in which a path can be searched using standard graph-search methods.
 
-Sampling-based motion planning methods are suitable for computing the trajectories of the ligand as they 
-can cope with many-DOF robots (objects) of arbitrary shapes.
-The flexibility of ligands (or even proteins) can be modeled using a multi-link kinematic chain, where torsional angles
-can change~\cite{songPFpath}.
-This is useful, e.g., in the protein folding studies~\cite{al2012motion,gipson2012computational,amato2002using,raveh2009rapid} or analysis of loop motions~\cite{cortes2004geometric}.
+%Sampling-based motion planning methods are suitable for computing the trajectories of the ligand as they 
+%can cope with many-DOF robots (objects) of arbitrary shapes.
+%The flexibility of ligands (or even proteins) can be modeled using a multi-link kinematic chain, where torsional angles
+%can change~\cite{songPFpath}.
+%This is useful, e.g., in the protein folding studies~\cite{al2012motion,gipson2012computational,amato2002using,raveh2009rapid} or analysis of loop motions~\cite{cortes2004geometric}.
 %This is useful, e.g., in the protein folding studies~\cite{al2012motion,gipson2012computational,amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro} or analysis of loop motions~\cite{cortes2004geometric}.
@@
 The algorithm terminates if the tree approaches the goal configuration close enough.
-%RRT is a suitable candidate for computing the trajectories of flexible ligands due to its ability to cope with many-DOF systems.
 
@@
 
-The well known issue of sampling-based motion planners is the narrow passage problem. 
-Many approaches have been proposed to improve performance in the presence of narrow passages.
-The performance of PRM-based methods can be improved by filtering out samples that are not located insider narrow passages.
-For example, in~\cite{hsuBridge}, a random sample is classified as..
+Sampling-based planners can provide poor performance in the presence of narrow passages, as many iterations
+are required in order to sample the narrow passages dense enough.
+PRM-based techniques solve the problem by increasing probablity of sampling in regions close to obstacles~\cite{hsuBridge,overmarsGauss}.
+These regions can be estimated e.g. based on amount of collision-free samples around a given sample~\cite{overmarsGauss} or e.g. based
+on workspace information represented by medial axis~\cite{amatoOBRRT,amato2002using}.
 
-In the Gaussian sampling strategy~\cite{overmarsGauss}, the samples are generated more frequently around obstacles.
-A random configuration $q_1\in\C$ is generated uniformly and another random configuration
-$q_2\in\C$ is drawn around using Normal distribution $N(q_1,\sigma^2)$.
-If both $q_1$ and $q_2$ are free or both are non-free, they are not added to the roadmap. 
-If only one of these configurations is free, then the free one is added to the roadmap. 
-The disadvantage of this strategy is the necessity to choose suitable value of the parameter $\sigma^2$, which depends
-on the used map and the shape of the robot.
-The Bridge-Test sampling~\cite{hsuBridge} employs the Gaussian strategy to generate two samples in order to compute their midpoint.
+%For example, in~\cite{hsuBridge}, a random sample is classified as..
+%In the Gaussian sampling strategy~\cite{overmarsGauss}, the samples are generated more frequently around obstacles.
+%A random configuration $q_1\in\C$ is generated uniformly and another random configuration
+%$q_2\in\C$ is drawn around using Normal distribution $N(q_1,\sigma^2)$.
+%If both $q_1$ and $q_2$ are free or both are non-free, they are not added to the roadmap. 
+%If only one of these configurations is free, then the free one is added to the roadmap. 
+%The disadvantage of this strategy is the necessity to choose suitable value of the parameter $\sigma^2$, which depends
+%on the used map and the shape of the robot.
+%The Bridge-Test sampling~\cite{hsuBridge} employs the Gaussian strategy to generate two samples in order to compute their midpoint.
 %The Gaussian strategy was improved in the Bridge-Test~\cite{hsuBridge}. % to generate samples inside a narrow passage.
@@
 %The midpoint $p$ on a segment $q,q'$ is constructed.
-If the midpoint is collision-free and both end configurations are non-free, the midpoint lies in a narrow passage, and it is added to the roadmap.
-Both Gaussian and Bridge-Test strategies have to be combined with the uniform sampling in order to ensure sampling of free-regions~\cite{sun2005narrow}.
-Despite the simplicity of these two modifications, they were proven to significantly improve performance of PRM~\cite{hsuOnProb,geraertsRA,geraertsPRMA,wang10adaptive}.
+%If the midpoint is collision-free and both end configurations are non-free, the midpoint lies in a narrow passage, and it is added to the roadmap.
+%Both Gaussian and Bridge-Test strategies have to be combined with the uniform sampling in order to ensure sampling of free-regions~\cite{sun2005narrow}.
+%Despite the simplicity of these two modifications, they were proven to significantly improve performance of PRM~\cite{hsuOnProb,geraertsRA,geraertsPRMA,wang10adaptive}.
+
+RRT-based planners cannot cope with the narrow passage problem simply by increasing the probability of sampling in them, as the
+growth of the configuration tree may be toward the random samples may be blocked by an obstacle~\cite{vonasekphd}.
+Instead, the random samples has to be generated in the dififcult regions only if the tree can approach them.
+This is the main idea of the guided sampling~\cite{vonasek2009rrt,denny2014marrt,denny2016dynamic}.
 
@@
 
-In~\cite{zhangRetraction}, the DISP metric~\cite{zhang2008fast}.
-
-The contact configuration $q$ is found on the line and its neighborhood is searched to find another contact configuration $q'$ such that 
-$\varrho(q',\qrand) < \varrho(q,\qrand)$.
-The retraction is terminated after several steps or if no new contact configuration can be found.
-The configuration trees built by RRT--Retraction can therefore contain more nodes than is the number of planning iterations,
-as the tree can be expanded by upto n contact configurations in each iteration.
-
-\cite{zhangRetraction}
-
-\cite{amatoOBRRT}
-
-dilatation
-\cite{hsu06multilevel}
+%In~\cite{zhangRetraction}, the DISP metric~\cite{zhang2008fast}.
 

### 2017-09-03 paper.tex
sota

@@
 
-energetic constraints are translated into geometric ones by considering a steric model of the molecules and applying collision detection
-\cite{deAngulo2005biocd}
-
-First the biredge test is applied to identify regions aeround narrow passages, optimization-based retraction operatoin sleectively only at those regions
-
-selective rrt~\cite{lee2012srrrt}
-
-Free-space dilation is an effective aproach for narrow passage sampling~\cite{hsu06multilevel}. 
-It is necessary to define proper method for dilating and determin the amount of dilation.
-
 \section{Related Work}
@@
 %The analysis of protein structure aiming to reveal the tunnels has been supported by different computational software tools which take the geometry of the protein as an input and explore the inner void space (e.g., CAVER 1.0~\cite{petrek2006caver} or MOLE~\cite{Petrek20071357}).
-Early methods for tunnel detection utilized a discretized 3D grid, where each cell is considered as occupied or free depending
-on the presence of atoms of the protein~\cite{petrek2006caver,Petrek20071357}.
-Tunnels can be then searched using standard graph-search methods, like Dijkstra's algorithm.
-Besides, the grid can be used to identify other relevant properties like 
-pockets, cavities, or channels~\cite{sehnal2013mole,petrek2006caver}.
+%Early methods for tunnel detection utilized a discretized 3D grid, where each cell is considered as occupied or free depending
+%on the presence of atoms of the protein~\cite{petrek2006caver,Petrek20071357}.
+
+The analysis of ligand traversability to/from an active site is usually based on so called tunnels.
+The tunnels are pathways computed for a single atom (spherical probe) using Voronio-diagrams~\cite{yaffe2008,caver3} or 
+grid-based approaches~\cite{sehnal2013mole,petrek2006caver}.
+The resulting tunnels are represented as a sequence of spheres and characterized by their bottleneck, length, curvature, and a list of surrounding residues, which are later used to estimate the interaction possibilities.
+The main disadvantage of both grid-based and VD-based methods is that the shape of the ligand is not taken into account during the tunnel detection and it is therefore not easy to estimate if (and how) a ligand might traverse the tunnels.
+In order to determine if a ligand can pass the tunnel, it is necessary to compute a trajectory considering the shape of the ligand.
+This can be formulated as a motion planning problem in a high-dimensional configuration space and solved using sampling-based planners.
+
+%Tunnels can be then searched using standard graph-search methods, like Dijkstra's algorithm.
+%Besides, the grid can be used to identify other relevant properties like 
+%pockets, cavities, or channels~\cite{sehnal2013mole,petrek2006caver}.
 %One of the first grid-based approaches to the detection of tunnels in protein is the CAVER 1.0 algorithm by Pet\v{r}ek et al.~\cite{citeulike:6257975}.
-The obvious disadvantage of the grid-based methods is the high memory demand and their dependency on the grid resolution.
+%The obvious disadvantage of the grid-based methods is the high memory demand and their dependency on the grid resolution.
 %Due to the high memory consumption, these methods are not suitable for tunnel detection in dynamic proteins and therefore they
 %are used primarily for analysis of static molecules or individual snapshots of molecular dynamics.
-Currently the most widely used approach to tunnel detection is based on ordinary Voronoi diagrams (VD) or Weighted Voronoi Diagrams (WVD). %~\cite{caver3,yaffe2008}.
-The ordinary VD is computed on points representing centers of all atoms, without considering the radii of atoms.
+%Currently the most widely used approach to tunnel detection is based on ordinary Voronoi diagrams (VD) or Weighted Voronoi Diagrams (WVD). %~\cite{caver3,yaffe2008}.
+%The ordinary VD is computed on points representing centers of all atoms, without considering the radii of atoms.
 %This may lead to detection of tunnels with incorrect bottlenecks. % i.e., incorrect radius of a smallest probe that can traverse the tunnels.
-To consider atoms with different radii, the weights of individual points are determined by the van der Walls radii of atoms in WVD.
-An alternative solution is to compute a non-weighted VD on an extended point set, where 
-each atom is approximated by several spheres with a small radius~\cite{yaffe2008,caver3}.
-VD-based methods are memory less demanding and also faster than the grid-based methods.
+%To consider atoms with different radii, the weights of individual points are determined by the van der Walls radii of atoms in WVD.
+%An alternative solution is to compute a non-weighted VD on an extended point set, where 
+%each atom is approximated by several spheres with a small radius~\cite{yaffe2008,caver3}.
+%VD-based methods are memory less demanding and also faster than the grid-based methods.
 %The extension of VD-based methods to dynamic molecules requires to construct VD in the frames being analyzed and 
@@
 %The existing approaches often use hierarchical clustering to match Voronoi vertices and edges from different frames, which is computationally demanding~\cite{lindow2012dynamic,caverDetails}.
-
-The tunnels detected by the above mentioned approaches are evaluated using basic characteristics like bottleneck, length, curvature, and a list of surrounding
-residues, which are later used to estimate the interaction possibilities.
-The biggest disadvantage of both grid-based and VD-based methods is that the shape of the ligand is not taken into account during the tunnel detection and it is therefore not easy to estimate if (and how) a ligand might traverse the tunnels.
-In order to determine if a ligand can pass the tunnel, it is necessary to compute a trajectory considering the shape of the ligand.
-This can be formulated as a motion planning problem in a high-dimensional configuration space and solved using sampling-based planners.
 
@@
 %Sampling-based motion planning methods can be used to search this high-dimensional configuration space~\cite{Lav06}.
-%The idea of sampling-based motion planning is to randomly sample $\C$ and classify the samples as free or non-free using collision detection.
-%The free samples are stored in a roadmap (a graph structure), in which a path can be searched using standard graph-search methods.
 
@@
 
+The idea of sampling-based motion planning is to randomly sample the configuration space $\C$ and classify the samples as free or non-free using collision detection.
+The free samples are stored in a roadmap (a graph structure), in which a path can be searched using standard graph-search methods.
 Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} is a single-query sampling-based motion planning method that 
@@
 
-Sampling-based planners can provide poor performance in the presence of narrow passages, as many iterations
-are required in order to sample the narrow passages dense enough.
-PRM-based techniques solve the problem by increasing probablity of sampling in regions close to obstacles~\cite{hsuBridge,overmarsGauss}.
-These regions can be estimated e.g. based on amount of collision-free samples around a given sample~\cite{overmarsGauss} or e.g. based
-on workspace information represented by medial axis~\cite{amatoOBRRT,amato2002using}.
+The well known issue of the sampling-based planners is the narrow passage problem.
+%Sampling-based planners can provide poor performance in the presence of narrow passages, as many iterations
+%are required in order to sample the narrow passages dense enough.
+The performance of PRM-based planners can be improved  by increasing probablity of sampling in regions close to obstacles. 
+These regions can be estimated using workspace information such as by medial axis~\cite{amatoOBRRT,amato2002using,wilmarthMAPRM} 
+or they can be identified e.g. based on number of collision-free samples around a given configurations~\cite{overmarsGauss,hsuBridge}.
+%.g. based on amount of collision-free samples around a given sample~\cite{overmarsGauss} or e.g. based
+%~\cite{hsuBridge,overmarsGauss}.
 
-%For example, in~\cite{hsuBridge}, a random sample is classified as..
 %In the Gaussian sampling strategy~\cite{overmarsGauss}, the samples are generated more frequently around obstacles.
@@
 Instead, the random samples has to be generated in the dififcult regions only if the tree can approach them.
-This is the main idea of the guided sampling~\cite{vonasek2009rrt,denny2014marrt,denny2016dynamic}.
+This is the main idea of the guided sampling~\cite{vonasek2009rrt,denny2014marrt,denny2016dynamic}, where a the configurations
+space is sampled with higher probability around a given list of configurations.
+These guiding configurations can be computed e.g. as a path in the workspace.
+Authors of~\cite{kardossRRTKK} proposed to place the guiding configurations between difficult and easy regions of the configuration space, but
+they did not proposed any strategy to find the waypoints.
 
 Another techniques to sample the narrow passages dense enough is to retract the random samples towards regions that are believed
-to be located in narrow passages.
+to be located in narrow passages~\cite{zhangRetraction,lee2012srrrt}.
 In the Retraction-based RRT~\cite{zhangRetraction}, the random sample $\qrand$ is connected to the tree if the line segment $\qnear,\qrand$ is collision-free, otherwise, the retraction-step is performed.
 The task of the retraction step is to find a contact configuration  that minimizes the distance to $\qrand$.
-It is necessary to use an apporpriated distance metric in the configuration space.
-Generally, it is hard to combine the translational and rotational components in the configuration space.
-
-%In~\cite{zhangRetraction}, the DISP metric~\cite{zhang2008fast}.
-
-\cite{saha2005finding}
-
-DISP metric 
-\cite{zhang2008fast}
-
-Motion planning for flexible ligands in protein tunnels brings two main issues: 
-the ligand flexibility increases the dimension of the configuration space, and the necessity to plan in protein tunnels
-leads to the narrow passage problem.
-A narrow passage is a region in the configuration space whose removal changes the connectivity of the free space~\cite{hannaWIS}.
-Narrow passages have smaller volume than other regions and it is therefore difficult to sample them dense enough using uniform distribution, that is used in the basic sampling-based planners.
-The presence of narrow passages in the configuration space, especially if they contain part of the solution, requires many iterations
-in order to put enough samples there and it consequently increases the planning time.
-Typical tunnels have bottlenecks smaller than $1.0$~\AA, so they can already be considered as narrow passages for ligands with more than two atoms.
-%The protein tunnels have the properties of the narrow passages, as they are very narrow
-To cope with the narrow passages, they have to be sampled more dense.
-For the family of RRT planners, it is useful to change the distribution of random samples according to the growth of the tree~\cite{vonasek2009rrt,vonasekphd,denny2016dynamic}.
-%about their 3D position, which is represented by the tunnel centerline.
-%The centerline can be used to guide the growth of the RRT tree through the configuration space~\cite{vonasek2009rrt,denny2016dynamic}.
-
-%The additional degrees of freedom required to model the flexibility increase the dimension of the configuration space.
-The kinematic chain representation used to model the ligand flexibility can generate all possible conformations but it also increases the dimension of the configuration space.
-Not all of conformations are however feasible and it is therefore necessary to verify the feasibility of a given conformation based on energy, which is time consuming.
-%RRT-based planners are sensitive to the employed metric, that should consider also the flexibility-related DOFs.
-An alternative solution is to employ a library of known conformations.
-The conformation changes of protein amino acids and ligand are represented by so called rotamers, stored in different rotamer libraries (e.g., Dunbrack library~\cite{dunbrack}).
-%Rotamers differ in the angles between their atoms.
-Possible rotamer conformations correspond to energetically and geometrically favorable positions.
-%This solution is common also in different types of calculations, e.g.~\cite{kellogg}.
+The retraction step is computationally intensive, as the computation of the closest contact point leads to computation
+of the generalized penetration depth, which has high complexity~\cite{zhang2008fast}.
+Therefore, many algorithms uses a heuristic to compute the samples near boundary of the configuration space~\cite{amatoOBPRM,amatoOBRRT}.
+The probability of sampling of the narrow passages can be increased by dilating the free-space, e.g. by shrinking the geometry of 
+the robot~\cite{hsuOnProb} or the obstacles~\cite{bayazitIRC}.
 
@@
 
-In this paper, we propose to analyze the traversability separately for each tunnel.
-This can bring several advantages for the biochemists.
-The tunnels in a protein being studied can be computed using standard tunnel detection tools for spherical probes and assessed
-according to their length, curvature, bottleneck, and biochemical properties (e.g., partial charge).
-These characteristics are already used in many research studies.
-Selected tunnels can further be analyzed with a given ligand to determine the traversability.
-According to the found trajectories, researches may decided of the given ligand can reach the active site, which helps to 
-organize subsequent MD simulations or other experiments.
+In this paper, we propose a novel modification of the RRT planner for computing trajectories insider protein tunnels.
+To cope with the narrow passage problem, the random samples are generated mainly around a tunnel computed using Voronoi diagrams and 
+not in the whole conifguration space as in~\cite{cortes2010simulating}.
+To further increase the probability of sampling inside the narrow passage, the ligand atoms are shrunk, which is similar
+to~\cite{cortes2010simulating}.
+The ligand flexibility is modeled using a predefied set of typical conformations, therefore the number of dimensions of the configuration space does not depend on the DOF of the ligand as in~\cite{cortes2010simulating}.
+Library of conformations are used also in different types of calculations, e.g.~\cite{kellogg}. 
+While classic 6D Euclidean metric is used in the nearest-neighbor search of RRT, we propose to use another metric during the expansion step.
+This metric measures distances between two configurations as the nearest distance of two atoms. 
+This enables the metric to consider the shape of the ligand and it support retraction of the ligand to the narrow passages.
 
-%It is not necessary to first run time consuming MD simulations with both protein and ligand being studied.
-%but it is suitable to detected tunnels in the protein only.
+%Using a library of predefined conformations has been used also in other tasks, e.g.~\cite{kellogg}.
+%This shrinking technique is also used in the MoMa-LigPath tool~\cite{cortes2005path}, which utilizes RRT to find pathways for unbinding a ligand  from a protein.
+%It is necessary to use an apporpriated distance metric in the configuration space.
+%Generally, it is hard to combine the translational and rotational components in the configuration space.
+%In~\cite{zhangRetraction}, the DISP metric~\cite{zhang2008fast}.
+%energetic constraints are translated into geometric ones by considering a steric model of the molecules and applying collision detection
+%\cite{deAngulo2005biocd}
+%First the biredge test is applied to identify regions aeround narrow passages, optimization-based retraction operatoin sleectively only at those regions selective rrt~\cite{lee2012srrrt}
+%Free-space dilation is an effective aproach for narrow passage sampling~\cite{hsu06multilevel}. 
+%It is necessary to define proper method for dilating and determin the amount of dilation.
+%\cite{saha2005finding}
+%DISP metric \cite{zhang2008fast}
 
-The proposed work employs the RRT planning principle with three extensions: 
-a) the configuration space is not sampled uniformly, but uses a guided sampling along the tunnel centerline,
-b) the radii of the ligand atoms can be decreased, and 
-c) ligand flexibility is modeled using a library of predefined conformations.
-The first two extensions are designed to cope with the narrow passage problem.
-By reducing the scale of the atomic radii, the ligand can move more without collisions, which widens the narrow passages and increases
-the probability of placing samples into them.
-The scaling-down technique have been used e.g. for motion planning of deformable objects~\cite{gayle2005path,alterovitz2008motion}, and
-they are also used in the most related MoMa-LigPath tool~\cite{cortes2005path}.
-By using the predefined set of conformations, the number of dimensions of the configuration space does not depend on the DOF of the ligand.
-Library of conformations are used also in different types of calculations, e.g.~\cite{kellogg}. 
-%Using a library of predefined conformations has been used also in other tasks, e.g.~\cite{kellogg}.
+%Motion planning for flexible ligands in protein tunnels brings two main issues: 
+%the ligand flexibility increases the dimension of the configuration space, and the necessity to plan in protein tunnels
+%leads to the narrow passage problem.
+%A narrow passage is a region in the configuration space whose removal changes the connectivity of the free space~\cite{hannaWIS}.
+%Narrow passages have smaller volume than other regions and it is therefore difficult to sample them dense enough using uniform distribution, that is used in the basic sampling-based planners.
+%The presence of narrow passages in the configuration space, especially if they contain part of the solution, requires many iterations
+%in order to put enough samples there and it consequently increases the planning time.
+%Typical tunnels have bottlenecks smaller than $1.0$~\AA, so they can already be considered as narrow passages for ligands with more than two atoms.
+%%The protein tunnels have the properties of the narrow passages, as they are very narrow
+%To cope with the narrow passages, they have to be sampled more dense.
+%For the family of RRT planners, it is useful to change the distribution of random samples according to the growth of the tree~\cite{vonasek2009rrt,vonasekphd,denny2016dynamic}.
+%%about their 3D position, which is represented by the tunnel centerline.
+%%The centerline can be used to guide the growth of the RRT tree through the configuration space~\cite{vonasek2009rrt,denny2016dynamic}.
+%
+%%The additional degrees of freedom required to model the flexibility increase the dimension of the configuration space.
+%The kinematic chain representation used to model the ligand flexibility can generate all possible conformations but it also increases the dimension of the configuration space.
+%Not all of conformations are however feasible and it is therefore necessary to verify the feasibility of a given conformation based on energy, which is time consuming.
+%%RRT-based planners are sensitive to the employed metric, that should consider also the flexibility-related DOFs.
+%An alternative solution is to employ a library of known conformations.
+%The conformation changes of protein amino acids and ligand are represented by so called rotamers, stored in different rotamer libraries (e.g., Dunbrack library~\cite{dunbrack}).
+%%Rotamers differ in the angles between their atoms.
+%Possible rotamer conformations correspond to energetically and geometrically favorable positions.
+%%This solution is common also in different types of calculations, e.g.~\cite{kellogg}.
 
@@
 
-%and the third modification
-%is introduced in order to decrease the dimension of the configuration space.
-
-%Ligands are small molecules interacting with the protein. 
 %Due to the interaction, the relative positions of the ligand's atoms can change.

### 2017-09-04 paper.tex
experiments reorganized

@@
 \begin{abstract}
-Chemical interactions of proteins and other molecules (ligands) take place in actives sites that are usually deeply burid inside the proteins.
-The active site are accesible throu one or more tunnels (pathways), which have been traditionally computed using Voronoi diagrams.
-The Voronoi-based tunnel computation is however not suitable for ligands of arbitrary shapes neither for flexible
-ligands, which can changes their conformation.
-In this paper, we propose to analyze the accessibility of the active sites using sampling-based motion planning.
-Thing brings three challenges: how to find path in a dense protein structures, where the tunnels behave like narrow passages,
-how to model the conformation changes of the ligands, and how to measure distance in the related configuration space.
-To cope with the narrow passage problem, we generate the random samples areound a Voronoi-based tunnel and we shrink
-tha atomic radii of the ligand.
+Chemical interactions of proteins and other molecules (ligands) take place in actives sites that are usually deeply buried inside the proteins.
+The active sites are accessible through one or more pathways (called tunnels).
+The knowledge about a tunnel (e.g. it's bottleneck or physico-chemical properties) can help chemists to estimate possible interactions
+between the protein and the ligand.
+The tunnels have been traditionally computed using Voronoi diagrams, which is suitable only for small, spherical and rigid ligands.
+We propose to analyze the accessibility of the active sites using sampling-based motion planning.
+This requires to cope with narrow passages and to model the flexibility of the ligands.
+To cope with the narrow passage problem, we generate the random samples around a Voronoi-based tunnel and we shrink
+the atomic radii of the ligand.
 The flexibility of the ligand is modeled using a fixed set of known conformations.
-The distance between the configurations is measured using the nearest distance between two atoms.
 \end{abstract}
@@
 A path in the roadmap then corresponds to a motion in the workspace.
+As the random samples are classified as free or non-free using collision detection, the sampling-based methods can found paths for robots of various shapes. 
+Another advantage of the sampling-based planners is that they can cope with robots with many degrees of freedom (DOF).
 The two most used sampling-based techniques are Probabilistic Roadmaps (PRM)~\cite{kavrakiForPP} and Rapidly Exploring Random Trees (RRT)~\cite{lavalleRRT}.
-Originally, both samples utilize uniform distribution to sample the configuration space.
-As the random samples are classify as free or non-free using collision detection, the sampling-based methods can found paths for robots of various shapes. 
-Another important advantage of the sampling-based planners is that they can cope with robots with many degrees of freedom (DOF).
-Thanks to these features, sampling-based planners has been utilized in wide range of applications in robotics, but also in 
+These planners and their variants have been utilized in wide range of applications in robotics, but also in 
 other research domain, namely in bio-chemistry 
@@
 The problem of tunnel detection is to find a tunnel (pathways) leading to a specific place (called active site) inside the protein.
-The knowledge about the existence of the tunnels and their properties (e.g., their bottlenect and chemical properties around the tunnel),
+The knowledge about the existence of the tunnels and their properties (e.g., their bottleneck and chemical properties around the tunnel),
 is important to understand interactions between proteins and other molecules (ligands).
 To get the ligand to the active site, there has to be a transportation path connecting the protein outer environment with the active site.
-These paths, called tunnels, have to be wide enough and the physico-chemical properties of the surrounding amino acids have to be compatible with the ligand (Fig.~\ref{fig::motiv}).
+The tunnels have to be wide enough and the physico-chemical properties of the surrounding amino acids have to be compatible with the ligand (Fig.~\ref{fig::motiv}).
 %These interactions are highly influenced by the ability to transport the small molecule, called ligand, to the protein active site.
@@
 a spherical ligand (probe) and they are represented as a sequence of spheres.
-However, ligands typically have non-spherical shape and it is therefore difficult to estimate their traversability through
-the Voronoi-based tunnels.
+However, as the ligands typically non-spherical and flexible,  it is difficult to estimate their traversability through the tunnels computed for only one sphere.
 
@@
 
-To provide biochemists  a better insight into the behavior of non-spherical ligands, it is necessary to compute the trajectories of the ligand considering its shape and possibly also its conformation changes.
-This can be formulated as the path planning problem where the ligand is considered as a robot moving amongst protein's atoms.
-Sampling-based planners are suitable candidates to solve this problem due to their ability to cope with many-DOF problems and objects with arbitrary shapes.
+To provide biochemists  a better insight into the behavior of non-spherical ligands, it is necessary to compute the trajectories of the ligand considering conformation changes.
+The computation of trajectories for such ligands can be formulated as a motion planning problem.
+RRT is a suitable candidate to solve this problem due to its ability to cope with many-DOF robots of arbitrary shapes.
+This motion planning however require to cope with the narrow passage problem, as the proteins are dense structures and the
+movements of the ligands are limited.
+A narrow passage is a small region in the configuration space whose removal changes the connectivity of the space.
+Due to its low volumes, the probability of sampling the narrow passages is low.
 
-The well known issues of the sampling-based planners are the narrow passage problem and the sensitiviy to distance metric in the configuration space.
-A narrow passage is a small region in the configuration space whose removal changes the connectivity of the space.
-Due to their low volumes, the probability of sampling the narrow passages is low.
 Consequently, many iterations are needed in order to put enough samples into the passages, which increases the computation time.
@@
 
-To cope with these issues, we propose to guide the sampling of the configuration space using the Voronoi-baed tunnels.
+To cope with these issues, we propose to guide the sampling of the configuration space using the Voronoi-based tunnels.
 The random samples are not generated from the whole configuration space, but mainly around the tunnel.
-To increse movements of the ligands in the narrow tunnels, the free-space is dilated by 
+To increase movements of the ligands in the narrow tunnels, the free-space is dilated by 
 shrinking the atom radii, similarly, e.g., to~\cite{cortes2005path,hsu06multilevel}.
-The ligand can be considered as a multi-link kinematic chain and by chaning the torsion angles, the flexibility can be modeled~\cite{songPFpath,cortes2005path}.
+The ligand can be considered as a multi-link kinematic chain and by changing the torsion angles, the flexibility can be modeled~\cite{songPFpath,cortes2005path}.
 This however increases the dimension of the configuration space.
@@
 
+TODO some words about MD and why tunnels are predered
+
 The analysis of ligand traversability to/from an active site is usually based on so called tunnels.
@@
 In order to determine if a ligand can pass the tunnel, it is necessary to compute a trajectory considering the shape of the ligand.
-This can be formulated as a motion planning problem in a high-dimensional configuration space and solved using sampling-based planners.
+This can be formulated as a motion planning problem in a high-dimensional configuration space $\C$ and solved using sampling-based planners.
+In this case, the configuration space $\C$ has at least $6+n$ dimensions (3D rotation + 3D translation + additional degrees of freedom
+caused by the ligand flexibility).
 
@@
 %However, not all conformations are feasible, which has to be checked using potential energy, which is time consuming.
+%The idea of sampling-based motion planning is to randomly sample the configuration space $\C$ and classify the samples as free or non-free using collision detection.
+%The free samples are stored in a roadmap (a graph structure), in which a path can be searched using standard graph-search methods.
 
-The idea of sampling-based motion planning is to randomly sample the configuration space $\C$ and classify the samples as free or non-free using collision detection.
-The free samples are stored in a roadmap (a graph structure), in which a path can be searched using standard graph-search methods.
 Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} is a single-query sampling-based motion planning method that 
@@
 %are required in order to sample the narrow passages dense enough.
-The performance of PRM-based planners can be improved  by increasing probablity of sampling in regions close to obstacles. 
+PRM-based planners can cope the narrow passage problem by increasing probability of sampling in given regions.
 These regions can be estimated using workspace information such as by medial axis~\cite{amatoOBRRT,amato2002using,wilmarthMAPRM} 
 or they can be identified e.g. based on number of collision-free samples around a given configurations~\cite{overmarsGauss,hsuBridge}.
+The narrow passage are however located in the configuration space and it is generally not easy to estimate their location
+based only on the features of the workspace~\cite{hannaWIS}.
 %.g. based on amount of collision-free samples around a given sample~\cite{overmarsGauss} or e.g. based
@@
 RRT-based planners cannot cope with the narrow passage problem simply by increasing the probability of sampling in them, as the
-growth of the configuration tree may be toward the random samples may be blocked by an obstacle~\cite{vonasekphd}.
-Instead, the random samples has to be generated in the dififcult regions only if the tree can approach them.
+growth of the configuration tree may be toward the random samples may be blocked by obstacles~\cite{vonasekphd}.
+Instead, the random samples has to be generated in the difficult regions only if the tree can approach them.
 This is the main idea of the guided sampling~\cite{vonasek2009rrt,denny2014marrt,denny2016dynamic}, where a the configurations
@@
 
-Another techniques to sample the narrow passages dense enough is to retract the random samples towards regions that are believed
-to be located in narrow passages~\cite{zhangRetraction,lee2012srrrt}.
-In the Retraction-based RRT~\cite{zhangRetraction}, the random sample $\qrand$ is connected to the tree if the line segment $\qnear,\qrand$ is collision-free, otherwise, the retraction-step is performed.
-The task of the retraction step is to find a contact configuration  that minimizes the distance to $\qrand$.
+Besides increasing probability of sampling in certain regions, the growth of the tree towards the regions can be reached by retracting
+the sampling to the regions.
+%Another techniques to sample the narrow passages dense enough is to retract the random samples towards regions that are believed
+%to be located in narrow passages~\cite{zhangRetraction,lee2012srrrt}.
+In Retraction-RRT~\cite{zhangRetraction}, the random sample $\qrand$ is connected to the tree if the line segment $\qnear,\qrand$ is collision-free, otherwise, the retraction-step is performed.
+The task of the retraction step is to find a contact configuration around $\qnear$ that minimizes the distance to $\qrand$.
 The retraction step is computationally intensive, as the computation of the closest contact point leads to computation
@@
 
-A RRT-based method for computing exit pathways for a small flexible molecule was presented in~\cite{cortes2010simulating}.
-To cope with many-DOF ligands, the RRT-ML variant~\cite{cortes2007mlrrt} was employed.
+The shrinking technique has been used also in~\cite{cortes2010simulating}, where exit pathways for a small flexible molecule are
+computed using a RRT-based method.
+To cope with many-DOF ligands, the RRT-ML variant~\cite{cortes2007mlrrt} is employed as the basic planner.
 RRT-ML expands the tree primarily using those DOF that are essential for achieving motion of the ligand (i.e., rotation
 and translation) and it employs the other DOFs (i.e., that are responsible for conformation changes) if they hinder the growth of the tree.
-The approach~\cite{cortes2010simulating} may however suffer from the narrow passage problem, as it aims to find some exit pathway for a given ligand, which requires to search the whole configuration space of the ligand/protein complex.
-Although there are usually multiple pathways from a given active site, not all of them are traversable by a given ligand.
-Many iterations are needed to find a solution, which increases the computational time.
 
-In this paper, we propose a novel modification of the RRT planner for computing trajectories insider protein tunnels.
-To cope with the narrow passage problem, the random samples are generated mainly around a tunnel computed using Voronoi diagrams and 
-not in the whole conifguration space as in~\cite{cortes2010simulating}.
-To further increase the probability of sampling inside the narrow passage, the ligand atoms are shrunk, which is similar
-to~\cite{cortes2010simulating}.
-The ligand flexibility is modeled using a predefied set of typical conformations, therefore the number of dimensions of the configuration space does not depend on the DOF of the ligand as in~\cite{cortes2010simulating}.
+\section{Proposed method}
+
+In this paper, we propose a modification of the RRT planner for computing trajectories through protein tunnels.
+The input is a set of tunnels computed by Voronoi diagrams and the task is to find trajectories inside (or along) the tunnels
+leading from outside the protein to the active site.
+
+Contrary to~\cite{cortes2010simulating}, which find any exit pathway, we propose to compute the trajectories only around/inside each tunnel
+separately.
+This allows us to decrease the volume of the configuration space to be search, which increases the relative volume of the narrow passages.
+To further cope with the narrow passage problem, the random samples are not generated in the whole configuration space, but only (mainly) around
+the tunnel in a user-defined distance.
+Therefore, the tunnels serve as a guiding path for the configuration tree~\cite{vonasek2009rrt}.
+
+As the input data are snapshots of the proteins without ligands, it can be expected, that the tunnels are narrow therefore it would be difficult
+to find trajectories for all ligands.
+In reality, the tunnels adapt to the ligands (and vice versa), so even narrow tunnels can serve as transportation path for ligands larger than
+the tunnel bottleneck TODO CITACE??.
+To emulate this behavior, we shrink the atom radii of the ligand, similarly to~\cite{cortes2010simulating}.
+
+The ligand flexibility is modeled using a predefined set of typical conformations, therefore the number of dimensions of the configuration space does not depend on the DOF of the ligand as in~\cite{cortes2010simulating}.
 Library of conformations are used also in different types of calculations, e.g.~\cite{kellogg}. 
@@
 
-%Using a library of predefined conformations has been used also in other tasks, e.g.~\cite{kellogg}.
-%This shrinking technique is also used in the MoMa-LigPath tool~\cite{cortes2005path}, which utilizes RRT to find pathways for unbinding a ligand  from a protein.
-%It is necessary to use an apporpriated distance metric in the configuration space.
-%Generally, it is hard to combine the translational and rotational components in the configuration space.
-%In~\cite{zhangRetraction}, the DISP metric~\cite{zhang2008fast}.
-%energetic constraints are translated into geometric ones by considering a steric model of the molecules and applying collision detection
-%\cite{deAngulo2005biocd}
-%First the biredge test is applied to identify regions aeround narrow passages, optimization-based retraction operatoin sleectively only at those regions selective rrt~\cite{lee2012srrrt}
-%Free-space dilation is an effective aproach for narrow passage sampling~\cite{hsu06multilevel}. 
-%It is necessary to define proper method for dilating and determin the amount of dilation.
-%\cite{saha2005finding}
-%DISP metric \cite{zhang2008fast}
+%approximate the distance metric using nonlinear parametric regression model 
+%\cite{palmieri2015Distance}
+%true cost-to-go function implies solving two-point boundary problem which can be as expensive as solving the planning query itself.
 
@@
 
-\subsection{Traversability Characteristics}
-
-For each initial configuration $\qinit \in \QI$, $K$ trajectories are computed, which results in the set of $K |\QI|$ trajectories.
-All these trajectories are used to compute the following properties of the tunnel $T$.
-A trajectory $P$ reaches the tunnel sphere $c_i \in T$, if the 3D Euclidean distance of the 
-nearest configuration $q \in P$ towards $c_i$ is less than $r_i$ (radius of the sphere $c_i$).
-Let $N(i)$ denote the number of trajectories that reached $i-$th sphere of the tunnel, $i=1,\ldots,n$.
-Three basic characteristics of the tunnel are computed from the trajectories: accessibility, throughput, and the scaling factor.
-
-The accessibility $A(i)=N(i)/N$ of the sphere $i$ is the probability of reaching the sphere $i$, where $N=K|\QI|$ is the total number of trajectories.
-The accessibility shows how probable it is to pass the tunnel up to the sphere $i$.
-Obviously, the most important is $A(n)$ of the last sphere of the tunnel, which can be considered as the overall difficulty of the tunnel.
-The  ligand passage may however be strongly affected by the first bottleneck, so the parts of the tunnel located behind
-the bottleneck have low accessibility.
-
-%To better measure the probability of passing a given tunnel part, we should take into account only trajectories that even have chance
-%to visit this part, e.g., the trajectories that reached the preceding tunnel part.
-The throughput $T(i)$ is the ratio of trajectories that passed sphere $i$ (i.e., visited sphere $i+1$) and reached the sphere $i$,
-i.e., $T(i) = N(i+1) / N(i)$. 
-The throughput is not computed for the last sphere ($i=N$).
-The throughput shows a local accessibility of tunnel parts and it can be used to detect places where most of the trajectories ends.
-
-%The probability of the ligand that it reaches a tunnel's part.
-%We employ the tunnel representation from CAVER 3.0 and we define a tunnel part as a sphere $P = (x, r)$ where $x \in \mathbb{R}^3$ is sphere center and $r \in \R$ is its radius.
-%Then, for each trajectory $t$ we define its ending tunnel part $P_t^e$ as the part whose center is closest to the end position of $t$.
-%Finally, the accessibility of $P$ is defined as a ratio of trajectories that passed $P$ or ended in $P$ over all trajectories.
-%On the other hand, the probability does not show any details about the ligand's passage, e.g., how complicated it is to pass through some part of the tunnel, which trajectory is the most used one, etc.
-%Moreover, it is not feasible to directly visualize and explore the resulting trajectories due to the visual clutter (see Fig.~\ref{fig:trajectories} left) caused by their high number and close proximity.
-
-The proposed planner is allowed to scale down the radii of ligand spheres up to the allowed scale $\smin$. 
-It can be expected that narrower parts of the tunnel are more often passed with a more scaled-down ligand than
-the wider parts.
-Ligand scale $L(i)$ at the tunnel sphere $i$ is the average scale of ligand that reaches the sphere $i$.
-%The ligand scale is computed as the average value of scales of the nearest configurations of the trajectories towards tunnel sphere~$i$.
-
-%--- The scale of the ligand when it passes a given tunnel's part.
-%For the trajectory $t$, the ligand scale $s(P_t)$ in the tunnel part $P_t$ is defined as the scale of such sample of $t$ whose position is the closest one to the center of $P_t$.
-%Then, the average ligand scale is defined as $s(P) = \frac{1}{|T|} \sum_{t \in T} s(P_t)$.
-%Similarly, the minimum and maximum ligand scales can be defined.
-
 \subsection{Visualization of Results}
@@
 
-Besides the color mapping, it is also necessary to show the computed trajectories.
-%This is useful, for example, to find out in which part of the tunnel the trajectories detour.
-Simple visualization of all trajectories could however be too slow for an interactive work.
-Therefore, the trajectories are first clustered and then only the clusters are visualized.
-Due to different lengths of the trajectories, they are first converted to a normalized form.
-Let $\cstart$ represent the average starting position of all trajectories and let $d_{max}$ be the 3D Euclidean distance
-of the most distant configuration from $\cstart$ among all trajectories.
-A set of $M$ spheres centered at $\cstart$ are created with radii $r'_i=i {d_{max} \over M}$, where $i=0,\ldots, M-1$.
-The trajectory $P=(q_1,\ldots,q_n)$ of length $n$ is represented by the normalized vector $v=(x_1,\ldots,x_M)$ of length $M$,
-where $x_i$ is the 3D position of the nearest configuration $q \in P$ to the surface of the $i$-th sphere with the radius $r'_i$.
-The distance between two normalized trajectories $v_i$ and $v_j$ is defined as
-$d(v_i, v_j) = \frac{1}{M} \sum_{1 \leq k \leq N} |x_k^i x_k^j|$.
-This distance is used in the UPGMA clustering technique~\cite{sokal1958statistical}.
-The trajectories can be visualized using a representative of each cluster.
-The number of trajectories in each cluster is represented by the width of the polyline (Fig.~\ref{fig:trajectories}).
-%In this manner, we are able to convey the information about all different trajectories together (see Fig.~\ref{fig:trajectories} right).
-
-\begin{figure}
-\centering
-\begin{tabular}{cc}
-\includegraphics[width=0.4\textwidth]{fig/trajectories-all} &
-\includegraphics[width=0.4\textwidth]{fig/trajectories-clustered-21} \\
-(a) & (b) \\                       
-\end{tabular}
-\caption{Visualization of the trajectories. The tunnel begins in the top left corner.
-(a) All trajectories ($\sim5000$) colored according to whether they reached the end of the tunnel (green) or not (red). 
-(b) Visualization using clusters of trajectories. 
-%Trajectories clustered according to their proximity colored by distinct colors.
-%The different line widths represent the number of trajectories in a cluster.
-\label{fig:trajectories}
-}
-\end{figure}
-
 \section{Experimental Verification}
 
-The tunnels in Haloalkane dehalogenase protein (PDB ID 1CQW) has been analyzed using the proposed approach.
-Three tunnels were detected using CAVER 3.0~\cite{caver3} for the spherical probe of radius 0.9~\AA (Fig.~\ref{fig::tunnel}a).
-The traversability was evaluated for 1-chlorpropan (denoted as $\LA$) with 11 atoms and 
+The proposed approach was used to analyze tunnel traversability towards to active site of Haloalkane dehalogenase protein (PDB ID 1CQW).
+The active site was defined using residues 38 (Asparagine), 102 (Histidine) and 103 (Aspartate).
+Molecular dynamics of the protein was computed using TODO. 
+First 100 frames were analyzed. 
+In each frame, tunnels for spherical probe 0.9~\AA were dected using CAVER 3.0~\cite{caver3}. 
+In most of the frames, two tunnels were detected.
+
+%has been analyzed using the proposed approach.
+%Three tunnels were detected using CAVER 3.0~\cite{caver3} for the spherical probe of radius 0.9~\AA (Fig.~\ref{fig::tunnel}a).
+%The traversability was evaluated for 1-chlorpropan (denoted as $\LA$) with 11 atoms and 
 % from David: m003 je "1-chlorpropan" a m004 "1-chlorbutan"
-and 1-chlorbutan (denoted as $\LB$) with 14 atoms.
-$\LA$ was represented by 12 conformations, and $\LB$ by a set of $114$ conformations.
-Examples of three conformations of $\LA$ and $\LB$ are depicted in Fig.~\ref{fig::tunnel}b.
-Further information about the experiments can be found at %{\url{http://mrs.felk.cvut.cz/isrr2017}}.
+%and 1-chlorbutan (denoted as $\LB$) with 14 atoms.
+%$\LA$ was represented by 12 conformations, and $\LB$ by a set of $114$ conformations.
+%Examples of three conformations of $\LA$ and $\LB$ are depicted in Fig.~\ref{fig::tunnel}b.
+%Further information about the experiments can be found at %{\url{http://mrs.felk.cvut.cz/isrr2017}}.
 
-Both tested ligands have more than 10 atoms and therefore they cannot fit into the tunnel of with 0.9~\AA, so
-the radii of ligands were scaled down. 
-Four different scaling-down factors were used: $\smin=\{0.3,0.4,0.5,0.6\}$.
-No trajectories were found for $\smin > 0.6$.
-For each scale, 50 collision-free initial configurations were found in the $\RI=2$\AA\ radius around the first sphere of the tunnel.
-For each ligand, each minimal scale $\smin$ and each initial configuration, 100 trajectories were computed using the proposed planner.
+%Both tested ligands have more than 10 atoms and therefore they cannot fit into the tunnel of with 0.9~\AA, so
+%the radii of ligands were scaled down. 
+Four different scaling-down factors were used: $\smin=\{0.4,0.5,0.6,0.7\}$.
+For each scale, 20 collision-free initial configurations were found in the $\RI=2$\AA\ radius around the first sphere of the tunnel (outside the protein).
+For each ligand, each minimal scale $\smin$ and each initial configuration, 10 trajectories were computed using the proposed planner.
 The parameters of the planner were: $\Imax=10,000$, $m=50$, $\gb=0.9$, $\dt=1.5$\AA, $\rv=2$~\AA.
@@
 
-The average runtimes, success rates and sizes of the built trees for the first tunnel are shown in Tab.~\ref{tab::rrt}.
-The highest average success rate was achieved for the most scaled-down ligands ($\smin=0.3$) and it decreases with the increasing $\smin$. 
-The difference between the minimal and maximal success rates indicates how the initial configurations of the ligand influence the ability to pass the tunnels.
-For example, the most difficult initial configuration of $\LA$ and $\smin=0.3$ lead to success rate 78~\% (the easiest 
-        initial configuration lead to success rate $100$~\%), but for the scale $\smin=0.4$, the worst success rate is
-only $42$~\% and the best 96~\%.
-The traversability of larger ligands ($\smin=0.4$ vs. $\smin=0.3$) is therefore more sensitive to the initial configuration.
-This shows the importance of testing the traversability from multiple initial configurations.
-The runtime is significantly higher for the $\LB$ ligand, which is caused by the larger number of conformations that need to be
-examined in each expansion.
-The average values of the runtimes, which are in the order of minutes, are negligible in the comparison of runtime of MD simulations.
-
-%The tree can be expanded by up to $|\L|$ conformations in each iterations, which gives theoretical
-%size of the tree $110\cdot 10^3$ for $\LA$ and $1140\cdot 10^3$ for $\LB$.
-%The constructed tree are however significantly smaller
-
 \def\tmpa{0.13\textwidth}
 
-\begin{figure}
-\centering
-\begin{tabular}{cc}
-%\includegraphics[width=0.38\textwidth]{fig/protein-tunnel-gray} &
-\includegraphics[width=0.3\textwidth]{fig/t05proteintunnels} &
-\includegraphics[width=0.45\textwidth]{fig/ligAll}  \\
-(a) & (b)                        
-\end{tabular}                       
-\caption{\label{fig::tunnel}
-    (a) The protein 1CQW visualized using the cartoon representation (gray) with 
-        three tunnels (red, green, blue) detected by CAVER 3.0.
-        The first tunnel is depicted red.
-%        its highest ranked tunnel (green) which was detected using CAVER 3.0.
-    (b) Examples of three conformations of $\LA$ (top) and $\LB$ (bottom)
-}
-\end{figure}
+%\begin{figure}
+%\centering
+%\begin{tabular}{cc}
+%\includegraphics[width=0.3\textwidth]{fig/t05proteintunnels} &
+%\includegraphics[width=0.45\textwidth]{fig/ligAll}  \\
+%(a) & (b)                        
+%\end{tabular}                       
+%\caption{\label{fig::tunnel}
+%    (a) The protein 1CQW visualized using the cartoon representation (gray) with 
+%        three tunnels (red, green, blue) detected by CAVER 3.0.
+%        The first tunnel is depicted red.
+%%        its highest ranked tunnel (green) which was detected using CAVER 3.0.
+%    (b) Examples of three conformations of $\LA$ (top) and $\LB$ (bottom)
+%}
+%\end{figure}
+%
 
@@
 Despite the low bottleneck of this tunnel, the ligand may reach its end using the alternative pathway.
+
+\section{Discussion}
+
+The computed trajectories cannot be considered as `real', they should rather be considered as a hint for the biochemists.
+One of the reason is that the trajectories are computed inside a static protein and static tunnels.
+Real proteins are dynamic structures and consequently also the tunnels are dynamic: they move, merge or even disappear due to motions of protein atoms.
+Contrary, usually static tunnels are used in protein engineering.
+Despite such a simplification, researches often used information about the static tunnels  to estimate the possibility of chemical reactions
+at the active sites.
+
+The second limitation is caused by the utilization of scaled-down ligands.
+The protein atoms fill the internal void space therefore the tunnels identified inside proteins without ligand tend
+to be very narrow.
+To enable at least some motion of the ligands inside such narrow tunnels, the scaling-down of the ligand atoms is necessary.
+It is used also in other related tool, e.g.~\cite{cortes2005path}.
+
+Due to these reasons, the computed trajectories can be considered either as too optimistic
+(e.g. because they are computed on a wide tunnel that can be in fact closed due to molecular dynamics),
+or too pessimistic (e.g. the trajectory is not found because of narrow passage around the tunnel, which can be 
+opened due to the molecular dynamics).
+Despite these limitations, testing the tunnel traversability using motion planning technique can provide chemists more information
+than the simple bottleneck radius, which is used nowadays.
+For example, possible detours from the tunnel can be identified, which indicates that even a tunnel with a small
+bottleneck can be traversed (Fig.~\ref{fig::detail}).
+
+%Most related works, e.g. those based on motion planning, are usually focused on protein folding, or finding arbitrary pathways
+%from proteins~\cite{cortes2010simulating,guieysse2008structure}.
+
+The proposed planner is very simple and it can be improved in many ways: by using a less aggressive scaling-down technique (in order
+to obtain better trajectories) or by using only a subset of conformations and a better expansion (in order to speed it up).
+Having a better motion planner (in the term of the speed or even success rate) however does not automatically mean that the resulting trajectories will help the biochemists more.
+The presented work shows possible advantages of the analysis based on ligand trajectories.
+By comparing the planned trajectories with trajectories observed in MD simulations, thresholds for the success rate or minimal tunnel throughput can be set, which will help biochemists to make decisions.
+
+%The runtime of the planner is currently given mainly by the collision-detection.
+%As the ligand is assumed to travel only along the tunnel centerline (in the distance defined by $\rv$),
+%only atoms located in the vicinity of the tunnel can be considered for collision detection.
+%Fast collision-detection can be achieved either using fast hierarchical data structures as OBB (implemented, e.g., in~\cite{ozcollide})
+%    or using collision-detection system developed directly for proteins~\cite{ruiz2005biocd}.
+%We assume that a conformation $l \in \L$ can be switched to all other conformations in $\L$.
+
+%parmas:
+%$\RI$
+%$\rv$
+%$\dt$
+%The presented expansion procedure favors the samples with maximal scale (they are tested first), the trajectory contains samples with higher scales at the easy part of the tunnels (e.i., wide areas), and the smaller scales are used only if the ligand has to fit into a narrow space.
+
+\section{Conclusion }
+The protein tunnels are transporting pathways for ligands to the  active sites.
+In this paper, we propose to analyze the traversability in the tunnels using motion planning, that can consider both shape
+of the protein and ligand.
+We proposed modifications to the RRT method to compute trajectories for a ligand represented by a library of common conformations.
+The corresponding configuration space is sampled using the guided sampling, where a virtual sphere moves along the tunnel and attracts growth
+of the tree towards it.
+To enable trajectory computation in narrow tunnels, atomic radii of the ligand are scaled down.
+Computed trajectories allows us to evaluate the traversability of the tunnels using
+accessibility, throughput and also by visualizing typical pathways found by the planner.
+These properties can help to better understand importance of the tunnels.
+
+\bibliographystyle{plain}
+\bibliography{paper}
+
+\end{document}
+
+% ============  unused ideas =========================
+
+%neni treba \cite{cheng01reducing}
+%Using a library of predefined conformations has been used also in other tasks, e.g.~\cite{kellogg}.
+%This shrinking technique is also used in the MoMa-LigPath tool~\cite{cortes2005path}, which utilizes RRT to find pathways for unbinding a ligand  from a protein.
+%It is necessary to use an apporpriated distance metric in the configuration space.
+%Generally, it is hard to combine the translational and rotational components in the configuration space.
+%In~\cite{zhangRetraction}, the DISP metric~\cite{zhang2008fast}.
+%energetic constraints are translated into geometric ones by considering a steric model of the molecules and applying collision detection
+%\cite{deAngulo2005biocd}
+%First the biredge test is applied to identify regions aeround narrow passages, optimization-based retraction operatoin sleectively only at those regions selective rrt~\cite{lee2012srrrt}
+%Free-space dilation is an effective aproach for narrow passage sampling~\cite{hsu06multilevel}. 
+%It is necessary to define proper method for dilating and determin the amount of dilation.
+%\cite{saha2005finding}
+%DISP metric \cite{zhang2008fast}
+
+\subsection{Traversability Characteristics}
+
+For each initial configuration $\qinit \in \QI$, $K$ trajectories are computed, which results in the set of $K |\QI|$ trajectories.
+All these trajectories are used to compute the following properties of the tunnel $T$.
+A trajectory $P$ reaches the tunnel sphere $c_i \in T$, if the 3D Euclidean distance of the 
+nearest configuration $q \in P$ towards $c_i$ is less than $r_i$ (radius of the sphere $c_i$).
+Let $N(i)$ denote the number of trajectories that reached $i-$th sphere of the tunnel, $i=1,\ldots,n$.
+Three basic characteristics of the tunnel are computed from the trajectories: accessibility, throughput, and the scaling factor.
+
+The accessibility $A(i)=N(i)/N$ of the sphere $i$ is the probability of reaching the sphere $i$, where $N=K|\QI|$ is the total number of trajectories.
+The accessibility shows how probable it is to pass the tunnel up to the sphere $i$.
+Obviously, the most important is $A(n)$ of the last sphere of the tunnel, which can be considered as the overall difficulty of the tunnel.
+The  ligand passage may however be strongly affected by the first bottleneck, so the parts of the tunnel located behind
+the bottleneck have low accessibility.
+
+The throughput $T(i)$ is the ratio of trajectories that passed sphere $i$ (i.e., visited sphere $i+1$) and reached the sphere $i$,
+i.e., $T(i) = N(i+1) / N(i)$. 
+The throughput is not computed for the last sphere ($i=N$).
+The throughput shows a local accessibility of tunnel parts and it can be used to detect places where most of the trajectories ends.
+
+The proposed planner is allowed to scale down the radii of ligand spheres up to the allowed scale $\smin$. 
+It can be expected that narrower parts of the tunnel are more often passed with a more scaled-down ligand than
+the wider parts.
+Ligand scale $L(i)$ at the tunnel sphere $i$ is the average scale of ligand that reaches the sphere $i$.
 
@@
 
-\section{Discussion}
+Besides the color mapping, it is also necessary to show the computed trajectories.
+%This is useful, for example, to find out in which part of the tunnel the trajectories detour.
+Simple visualization of all trajectories could however be too slow for an interactive work.
+Therefore, the trajectories are first clustered and then only the clusters are visualized.
+Due to different lengths of the trajectories, they are first converted to a normalized form.
+Let $\cstart$ represent the average starting position of all trajectories and let $d_{max}$ be the 3D Euclidean distance
+of the most distant configuration from $\cstart$ among all trajectories.
+A set of $M$ spheres centered at $\cstart$ are created with radii $r'_i=i {d_{max} \over M}$, where $i=0,\ldots, M-1$.
+The trajectory $P=(q_1,\ldots,q_n)$ of length $n$ is represented by the normalized vector $v=(x_1,\ldots,x_M)$ of length $M$,
+where $x_i$ is the 3D position of the nearest configuration $q \in P$ to the surface of the $i$-th sphere with the radius $r'_i$.
+The distance between two normalized trajectories $v_i$ and $v_j$ is defined as
+$d(v_i, v_j) = \frac{1}{M} \sum_{1 \leq k \leq N} |x_k^i x_k^j|$.
+This distance is used in the UPGMA clustering technique~\cite{sokal1958statistical}.
+The trajectories can be visualized using a representative of each cluster.
+The number of trajectories in each cluster is represented by the width of the polyline (Fig.~\ref{fig:trajectories}).
+%In this manner, we are able to convey the information about all different trajectories together (see Fig.~\ref{fig:trajectories} right).
 
-%Trajectories computed for a ligand moving in a tunnel can reveal such properties that cannot be computed
-%only based on the geometry of the tunnels.
-
-The computed trajectories cannot be considered as `real', they should rather be considered as a hint for the biochemists.
-One of the reason is that the trajectories are computed inside a static protein and static tunnels.
-Real proteins are dynamic structures and consequently also the tunnels are dynamic: they move, merge or even disappear due to motions of protein atoms.
-Contrary, usually static tunnels are used in protein engineering.
-%multiple tunnels are detected separately in each frame of molecular dynamics and the tunnels are then aggregated to a single one.
-Despite such a simplification, researches often used information about the static tunnels  to estimate the possibility of chemical reactions
-at the active sites.
-
-The second limitation is caused by the utilization of scaled-down ligands.
-The protein atoms fill the internal void space therefore the tunnels identified inside proteins without ligand tend
-to be very narrow.
-To enable at least some motion of the ligands inside such narrow tunnels, the scaling-down of the ligand atoms is necessary.
-It is used also in other related tool, e.g.~\cite{cortes2005path}.
-%By allowing more aggressive scaling-down, more easier a trajectory can be found.
-%The tunnels therefore should be always evaluated considering various minimal scales.
-
-Due to these reasons, the computed trajectories can be considered either as too optimistic
-(e.g. because they are computed on a wide tunnel that can be in fact closed due to molecular dynamics),
-or too pessimistic (e.g. the trajectory is not found because of narrow passage around the tunnel, which can be 
-opened due to the molecular dynamics).
-Despite these limitations, testing the tunnel traversability using motion planning technique can provide chemists more information
-than the simple bottleneck radius, which is used nowadays.
-For example, possible detours from the tunnel can be identified, which indicates that even a tunnel with a small
-bottleneck can be traversed (Fig.~\ref{fig::detail}).
-
-%To better utilize the results of motion planning, it is necessary to observe behavior of ligands in many simulations with various proteins.
-%To the best knowledge of the authors, this is the first work attempt to analyze traversability of ligands in protein tunnels.
-%Most related works, e.g. those based on motion planning, are usually focused on protein folding, or finding arbitrary pathways
-%from proteins~\cite{cortes2010simulating,guieysse2008structure}.
-
-The proposed planner is very simple and it can be improved in many ways: by using a less aggressive scaling-down technique (in order
-to obtain better trajectories) or by using only a subset of conformations and a better expansion (in order to speed it up).
-Having a better motion planner (in the term of the speed or even success rate) however does not automatically mean that the resulting trajectories will help the biochemists more.
-The presented work shows possible advantages of the analysis based on ligand trajectories.
-By comparing the planned trajectories with trajectories observed in MD simulations, thresholds for the success rate or minimal tunnel throughput can be set, which will help biochemists to make decisions.
-
-%The runtime of the planner is currently given mainly by the collision-detection.
-%As the ligand is assumed to travel only along the tunnel centerline (in the distance defined by $\rv$),
-%only atoms located in the vicinity of the tunnel can be considered for collision detection.
-%Fast collision-detection can be achieved either using fast hierarchical data structures as OBB (implemented, e.g., in~\cite{ozcollide})
-%    or using collision-detection system developed directly for proteins~\cite{ruiz2005biocd}.
-%We assume that a conformation $l \in \L$ can be switched to all other conformations in $\L$.
-
-%parmas:
-%$\RI$
-%$\rv$
-%$\dt$
-%The presented expansion procedure favors the samples with maximal scale (they are tested first), the trajectory contains samples with higher scales at the easy part of the tunnels (e.i., wide areas), and the smaller scales are used only if the ligand has to fit into a narrow space.
-
-\section{Conclusion }
-%In the future, we will focus on the following extensions proposed by the domain experts.
-%Only ligand flexibility is included in the current algorithm but the flexibility of amino acid side-chains would greatly improve the prediction. 
-%Moreover, incorporation of some scoring function for energy evaluation would introduce the characterization of not only geometrical but also physico-chemical properties and would further improve the prediction ability of our approach.
-
-The protein tunnels are transporting pathways for ligands to the  active sites.
-In this paper, we propose to analyze the traversability in the tunnels using motion planning, that can consider both shape
-of the protein and ligand.
-We proposed modifications to the RRT method to compute trajectories for a ligand represented by a library of common conformations.
-The corresponding configuration space is sampled using the guided sampling, where a virtual sphere moves along the tunnel and attracts growth
-of the tree towards it.
-To enable trajectory computation in narrow tunnels, atomic radii of the ligand are scaled down.
-Computed trajectories allows us to evaluate the traversability of the tunnels using
-accessibility, throughput and also by visualizing typical pathways found by the planner.
-These properties can help to better understand importance of the tunnels.
-
-%which can be useful in the virtual screening
-
-\bibliographystyle{plain}
-\bibliography{paper}
-
-\end{document}
+\begin{figure}
+\centering
+\begin{tabular}{cc}
+\includegraphics[width=0.4\textwidth]{fig/trajectories-all} &
+\includegraphics[width=0.4\textwidth]{fig/trajectories-clustered-21} \\
+(a) & (b) \\                       
+\end{tabular}
+\caption{Visualization of the trajectories. The tunnel begins in the top left corner.
+(a) All trajectories ($\sim5000$) colored according to whether they reached the end of the tunnel (green) or not (red). 
+(b) Visualization using clusters of trajectories. 
+%Trajectories clustered according to their proximity colored by distinct colors.
+%The different line widths represent the number of trajectories in a cluster.
+\label{fig:trajectories}
+}
+\end{figure}

### 2017-09-05 paper.tex
sota

@@
 growth of the configuration tree may be toward the random samples may be blocked by obstacles~\cite{vonasekphd}.
-Instead, the random samples has to be generated in the difficult regions only if the tree can approach them.
+DD-RRT~\cite{yershovaDDRRT} limits the selection of nodes for expansion to a small ball. 
+The radius of a ball is defaulty infinity, but if the node cannot be expanded, it is set to a predefied (small) value.
+This small radius can be adapted according to the success rate of the expansion in ADD-RRT~\cite{jailletADRRT}.
+To attract the tree towards a region, the random samples have to be generate there only if the tree can expand to the region.
 This is the main idea of the guided sampling~\cite{vonasek2009rrt,denny2014marrt,denny2016dynamic}, where a the configurations
@@
 %the radii of ligands were scaled down. 
-Four different scaling-down factors were used: $\smin=\{0.4,0.5,0.6,0.7\}$.
+Four different scaling-down factors were used: $\smin=\{0.5,0.6,0.7,0.8\}$.
 For each scale, 20 collision-free initial configurations were found in the $\RI=2$\AA\ radius around the first sphere of the tunnel (outside the protein).
@@
 The parameters of the planner were: $\Imax=10,000$, $m=50$, $\gb=0.9$, $\dt=1.5$\AA, $\rv=2$~\AA.
+
 For each initial configuration, the success rate is computed as the ratio of trajectories that reached the end of the tunnel (to the distance 
@@
 
+The results are shown in Tab.~\ref{tab::main}.
+
 \begin{table}
 \centering
-\caption{\label{tab::rrt}
-    Runtime and success ratio for ligand $\LA$ (left) and $\LB$ (right).
-}
-{
+\caption{\label{tab::main}Traversability of the tunnels}
 \small
 \renewcommand{\tabcolsep}{1pt}
-\input table.003.tex
-\hskip 4pt   
-\input table.004.tex
+{\small
+\input graph.tex
 }
 \end{table}
-
-Trajectories for $\LA$ in all detected tunnels were classified as successful if they reached the last sphere of the tunnel
-to distance $\dt=2$~\AA and they were considered unsuccessful otherwise.
-The throughput computed from the trajectories shows that the tunnels are difficult not only around bottlenecks, but
-also in other places.
-The comparison of the classic bottleneck (i.e., measured by the radius of spherical probe) and throughput is depicted in Fig.~\ref{fig::tunnel2}a.
-The trajectories for $\smin=0.4$ are depicted in Fig.~\ref{fig::tunnel2}b and for $\smin=0.5$ in Fig.~\ref{fig::tunnel2}c.
-The successful trajectories for $\smin=0.4$ reveal that the end of the tunnel No. 3 can be approached by two different pathways (one in the tunnel and another one outside the tunnel).
-The detail is depicted in Fig.~\ref{fig::detail}.
-Despite the low bottleneck of this tunnel, the ligand may reach its end using the alternative pathway.
 
@@
 \end{figure}
+
+\begin{table}
+\centering
+\caption{\label{tab::rrt}
+    Runtime and success ratio for ligand $\LA$ (left) and $\LB$ (right).
+}
+{
+\small
+\renewcommand{\tabcolsep}{1pt}
+\input table.003.tex
+\hskip 4pt   
+\input table.004.tex
+}
+\end{table}
+
+Trajectories for $\LA$ in all detected tunnels were classified as successful if they reached the last sphere of the tunnel
+to distance $\dt=2$~\AA and they were considered unsuccessful otherwise.
+The throughput computed from the trajectories shows that the tunnels are difficult not only around bottlenecks, but
+also in other places.
+The comparison of the classic bottleneck (i.e., measured by the radius of spherical probe) and throughput is depicted in Fig.~\ref{fig::tunnel2}a.
+The trajectories for $\smin=0.4$ are depicted in Fig.~\ref{fig::tunnel2}b and for $\smin=0.5$ in Fig.~\ref{fig::tunnel2}c.
+The successful trajectories for $\smin=0.4$ reveal that the end of the tunnel No. 3 can be approached by two different pathways (one in the tunnel and another one outside the tunnel).
+The detail is depicted in Fig.~\ref{fig::detail}.
+Despite the low bottleneck of this tunnel, the ligand may reach its end using the alternative pathway.

### 2017-09-06 paper.tex
added first tables from the experiment
experiments + abstract

@@
 
+\def\RA{A$_{1}$}
+\def\RB{A$_{2}$}
+
 % ==================================================================================
@@
 The tunnels have been traditionally computed using Voronoi diagrams, which is suitable only for small, spherical and rigid ligands.
-We propose to analyze the accessibility of the active sites using sampling-based motion planning.
-This requires to cope with narrow passages and to model the flexibility of the ligands.
+Sampling-based motion planning can be used to analyze the traversability of the tunnels for non-spherical and flexible ligands.
+Motion planning inside protein structures leads to the narrow passage problem while considering flexible ligands increasing
+the degrees of freedom, and consequently the dimension of the configuration space.
+Moreover, as the flexible ligands have various shapes, the selection of a proper metric in the configuration space is crucial.
+In this paper, we propose a modification to the Rapidly Exploring Random Tree planner.
 To cope with the narrow passage problem, we generate the random samples around a Voronoi-based tunnel and we shrink
 the atomic radii of the ligand.
-The flexibility of the ligand is modeled using a fixed set of known conformations.
+The flexibility of the ligand is modeled using a fixed set of known conformations, so the dimension of the configuration space is not increased.
+The proposed method has been evaluated in a virtual screening scenario.
 \end{abstract}
@@
 A proper distance metric in the configuration space is necessary in order to determine which samples should be connected together (in the case of PRM), or which node of the tree should be expanded (in the case of RRT).
-Generally, it is not easy to combine the rotational and translational components of the configuration space.
+Generally, it is not easy to combine the rotational and transnational components of the configuration space.
 Moreover, the selection of the metric is also influence by the shape of the objects (robots).
@@
 DD-RRT~\cite{yershovaDDRRT} limits the selection of nodes for expansion to a small ball. 
-The radius of a ball is defaulty infinity, but if the node cannot be expanded, it is set to a predefied (small) value.
+The radius of a ball is default infinity, but if the node cannot be expanded, it is set to a predefined (small) value.
 This small radius can be adapted according to the success rate of the expansion in ADD-RRT~\cite{jailletADRRT}.
@@
 
-\begin{figure}[h]
+\begin{figure*}[h]
 \centering
@@
 }
-\end{figure}
+\end{figure*}
 
@@
 For each ligand, each minimal scale $\smin$ and each initial configuration, 10 trajectories were computed using the proposed planner.
-The parameters of the planner were: $\Imax=10,000$, $m=50$, $\gb=0.9$, $\dt=1.5$\AA, $\rv=2$~\AA.
 
@@
 
-\def\tmpa{0.13\textwidth}
+The proposed algorithm (Alg.~\ref{alg::main} and Alg.~\ref{alg::expand}), denoted as \RA\ in the rest of the paper, was run
+with the parameters $\Imax=10,000$, $m=50$, $\gb=0.9$, $\dt=1.5$\AA, $\rv=2$~\AA.
+This method was compared to RRT-Retraction~\cite{zhangRetraction} which is denoted as \RB\ in the rest of the text.
+TODO parametry.
+
+Both methods utilized the same data structures for nearest-neighbor search and collision detection.
+Parameters of the methods were chosen in order to use similar (in the term of order) number of collision-detection queries.
+
+\subsection{Traversability of one atom}
+
+The aim of the first experiment is to measure performance of both planner in the simplest scenario: finding path for one sphere towards the active site.
+The radius of the sphere is $1.7$~\AA.
+
+The results are shown in Tab.~\ref{tab::one} which shows percentage of MD frames where the sphere traversed the whole tunnel 
+(i.e., when it reached the active site).
+As the bottlenecks of the tunnels in each frame is known, we can compute the percentage of frames that have the bottleneck larger
+than the tested sphere radius.
+The first line in the Tab.~\ref{tab::one} therefore show the maximum percentage of traversable frames, while the second
+line (denote Planners) shows how both planners performed in this scenario.
+We can conclude, that the performance of both methods is similar in the case of first tunnel.
+In the case of the second tunnel, \RA\ was able to find trajectory in more frames than \RB.
+
+\begin{table}
+\centering
+\caption{\label{tab::one}Traversability for one-atome}
+\small
+\renewcommand{\tabcolsep}{3.5pt}
+{\small
+\input graphProbe.tex
+}
+\end{table}
 
@@
 
+\subsection{Traversability of ligands}
+
+The aim of the second experiment is to determine which ligands are
+
 The results are shown in Tab.~\ref{tab::main}.
@@
 \small
-\renewcommand{\tabcolsep}{1pt}
+\renewcommand{\tabcolsep}{3pt}
 {\small
@@
 \end{table}
+
+\subsection{Influence of conformation selection}
+
+The proposed planner utilizes a set of predefined ligand's conformations.
+As each of them is tested in the expansion step, this number of influences both runtime and the success rate.
+By using more conformations, there is a bigger chance that at least one of them can expand the tree towards the random sample $\qrand$.
+However, using large number of conformations also increase the time complexity of the planner.
+Therefore, it is important so select a `good' set of conformations.
+Depending on the ligand, some conformations are more packed i.e., the atoms in that conformations are close to each other, but 
+conformations where the atoms form a longer molecule are also possible.
+
+From motion planning point of view, both types of conformations are useful.
+The `longer' conformations can better fit in a narrow passage, but their rotation inside the narrow passage is problematic.
+On the other hand, the `packed' conformations cannot fit the narrow passage, but they can more likely rotate.
+
+In the third experiment, we tested how the selection of conformations influences the ability of the planners to find a solution.
+Two strategies were used to select the conformations.
+In the first strategy, denoted as Packed, $n$ `packed' conformations are selected from the available set of conformations.
+The second strategy, denoted as Various, selects $n$ conformations as follows.
+For each conformation, the geometric center is calculated and the deviation of the distances of the atoms from the center is calculated.
+The conformations are sorted according to this deviation and then $n$ conformations are selected from this sorted list uniformly (i.e., the
+        conformation number $1$, then $1+n$, etc.).
 

### 2017-09-06 paper.tex
intro, sota, algorithm

@@
 
-\def\L{L}
+\def\L{\mathcal{L}}
+\def\S{\mathcal{S}}
+
 \def\LA{L_1}
@@
 The active sites are accessible through one or more pathways (called tunnels).
-The knowledge about a tunnel (e.g. it's bottleneck or physico-chemical properties) can help chemists to estimate possible interactions
+The knowledge about a tunnel (e.g. the length, bottleneck or physico-chemical properties) can help chemists to estimate possible interactions
 between the protein and the ligand.
-The tunnels have been traditionally computed using Voronoi diagrams, which is suitable only for small, spherical and rigid ligands.
-Sampling-based motion planning can be used to analyze the traversability of the tunnels for non-spherical and flexible ligands.
-Motion planning inside protein structures leads to the narrow passage problem while considering flexible ligands increasing
-the degrees of freedom, and consequently the dimension of the configuration space.
-Moreover, as the flexible ligands have various shapes, the selection of a proper metric in the configuration space is crucial.
-In this paper, we propose a modification to the Rapidly Exploring Random Tree planner.
+Voronoi-diagrams are usually used to detect the tunnels, but this is useful only for estimate traversability of small spherical ligands.
+Computing motions of non-spherical ligands including their possible conformation changes (i.e., when atoms of the ligand move relatively to each other) leads to motion planning in high-dimensional configuration space, which can be solved using sampling-based planning methods.
+Such a planning problem necessarily includes narrow passages, as the movements of ligands are limited by many protein atoms.
+Moreover, considering flexible ligands increase the degrees of freedom, and consequently the dimension of the configuration space.
+%Moreover, as the flexible ligands have various shapes, the selection of a proper metric in the configuration space is crucial.
+In this paper, we propose a modification to the Rapidly Exploring Random Tree planner for the purpose of trajectory computations of ligands along protein tunnels.
 To cope with the narrow passage problem, we generate the random samples around a Voronoi-based tunnel and we shrink
@@
 A path in the roadmap then corresponds to a motion in the workspace.
-As the random samples are classified as free or non-free using collision detection, the sampling-based methods can found paths for robots of various shapes. 
-Another advantage of the sampling-based planners is that they can cope with robots with many degrees of freedom (DOF).
+The random samples are classified as free or non-free using black-box collision detection which allows the sampling-based methods to find paths for robots of various shapes. 
+The sampling-based planners is that they can cope with robots with many degrees of freedom (DOF).
 The two most used sampling-based techniques are Probabilistic Roadmaps (PRM)~\cite{kavrakiForPP} and Rapidly Exploring Random Trees (RRT)~\cite{lavalleRRT}.
-These planners and their variants have been utilized in wide range of applications in robotics, but also in 
-other research domain, namely in bio-chemistry 
+Besides many applications in robotics, sampling-based planners have been used e.g. in bio-chemistry 
 %Sampling-based motion planners has been intensively studied in robotics, but they have been applied also in other research fields.
@@
 %protein folding~\cite{amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},
-protein folding~\cite{raveh2009rapid,novinskaya2015improving},
-or for tunnel detection~\cite{vonasek2017tunnel}.
+or for protein folding~\cite{raveh2009rapid,novinskaya2015improving}.
+%or for tunnel detection~\cite{vonasek2017tunnel}.
 
+One of the problems in protein engineering is the tunnel detection, where the task is to find a tunnel (pathway) 
+leading to a specific place (called active site) inside the protein (MAME CITACI?)
 %or protein folding combined with ligand diffusion~\cite{cortes2010simulating}.
@@
 %Understanding of interactions between proteins and other small molecules is crucial in many research fields, including drug design and protein engineering. 
-The problem of tunnel detection is to find a tunnel (pathways) leading to a specific place (called active site) inside the protein.
 The knowledge about the existence of the tunnels and their properties (e.g., their bottleneck and chemical properties around the tunnel),
-is important to understand interactions between proteins and other molecules (ligands).
-To get the ligand to the active site, there has to be a transportation path connecting the protein outer environment with the active site.
-The tunnels have to be wide enough and the physico-chemical properties of the surrounding amino acids have to be compatible with the ligand (Fig.~\ref{fig::motiv}).
+is important to understand interactions between proteins and other molecules (ligands)~\cite{Koudelakova2013}.
+To get the ligand to the active site, a transportation tunnel connecting the protein outer environment with the active site has to exists.
+Such a  tunnel has to be wide enough and the physico-chemical properties of the surrounding amino acids have to be compatible with the ligand (Fig.~\ref{fig::motiv}).
 %These interactions are highly influenced by the ability to transport the small molecule, called ligand, to the protein active site.
@@
 
-To provide biochemists  a better insight into the behavior of non-spherical ligands, it is necessary to compute the trajectories of the ligand considering conformation changes.
+To provide biochemists  a better insight into the behavior of non-spherical ligands, it is necessary to compute the trajectories of the ligand considering its shape and also conformation changes (flexible ligands).
 The computation of trajectories for such ligands can be formulated as a motion planning problem.
 RRT is a suitable candidate to solve this problem due to its ability to cope with many-DOF robots of arbitrary shapes.
-This motion planning however require to cope with the narrow passage problem, as the proteins are dense structures and the
+This motion planning however requires to cope with the narrow passage problem, as the proteins are dense structures and the
 movements of the ligands are limited.
-A narrow passage is a small region in the configuration space whose removal changes the connectivity of the space.
-Due to its low volumes, the probability of sampling the narrow passages is low.
+The flexible ligands can be modeled as a multi-link kinematic chain~\cite{songPFpath,cortes2005path}, but it significantly
+increases the number of DOFs and consequently the dimension of the configuration space.
+Moreover, by considering flexible ligands, their shape differ in each conformation, which makes the selection of a suitable metric difficult.
 
-Consequently, many iterations are needed in order to put enough samples into the passages, which increases the computation time.
-A proper distance metric in the configuration space is necessary in order to determine which samples should be connected together (in the case of PRM), or which node of the tree should be expanded (in the case of RRT).
-Generally, it is not easy to combine the rotational and transnational components of the configuration space.
-Moreover, the selection of the metric is also influence by the shape of the objects (robots).
-
-To cope with these issues, we propose to guide the sampling of the configuration space using the Voronoi-based tunnels.
-The random samples are not generated from the whole configuration space, but mainly around the tunnel.
+In this paper, we propose a modification to RRT to cope with these issues.
+In the proposed method, the random samples are not generated from the whole configuration space, but the sampling is guided using
+the spheres of the tunnel.
+This allows us to find trajectories along the given tunnel, but it also decreases the volume of the configuration space to be searched.
 To increase movements of the ligands in the narrow tunnels, the free-space is dilated by 
 shrinking the atom radii, similarly, e.g., to~\cite{cortes2005path,hsu06multilevel}.
-The ligand can be considered as a multi-link kinematic chain and by changing the torsion angles, the flexibility can be modeled~\cite{songPFpath,cortes2005path}.
-This however increases the dimension of the configuration space.
-Instead, we model the flexibility using a fixed set of known conformations.
+We model the ligand flexibility using a fixed set of predefined conformations.
 
@@
 
-TODO some words about MD and why tunnels are predered
+TODO some words about MD and why it better to analyze stuff with tunnels than with MD
 
 The analysis of ligand traversability to/from an active site is usually based on so called tunnels.
-The tunnels are pathways computed for a single atom (spherical probe) using Voronio-diagrams~\cite{yaffe2008,caver3} or 
-grid-based approaches~\cite{sehnal2013mole,petrek2006caver}.
+Voronoi diagrams~\cite{yaffe2008,caver3} or grid-based methods~\cite{sehnal2013mole} are often used for tunnel computation.
+%The tunnels are pathways computed for a single atom (spherical probe) using Voronio-diagrams~\cite{yaffe2008,caver3} or 
+%grid-based approaches~\cite{sehnal2013mole,petrek2006caver}.
 The resulting tunnels are represented as a sequence of spheres and characterized by their bottleneck, length, curvature, and a list of surrounding residues, which are later used to estimate the interaction possibilities.
-The main disadvantage of both grid-based and VD-based methods is that the shape of the ligand is not taken into account during the tunnel detection and it is therefore not easy to estimate if (and how) a ligand might traverse the tunnels.
-In order to determine if a ligand can pass the tunnel, it is necessary to compute a trajectory considering the shape of the ligand.
-This can be formulated as a motion planning problem in a high-dimensional configuration space $\C$ and solved using sampling-based planners.
-In this case, the configuration space $\C$ has at least $6+n$ dimensions (3D rotation + 3D translation + additional degrees of freedom
-caused by the ligand flexibility).
+The main disadvantage of both grid-based and VD-based methods is that the shape of the ligand is not taken into account during the tunnel detection and it is therefore not easy to estimate if (and how) a non-spherical ligand might traverse the tunnels.
+Moreover, ligand flexibility cannot be principally considered by these approaches.
+In order to determine if a ligand can pass the tunnel, it is necessary to compute a trajectory considering the shape of the ligand and its 3D translation/rotation.
+%This can be formulated as a motion planning problem in a high-dimensional configuration space $\C$.
+%In this case, the configuration space $\C$ has at least $6+n$ dimensions (3D rotation + 3D translation + additional degrees of freedom
+%caused by the ligand flexibility).
 
@@
 The well known issue of the sampling-based planners is the narrow passage problem.
+A narrow passage is a small region in the configuration space whose removal changes the connectivity of the space.
+Due to its low volume, the probability of sampling the narrow passages is low.
+Consequently, many iterations are needed in order to put enough samples into the passages, which increases the computation time.
 %Sampling-based planners can provide poor performance in the presence of narrow passages, as many iterations
 %are required in order to sample the narrow passages dense enough.
-PRM-based planners can cope the narrow passage problem by increasing probability of sampling in given regions.
+PRM-based planners can cope with the narrow passage problem by increasing probability of sampling in given regions.
 These regions can be estimated using workspace information such as by medial axis~\cite{amatoOBRRT,amato2002using,wilmarthMAPRM} 
@@
 RRT-based planners cannot cope with the narrow passage problem simply by increasing the probability of sampling in them, as the
-growth of the configuration tree may be toward the random samples may be blocked by obstacles~\cite{vonasekphd}.
-DD-RRT~\cite{yershovaDDRRT} limits the selection of nodes for expansion to a small ball. 
-The radius of a ball is default infinity, but if the node cannot be expanded, it is set to a predefined (small) value.
-This small radius can be adapted according to the success rate of the expansion in ADD-RRT~\cite{jailletADRRT}.
+growth of the configuration tree toward the random samples may be blocked by obstacles~\cite{vonasekphd}.
+To prevent this blocking, DD-RRT~\cite{yershovaDDRRT} limits the selection of nodes for expansion to a small ball. 
+The radius of the ball is set to infinity for new nodes, and decreases to a predefined radius later if the node cannot be successfully expanded.
+The disadvantage of DD-RRT is the sensitivity to the predefined radius.
+Therefore, authors of ADD-RRT~\cite{jailletADRRT} proposed to adapt the radius  according to the success rate of the expansion.
+This is the main idea of the guided sampling~\cite{vonasek2009rrt,denny2014marrt,denny2016dynamic}, where a the configurations
 To attract the tree towards a region, the random samples have to be generate there only if the tree can expand to the region.
-This is the main idea of the guided sampling~\cite{vonasek2009rrt,denny2014marrt,denny2016dynamic}, where a the configurations
-space is sampled with higher probability around a given list of configurations.
-These guiding configurations can be computed e.g. as a path in the workspace.
-Authors of~\cite{kardossRRTKK} proposed to place the guiding configurations between difficult and easy regions of the configuration space, but
-they did not proposed any strategy to find the waypoints.
+In~\cite{kardossRRTKK}, probability of sampling is increased in several waypoints close to narrow passages, but without specifying
+how to find these waypoints.
+The generalization of~\cite{kardossRRTKK} is the guided sampling, where a whole path (computed e.g. in the workspace) is used
+to generate samples in the configuration space~\cite{vonasek2009rrt,denny2016dynamic,denny2014marrt}.
 
@@
 %to be located in narrow passages~\cite{zhangRetraction,lee2012srrrt}.
-In Retraction-RRT~\cite{zhangRetraction}, the random sample $\qrand$ is connected to the tree if the line segment $\qnear,\qrand$ is collision-free, otherwise, the retraction-step is performed.
+In Retraction-RRT~\cite{zhangRetraction}, the random sample $\qrand$ is connected to the tree if the line segment 
+from $\qnear$ to $\qrand$ is collision-free.
+Otherwise, the retraction-step is performed.
 The task of the retraction step is to find a contact configuration around $\qnear$ that minimizes the distance to $\qrand$.
 The retraction step is computationally intensive, as the computation of the closest contact point leads to computation
-of the generalized penetration depth, which has high complexity~\cite{zhang2008fast}.
+of the generalized penetration depth, which has high complexity~\cite{zhang2008fast,he2016efficient}. %\cite{he2016efficient} in RAL!!
 Therefore, many algorithms uses a heuristic to compute the samples near boundary of the configuration space~\cite{amatoOBPRM,amatoOBRRT}.
@@
 
+TODO NEKAM PRESUNU/SMAZU: 
+A proper distance metric in the configuration space is necessary in order to determine which samples should be connected together (in the case of PRM), or which node of the tree should be expanded (in the case of RRT).
+Generally, it is not easy to combine the rotational and transnational components of the configuration space.
+Moreover, the selection of the metric is also influence by the shape of the objects (robots).
+
 \section{Proposed method}
@@
 
-Contrary to~\cite{cortes2010simulating}, which find any exit pathway, we propose to compute the trajectories only around/inside each tunnel
+In contrast to~\cite{cortes2010simulating}, which find any exit pathway, we propose to compute the trajectories only around/inside each tunnel
 separately.
@@
 the tunnel bottleneck TODO CITACE??.
-To emulate this behavior, we shrink the atom radii of the ligand, similarly to~\cite{cortes2010simulating}.
+To emulate this behavior, we shrink the atom radii of the ligand, similarly to~\cite{cortes2010simulating,guieysse2008structure}.
 
@@
 To enable motion of ligands in the narrow tunnels, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
-A discrete set of scales is used, i.e., $s \in \{\smin, \smin+\sdelta, \ldots, \smax\}$, where 
+A discrete set of scales is used, i.e., $s \in \S=\{\smin, \smin+\sdelta, \ldots, \smax\}$, where 
 $\smin$ is the minimal allowed scale, $\smax=1$ is the maximal allowed scale and $\sdelta$ is the minimal difference between two scales.
@@
 by the 3D position $(x,y,z)$ of the reference point of the ligand (e.g., its geometric center), rotation around $x$, $y$, and $z$ axes,
-index of the conformation $l\in \L$ and the actual scale $s$.
-All possible configurations form the configuration space $\C$. 
-A configuration is collision-free if none of the ligand atoms scaled by $s$ and placed at the
+index of the conformation $l\in \L$ and the actual scale $s \in \S$.
+All possible configurations form the configuration space $\C$ and the collision-free configurations
+form the subspace $\CF \subseteq \C$.
+A configuration $q$ is collision-free if none of the ligand atoms scaled by $s$ and placed at the
 position defined by $q$ collides with the protein atoms.
@@
 
-\subsection{Computing Initial Configuration}
+\subsection{Computing Initial Configurations}
 
-The analysis of tunnel traversability is based on computation of multiple trajectories of the ligand inside the tunnel, which
-requires to generate a set $\QI$ of collision-free initial configurations.
-In this paper we assume that the ligand has to travel from the beginning to the end of the tunnel.
-As the tunnels are computed for a spherical probe, the ligand may not fit exactly to the first sphere of the tunnel.
-The initial configurations have to be searched around the first sphere of the tunnel.
+Depending on the chemical interactions between the ligand and the protein surface, the ligand may enter a tunnel with various rotations and in various conformations.
+The traversability of the tunnel should therefore be evaluated using various starting configurations $\QI$.
+The initial configurations are have to be searched around the beginning of the tunnel.
 To find a new initial configuration, a random sample $q$ is generated around $c_i$ in the distance $\RI$ (translation and rotation
@@
 If the sample $q$ is collision-free, it can be considered as a new starting configuration, so $q$ is added to $\QI$.
-Similarly, a single goal configuration $\qgoal$ is found around the end of the tunnel. 
+%Similarly, a single goal configuration $\qgoal$ is found around the end of the tunnel. 
 For each starting configuration $\qinit \in \QI$,  $K$ trajectories are created.
@@
 
-\subsection{Computing Single Trajectory of the Ligand}
+\subsection{RRT for computing ligands trajectory inside a tunnel}
 
-The task of the trajectory computation is to find a trajectory for a ligand in the given tunnel. 
-% which can be solved using RRT~\cite{lavalleRRT}.
-Here, the original RRT is extended to cope with the specific requirements needed for the traversability of ligands.
-First, the trajectory has to be found around the given tunnel, but deflection from the tunnel centerline is allowed.
-%The deviation of the trajectory from the tunnel centerline is influenced by the parameter $\rv$.
-%We assume, that the distance of the ligand from the tunnel centerline is at most $\rv$.
-Due to this requirement, sampling process of RRT has to be adapted in order to follow the tunnel, and to prevent
-construction of trajectories in the rest of the protein.
-%The other modifications aim to cope with various scales of the ligand and multiple conformations,
-
-The main loop of the proposed method is described in Alg.~\ref{alg::main}.
+The proposed motion planning is based on the principle of RRT~\cite{lavalleRRT} planner.
+The main loop of planner, which is described in Alg.~\ref{alg::main} is similar to the original RRT planner.
 In each iteration, a random sample $\qrand$ is generated and its nearest node $\qnear\in\T$ in the tree is found.
@@
 
+The aim of the first modification is to sample the configuration space along the tunnel being analyzed.
 To guide the growth of the tree through the tunnel, a moving virtual goal is used~\cite{vonasek2009rrt}.
@@
 The random samples $\qrand$ are generated around the sphere $c_v \in T$ with the probability $\gb$, and from the whole $\C$ otherwise.
-%with probability $1-\gb$ from $\C$, otherwise they are generated around the sphere $c_v \in T$.
 After the tree reaches the sphere $c_v$, i.e., the distance of the tree to $c_v$ is
@@
 
-The core of the proposed planner is the expansion procedure (Alg.~\ref{alg::expand}) which generates new collision-free nodes around $\qnear\in\T$.
+The second modification to the RRT is the novel expansion procedure (Alg.~\ref{alg::expand}), which generates new collision-free nodes around $\qnear\in\T$.
 For each ligand conformation $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with a maximal scale.
-First, the maximal scale $\smax$ is tested and $m$ random samples are generated around $\qrand$ and tested for collision.
+First, the maximal scale $\smax \in \S$ is tested and $m$ random samples are generated around $\qrand$ and tested for collision.
 The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
@@
 
-The sampling-based methods are sensitive to the employed metric, especially if the objects are not symmetrical, which is the
-case of the non-spherical ligands.
-To consider the actual shape of the ligand (which is different in each conformation) and to support finding such configurations
-that maximally approach $\qrand$, the distance between newly generated configurations and $\qrand$ is measured as the smallest
-3D distance between an atom of the ligand placed at $q$ and the 3D position of $\qrand$ ($\da(q,\qrand)$ on line~\ref{alg::expand:a} in
-Alg.~\ref{alg::expand}).
-By computing the distance between the nearest atoms, the shape of the ligand is actually considered.
+The third modification is the utilization of another metric during the expansion procedure.
+While the nearest neighbors are search using 6D Euclidean metric, which can be efficiently realized using KD-trees, the expansion
+procedure utilize a metric that is sensitive to the shape of the ligands.
+This is necessary as the ligands adopt various shapes according to the actual conformations.
+To minimize the distance between $\qrand$ and the ligand, the shape of the ligand should be considered.
+Therefore, we propose to measure the distance as the smallest 3D Euclidean distance between atoms placed at $\qrand$ and a configuration
+$q$ being tested.
+This metric is denoted ($\da(\cdot,\cdot)$ on line~\ref{alg::expand:a} in Alg.~\ref{alg::expand}).
+
 This metric supports retraction of the ligand towards $\qrand$.
-%As the expansion routine attempts to find such sample that minimizes this distance to $\qrand$, it supports retraction of the ligand towards $\qrand$.
+As the expansion routine attempts to find such sample that minimizes this distance to $\qrand$, it supports retraction of the ligand towards $\qrand$.
 

### 2017-09-07 paper.tex
added images, new intro image

@@
 \includegraphics[width=0.16\textwidth]{fig/motiv3}  \\
+Protein 1CQW & Active site & Detected tunnels \\ %& Example of a  \\
+             &            & (orange)         \\  %& trajectory
+\multicolumn{3}{c}{%
+\includegraphics[width=0.16\textwidth]{fig/renderDCP} 
+\includegraphics[width=0.16\textwidth]{fig/render37t}} \\ 
 %\hbox{
@@
 %}
-Protein 1CQW & Active site & Detected tunnels \\ %& Example of a  \\
-             &            & (orange)          %& trajectory
 \end{tabular}
@@
 A path in the roadmap then corresponds to a motion in the workspace.
-The random samples are classified as free or non-free using black-box collision detection which allows the sampling-based methods to find paths for robots of various shapes. 
-The sampling-based planners is that they can cope with robots with many degrees of freedom (DOF).
+%The random samples are classified as free or non-free using black-box collision detection which allows the sampling-based methods to find paths for robots of various shapes. 
+%The sampling-based planners can cope with robots with many degrees of freedom (DOF).
 The two most used sampling-based techniques are Probabilistic Roadmaps (PRM)~\cite{kavrakiForPP} and Rapidly Exploring Random Trees (RRT)~\cite{lavalleRRT}.
+The sampling-based planners can cope with many-DOF (Degrees of Freedom) robots of arbitrary shape and therefore they have been used for motion planning of varous robotic systems.
 Besides many applications in robotics, sampling-based planners have been used e.g. in bio-chemistry 
@@
 One of the problems in protein engineering is the tunnel detection, where the task is to find a tunnel (pathway) 
-leading to a specific place (called active site) inside the protein (MAME CITACI?)
+leading to a specific place (called active site) inside the protein~\cite{gora2013gates}.
 %or protein folding combined with ligand diffusion~\cite{cortes2010simulating}.

### 2017-09-07 paper.tex
affilications, intro

@@
 %\documentclass[a4paper, 10pt]{IEEEtran}      % for icra/RAL final
-\documentclass[letterpaper, 10 pt, conference]{ieeeconf} % initial submission to RAL/ICRA
-
+\documentclass[usletter, 10pt, conference]{ieeeconf} % initial submission to RAL/ICRA
 \usepackage{graphicx}
@@
 \usepackage{multirow}
+\usepackage{flushend}
 %\usepackage[bookmarks=false]{hyperref}
@@
 
-%\IEEEoverridecommandlockouts                              % This command is only
-%\overrideIEEEmargins
+\IEEEoverridecommandlockouts                              % This command is only
+\overrideIEEEmargins
 
@@
 \thanks{$^{2}$        
-Faculty of Informatics,   Masaryk University, 
-Botanick\'a 68a, 602~00 Brno, Czech Republic,\\
-The presented work has been supported by the Czech Science Foundation (GA{\v C}R) under research project No. 17-07690S.
-Access to computing and storage facilities owned by parties and projects contributing to the National Grid Infrastructure MetaCentrum, provided under the programme "Projects of Large Infrastructure for Research, Development, and Innovations" (LM2010005), is greatly appreciated.}}
+Faculty of Informatics,   Masaryk University, Botanick\'a 68a, 602~00 Brno, Czech Republic,
+%The presented work has been supported by the Czech Science Foundation (GA{\v C}R) under research project No. 17-07690S.
+%Access to computing and storage facilities owned by parties and projects contributing to the National Grid Infrastructure MetaCentrum, provided under the programme "Projects of Large Infrastructure for Research, Development, and Innovations" (LM2010005), is greatly appreciated.
+%}%
+}
+}
 
@@
 
-%\frontmatter          % for the preliminaries
-
 \begin{abstract}
@@
 Such a planning problem necessarily includes narrow passages, as the movements of ligands are limited by many protein atoms.
-Moreover, considering flexible ligands increase the degrees of freedom, and consequently the dimension of the configuration space.
-%Moreover, as the flexible ligands have various shapes, the selection of a proper metric in the configuration space is crucial.
+Considering flexible ligands increases the degrees of freedom, and consequently the dimension of the configuration space.
 In this paper, we propose a modification to the Rapidly Exploring Random Tree planner for the purpose of trajectory computations of ligands along protein tunnels.
@@
 
-\begin{figure}[b]
+\begin{figure}[t]
 \centering
@@
 \includegraphics[width=0.16\textwidth]{fig/renderDCP} 
-\includegraphics[width=0.16\textwidth]{fig/render37t}} \\ 
+\includegraphics[width=0.14\textwidth]{fig/render37t}} \\ 
 %\hbox{
@@
 To provide biochemists  a better insight into the behavior of non-spherical ligands, it is necessary to compute the trajectories of the ligand considering its shape and also conformation changes (flexible ligands).
-The computation of trajectories for such ligands can be formulated as a motion planning problem.
-RRT is a suitable candidate to solve this problem due to its ability to cope with many-DOF robots of arbitrary shapes.
-This motion planning however requires to cope with the narrow passage problem, as the proteins are dense structures and the
-movements of the ligands are limited.
-The flexible ligands can be modeled as a multi-link kinematic chain~\cite{songPFpath,cortes2005path}, but it significantly
-increases the number of DOFs and consequently the dimension of the configuration space.
-Moreover, by considering flexible ligands, their shape differ in each conformation, which makes the selection of a suitable metric difficult.
+This can be formulated as a motion planning problem.
+It is necessary to cope with the narrow passage problem, as the proteins are dense structures and the movements of the ligands are limited, and to efficiently model the flexible ligands.
 
@@
 computed using a RRT-based method.
+The flexible ligands is modeled as a kinematic chain, which increases dimension of the configuration space.
 To cope with many-DOF ligands, the RRT-ML variant~\cite{cortes2007mlrrt} is employed as the basic planner.
@@
 
-TODO NEKAM PRESUNU/SMAZU: 
-A proper distance metric in the configuration space is necessary in order to determine which samples should be connected together (in the case of PRM), or which node of the tree should be expanded (in the case of RRT).
-Generally, it is not easy to combine the rotational and transnational components of the configuration space.
-Moreover, the selection of the metric is also influence by the shape of the objects (robots).
+%TODO NEKAM PRESUNU/SMAZU: 
+%A proper distance metric in the configuration space is necessary in order to determine which samples should be connected together (in the case of PRM), or which node of the tree should be expanded (in the case of RRT).
+%Generally, it is not easy to combine the rotational and transnational components of the configuration space.
+%Moreover, the selection of the metric is also influence by the shape of the objects (robots).
+%The flexible ligands can be modeled as a multi-link kinematic chain~\cite{songPFpath,cortes2005path}, but it significantly
+%increases the number of DOFs and consequently the dimension of the configuration space.
+%Moreover, by considering flexible ligands, their shape differ in each conformation, which makes the selection of a suitable metric difficult.
 

### 2017-09-08 paper.tex
all text

@@
 The active sites are accessible through one or more pathways (called tunnels).
-The knowledge about a tunnel (e.g. the length, bottleneck or physico-chemical properties) can help chemists to estimate possible interactions
-between the protein and the ligand.
-Voronoi-diagrams are usually used to detect the tunnels, but this is useful only for estimate traversability of small spherical ligands.
-Computing motions of non-spherical ligands including their possible conformation changes (i.e., when atoms of the ligand move relatively to each other) leads to motion planning in high-dimensional configuration space, which can be solved using sampling-based planning methods.
-Such a planning problem necessarily includes narrow passages, as the movements of ligands are limited by many protein atoms.
-Considering flexible ligands increases the degrees of freedom, and consequently the dimension of the configuration space.
+Based on the tunnel properties like it's length and bottleneck, the chemists can estimate possibility of reaching the active site and therefore estimate the possibility of the interactions.
+%The knowledge about a tunnel (e.g. the length, bottleneck or physico-chemical properties) can help chemists to estimate possible interactions between the protein and the ligand.
+Nowadays, tunnels are computed using Voronoi diagrams, which helps to predict interactions only with small spherical ligands.
+Computing motions of non-spherical ligands including conformation changes (i.e., when atoms of the ligand move relatively to each other) leads to motion planning in high-dimensional configuration space, which can be solved using sampling-based planning methods.
+This planning problem necessarily includes narrow passages, as the movements of ligands are limited by many protein atoms.
+Considering flexible ligands increases the numer of degrees of freedom, and consequently the dimension of the configuration space.
 In this paper, we propose a modification to the Rapidly Exploring Random Tree planner for the purpose of trajectory computations of ligands along protein tunnels.
-To cope with the narrow passage problem, we generate the random samples around a Voronoi-based tunnel and we shrink
-the atomic radii of the ligand.
-The flexibility of the ligand is modeled using a fixed set of known conformations, so the dimension of the configuration space is not increased.
-The proposed method has been evaluated in a virtual screening scenario.
+To cope with the narrow passage problem, the random samples are generated around a Voronoi-based tunnel.
+The flexibility of the ligand is modeled using a fixed set of known conformations which does not increase the number of dimensions of the configuration space.
+The proposed method has been evaluated in a virtual screening scenario to estimate traversability of tunnels for several ligands.
 \end{abstract}
@@
 \multicolumn{3}{c}{%
-\includegraphics[width=0.16\textwidth]{fig/renderDCP} 
+\includegraphics[width=0.16\textwidth]{fig/renderDCP}  \hskip 15pt
 \includegraphics[width=0.14\textwidth]{fig/render37t}} \\ 
@@
 \caption{\label{fig::motiv}
-    Tunnels in haloalkane dehalogenase with a possible trajectory of 1-chlorpropan ligand.
+    Tunnels in haloalkane dehalogenase (1CQW) with possible trajectories of 1-chlorpropan ligand (bottom left) and 1,2-Dichloroethane (bottom right).
 }
@@
 The sampling-based planners can cope with many-DOF (Degrees of Freedom) robots of arbitrary shape and therefore they have been used for motion planning of varous robotic systems.
-Besides many applications in robotics, sampling-based planners have been used e.g. in bio-chemistry 
+Besides many applications in robotics, sampling-based planners have been used also in bio-chemistry 
 %Sampling-based motion planners has been intensively studied in robotics, but they have been applied also in other research fields.
@@
 %protein folding~\cite{amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},
-or for protein folding~\cite{raveh2009rapid,novinskaya2015improving}.
+and for protein folding~\cite{raveh2009rapid,novinskaya2015improving}.
 %or for tunnel detection~\cite{vonasek2017tunnel}.
@@
 %Understanding of interactions between proteins and other small molecules is crucial in many research fields, including drug design and protein engineering. 
-The knowledge about the existence of the tunnels and their properties (e.g., their bottleneck and chemical properties around the tunnel),
-is important to understand interactions between proteins and other molecules (ligands)~\cite{Koudelakova2013}.
-To get the ligand to the active site, a transportation tunnel connecting the protein outer environment with the active site has to exists.
-Such a  tunnel has to be wide enough and the physico-chemical properties of the surrounding amino acids have to be compatible with the ligand (Fig.~\ref{fig::motiv}).
-%These interactions are highly influenced by the ability to transport the small molecule, called ligand, to the protein active site.
-%The active site can be considered as a deeply buried inner cavity with the ability to interact with the incoming ligand.
-The tunnels are usually detected using Voronoi diagrams~\cite{yaffe2008,caver3} assuming 
+%The knowledge about the existence of the tunnels and their properties (e.g., their bottleneck and chemical properties around the tunnel),
+%is important to understand interactions between proteins and other molecules (ligands)~\cite{Koudelakova2013}.
+As the interaction takes place in the active site, the existence of a tunnel connection the active site with the outer environment
+is necessary for the interaction.
+%To get the ligand to the active site, a transportation tunnel connecting the protein outer environment with the active site has to exists.
+Such a  tunnel has to be wide enough and the physico-chemical properties of the surrounding amino acids have to be compatible with the ligand.
+Chemists are using the tunnel detection method in order to estimate possibility of interactions between a protein and a ligand without need to perform long Molecular Dynamics (MD) simulations or even real experiments (Fig.~\ref{fig::motiv}).
+The tunnels are detected mainly using Voronoi diagrams~\cite{yaffe2008,caver3} assuming 
 a spherical ligand (probe) and they are represented as a sequence of spheres.
 However, as the ligands typically non-spherical and flexible,  it is difficult to estimate their traversability through the tunnels computed for only one sphere.
-
 %Biochemists decide if a tunnel can be used to transport a given ligand mainly based on the tunnel length and bottleneck (i.e., the radius of the smallest sphere that forms the tunnel).
@@
 
-The analysis of ligand traversability to/from an active site is usually based on so called tunnels.
-Voronoi diagrams~\cite{yaffe2008,caver3} or grid-based methods~\cite{sehnal2013mole} are often used for tunnel computation.
+Most of the available tools for tunnel detection are based on Voronoi diagrams~\cite{yaffe2008,caver3} or they utilize grid-based computation~\cite{sehnal2013mole}.
 %The tunnels are pathways computed for a single atom (spherical probe) using Voronio-diagrams~\cite{yaffe2008,caver3} or 
@@
 A narrow passage is a small region in the configuration space whose removal changes the connectivity of the space.
-Due to its low volume, the probability of sampling the narrow passages is low.
+Due to its low volume, the probability of sampling the narrow passage is low.
 Consequently, many iterations are needed in order to put enough samples into the passages, which increases the computation time.
@@
 Therefore, authors of ADD-RRT~\cite{jailletADRRT} proposed to adapt the radius  according to the success rate of the expansion.
-This is the main idea of the guided sampling~\cite{vonasek2009rrt,denny2014marrt,denny2016dynamic}, where a the configurations
 To attract the tree towards a region, the random samples have to be generate there only if the tree can expand to the region.
@@
 how to find these waypoints.
-The generalization of~\cite{kardossRRTKK} is the guided sampling, where a whole path (computed e.g. in the workspace) is used
-to generate samples in the configuration space~\cite{vonasek2009rrt,denny2016dynamic,denny2014marrt}.
+The generalization of~\cite{kardossRRTKK} is the guided sampling, where a whole path is used
+to generate samples in the configuration space.
+The path can be computed in the workspace~\cite{vonasek2009rrt}, or iteratively refined in the configuration space based on solution of a relaxed version of the problem~\cite{bayazitIRC}.
 
-Besides increasing probability of sampling in certain regions, the growth of the tree towards the regions can be reached by retracting
-the sampling to the regions.
+%~\cite{vonasek2009rrt,denny2016dynamic,denny2014marrt}.
+%This is the main idea of the guided sampling~\cite{vonasek2009rrt,denny2014marrt,denny2016dynamic}, where a the configurations
+
 %Another techniques to sample the narrow passages dense enough is to retract the random samples towards regions that are believed
 %to be located in narrow passages~\cite{zhangRetraction,lee2012srrrt}.
-In Retraction-RRT~\cite{zhangRetraction}, the random sample $\qrand$ is connected to the tree if the line segment 
-from $\qnear$ to $\qrand$ is collision-free.
-Otherwise, the retraction-step is performed.
-The task of the retraction step is to find a contact configuration around $\qnear$ that minimizes the distance to $\qrand$.
-The retraction step is computationally intensive, as the computation of the closest contact point leads to computation
-of the generalized penetration depth, which has high complexity~\cite{zhang2008fast,he2016efficient}. %\cite{he2016efficient} in RAL!!
-Therefore, many algorithms uses a heuristic to compute the samples near boundary of the configuration space~\cite{amatoOBPRM,amatoOBRRT}.
+%the random sample $\qrand$ is connected to the tree if the line segment  from $\qnear$ to $\qrand$ is collision-free. Otherwise, the retraction-step is performed.
+%The task of the retraction step is to find a contact configuration around $\qnear$ that minimizes the distance to $\qrand$.
+In Retraction-based RRT~\cite{zhangRetraction}, the tree is retracted along the boundary of the obstacles in the configuration space.
+This requires the analysis of contact space, which leads to the generalized penetration depth computation of high complecity~\cite{zhang2008fast,he2016efficient}. %\cite{he2016efficient} in RAL!!
+Selective Retraction-based RRT~\cite{lee2012srrrt} avoid expensive growth in the open areas of the configuration space and focus
+the retractions only to narrow passages.
+Obstacle-based RRT~\cite{amatoOBRRT} utilizes several expansion procedures based on e.g. random vectors, obstacles vectors or medial axis in order to improve growth in the narrow passages.
+In MARRT~\cite{denny2014marrt}, the newly generated samples are pushed toward the medial axis.
+%Therefore, many algorithms uses a heuristic to compute the samples near boundary of the configuration space~\cite{amatoOBPRM,amatoOBRRT}.
+
 The probability of sampling of the narrow passages can be increased by dilating the free-space, e.g. by shrinking the geometry of 
 the robot~\cite{hsuOnProb} or the obstacles~\cite{bayazitIRC}.
-
 The shrinking technique has been used also in~\cite{cortes2010simulating}, where exit pathways for a small flexible molecule are
@@
 In this paper, we propose a modification of the RRT planner for computing trajectories through protein tunnels.
-The input is a set of tunnels computed by Voronoi diagrams and the task is to find trajectories inside (or along) the tunnels
+The input is a set of tunnels computed by Voronoi diagrams and the task is to find trajectories along the given tunnels
 leading from outside the protein to the active site.
 
-In contrast to~\cite{cortes2010simulating}, which find any exit pathway, we propose to compute the trajectories only around/inside each tunnel
-separately.
-This allows us to decrease the volume of the configuration space to be search, which increases the relative volume of the narrow passages.
+In contrast to~\cite{cortes2010simulating}, which finds any exit pathway, we compute the trajectories onlyalong each tunnel separately.
+This is motivated by the practical needs of chemists that are interested in the traversability of a selected tunnel.
+Computing trajectories along a single tunnel also brings advantages from the motion planning point of view.
+It decreases the volume of the configuration space to be search and increases the relative volume of the narrow passages.
 To further cope with the narrow passage problem, the random samples are not generated in the whole configuration space, but only (mainly) around
@@
 The flexibility of a ligand is modeled using the set $\L$ of its conformations.
-The conformations $L$ are used from a library (e.g.,~\cite{dunbrack}), or they can be prepared considering the potential energy.
+The conformations $\L$ are used from a library (e.g.,~\cite{dunbrack}), or they can be prepared considering the potential energy.
 
@@
 $c_i \in \R^3$ is their 3D position and $r_i > 0$ denotes the maximum collision-free radius of a sphere centered at $c_i$. 
-The tunnels can be found by tunnel detection tools like CAVER 3.0~\cite{caver3}.
+The tunnels can be found by the tunnel detection tools like CAVER 3.0~\cite{caver3}.
 
@@
 The initial configurations are have to be searched around the beginning of the tunnel.
-To find a new initial configuration, a random sample $q$ is generated around $c_i$ in the distance $\RI$ (translation and rotation
+To find a new initial configuration, a random sample $q$ is generated around $c_1$ in the distance $\RI$ (translation and rotation
  parts of $q$ are generated randomly, the scale is set to $\smin$ and the conformation index is set randomly).
@@
 The proposed motion planning is based on the principle of RRT~\cite{lavalleRRT} planner.
-The main loop of planner, which is described in Alg.~\ref{alg::main} is similar to the original RRT planner.
+The main loop of planner (Alg.~\ref{alg::main}) is similar to the original RRT planner.
 In each iteration, a random sample $\qrand$ is generated and its nearest node $\qnear\in\T$ in the tree is found.
@@
 The virtual goal $v, 1\le v \le n$, is the index of a sphere of the tunnel.
+At the begining of the algorithm, the virtual goal is $v=1$.
 The random samples $\qrand$ are generated around the sphere $c_v \in T$ with the probability $\gb$, and from the whole $\C$ otherwise.
@@
 Setting the virtual goal to this successor allows the tree to avoid such parts of the tunnels that are not traversable or reachable by the ligand.
-This is necessary in the dense protein structures, where it is not always possible to exactly follow the tunnel.
+%This is necessary in the dense protein structures, where it is not always possible to exactly follow the tunnel.
+This allows the tree to slightly detour from the tunnel, which improves performance around narrow tunnels.
 The algorithm terminates after a predefined number of planning trials $\Imax$ or if the tree reaches
@@
 
-The second modification to the RRT is the novel expansion procedure (Alg.~\ref{alg::expand}), which generates new collision-free nodes around $\qnear\in\T$.
+The second proposed modification of RRT is the expansion procedure (Alg.~\ref{alg::expand}), which generates new collision-free nodes around $\qnear\in\T$.
 For each ligand conformation $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with a maximal scale.
-First, the maximal scale $\smax \in \S$ is tested and $m$ random samples are generated around $\qrand$ and tested for collision.
+First, the maximal scale $\smax \in \S$ is selectd and $m$ random samples are generated around $\qrand$ and tested for collisions.
 The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
@@
 While the nearest neighbors are search using 6D Euclidean metric, which can be efficiently realized using KD-trees, the expansion
-procedure utilize a metric that is sensitive to the shape of the ligands.
+procedure utilizes a metric that is sensitive to the shape of the ligands.
 This is necessary as the ligands adopt various shapes according to the actual conformations.
 To minimize the distance between $\qrand$ and the ligand, the shape of the ligand should be considered.
+
 Therefore, we propose to measure the distance as the smallest 3D Euclidean distance between atoms placed at $\qrand$ and a configuration
 $q$ being tested.
-This metric is denoted ($\da(\cdot,\cdot)$ on line~\ref{alg::expand:a} in Alg.~\ref{alg::expand}).
-
+This metric is denoted $\da(\cdot,\cdot)$ on line~\ref{alg::expand:a} in Alg.~\ref{alg::expand}.
 This metric supports retraction of the ligand towards $\qrand$.

### 2017-09-08 paper.tex
added images (not in repo yet)
conformation selection

@@
 This metric is denoted $\da(\cdot,\cdot)$ on line~\ref{alg::expand:a} in Alg.~\ref{alg::expand}.
-This metric supports retraction of the ligand towards $\qrand$.
 As the expansion routine attempts to find such sample that minimizes this distance to $\qrand$, it supports retraction of the ligand towards $\qrand$.
+The fast evaluation of this metric with complexity $O(\log n)$, where $n$ is the number of atoms of the ligand, can be achieved using KD-trees.
+For each conformation $\l \in L$ of the ligand, a KD-tree is build for the ligand at configuration $q=(0,0,0,0,0,0,l,\smax)$
+(i.e., for ligand with no translation and no rotation).
+To find a nearest atom to $\qrand$ for a ligand at the configuration $q=(x,y,z,r_x,r_y,r_z,l,s)$, the translation vector $T=(x,y,z)$ and
+rotation matrix $R$ given by $r_x$, $r_y$ and $r_z$ are computed.
+Then, $\qrand$ is translated to $\qrand' = R^{-1}(\qrand-T)$ and the KD-tree search is used to find nearest point towards $\qrand'$, which
+is also the nearest neighbor of the ligand at $q$ to $\qrand$.
 
@@
 
+\subsection{Selection of conformations for motion planing}
+\label{sec::strat}
+
+The proposed planner utilizes a set of predefined ligand's conformations.
+As each of them is tested in the expansion step, this number of influences both runtime and the success rate.
+By using more conformations, there is a bigger chance that at least one of them can expand the tree towards the random sample $\qrand$.
+However, using large number of conformations also increase the time complexity of the planner.
+Therefore, it is important so select a `good' set of conformations.
+Depending on the ligand, some conformations are more packed i.e., the atoms in that conformations are close to each other, but 
+conformations where the atoms form a longer molecule are also possible.
+
+From motion planning point of view, both types of conformations are useful.
+The `longer' conformations can better fit in a narrow passage, while the `packed' conformations can be better rotated at place.
+To select the suitable conformations, a simple technique is proposed.  
+For each conformation, the geometric center is calculated and the deviation of the distances of the atoms from the center is calculated.
+The conformations are sorted according to this deviation and then the desired number of conformations are selected from this sorted list uniformly.
+
 \subsection{Visualization of Results}
@@
 
+TODO Bara: prepsat/motivace/jak se visuazlizuje
+
 The proposed approach was used to analyze tunnel traversability towards to active site of Haloalkane dehalogenase protein (PDB ID 1CQW).
@@
 \end{table}
+
+{\def\a{270}
+\begin{figure}
+\centering
+\renewcommand{\tabcolsep}{0pt}
+\begin{tabular}{cccc}
+\includegraphics[width=0.15\textwidth,angle=\a]{fig/m001-oneAtom-cdCount} &
+\includegraphics[width=0.15\textwidth,angle=\a]{fig/m001-oneAtom-rrt_path_time} &
+\includegraphics[width=0.15\textwidth,angle=\a]{fig/m037t-cdCount} & 
+\includegraphics[width=0.15\textwidth,angle=\a]{fig/m037t-rrt_path_time} \\
+m01 & m01 & m37 & m37 \\
+
+\end{tabular}
+\end{figure}
+}
 
@@
 
+\subsection{Influence of conformation selection}
+
+The aim of this experiment is to verify the procedure to select ligand conformations for the motion planning.
+The experiment was performed with the ligand DCP, for which $1300$ conformations where available and for the ligand
+M040 with $7560$ available conformations.
+Two strategies were tested: the strategy described in Sec.~\ref{sec::strat} (denoted `S') and a strategy,
+that prefers smaller conformations (denoted `P').
+In the second method, deviation of the atomic distance from the geometric center are computed and the conformations
+with the lowest deviations are selected.
+The results are summarized in Tab.~\ref{tabl::selection}, which shows the percentage number of frames that are traversable by the tested ligand.
+For each tested ligand, 5 and 50 conformations from the available pool of conformations were selected using the strategies.
+In the case of both tested ligands and 50 selected conformations, the `S' strategy leads to higher traversability rate than the `P' strategy.
+For the ligand DCP, a higher traversability rates was achieved when 50 conformations are selected.
+This experiments shows that selecting 50 conformations using the `S' strategy (Sec.~\ref{sec::strat}) leads to better results
+than using less number of conformations or when preferring the packed conformations.
+
+\begin{table}
+\centering
+\caption{\label{tab::selection}Traversability of the tunnels depending on conformation selection}
+\small
+\renewcommand{\tabcolsep}{3pt}
+{\small
+\input graphSel.tex
+}
+\end{table}
+
 \subsection{Traversability of ligands}
@@
 \end{table}
-
-\subsection{Influence of conformation selection}
-
-The proposed planner utilizes a set of predefined ligand's conformations.
-As each of them is tested in the expansion step, this number of influences both runtime and the success rate.
-By using more conformations, there is a bigger chance that at least one of them can expand the tree towards the random sample $\qrand$.
-However, using large number of conformations also increase the time complexity of the planner.
-Therefore, it is important so select a `good' set of conformations.
-Depending on the ligand, some conformations are more packed i.e., the atoms in that conformations are close to each other, but 
-conformations where the atoms form a longer molecule are also possible.
-
-From motion planning point of view, both types of conformations are useful.
-The `longer' conformations can better fit in a narrow passage, but their rotation inside the narrow passage is problematic.
-On the other hand, the `packed' conformations cannot fit the narrow passage, but they can more likely rotate.
-
-In the third experiment, we tested how the selection of conformations influences the ability of the planners to find a solution.
-Two strategies were used to select the conformations.
-In the first strategy, denoted as Packed, $n$ `packed' conformations are selected from the available set of conformations.
-The second strategy, denoted as Various, selects $n$ conformations as follows.
-For each conformation, the geometric center is calculated and the deviation of the distances of the atoms from the center is calculated.
-The conformations are sorted according to this deviation and then $n$ conformations are selected from this sorted list uniformly (i.e., the
-        conformation number $1$, then $1+n$, etc.).
 

### 2017-09-08 paper.tex
missing images

@@
 From motion planning point of view, both types of conformations are useful.
-The `longer' conformations can better fit in a narrow passage, while the `packed' conformations can be better rotated at place.
+The `longer' conformations can better fit in a narrow passage, while the `packed' conformations can be better rotated at place 
+(Fig.~\ref{fig::m040c}.
 To select the suitable conformations, a simple technique is proposed.  
@@
 The conformations are sorted according to this deviation and then the desired number of conformations are selected from this sorted list uniformly.
+
+\begin{figure}
+\centering
+\includegraphics[width=0.3\textwidth]{fig/m040-conformations}
+\caption{\label{fig::m040c}
+ Examples of `long' conformation (left) and a `packed' one for the ligand M040 (1,5-Dichloropentane).
+}
+\end{figure}
 
@@
 \includegraphics[width=0.15\textwidth,angle=\a]{fig/m037t-rrt_path_time} \\
-m01 & m01 & m37 & m37 \\
-
+\multicolumn{2}{c}{One-atom} &
+\multicolumn{2}{c}{m037}  \\
+\multicolumn{4}{c}{%
+\includegraphics[width=0.15\textwidth,angle=\a]{fig/m040-10-cdCount} 
+\includegraphics[width=0.15\textwidth,angle=\a]{fig/m040-10-rrt_path_time} 
+\includegraphics[width=0.15\textwidth,angle=\a]{fig/m040-10-trajDistToGoal}} \\
+\multicolumn{4}{c}{m040} 
 \end{tabular}

### 2017-09-08 paper.tex
conclusion, table

@@
 This planning problem necessarily includes narrow passages, as the movements of ligands are limited by many protein atoms.
-Considering flexible ligands increases the numer of degrees of freedom, and consequently the dimension of the configuration space.
+Considering flexible ligands increases the number of degrees of freedom, and consequently the dimension of the configuration space.
 In this paper, we propose a modification to the Rapidly Exploring Random Tree planner for the purpose of trajectory computations of ligands along protein tunnels.
@@
 The two most used sampling-based techniques are Probabilistic Roadmaps (PRM)~\cite{kavrakiForPP} and Rapidly Exploring Random Trees (RRT)~\cite{lavalleRRT}.
-The sampling-based planners can cope with many-DOF (Degrees of Freedom) robots of arbitrary shape and therefore they have been used for motion planning of varous robotic systems.
+The sampling-based planners can cope with many-DOF (Degrees of Freedom) robots of arbitrary shape and therefore they have been used for motion planning of various robotic systems.
 Besides many applications in robotics, sampling-based planners have been used also in bio-chemistry 
@@
 In Retraction-based RRT~\cite{zhangRetraction}, the tree is retracted along the boundary of the obstacles in the configuration space.
-This requires the analysis of contact space, which leads to the generalized penetration depth computation of high complecity~\cite{zhang2008fast,he2016efficient}. %\cite{he2016efficient} in RAL!!
+This requires the analysis of contact space, which leads to the generalized penetration depth computation of high complexity~\cite{zhang2008fast,he2016efficient}. %\cite{he2016efficient} in RAL!!
 Selective Retraction-based RRT~\cite{lee2012srrrt} avoid expensive growth in the open areas of the configuration space and focus
@@
 
-In contrast to~\cite{cortes2010simulating}, which finds any exit pathway, we compute the trajectories onlyalong each tunnel separately.
+In contrast to~\cite{cortes2010simulating}, which finds any exit pathway, we compute the trajectories only along each tunnel separately.
 This is motivated by the practical needs of chemists that are interested in the traversability of a selected tunnel.
@@
 The virtual goal $v, 1\le v \le n$, is the index of a sphere of the tunnel.
-At the begining of the algorithm, the virtual goal is $v=1$.
+At the beginning of the algorithm, the virtual goal is $v=1$.
 The random samples $\qrand$ are generated around the sphere $c_v \in T$ with the probability $\gb$, and from the whole $\C$ otherwise.
@@
 For each ligand conformation $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with a maximal scale.
-First, the maximal scale $\smax \in \S$ is selectd and $m$ random samples are generated around $\qrand$ and tested for collisions.
+First, the maximal scale $\smax \in \S$ is selected and $m$ random samples are generated around $\qrand$ and tested for collisions.
 The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
@@
 
+TODO Bara: prepsat/motivace/jak se visuazlizuje
+toto je cast textu z toho posledniho paperu. Kdyztak to vyhod.
+
 %In order to fursupport the tunnel exploration process performed by biochemists.
@@
 
-TODO Bara: prepsat/motivace/jak se visuazlizuje
-
 The proposed approach was used to analyze tunnel traversability towards to active site of Haloalkane dehalogenase protein (PDB ID 1CQW).
 The active site was defined using residues 38 (Asparagine), 102 (Histidine) and 103 (Aspartate).
-Molecular dynamics of the protein was computed using TODO. 
-First 100 frames were analyzed. 
-In each frame, tunnels for spherical probe 0.9~\AA were dected using CAVER 3.0~\cite{caver3}. 
+Molecular dynamics of the protein was computed using Amber tool. 
+First 100 frames were analyzed with the aim to calculate percentage of frames where the active site was accessible through the tunnels.
+In each frame, tunnels for spherical probe 0.9~\AA were detected using CAVER 3.0~\cite{caver3}. 
 In most of the frames, two tunnels were detected.
@@
 %the radii of ligands were scaled down. 
-Four different scaling-down factors were used: $\smin=\{0.5,0.6,0.7,0.8\}$.
+Four different scaling-down factors were used: $\S=\{0.5,0.6,0.7,0.8\}$.
 For each scale, 20 collision-free initial configurations were found in the $\RI=2$\AA\ radius around the first sphere of the tunnel (outside the protein).
 For each ligand, each minimal scale $\smin$ and each initial configuration, 10 trajectories were computed using the proposed planner.
-
-For each initial configuration, the success rate is computed as the ratio of trajectories that reached the end of the tunnel (to the distance 
- $\rv=2$~\AA or less) over 100 trials.
 
@@
 This method was compared to RRT-Retraction~\cite{zhangRetraction} which is denoted as \RB\ in the rest of the text.
+
 TODO parametry.
@@
 
-\section{Conclusion }
-The protein tunnels are transporting pathways for ligands to the  active sites.
-In this paper, we propose to analyze the traversability in the tunnels using motion planning, that can consider both shape
-of the protein and ligand.
-We proposed modifications to the RRT method to compute trajectories for a ligand represented by a library of common conformations.
-The corresponding configuration space is sampled using the guided sampling, where a virtual sphere moves along the tunnel and attracts growth
-of the tree towards it.
-To enable trajectory computation in narrow tunnels, atomic radii of the ligand are scaled down.
-Computed trajectories allows us to evaluate the traversability of the tunnels using
-accessibility, throughput and also by visualizing typical pathways found by the planner.
-These properties can help to better understand importance of the tunnels.
+\section{Conclusion}
+
+In this paper, we presented a modification of the RRT planner for trajectory computation of ligands inside protein tunnels.
+The tunnels are pathways leading from an outer environment toward the active sites, where chemical interactions may undergo.
+So far, the tunnels have been detected using Voronoi diagrams and analyzed based on their length, bottleneck and the list of residues along them.
+Such an analysis is however useful only for small spherical ligands and it is difficult to estimate behavior of a large, non-spherical ligands.
+In the proposed work, we compute trajectories using modified RRT for whole ligands and we model the conformation changes of the
+ligands using a predefined set of conformations.
+The main idea of the proposed planner is to generate the random samples around the Voronoi-based tunnels.
+The probability of sampling is increased around so called virtual goal, which is a point moving along the tunnel. 
+The virtual goal is moved forward, if the tree approaches it close enough.
+To enable movements in narrow tunnels, the atomic radii of the ligands are scaled down.
 

### 2017-09-09 paper.tex
graphs for comparing methods a1,a2, ligand images, expeirments

@@
 The fast evaluation of this metric with complexity $O(\log n)$, where $n$ is the number of atoms of the ligand, can be achieved using KD-trees.
-For each conformation $\l \in L$ of the ligand, a KD-tree is build for the ligand at configuration $q=(0,0,0,0,0,0,l,\smax)$
+For each conformation $\l \in \L$ of the ligand, a KD-tree is build for the ligand at configuration $q=(0,0,0,0,0,0,l,\smax)$
 (i.e., for ligand with no translation and no rotation).
@@
 \centering
-\includegraphics[width=0.3\textwidth]{fig/m040-conformations}
+%\includegraphics[width=0.25\textwidth]{fig/m040-conformations}
+\includegraphics[width=0.15\textwidth]{fig/m040-conf1} \hskip 25pt
+\includegraphics[width=0.12\textwidth]{fig/m040-conf2}
 \caption{\label{fig::m040c}
- Examples of `long' conformation (left) and a `packed' one for the ligand M040 (1,5-Dichloropentane).
+ Examples of `long' conformation (left) and a `packed' one for the ligand m040 (1,5-Dichloropentane).
 }
@@
 
-{\def\a{270}
-\begin{figure}
-\centering
-\renewcommand{\tabcolsep}{0pt}
-\begin{tabular}{cccc}
-\includegraphics[width=0.15\textwidth,angle=\a]{fig/m001-oneAtom-cdCount} &
-\includegraphics[width=0.15\textwidth,angle=\a]{fig/m001-oneAtom-rrt_path_time} &
-\includegraphics[width=0.15\textwidth,angle=\a]{fig/m037t-cdCount} & 
-\includegraphics[width=0.15\textwidth,angle=\a]{fig/m037t-rrt_path_time} \\
-\multicolumn{2}{c}{One-atom} &
-\multicolumn{2}{c}{m037}  \\
-\multicolumn{4}{c}{%
-\includegraphics[width=0.15\textwidth,angle=\a]{fig/m040-10-cdCount} 
-\includegraphics[width=0.15\textwidth,angle=\a]{fig/m040-10-rrt_path_time} 
-\includegraphics[width=0.15\textwidth,angle=\a]{fig/m040-10-trajDistToGoal}} \\
-\multicolumn{4}{c}{m040} 
-\end{tabular}
-\end{figure}
-}
+%
+%{\def\a{270}
+%\begin{figure}
+%\centering
+%\renewcommand{\tabcolsep}{0pt}
+%\begin{tabular}{cccc}
+%\includegraphics[width=0.15\textwidth,angle=\a]{fig/m001-oneAtom-cdCount} &
+%\includegraphics[width=0.15\textwidth,angle=\a]{fig/m001-oneAtom-rrt_path_time} &
+%\includegraphics[width=0.15\textwidth,angle=\a]{fig/m037t-cdCount} & 
+%\includegraphics[width=0.15\textwidth,angle=\a]{fig/m037t-rrt_path_time} \\
+%\multicolumn{2}{c}{One-atom} &
+%\multicolumn{2}{c}{m037}  \\
+%\multicolumn{4}{c}{%
+%\includegraphics[width=0.15\textwidth,angle=\a]{fig/m040-10-cdCount} 
+%\includegraphics[width=0.15\textwidth,angle=\a]{fig/m040-10-rrt_path_time} 
+%\includegraphics[width=0.15\textwidth,angle=\a]{fig/m040-10-trajDistToGoal}} \\
+%\multicolumn{4}{c}{m040} 
+%\end{tabular}
+%\end{figure}
+%}
+%
 
@@
 The aim of this experiment is to verify the procedure to select ligand conformations for the motion planning.
-The experiment was performed with the ligand DCP, for which $1300$ conformations where available and for the ligand
-M040 with $7560$ available conformations.
-Two strategies were tested: the strategy described in Sec.~\ref{sec::strat} (denoted `S') and a strategy,
-that prefers smaller conformations (denoted `P').
-In the second method, deviation of the atomic distance from the geometric center are computed and the conformations
-with the lowest deviations are selected.
-The results are summarized in Tab.~\ref{tabl::selection}, which shows the percentage number of frames that are traversable by the tested ligand.
-For each tested ligand, 5 and 50 conformations from the available pool of conformations were selected using the strategies.
+The experiment was performed with the ligand DCP (Dichloropropanol), for which $1300$ conformations where available and for the ligand
+m040 (1,5-Dichloropentane) with $7560$ available conformations.
+Two strategies were tested: the strategy described in Sec.~\ref{sec::strat} (denoted `S') and a strategy that prefers smaller conformations (denoted `P').
+The `P' method selects such conformations where the deviation of the atomic distance from the geometric center are lowest, which leads 
+to selection of the packed conformations (Fig.~\ref{m040c}).
+
+The percentage of frames that were traversible by the ligands are shown in Tab.~\ref{tab::selection}, where the subscript of each ligand
+name shows the number of selected conformations (5 or 50) and the used method (S or P).
+The number after the slash in the ligand's name denote the number of atoms of the ligand.
 In the case of both tested ligands and 50 selected conformations, the `S' strategy leads to higher traversability rate than the `P' strategy.
@@
 \centering
-\caption{\label{tab::selection}Traversability of the tunnels depending on conformation selection}
+\caption{\label{tab::selection}
+    Traversability of the tunnels depending on conformation selection.
+    The number after $/$ denotes the number of atoms.
+}
 \small
@@
 
-The aim of the second experiment is to determine which ligands are
-
-The results are shown in Tab.~\ref{tab::main}.
+Based on the results of the previous experiment, 50 conformations using the `S' strategy were selected for the following
+ligands: m037t (1,2-Dichloroethane), m038t (1,3-Dichloropropane), m056 ( 2-Chlorobutane) and m080 (Trichloropropane), and
+the task was to compute trajectories for these ligands in the MD sequence.
+
+The results are summarized in Tab.~\ref{tab::main}, which shows percentage of frames in which the tested planners \RA\ and \RB\ found
+trajecotries close to the active site. 
+The active site was considered as reached if the distance from the center of the ligand to the active site was less than 2~\AA.
+
+The analyzes sequence of MD  snapshots contains two tunnels. 
+The average bottleneck of the tunnels are $1.48$~\AA and $1.2$~\AA for the first, resp. the second tunnel.
+The tables shows percentages of traversable MD snapshots only for the first tunnel, as no trajectories were found in the second tunnel
+for most of the tested ligand.
+The only except is the ligand m037t, which was the smallest tested ligand (with 8 atoms), and therefore it can fit even inside the second tunnel.
+
+The proposed \RA\ planner outperforms the \RB\ planner in the number of frames where a solution was found.
+The distances of the configuration trees (measured using 3D Euclidean metric) to the active site and the runtimes in each frame
+are shown in Fig.~\ref{fig::compared}.
+The runtimes of both algorithms increases in those frames, where the tunnel is not traversable, i.e., where the distance
+to the active site is more than 2~\AA (e.g., the frame number 60).
 
@@
 \centering
-\caption{\label{tab::main}Traversability of the tunnels}
+\caption{\label{tab::main}
+    Traversability of the tunnels for ligands with 50 conformations. 
+    The number after $/$ denotes the number of atoms.
+}
 \small
@@
 \end{table}
+
+{\def\a{270}
+\begin{figure}
+\centering
+\includegraphics[width=0.25\textwidth,angle=\a]{fig/m037tmins-03tunnel-1rrt_path_time-2}
+\includegraphics[width=0.25\textwidth,angle=\a]{fig/m037tmins-03tunnel-1trajDistToGoal-2}
+\caption{\label{fig::comparison}
+    Runtimes and distance to the active site for the m037 ligand at scale 0.4.
+    The values are computed from all 200 trials (20 initial configurations $\times$ 10 trials).
+    Only each second frame is shown in the graph.
+}
+\end{figure}
+}
 

### 2017-09-09 paper.tex
info about rrt retraction

@@
     \ForEach{$s \in (\smax,\smax-\sdelta, \ldots, \smin+\sdelta, \smin)$}{
-        $\qnew = \emptyset$; // empty configuration\\
+        $\qnew = \emptyset$; // empty configuration \nllabel{alg::expa} \\
         \For{$i = 1,\ldots,m$}{
@@
         }
+        \nllabel{alg::expb}
         \If{$\qnew \ne \emptyset$} {
@@
 
-\subsection{Selection of conformations for motion planing}
+\subsection{Selection of Conformations for Motion Planning}
 \label{sec::strat}
@@
 Molecular dynamics of the protein was computed using Amber tool.
-First 100 frames were analyzed with the aim to calculate percentage of frames where the active site was accessible through the tunnels.
-In each frame, tunnels for spherical probe 0.9~\AA were detected using CAVER 3.0~\cite{caver3}. 
+%100 frames of the dynamics were analyzed with the aim to calculate percentage of frames where the active site was accessible through the tunnels.
+In each frame, tunnels for the spherical probe 0.9~\AA were detected using CAVER 3.0~\cite{caver3}. 
 In most of the frames, two tunnels were detected.
-
-%has been analyzed using the proposed approach.
-%Three tunnels were detected using CAVER 3.0~\cite{caver3} for the spherical probe of radius 0.9~\AA (Fig.~\ref{fig::tunnel}a).
-%The traversability was evaluated for 1-chlorpropan (denoted as $\LA$) with 11 atoms and 
-% from David: m003 je "1-chlorpropan" a m004 "1-chlorbutan"
-%and 1-chlorbutan (denoted as $\LB$) with 14 atoms.
-%$\LA$ was represented by 12 conformations, and $\LB$ by a set of $114$ conformations.
-%Examples of three conformations of $\LA$ and $\LB$ are depicted in Fig.~\ref{fig::tunnel}b.
-%Further information about the experiments can be found at %{\url{http://mrs.felk.cvut.cz/isrr2017}}.
-
-%Both tested ligands have more than 10 atoms and therefore they cannot fit into the tunnel of with 0.9~\AA, so
-%the radii of ligands were scaled down. 
+The average bottleneck of the tunnels are $1.48$~\AA and $1.2$~\AA for the first, resp. the second tunnel.
+%The second tunnel is more dynamic, i.e., it fluctuatest more
+
 Four different scaling-down factors were used: $\S=\{0.5,0.6,0.7,0.8\}$.
 For each scale, 20 collision-free initial configurations were found in the $\RI=2$\AA\ radius around the first sphere of the tunnel (outside the protein).
-For each ligand, each minimal scale $\smin$ and each initial configuration, 10 trajectories were computed using the proposed planner.
-
-The proposed algorithm (Alg.~\ref{alg::main} and Alg.~\ref{alg::expand}), denoted as \RA\ in the rest of the paper, was run
-with the parameters $\Imax=10,000$, $m=50$, $\gb=0.9$, $\dt=1.5$\AA, $\rv=2$~\AA.
-This method was compared to RRT-Retraction~\cite{zhangRetraction} which is denoted as \RB\ in the rest of the text.
-
-TODO parametry.
-
-Both methods utilized the same data structures for nearest-neighbor search and collision detection.
-Parameters of the methods were chosen in order to use similar (in the term of order) number of collision-detection queries.
+For each ligand, each minimal scale $\smin$ and each initial configuration, 10 trajectories were computed.
+
+Two motion planners were utilized: the proposed planner (Alg.~\ref{alg::main} and Alg.~\ref{alg::expand}), denoted as \RA\ in the rest of the paperm, and the Retraction-based RRT~\cite{zhangRetraction} denoted as \RB\ in the rest of the paper.
+The \RA\ method was run with the following parameters:
+$\Imax=10,000$, $m=100$, $\gb=0.9$, $\dt=1.5$\AA, $\rv=2$~\AA.
+Both methods were implemented in the same software framework and they utilized the same data structures and routines for collision detection and nearest-neighbor search.
+The Retraction-based RRT was implemented using 6D Euclidean metric.
+The tested methods differ in the expansion step (lines \ref{alg::expa}--\ref{alg::expb} in Alg.~\ref{alg::expand}).
+In these lines, the \RB\ method employs the retraction step in order to find a new collision-free configurations $\qnew$ that approaches $\qrand$.
+The retraction step was realized by sampling the vicinity of the configuration $\qnear$ in the distance $0.1$~\AA\ using 20 samples.
+A collision-free sample that maximally approached $\qrand$ was selected and added to the tree, and the retraction continued from this sample.
+The retraction step was repeated 5 times.
+The \RB\ method can therefore expand the tree by at most 5 new nodes per conformation, while \RA\ extends the tree only by one configuration.
+The larger number of nodes in \RB\ method is not problematic, as the KD-trees were used for nearest-neighbor search and the time to solve these queries is significantly less than computing collision detection.
+Parameters of the methods were selected in such a way that the number of collision detection queries in each expansion step is same.
+For each conformation and tested scale, the \RA\ methods attempts to find $m=100$ collision-free configurations, and \RB\ attempts to perform $5 \times 20$ collision-detection queries.
 
@@
 
-The analyzes sequence of MD  snapshots contains two tunnels. 
-The average bottleneck of the tunnels are $1.48$~\AA and $1.2$~\AA for the first, resp. the second tunnel.
 The tables shows percentages of traversable MD snapshots only for the first tunnel, as no trajectories were found in the second tunnel
@@
 The distances of the configuration trees (measured using 3D Euclidean metric) to the active site and the runtimes in each frame
-are shown in Fig.~\ref{fig::compared}.
+are shown in Fig.~\ref{fig::comparison}.
 The runtimes of both algorithms increases in those frames, where the tunnel is not traversable, i.e., where the distance

### 2017-09-09 paper.tex
text

@@
 This requires the analysis of contact space, which leads to the generalized penetration depth computation of high complexity~\cite{zhang2008fast,he2016efficient}. %\cite{he2016efficient} in RAL!!
-Selective Retraction-based RRT~\cite{lee2012srrrt} avoid expensive growth in the open areas of the configuration space and focus
-the retractions only to narrow passages.
+Selective Retraction-based RRT~\cite{lee2012srrrt} avoids the expensive growth in the open areas of the configuration space and focuses 
+the retractions only to the narrow passages.
 Obstacle-based RRT~\cite{amatoOBRRT} utilizes several expansion procedures based on e.g. random vectors, obstacles vectors or medial axis in order to improve growth in the narrow passages.
@@
 computed using a RRT-based method.
-The flexible ligands is modeled as a kinematic chain, which increases dimension of the configuration space.
-To cope with many-DOF ligands, the RRT-ML variant~\cite{cortes2007mlrrt} is employed as the basic planner.
+The flexible ligand is modeled as a kinematic chain, which increases the dimension of the configuration space.
+To cope with these many-DOF ligands, RRT-ML~\cite{cortes2007mlrrt} is employed as the basic planner.
 RRT-ML expands the tree primarily using those DOF that are essential for achieving motion of the ligand (i.e., rotation
@@
 leading from outside the protein to the active site.
+The tunnels are detected in a sequence of protein dynamics that is computed using MD simulations.
 
@@
 Computing trajectories along a single tunnel also brings advantages from the motion planning point of view.
-It decreases the volume of the configuration space to be search and increases the relative volume of the narrow passages.
+It decreases the volume of the configuration space to be searched  and increases the relative volume of the narrow passages.
 To further cope with the narrow passage problem, the random samples are not generated in the whole configuration space, but only (mainly) around
@@
 
-As the input data are snapshots of the proteins without ligands, it can be expected, that the tunnels are narrow therefore it would be difficult
-to find trajectories for all ligands.
-In reality, the tunnels adapt to the ligands (and vice versa), so even narrow tunnels can serve as transportation path for ligands larger than
-the tunnel bottleneck TODO CITACE??.
-To emulate this behavior, we shrink the atom radii of the ligand, similarly to~\cite{cortes2010simulating,guieysse2008structure}.
-
-The ligand flexibility is modeled using a predefined set of typical conformations, therefore the number of dimensions of the configuration space does not depend on the DOF of the ligand as in~\cite{cortes2010simulating}.
+Due to the protein dynamics, the tunnel changes, move and merge, and their bottleneck changes in each frame.
+It can therefore be expected, that the detected tunnels are rather narrow.
+Based on the observation from MD simulations, the tunnels adapt to the ligands (and vice versa), so even narrow tunnels can serve as transportation path for thhe ligands TODO CITACE??.
+To enable computing of trajectories in the narrow tunnels, we shrink the atom radii of the ligand, similarly to~\cite{cortes2010simulating,guieysse2008structure}.
+
+The ligand flexibility is modeled using a predefined set of typical conformations, therefore the number of the dimensions of the configuration space does not depend on the DOF of the ligand as in~\cite{cortes2010simulating}.
 Library of conformations are used also in different types of calculations, e.g.~\cite{kellogg}. 
@@
 Proteins and ligands are represented by the hard sphere model, where the radius of each sphere (atom) is given by its van der Waals radius.
-The flexibility of a ligand is modeled using the set $\L$ of its conformations.
-The conformations $\L$ are used from a library (e.g.,~\cite{dunbrack}), or they can be prepared considering the potential energy.
-
-%or downloaded from a library~\cite{dunbrack}.
-%We assume that a conformation $l \in \L$ can be switched to all other conformations in $\L$.
-
-%Another approach is to model the ligand flexibility using a predefined set of feasible conformations $\L$
-%To decrease the computational burden, a set of feasible conformations $\L$ is prepared, e.g. using Rosseta tool.
-%The conformations can be prepared using Molecular Dynamics software such as Rosetta.
-%For each conformation $c\in \L$ an energy potential is also calculated.
-%Let $\L_i$ denote the set of conformations that 
-%The transition matrix $M$ defines the possible changes of conformations, i.e., $M(i,j)=1$ if the conformation $c_i \in C$
-%can be switched to the confomration $c_j \in C$.
 
@@
 %Depending on the protein, the tunnel bottlenecks can be even less than $1$~\AA, which is too narrow for ligands with more than 2 atoms.
+The flexibility of a ligand is modeled using the set $\L$ of its conformations.
+We assume that a library of all possible (in the considered resolution) conformations is available (e.g., ~\cite{dunbrack}) and we later show how to select a subset of conformations suitable for motion planning.
+%The conformations $\L$ are used from a library (e.g.,~\cite{dunbrack}), or they can be prepared considering the potential energy.
+
 To enable motion of ligands in the narrow tunnels, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
@@
 A configuration of the ligand $q=(x,y,z,r_x,r_y,r_z,l,s)$  is described
-by the 3D position $(x,y,z)$ of the reference point of the ligand (e.g., its geometric center), rotation around $x$, $y$, and $z$ axes,
-index of the conformation $l\in \L$ and the actual scale $s \in \S$.
+by the 3D position $(x,y,z)$ of the reference point of the ligand (the geometric center), the rotation around $x$, $y$, and $z$ axes,
+the index of the conformation $l\in \L$ and the actual scale $s \in \S$.
 All possible configurations form the configuration space $\C$ and the collision-free configurations
@@
 The traversability of the tunnel should therefore be evaluated using various starting configurations $\QI$.
-The initial configurations are have to be searched around the beginning of the tunnel.
-To find a new initial configuration, a random sample $q$ is generated around $c_1$ in the distance $\RI$ (translation and rotation
- parts of $q$ are generated randomly, the scale is set to $\smin$ and the conformation index is set randomly).
-If the sample $q$ is collision-free, it can be considered as a new starting configuration, so $q$ is added to $\QI$.
+The initial configurations are searched around the beginning of the tunnel.
+To find a new initial configuration, a random sample $q$ is generated around the first sphere of the tunnel 
+$c_1$ in the distance $\RI$ (the translation and rotation  parts of $q$ are generated randomly, the scale is set to $\smin$ and the conformation index is set randomly).
+If the sample $q$ is collision-free, it can be considered as a new starting configuration and it is added to $\QI$.
 %Similarly, a single goal configuration $\qgoal$ is found around the end of the tunnel. 
-For each starting configuration $\qinit \in \QI$,  $K$ trajectories are created.
-Each trajectory is computed using a modified RRT, which is introduced in the following section.
+For each starting configuration $\qinit \in \QI$, the trajectories are computed using a modified RRT, which is introduced in the following section.
 
@@
 The aim of the first modification is to sample the configuration space along the tunnel being analyzed.
-To guide the growth of the tree through the tunnel, a moving virtual goal is used~\cite{vonasek2009rrt}.
+To guide the growth of the tree using the tunnel, a moving virtual goal is used~\cite{vonasek2009rrt}.
 The virtual goal $v, 1\le v \le n$, is the index of a sphere of the tunnel.
-At the beginning of the algorithm, the virtual goal is $v=1$.
+At the beginning of the algorithm, $v=1$.
 The random samples $\qrand$ are generated around the sphere $c_v \in T$ with the probability $\gb$, and from the whole $\C$ otherwise.
@@
 To generate the samples $\qrand$ around the virtual goal $v$, the translation part $(x,y,z)$ of $\qrand$ is generated
-from $N(c_v,\Sigma)$, where $\Sigma$ is the diagonal matrix with diagonal entries equal to the parameter $\rv$, and the rotational
+from the Gaussian distribution $N(c_v,\Sigma)$, where $\Sigma$ is the diagonal matrix with diagonal entries equal to the parameter $\rv$, and the rotational
 part of $\qrand$ is generated using techniques described in~\cite{kuffnerES}.
@@
 The random samples are generated similarly as in the case of $\qrand$ samples, only their translation
-part is generated around $\qnear$.
+part $(x,y,z)$ is generated around $\qnear$.
 
@@
 The fast evaluation of this metric with complexity $O(\log n)$, where $n$ is the number of atoms of the ligand, can be achieved using KD-trees.
-For each conformation $\l \in \L$ of the ligand, a KD-tree is build for the ligand at configuration $q=(0,0,0,0,0,0,l,\smax)$
+For each conformation $\l \in \L$ of the ligand, a KD-tree is built for the ligand at the configuration $q=(0,0,0,0,0,0,l,\smax)$
 (i.e., for ligand with no translation and no rotation).
-To find a nearest atom to $\qrand$ for a ligand at the configuration $q=(x,y,z,r_x,r_y,r_z,l,s)$, the translation vector $T=(x,y,z)$ and
+To find the nearest atom to $\qrand$ for a ligand at the configuration $q=(x,y,z,r_x,r_y,r_z,l,s)$, the translation vector $T=(x,y,z)$ and
 rotation matrix $R$ given by $r_x$, $r_y$ and $r_z$ are computed.
@@
 between $\qinit$ (root of the tree) and $q'$ is found, where
- $q'$ is the nearest node in the tree towards $\qgoal$ (using 3D Euclidean metric). 
-The path $P=(q_i), q_i \in \C$ is represented as a sequence of collision-free configurations.
+ $q'$ is the nearest node to the end of the tunnel (measured using the 3D Euclidean metric). 
+%The path $P=(q_i), q_i \in \C$ is represented as a sequence of collision-free configurations.
 The path is found in the tree even if the tree does not approach $\qgoal$ close enough.
@@
 
-The proposed planner utilizes a set of predefined ligand's conformations.
-As each of them is tested in the expansion step, this number of influences both runtime and the success rate.
+In the expansion step, each conformation from $\L$ is tested in order to approach $\qrand$.
 By using more conformations, there is a bigger chance that at least one of them can expand the tree towards the random sample $\qrand$.
-However, using large number of conformations also increase the time complexity of the planner.
-Therefore, it is important so select a `good' set of conformations.
-Depending on the ligand, some conformations are more packed i.e., the atoms in that conformations are close to each other, but 
-conformations where the atoms form a longer molecule are also possible.
+On the other hand, using a large number of conformations also increases the time complexity of the expansion step.
+Selection of a suitable set of conformations $\L$ is therefore crucial.
+Depending on the ligand, some conformations can be considered as more packed i.e., the atoms in such conformations are close to each other, but 
+the atoms can also form a more longer shape (Fig.~\ref{fig::m040c}).
 
@@
 The `longer' conformations can better fit in a narrow passage, while the `packed' conformations can be better rotated at place 
-(Fig.~\ref{fig::m040c}.
 To select the suitable conformations, a simple technique is proposed.  
@@
 \caption{\label{fig::m040c}
- Examples of `long' conformation (left) and a `packed' one for the ligand m040 (1,5-Dichloropentane).
+ The examples of a `long' conformation (left) and a `packed' one for the ligand m040 (1,5-Dichloropentane).
 }
@@
 
-The proposed approach was used to analyze tunnel traversability towards to active site of Haloalkane dehalogenase protein (PDB ID 1CQW).
+The proposed approach was used to analyze tunnel traversability towards the active site of Haloalkane dehalogenase protein (PDB ID 1CQW).
 The active site was defined using residues 38 (Asparagine), 102 (Histidine) and 103 (Aspartate).
-Molecular dynamics of the protein was computed using Amber tool.
+The molecular dynamics of the protein was computed using the Amber tool.
 %100 frames of the dynamics were analyzed with the aim to calculate percentage of frames where the active site was accessible through the tunnels.
-In each frame, tunnels for the spherical probe 0.9~\AA were detected using CAVER 3.0~\cite{caver3}. 
+In each frame, tunnels for the spherical probe 0.9~\AA\ were detected using CAVER 3.0~\cite{caver3}. 
 In most of the frames, two tunnels were detected.
-The average bottleneck of the tunnels are $1.48$~\AA and $1.2$~\AA for the first, resp. the second tunnel.
+The average bottleneck of the tunnels are $1.48$~\AA\ and $1.2$~\AA\ for the first and the second tunnel, resp.
 %The second tunnel is more dynamic, i.e., it fluctuatest more
@@
 
-Two motion planners were utilized: the proposed planner (Alg.~\ref{alg::main} and Alg.~\ref{alg::expand}), denoted as \RA\ in the rest of the paperm, and the Retraction-based RRT~\cite{zhangRetraction} denoted as \RB\ in the rest of the paper.
+Two motion planners were utilized: the proposed planner (described in Alg.~\ref{alg::main} and Alg.~\ref{alg::expand}), 
+    denoted as \RA\ in the rest of the paper, and the Retraction-based RRT~\cite{zhangRetraction} denoted as \RB\ in the rest of the paper.
 The \RA\ method was run with the following parameters:
-$\Imax=10,000$, $m=100$, $\gb=0.9$, $\dt=1.5$\AA, $\rv=2$~\AA.
+$\Imax=5,000$, $m=100$, $\gb=0.9$, $\dt=1.5$\AA, $\rv=2$~\AA.
 Both methods were implemented in the same software framework and they utilized the same data structures and routines for collision detection and nearest-neighbor search.
@@
 
+In the following experiments, we measure performance of the methods using the traversability rate, which is the percentage of the protein dynamic frames that are traversable.
+A frame is considered traversable, if at least one trajectory out of 20 (the number of initial configurations $|\QI|$) \times 10 trials approaches the active site to the distance less than 2~\AA.
+
 \subsection{Traversability of one atom}
@@
 
-The results are shown in Tab.~\ref{tab::one} which shows percentage of MD frames where the sphere traversed the whole tunnel 
-(i.e., when it reached the active site).
-As the bottlenecks of the tunnels in each frame is known, we can compute the percentage of frames that have the bottleneck larger
-than the tested sphere radius.
-The first line in the Tab.~\ref{tab::one} therefore show the maximum percentage of traversable frames, while the second
-line (denote Planners) shows how both planners performed in this scenario.
-We can conclude, that the performance of both methods is similar in the case of first tunnel.
-In the case of the second tunnel, \RA\ was able to find trajectory in more frames than \RB.
+The results are shown in Tab.~\ref{tab::one} which shows traversability rate. 
+The maximum percentage of traversable frames (the row Maximum in Tab.~\ref{tab::one}) is computed as the percentage of tunnels with bottleneck larger than the tested radius.
+The row Planners shows traversable rate detected by the planners.
+This rate decreases with the increasing radius and it is lower than the maximum traversable rate.
+Higher traversability rate can be achieved by increasing the maximum number of planning iterations $\Imax$, which would also
+increases the planning time.
+For the purpose of the following experiments, we used $\Imax=5,000$ as a compromised between the performance and the runtime.
+The rates provided by \RA\ and \RB\ are similar in this case.
 
@@
 The `P' method selects such conformations where the deviation of the atomic distance from the geometric center are lowest, which leads 
-to selection of the packed conformations (Fig.~\ref{m040c}).
-
-The percentage of frames that were traversible by the ligands are shown in Tab.~\ref{tab::selection}, where the subscript of each ligand
+to selection of the packed conformations (Fig.~\ref{fig::m040c}).
+
+The traversability rates are shown in Tab.~\ref{tab::selection}, where the subscript of each ligand
 name shows the number of selected conformations (5 or 50) and the used method (S or P).
-The number after the slash in the ligand's name denote the number of atoms of the ligand.
+The number after the slash in the ligand's name denotes the number of atoms of the ligand.
 In the case of both tested ligands and 50 selected conformations, the `S' strategy leads to higher traversability rate than the `P' strategy.
-For the ligand DCP, a higher traversability rates was achieved when 50 conformations are selected.
+For the ligand DCP, higher traversability rates was achieved when 50 conformations were selected.
 This experiments shows that selecting 50 conformations using the `S' strategy (Sec.~\ref{sec::strat}) leads to better results
@@
 
-The results are summarized in Tab.~\ref{tab::main}, which shows percentage of frames in which the tested planners \RA\ and \RB\ found
-trajecotries close to the active site. 
-The active site was considered as reached if the distance from the center of the ligand to the active site was less than 2~\AA.
-
+The results are summarized in Tab.~\ref{tab::main}.
 The tables shows percentages of traversable MD snapshots only for the first tunnel, as no trajectories were found in the second tunnel
 for most of the tested ligand.
-The only except is the ligand m037t, which was the smallest tested ligand (with 8 atoms), and therefore it can fit even inside the second tunnel.
-
-The proposed \RA\ planner outperforms the \RB\ planner in the number of frames where a solution was found.
+The only exception is the ligand m037t, which was the smallest tested ligand (with 8 atoms), and therefore it can fit even inside the second tunnel.
+
+The proposed \RA\ planner provides higher traversability rate than \RB.
 The distances of the configuration trees (measured using 3D Euclidean metric) to the active site and the runtimes in each frame
 are shown in Fig.~\ref{fig::comparison}.
-The runtimes of both algorithms increases in those frames, where the tunnel is not traversable, i.e., where the distance
+The runtimes of both algorithms increase in those frames, where the tunnel is not traversable, i.e., where the distance
 to the active site is more than 2~\AA (e.g., the frame number 60).
+
+The most time consuming part of the planners is the collision detection. 
+Although both methods can utilize same amount of collision detection queries in the expansion step, \RB\ is slower than \RA.
+The reason is, that \RA\ expands the trees more closer to the random samples $\qrand$ than \RB.
+This results in faster exploration of the conifguration space and therefore, \RA\ needs less number of iterations to reach the goal
+configuration.
+We explain this behavior
 

### 2017-09-09 paper.tex
missing $$ for times

@@
 In the following experiments, we measure performance of the methods using the traversability rate, which is the percentage of the protein dynamic frames that are traversable.
-A frame is considered traversable, if at least one trajectory out of 20 (the number of initial configurations $|\QI|$) \times 10 trials approaches the active site to the distance less than 2~\AA.
+A frame is considered traversable, if at least one trajectory out of 20 (the number of initial configurations $|\QI|$) $\times$ 10 trials approaches the active site to the distance less than 2~\AA.
 
@@
     Traversability of the tunnels for ligands with 50 conformations. 
-    The number after $/$ denotes the number of atoms.
+    The number after '$/$' denotes the number of atoms.
 }

### 2017-09-09 paper.tex
experiments, removed tunnel number from table

@@
 It can therefore be expected, that the detected tunnels are rather narrow.
-Based on the observation from MD simulations, the tunnels adapt to the ligands (and vice versa), so even narrow tunnels can serve as transportation path for thhe ligands TODO CITACE??.
+Based on the observation from MD simulations, the tunnels adapt to the ligands (and vice versa), so even narrow tunnels can serve as transportation path for thee ligands TODO CITACE??.
 To enable computing of trajectories in the narrow tunnels, we shrink the atom radii of the ligand, similarly to~\cite{cortes2010simulating,guieysse2008structure}.
@@
 $q$ being tested.
-This metric is denoted $\da(\cdot,\cdot)$ on line~\ref{alg::expand:a} in Alg.~\ref{alg::expand}.
+This metric is denoted $\da(\cdot,\cdot)$ on line~\ref{alg::expand:a} in Alg.~\ref{alg::expand} and it is referred to as atom-based metric in the rest of the paper.
 As the expansion routine attempts to find such sample that minimizes this distance to $\qrand$, it supports retraction of the ligand towards $\qrand$.
@@
 Both methods were implemented in the same software framework and they utilized the same data structures and routines for collision detection and nearest-neighbor search.
-The Retraction-based RRT was implemented using 6D Euclidean metric.
+The nearest-neighbor search of both planners is based on the 6D Euclidean metric with the same weights for rotation and translation parts.
+This metric was selected based on preliminary experiments with both planners as it resulted in the higher success rates.
+
 The tested methods differ in the expansion step (lines \ref{alg::expa}--\ref{alg::expb} in Alg.~\ref{alg::expand}).
@@
 For each conformation and tested scale, the \RA\ methods attempts to find $m=100$ collision-free configurations, and \RB\ attempts to perform $5 \times 20$ collision-detection queries.
+Other parameters of the \RB\ planner were same as for \RA.
+Planning was terminated after $\Imax$ iterations, or if the tree approached the active site to less than 0.5~\AA.
 
@@
 \centering
-\caption{\label{tab::one}Traversability for one-atome}
+\caption{\label{tab::one}Traversability for one atom.}
 \small
@@
 \caption{\label{tab::selection}
-    Traversability of the tunnels depending on conformation selection.
-    The number after $/$ denotes the number of atoms.
+    Influence of the conformation selection method to the traversability rates in the first tunnel.
+    The number after~`$/$' denotes the number of atoms.
 }
@@
 The only exception is the ligand m037t, which was the smallest tested ligand (with 8 atoms), and therefore it can fit even inside the second tunnel.
+The traversability rate is higher for the most scaled-down ligands ($\smin=0.5$) and it decreases with the increasing scale.
+For the scale $\smin=0.8$, the most traversable frames were detected for ligands DCP (8~\%), m037t (20~\%) and m038t (16~\%).
+
+{\bf doufam, ze toto neni moc drze:}
+The traversability of the DCP ligand to active site was approved by the MD simulation.
+The observed trajectory of the DCP ligand followed the first tunnel.
+From these results, we expect that at least ligands m037t and m038t can also reach the active site, but the MD simulations
+haven't been performed yet.
 
@@
 Although both methods can utilize same amount of collision detection queries in the expansion step, \RB\ is slower than \RA.
-The reason is, that \RA\ expands the trees more closer to the random samples $\qrand$ than \RB.
-This results in faster exploration of the conifguration space and therefore, \RA\ needs less number of iterations to reach the goal
-configuration.
-We explain this behavior
+The reason is, that \RA\ expands the tree more closer to the random samples $\qrand$ than \RB.
+This results in a faster exploration of the configuration space and therefore, \RA\ needs less number of iterations to reach the active site.
+The better performance of \RA\ is caused by the proposed atom-based metric, because this distance between two configurations depends on the actual shape of the ligand.
+This shape is different in each configuration.
+Contrary, the \RB\ utilizes only the single 6D Euclidean metric which is same for all conformations.
+To improve performance of \RB, the weights of the 6D Euclidean metric should differ for each conformation, but it would
+prevent to use KD-tree for the nearest-neighbor search.
 
@@
 \caption{\label{fig::comparison}
-    Runtimes and distance to the active site for the m037 ligand at scale 0.4.
+    Runtimes and distance to the active site for the m037 ligand at scale 0.6.
     The values are computed from all 200 trials (20 initial configurations $\times$ 10 trials).
@@
 
-\section{Discussion}
-
-The computed trajectories cannot be considered as `real', they should rather be considered as a hint for the biochemists.
-One of the reason is that the trajectories are computed inside a static protein and static tunnels.
-Real proteins are dynamic structures and consequently also the tunnels are dynamic: they move, merge or even disappear due to motions of protein atoms.
-Contrary, usually static tunnels are used in protein engineering.
-Despite such a simplification, researches often used information about the static tunnels  to estimate the possibility of chemical reactions
-at the active sites.
+\section{Discussion - TODO prepsat}
+
+%The computed trajectories cannot be considered as `real', they should rather be considered as a hint for the biochemists.
+%One of the reason is that the trajectories are computed inside a static protein and static tunnels.
+%Real proteins are dynamic structures and consequently also the tunnels are dynamic: they move, merge or even disappear due to motions of protein atoms.
+%Contrary, usually static tunnels are used in protein engineering.
+%Despite such a simplification, researches often used information about the static tunnels  to estimate the possibility of chemical reactions
+%at the active sites.
 

### 2017-09-09 paper.tex
link to web page

@@
 \usepackage{flushend}
-%\usepackage[bookmarks=false]{hyperref}
+\usepackage[bookmarks=false]{hyperref}
 %\usepackage{setspace}
@@
 %is important to understand interactions between proteins and other molecules (ligands)~\cite{Koudelakova2013}.
-As the interaction takes place in the active site, the existence of a tunnel connection the active site with the outer environment
+As the interaction takes place in the active site, the existence of a tunnel connecting the active site with the outer environment
 is necessary for the interaction.
@@
 In this paper, we propose a modification to RRT to cope with these issues.
-In the proposed method, the random samples are not generated from the whole configuration space, but the sampling is guided using
+The random samples are not generated from the whole configuration space, but the sampling is guided using
 the spheres of the tunnel.
-This allows us to find trajectories along the given tunnel, but it also decreases the volume of the configuration space to be searched.
+This allows us to find trajectories along the given tunnel, which is faster than searching any trajectory in the whole protein.
 To increase movements of the ligands in the narrow tunnels, the free-space is dilated by 
 shrinking the atom radii, similarly, e.g., to~\cite{cortes2005path,hsu06multilevel}.
-We model the ligand flexibility using a fixed set of predefined conformations.
+The ligand flexibility is modelled using a fixed set of predefined conformations.
 
@@
 In each frame, tunnels for the spherical probe 0.9~\AA\ were detected using CAVER 3.0~\cite{caver3}. 
-In most of the frames, two tunnels were detected.
+In most of the frames, two tunnels were detected.~\footnote{%
+We refer to~\url{http://mrs.felk.cvut.cz/icra2018protein} for additional images.
+}
 The average bottleneck of the tunnels are $1.48$~\AA\ and $1.2$~\AA\ for the first and the second tunnel, resp.
-%The second tunnel is more dynamic, i.e., it fluctuatest more
 
 Four different scaling-down factors were used: $\S=\{0.5,0.6,0.7,0.8\}$.
-For each scale, 20 collision-free initial configurations were found in the $\RI=2$\AA\ radius around the first sphere of the tunnel (outside the protein).
+For each scale, 20 collision-free initial configurations were found in the $\RI=2$~\AA\ radius around the first sphere of the tunnel (outside the protein).
 For each ligand, each minimal scale $\smin$ and each initial configuration, 10 trajectories were computed.
@@
 The \RA\ method was run with the following parameters:
-$\Imax=5,000$, $m=100$, $\gb=0.9$, $\dt=1.5$\AA, $\rv=2$~\AA.
+$\Imax=5,000$, $m=100$, $\gb=0.9$, $\dt=1.5$~\AA, $\rv=2$~\AA.
 Both methods were implemented in the same software framework and they utilized the same data structures and routines for collision detection and nearest-neighbor search.
@@
 increases the planning time.
-For the purpose of the following experiments, we used $\Imax=5,000$ as a compromised between the performance and the runtime.
+For the purpose of the following experiments, we used $\Imax=5,000$ as a compromise between the performance and the runtime.
 The rates provided by \RA\ and \RB\ are similar in this case.
@@
 \caption{\label{tab::one}Traversability for one atom.}
-\small
 \renewcommand{\tabcolsep}{3.5pt}
@@
 In the case of both tested ligands and 50 selected conformations, the `S' strategy leads to higher traversability rate than the `P' strategy.
-For the ligand DCP, higher traversability rates was achieved when 50 conformations were selected.
+For the ligand DCP, a higher traversability rate was achieved when 50 conformations were selected.
 This experiments shows that selecting 50 conformations using the `S' strategy (Sec.~\ref{sec::strat}) leads to better results
@@
 The most time consuming part of the planners is the collision detection. 
-Although both methods can utilize same amount of collision detection queries in the expansion step, \RB\ is slower than \RA.
+Although the both methods can utilize the same number of collision detection queries in the expansion step, \RB\ is slower than \RA.
 The reason is, that \RA\ expands the tree more closer to the random samples $\qrand$ than \RB.
@@
 To improve performance of \RB, the weights of the 6D Euclidean metric should differ for each conformation, but it would
-prevent to use KD-tree for the nearest-neighbor search.
+prevent the usage KD-tree for the nearest-neighbor search.
 

### 2017-09-10 paper.tex
image of dcp trajectories + tunnel from Adam
added colormap form Adam
atom-based metric

@@
 In Retraction-based RRT~\cite{zhangRetraction}, the tree is retracted along the boundary of the obstacles in the configuration space.
-This requires the analysis of contact space, which leads to the generalized penetration depth computation of high complexity~\cite{zhang2008fast,he2016efficient}. %\cite{he2016efficient} in RAL!!
+This requires the analysis of contact space, which leads to the generalized penetration depth computation of high complexity~\cite{he2016efficient}. %\cite{he2016efficient} in RAL!! + zhang2008fast
 Selective Retraction-based RRT~\cite{lee2012srrrt} avoids the expensive growth in the open areas of the configuration space and focuses 
@@
 
-In this paper, we propose a modification of the RRT planner for computing trajectories through protein tunnels.
-The input is a set of tunnels computed by Voronoi diagrams and the task is to find trajectories along the given tunnels
+We propose a modification of the RRT planner for computing trajectories through protein tunnels.
+The input is a set of tunnels computed using Voronoi diagrams and the task is to find trajectories along the given tunnels
 leading from outside the protein to the active site.
@@
 This metric measures distances between two configurations as the nearest distance of two atoms. 
-This enables the metric to consider the shape of the ligand and it support retraction of the ligand to the narrow passages.
+This enables the metric to consider the shape of the ligand and it supports retraction of the ligand to the narrow passages.
 
@@
 Proteins and ligands are represented by the hard sphere model, where the radius of each sphere (atom) is given by its van der Waals radius.
-
 A protein tunnel is described by a sequence of collision-free spheres 
@@
 by the 3D position $(x,y,z)$ of the reference point of the ligand (the geometric center), the rotation around $x$, $y$, and $z$ axes,
-the index of the conformation $l\in \L$ and the actual scale $s \in \S$.
+the index of the conformation $l\in \L$ and the scale $s \in \S$.
 All possible configurations form the configuration space $\C$ and the collision-free configurations
@@
 Depending on the chemical interactions between the ligand and the protein surface, the ligand may enter a tunnel with various rotations and in various conformations.
-The traversability of the tunnel should therefore be evaluated using various starting configurations $\QI$.
+The traversability of the tunnel should therefore be evaluated using multiple starting configurations $\QI$.
 The initial configurations are searched around the beginning of the tunnel.
@@
 
-\subsection{RRT for computing ligands trajectory inside a tunnel}
-
-The proposed motion planning is based on the principle of RRT~\cite{lavalleRRT} planner.
-The main loop of planner (Alg.~\ref{alg::main}) is similar to the original RRT planner.
+\subsection{RRT for computing ligands trajectory along a tunnel}
+
+The proposed motion planning is based on the principle of the RRT~\cite{lavalleRRT} planner.
 In each iteration, a random sample $\qrand$ is generated and its nearest node $\qnear\in\T$ in the tree is found.
-The nearest-neighbor search between $\qrand$ and the tree is performed using the weighted 6D Euclidean metric considering
+The nearest-neighbor search between $\qrand$ and the tree is performed using the weighted 6D Euclidean metric computed using
 both 3D rotation and 3D translation.
+The main loop of the planner (Alg.~\ref{alg::main}) is similar to original RRT, but we propose three modifications.
 
@@
 Setting the virtual goal to this successor allows the tree to avoid such parts of the tunnels that are not traversable or reachable by the ligand.
-%This is necessary in the dense protein structures, where it is not always possible to exactly follow the tunnel.
-This allows the tree to slightly detour from the tunnel, which improves performance around narrow tunnels.
+This allows the tree to slightly detour from the tunnel and overcome its difficult (narrow) parts by finding an alternative trajectory nearby.
 The algorithm terminates after a predefined number of planning trials $\Imax$ or if the tree reaches
@@
 metric for the nearest-neighbor search.
-The parameter $\rv$ influences the distribution of samples around the tunnel centerline. 
+The parameter $\rv$ influences the distribution of the random samples around the tunnel centerline. 
 By setting $\rv$ to a small value, the planner attempts to find the trajectories inside the tunnel, while higher values
 of $\rv$ cause  exploration of paths around the tunnel.
-We propose to set this parameter to the average tunnel width.
+We propose to set this parameter to the average width of the tunnel.
 
@@
 
-The second proposed modification of RRT is the expansion procedure (Alg.~\ref{alg::expand}), which generates new collision-free nodes around $\qnear\in\T$.
+The second proposed modification is the expansion procedure (Alg.~\ref{alg::expand}), which expands the tree from $\qnear$ towards $\qrand$.
+The expansion towards $\qrand$ is crucial for the growth of the tree towards unexplored areas of the configuration space.
+Traditionally, a line is constructed from $\qnear$ to $\qrand$ and the tree is expanded by a configuration located on this line.
+This would however mostly lead to a collision in the case of ligands.
+
+Instead, we propose to expand the tree by considering all conformations and also by expanding the tree using the maximal possible scale that
+does not lead to a collision.
 For each ligand conformation $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with a maximal scale.
-First, the maximal scale $\smax \in \S$ is selected and $m$ random samples are generated around $\qrand$ and tested for collisions.
+First, the maximal scale $\smax \in \S$ is selected and $m$ random samples are generated around $\qnear$ and tested for collisions.
 The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
@@
 
-The third modification is the utilization of another metric during the expansion procedure.
-While the nearest neighbors are search using 6D Euclidean metric, which can be efficiently realized using KD-trees, the expansion
-procedure utilizes a metric that is sensitive to the shape of the ligands.
-This is necessary as the ligands adopt various shapes according to the actual conformations.
-To minimize the distance between $\qrand$ and the ligand, the shape of the ligand should be considered.
-
-Therefore, we propose to measure the distance as the smallest 3D Euclidean distance between atoms placed at $\qrand$ and a configuration
-$q$ being tested.
-This metric is denoted $\da(\cdot,\cdot)$ on line~\ref{alg::expand:a} in Alg.~\ref{alg::expand} and it is referred to as atom-based metric in the rest of the paper.
-As the expansion routine attempts to find such sample that minimizes this distance to $\qrand$, it supports retraction of the ligand towards $\qrand$.
-The fast evaluation of this metric with complexity $O(\log n)$, where $n$ is the number of atoms of the ligand, can be achieved using KD-trees.
+The third modification is the utilization of another metric during the expansion procedure (line~\ref{alg::expand:a} in Alg.~\ref{alg::expand}).
+This metric is denoted $\da(\cdot,\cdot)$ and it is referred to as atom-based metric in the rest of the paper.
+To evaluate $\da(q_1,q_2)$, the absolute positions of atoms of the ligand defined by $q_1$ and $q_2$ are computed.
+The minimal 3D Euclidean distance from the first set of atoms (defined by $q_1$) to the second set of atoms (defined by $q_2$) is the value of 
+$\da(q_1,q_2)$.
+By computing the atom-based metric using the absolute positions of atoms, this metric considers the shape of the ligands, which
+is useful when multiple conformations are utilized.
+By employing this metric in the expansion procedure, the tree is expanded by such a new configurations, whose atoms maximally approach
+the atoms of the configuration $\qrand$.
+We observed that this leads to a better construction of the configuration tree inside the proteins and it increases probability
+of reaching the active site.
+
+%While the nearest neighbors are searched using 6D Euclidean metric, which can be efficiently realized using KD-trees, the expansion
+%procedure utilizes a metric that is sensitive to the shape of the ligands.
+%This is necessary as the ligands adopt various shapes according to the actual conformations.
+%To minimize the distance between $\qrand$ and the ligand, the shape of the ligand should be considered.
+
+%Therefore, we propose to measure the distance as the smallest 3D Euclidean distance between atoms placed at $\qrand$ and a configuration $q$ being tested.
+The atom-based metric can be evaluated fast with complexity $O(\log n)$, where $n$ is the number of atoms of the ligand, using KD-trees.
 For each conformation $\l \in \L$ of the ligand, a KD-tree is built for the ligand at the configuration $q=(0,0,0,0,0,0,l,\smax)$
@@
 rotation matrix $R$ given by $r_x$, $r_y$ and $r_z$ are computed.
-Then, $\qrand$ is translated to $\qrand' = R^{-1}(\qrand-T)$ and the KD-tree search is used to find nearest point towards $\qrand'$, which
-is also the nearest neighbor of the ligand at $q$ to $\qrand$.
+Then, $\qrand$ is translated to $\qrand' = R^{-1}(\qrand-T)$ and the KD-tree search is used to find the nearest point towards $\qrand'$, which
+is also the nearest point of the ligand at $q$ to $\qrand$.
 
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=0.4\textwidth]{fig/dcp-image1Label}\\
+\includegraphics[width=0.2\textwidth]{fig/colormap}
+\caption{\label{fig::dcp}
+Visualization of trajectories for the ligand DCP   in the two examined tunnels.  
+    The color of the trajectories corresponds to the scale, that was needed to pass the given part of the tunnel.
+    For the purpose of this image, also scales $\smin=0.3$ and $\smin=0.4$ where tested.
+}
+\end{figure}
+
 \section{Discussion - TODO prepsat}

### 2017-09-10 paper.tex
selection of conformations

@@
     \eIf{$rand() < \gb$}{
-        $\qrand$ = random sample around actual virtual goal $c_v \in T$\;
+        $\qrand$ = random sample around virtual goal $c_v\!\in\!T$\;
     }{
@@
 This metric is denoted $\da(\cdot,\cdot)$ and it is referred to as atom-based metric in the rest of the paper.
-To evaluate $\da(q_1,q_2)$, the absolute positions of atoms of the ligand defined by $q_1$ and $q_2$ are computed.
+To evaluate $\da(q_1,q_2)$, the absolute positions of the atoms of the ligand defined by $q_1$ and $q_2$ are computed.
 The minimal 3D Euclidean distance from the first set of atoms (defined by $q_1$) to the second set of atoms (defined by $q_2$) is the value of 
@@
 is useful when multiple conformations are utilized.
-By employing this metric in the expansion procedure, the tree is expanded by such a new configurations, whose atoms maximally approach
+By employing this metric in the expansion procedure, the tree is expanded by such a new configuration, which atoms maximally approach
 the atoms of the configuration $\qrand$.
-We observed that this leads to a better construction of the configuration tree inside the proteins and it increases probability
+We observed that this leads to a better construction of the configuration tree inside the proteins and it increases the probability
 of reaching the active site.
@@
 %Therefore, we propose to measure the distance as the smallest 3D Euclidean distance between atoms placed at $\qrand$ and a configuration $q$ being tested.
-The atom-based metric can be evaluated fast with complexity $O(\log n)$, where $n$ is the number of atoms of the ligand, using KD-trees.
+The atom-based metric can be evaluated fast using KD-trees with complexity $\mathcal{O}(\log n)$, where $n$ is the number of atoms of the ligand.
 For each conformation $\l \in \L$ of the ligand, a KD-tree is built for the ligand at the configuration $q=(0,0,0,0,0,0,l,\smax)$
@@
 Considering also these non-feasible solutions is necessary to evaluate difficult areas of the tunnels, e.g. bottlenecks.
-The utilization of all computed paths for evaluation of tunnel difficulty is described in the next section.
+%The utilization of all computed paths for evaluation of tunnel difficulty is described in the next section.
 
@@
 In the expansion step, each conformation from $\L$ is tested in order to approach $\qrand$.
-By using more conformations, there is a bigger chance that at least one of them can expand the tree towards the random sample $\qrand$.
+By using more conformations, there is a higher chance that at least one of them can expand the tree towards the random sample $\qrand$.
 On the other hand, using a large number of conformations also increases the time complexity of the expansion step.
@@
 From motion planning point of view, both types of conformations are useful.
-The `longer' conformations can better fit in a narrow passage, while the `packed' conformations can be better rotated at place 
-To select the suitable conformations, a simple technique is proposed.  
+The longer conformations can better fit in a narrow passage, while the packed conformations can be rotated more at a place.
+In the narrow parts of the tunnels, the tree can be extended by the longer conformations, while in the more open regions, also the
+packed configurations can be used.
+
+To select the suitable set $\L$ of conformations from a pool of available conformations, a simple technique is proposed.  
 For each conformation, the geometric center is calculated and the deviation of the distances of the atoms from the center is calculated.
-The conformations are sorted according to this deviation and then the desired number of conformations are selected from this sorted list uniformly.
+The low value of the deviation is characteristic for the packed conformations, while larger deviations correspond to the longer conformations.
+To select $n$ conformations from all $m$ available conformations, they are sorted according to the deviation.
+Each $k-$th conformation, where $k={m\over n}$ is added to $\L$.
 

### 2017-09-10 paper.tex
experiments text updated, graphProbe regenerated

@@
 The molecular dynamics of the protein was computed using the Amber tool.
-%100 frames of the dynamics were analyzed with the aim to calculate percentage of frames where the active site was accessible through the tunnels.
 In each frame, tunnels for the spherical probe 0.9~\AA\ were detected using CAVER 3.0~\cite{caver3}. 
-In most of the frames, two tunnels were detected.~\footnote{%
-We refer to~\url{http://mrs.felk.cvut.cz/icra2018protein} for additional images.
-}
+In most of the frames, two tunnels were detected.
 The average bottleneck of the tunnels are $1.48$~\AA\ and $1.2$~\AA\ for the first and the second tunnel, resp.
+The tunnel traversability was analyzed in 100 frames.
 
 Four different scaling-down factors were used: $\S=\{0.5,0.6,0.7,0.8\}$.
-For each scale, 20 collision-free initial configurations were found in the $\RI=2$~\AA\ radius around the first sphere of the tunnel (outside the protein).
-For each ligand, each minimal scale $\smin$ and each initial configuration, 10 trajectories were computed.
+For each scale, $|\QI|=20$ collision-free initial configurations were found in the $\RI=2$~\AA\ radius around the first sphere of the tunnel (outside the protein).
+For each frame, minimal scale $\smin$ and each initial configuration, 10 trajectories were computed, which resulted
+in $100 \times 4 \times 20 \times 10 = 80,000$ trajectories per ligand.
 
@@
 The \RA\ method was run with the following parameters:
-$\Imax=5,000$, $m=100$, $\gb=0.9$, $\dt=1.5$~\AA, $\rv=2$~\AA.
+$\Imax=5,000$, $m=100$, $\gb=0.9$, $\dt=1.5$~\AA, $\rv=3$~\AA.
 Both methods were implemented in the same software framework and they utilized the same data structures and routines for collision detection and nearest-neighbor search.
@@
 In these lines, the \RB\ method employs the retraction step in order to find a new collision-free configurations $\qnew$ that approaches $\qrand$.
-The retraction step was realized by sampling the vicinity of the configuration $\qnear$ in the distance $0.1$~\AA\ using 20 samples.
-A collision-free sample that maximally approached $\qrand$ was selected and added to the tree, and the retraction continued from this sample.
-The retraction step was repeated 5 times.
+The retraction step is realized by sampling the vicinity of the configuration $\qnear$ in the distance $0.1$~\AA\ using 20 samples.
+A collision-free sample that maximally approaches $\qrand$ is selected and added to the tree, and the retraction continues from this sample.
+The retraction step is repeated 5 times.
 The \RB\ method can therefore expand the tree by at most 5 new nodes per conformation, while \RA\ extends the tree only by one configuration.
@@
 
-In the following experiments, we measure performance of the methods using the traversability rate, which is the percentage of the protein dynamic frames that are traversable.
-A frame is considered traversable, if at least one trajectory out of 20 (the number of initial configurations $|\QI|$) $\times$ 10 trials approaches the active site to the distance less than 2~\AA.
+In the following experiments, we measure performance of the methods using the traversability rate, which is the percentage of frames of the protein dynamic that are traversable by the tested ligand.
+A frame is considered traversable, if at least one trajectory out of 20 (the number of initial configurations $|\QI|$) $\times$ 10 trials approaches the active site to the distance less than 2~\AA.~\footnote{We refer to~\url{http://mrs.felk.cvut.cz/icra2018protein} for additional material}
 
@@
 This rate decreases with the increasing radius and it is lower than the maximum traversable rate.
+The rates provided by \RA\ and \RB\ are similar which is a good indicator that the methods have been programmed correctly.
 Higher traversability rate can be achieved by increasing the maximum number of planning iterations $\Imax$, which would also
-increases the planning time.
+increase the planning time.
 For the purpose of the following experiments, we used $\Imax=5,000$ as a compromise between the performance and the runtime.
-The rates provided by \RA\ and \RB\ are similar in this case.
 
@@
 
-The traversability rates are shown in Tab.~\ref{tab::selection}, where the subscript of each ligand
+The traversability rates are shown in Tab.~\ref{tab::selection}, where the subscript of each ligand's 
 name shows the number of selected conformations (5 or 50) and the used method (S or P).
@@
 The results are summarized in Tab.~\ref{tab::main}.
-The tables shows percentages of traversable MD snapshots only for the first tunnel, as no trajectories were found in the second tunnel
-for most of the tested ligand.
+The table shows the traversability rate only for the first tunnel, as no trajectory was found in the second tunnel for most of the tested ligands.
 The only exception is the ligand m037t, which was the smallest tested ligand (with 8 atoms), and therefore it can fit even inside the second tunnel.
@@
 The traversability of the DCP ligand to active site was approved by the MD simulation.
-The observed trajectory of the DCP ligand followed the first tunnel.
-From these results, we expect that at least ligands m037t and m038t can also reach the active site, but the MD simulations
-haven't been performed yet.
+The observed trajectory of the DCP ligand in the MD simulation followed the first tunnel.
+Based on the traversability rates with $\smin=0.8$, we expect that at least ligands m037t and m038t can also reach the active site, as
+they have same (or even larger) traversability rate than DCP.
+However, this hypothesis hasn't been tested agains MD simulations yet.
 

### 2017-09-10 paper.tex
final image

@@
 The color of the trajectories corresponds to the scale, that was needed to pass the given part of the tunnel.
-    For the purpose of this image, also scales $\smin=0.3$ and $\smin=0.4$ where tested.
+The ligand can enter the first tunnel with a high scale (almost 1=100~\%) (A), but then it has to scale-down (to $\sim75$~\%) to pass the
+next part of the tunnel (B). After this narrow passage, the tunnel is larger and the ligand can traverse to the narrow passage with
+scale $>75$~\% (C).
+The second tunnel can also be entered with the high scale (green), but then the narrow part is traversable only with a low scale ($\sim50$~\%) (D). 
+For the purpose of the visualization, even more scale-down to 0.3=30~\% was allowed (E) to see how the ligand pass the tunnel.
+However, the ligand can reach the active site only with very low scale ($\sim0.4=40$~\%) using the second tunnel.
 }

### 2017-09-10 paper.tex
intro

--- 2017-09-10 paper.tex
@@
 \begin{abstract}
-Chemical interactions of proteins and other molecules (ligands) take place in actives sites that are usually deeply buried inside the proteins.
-The active sites are accessible through one or more pathways (called tunnels).
-Based on the tunnel properties like it's length and bottleneck, the chemists can estimate possibility of reaching the active site and therefore estimate the possibility of the interactions.
+Chemical interactions between proteins and other molecules (ligands) take place in specific locations, called active sites, that are often deeply buried inside the protein structure.
+These active sites are accessible through one or more void paths, called tunnels.
+The passage of a ligand through a tunnel is directly influenced by the tunnel geometric properties, such as its length, width, and bottleneck.
+The task of the biochemists is to evaluate these properties and thus estimate the reactivity of the protein, i.e., the possibility of the ligand to reach the protein active site.
 %The knowledge about a tunnel (e.g. the length, bottleneck or physico-chemical properties) can help chemists to estimate possible interactions between the protein and the ligand.
-Nowadays, tunnels are computed using Voronoi diagrams, which helps to predict interactions only with small spherical ligands.
-Computing motions of non-spherical ligands including conformation changes (i.e., when atoms of the ligand move relatively to each other) leads to motion planning in high-dimensional configuration space, which can be solved using sampling-based planning methods.
-This planning problem necessarily includes narrow passages, as the movements of ligands are limited by many protein atoms.
-Considering flexible ligands increases the number of degrees of freedom, and consequently the dimension of the configuration space.
-In this paper, we propose a modification to the Rapidly Exploring Random Tree planner for the purpose of trajectory computations of ligands along protein tunnels.
-To cope with the narrow passage problem, the random samples are generated around a Voronoi-based tunnel.
-The flexibility of the ligand is modeled using a fixed set of known conformations which does not increase the number of dimensions of the configuration space.
-The proposed method has been evaluated in a virtual screening scenario to estimate traversability of tunnels for several ligands.
+Nowadays, tunnels are mostly computed using Voronoi diagrams. 
+This approach helps to predict the interactions between proteins and small ligands, whose shape is approximated by a bounding sphere.
+Such approximation usually leads to highly imprecise results.
+Computing the transportation path of non-spherical ligands including their conformation changes (i.e., when the atoms of the ligand move relatively to each other) leads to the motion planning task in high-dimensional configuration space, which can be solved using sampling-based planning methods.
+This planning problem necessarily includes narrow passages, as the movements of ligands are limited by the protein atoms surrounding the ligand path.
+Considering flexible ligands increases the number of degrees of freedom, and consequently also the dimension of the configuration space.
+To address this problem, in this paper we propose a modification of the Rapidly Exploring Random Tree planner for the purpose of trajectory computations of flexible ligands, heading from the protein outside environment to its active site.
+The detected trajectory then defines the centerline of the searched tunnel.
+To cope with the narrow passage problem, the random samples are generated around a Voronoi-based tunnel, which serves us for better localization of the searched space.
+The flexibility of the ligand is modeled using a fixed set of known conformations, which does not increase the number of dimensions of the configuration space.
+The proposed method has been evaluated in a virtual screening scenario to estimate the traversability of tunnels for several ligands.
 \end{abstract}
@@
 \caption{\label{fig::motiv}
-    Tunnels in haloalkane dehalogenase (1CQW) with possible trajectories of 1-chlorpropan ligand (bottom left) and 1,2-Dichloroethane (bottom right).
+    Tunnels in the DhaA haloalkane dehalogenase protein (PDB ID 1CQW) with possible trajectories of 1-Chlorpropan ligand (bottom left) and 1,2-Dichloroethane (bottom right).
 }
@@
 
-The idea of sampling-based motion planning is to randomly sample the configuration space of the robot and connect the collision-free samples
+Reactivity of proteins with other molecules plays a crucial role in many research disciplines, including protein engineering and drug design.
+Protein reactivity is tightly connected with the presence of the void space between its amino acids. 
+This space, forming a path called tunnel, can be utilized by a small molecule (ligand) which can enter the protein inner space and react with its amino acids surrounding a specific site, called the active site~\cite{gora2013gates}.
+The size and location of tunnels determine the ligand size and shape which can enter the active site.
+%The tunnel has to be wide enough and also the physico-chemical properties of the surrounding amino acids have to be compatible with the ligand.
+Biochemists are currently using computational methods for tunnel detection in order to estimate the possibility of interactions between a protein and a ligand without the need to perform laboratory experiments (Fig.~\ref{fig::motiv}).
+The most common methods detect tunnels using Voronoi diagrams~\cite{yaffe2008,caver3} assuming a spherical ligand (probe), i.e., the ligand is approximated by a bounding sphere.
+However, as the ligands are typically non-spherical and flexible, these methods can often skip possible candidate tunnels.
+
+To provide the biochemists with a solution which detects also candidate tunnels omitted by the traditional methods, it is necessary take into account also the ligand shape and its conformation changes when the ligand moves, i.e., changes its conformation.
+Such a ligand changing the mutual position of its atoms within a molecular dynamics is denoted as a flexible ligand.
+To address this, we can formulate the task as a motion planning problem.
+It is also necessary to cope with the narrow passage problem, as the proteins are dense structures and the movements of the ligand are limited by the surrounding amino acids.
+
+The idea of the sampling-based motion planning is to randomly sample the configuration space of the robot and to connect the collision-free samples
 into a graph structure (roadmap).
@@
 %The sampling-based planners can cope with robots with many degrees of freedom (DOF).
-The two most used sampling-based techniques are Probabilistic Roadmaps (PRM)~\cite{kavrakiForPP} and Rapidly Exploring Random Trees (RRT)~\cite{lavalleRRT}.
-The sampling-based planners can cope with many-DOF (Degrees of Freedom) robots of arbitrary shape and therefore they have been used for motion planning of various robotic systems.
-Besides many applications in robotics, sampling-based planners have been used also in bio-chemistry 
+Two most commonly used sampling-based techniques are Probabilistic Roadmaps (PRM)~\cite{kavrakiForPP} and Rapidly Exploring Random Trees (RRT)~\cite{lavalleRRT}.
+The sampling-based planners can cope with many-DOF (Degrees of Freedom) robots of an arbitrary shape.
+Therefore, they have been used for motion planning of various robotic systems.
+Besides many applications in robotics, the sampling-based planners have been already used in biochemistry as well 
 %Sampling-based motion planners has been intensively studied in robotics, but they have been applied also in other research fields.
 %In bio-chemistry, the planners have been used to study
-to study loop motions~\cite{cortes2004geometric},
+to study loop motions~\cite{cortes2004geometric}
 %protein folding~\cite{amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},
@@
 
-One of the problems in protein engineering is the tunnel detection, where the task is to find a tunnel (pathway) 
-leading to a specific place (called active site) inside the protein~\cite{gora2013gates}.
+In this paper, we propose a modification to RRT to address the problem of planning a non-spherical flexible ligand passage through a given protein.
+We localize the generation of the random samples, i.e., instead of sampling the whole configuration space we focus only on the space surrounding a potential tunnel detected by the Voronoi-based method (using a smaller ligand probe in order to avoid omitting possibly feasible tunnel candidates).
+The potential tunnel is represented by a set of consecutive spheres surrounding the tunnel centerline.
+This optimization allows us to find trajectories along the given tunnel, which is faster than searching for any trajectory in the whole protein.
+To enable the movement of the ligand in the narrow parts of the tunnel, the free-space can be dilated by shrinking the atom radii, similarly, e.g., to~\cite{cortes2005path,hsu06multilevel}. 
+The ligand flexibility is modeled using a fixed set of predefined configurations which the ligand can occupy.
+
+Our proposed method was tested by the biochemists who compared the results with trajectories of ligands determined experimentally in the lab.
+
 %or protein folding combined with ligand diffusion~\cite{cortes2010simulating}.
@@
 %is important to understand interactions between proteins and other molecules (ligands)~\cite{Koudelakova2013}.
-As the interaction takes place in the active site, the existence of a tunnel connecting the active site with the outer environment
-is necessary for the interaction.
+%As the interaction takes place in the active site, the existence of a tunnel connecting the active site with the outer environment
+%is necessary for the interaction.
 %To get the ligand to the active site, a transportation tunnel connecting the protein outer environment with the active site has to exists.
-Such a  tunnel has to be wide enough and the physico-chemical properties of the surrounding amino acids have to be compatible with the ligand.
-Chemists are using the tunnel detection method in order to estimate possibility of interactions between a protein and a ligand without need to perform long Molecular Dynamics (MD) simulations or even real experiments (Fig.~\ref{fig::motiv}).
-The tunnels are detected mainly using Voronoi diagrams~\cite{yaffe2008,caver3} assuming 
-a spherical ligand (probe) and they are represented as a sequence of spheres.
-However, as the ligands typically non-spherical and flexible,  it is difficult to estimate their traversability through the tunnels computed for only one sphere.
 %Biochemists decide if a tunnel can be used to transport a given ligand mainly based on the tunnel length and bottleneck (i.e., the radius of the smallest sphere that forms the tunnel).
@@
 %The decision based only on spherical tunnels requires previous expertise in the domain and yet it may be imprecise.
-
-To provide biochemists  a better insight into the behavior of non-spherical ligands, it is necessary to compute the trajectories of the ligand considering its shape and also conformation changes (flexible ligands).
-This can be formulated as a motion planning problem.
-It is necessary to cope with the narrow passage problem, as the proteins are dense structures and the movements of the ligands are limited, and to efficiently model the flexible ligands.
-
-In this paper, we propose a modification to RRT to cope with these issues.
-The random samples are not generated from the whole configuration space, but the sampling is guided using
-the spheres of the tunnel.
-This allows us to find trajectories along the given tunnel, which is faster than searching any trajectory in the whole protein.
-To increase movements of the ligands in the narrow tunnels, the free-space is dilated by 
-shrinking the atom radii, similarly, e.g., to~\cite{cortes2005path,hsu06multilevel}.
-The ligand flexibility is modelled using a fixed set of predefined conformations.
 
@@
     \eIf{$rand() < \gb$}{
-        $\qrand$ = random sample around virtual goal $c_v\!\in\!T$\;
+        $\qrand$ = random sample around actual virtual goal $c_v \in T$\;
     }{
@@
 This metric is denoted $\da(\cdot,\cdot)$ and it is referred to as atom-based metric in the rest of the paper.
-To evaluate $\da(q_1,q_2)$, the absolute positions of the atoms of the ligand defined by $q_1$ and $q_2$ are computed.
+To evaluate $\da(q_1,q_2)$, the absolute positions of atoms of the ligand defined by $q_1$ and $q_2$ are computed.
 The minimal 3D Euclidean distance from the first set of atoms (defined by $q_1$) to the second set of atoms (defined by $q_2$) is the value of 
@@
 is useful when multiple conformations are utilized.
-By employing this metric in the expansion procedure, the tree is expanded by such a new configuration, which atoms maximally approach
+By employing this metric in the expansion procedure, the tree is expanded by such a new configurations, whose atoms maximally approach
 the atoms of the configuration $\qrand$.
-We observed that this leads to a better construction of the configuration tree inside the proteins and it increases the probability
+We observed that this leads to a better construction of the configuration tree inside the proteins and it increases probability
 of reaching the active site.
@@
 %Therefore, we propose to measure the distance as the smallest 3D Euclidean distance between atoms placed at $\qrand$ and a configuration $q$ being tested.
-The atom-based metric can be evaluated fast using KD-trees with complexity $\mathcal{O}(\log n)$, where $n$ is the number of atoms of the ligand.
+The atom-based metric can be evaluated fast with complexity $O(\log n)$, where $n$ is the number of atoms of the ligand, using KD-trees.
 For each conformation $\l \in \L$ of the ligand, a KD-tree is built for the ligand at the configuration $q=(0,0,0,0,0,0,l,\smax)$
@@
 Considering also these non-feasible solutions is necessary to evaluate difficult areas of the tunnels, e.g. bottlenecks.
-%The utilization of all computed paths for evaluation of tunnel difficulty is described in the next section.
+The utilization of all computed paths for evaluation of tunnel difficulty is described in the next section.
 
@@
 In the expansion step, each conformation from $\L$ is tested in order to approach $\qrand$.
-By using more conformations, there is a higher chance that at least one of them can expand the tree towards the random sample $\qrand$.
+By using more conformations, there is a bigger chance that at least one of them can expand the tree towards the random sample $\qrand$.
 On the other hand, using a large number of conformations also increases the time complexity of the expansion step.
@@
 From motion planning point of view, both types of conformations are useful.
-The longer conformations can better fit in a narrow passage, while the packed conformations can be rotated more at a place.
-In the narrow parts of the tunnels, the tree can be extended by the longer conformations, while in the more open regions, also the
-packed configurations can be used.
-
-To select the suitable set $\L$ of conformations from a pool of available conformations, a simple technique is proposed.  
+The `longer' conformations can better fit in a narrow passage, while the `packed' conformations can be better rotated at place 
+To select the suitable conformations, a simple technique is proposed.  
 For each conformation, the geometric center is calculated and the deviation of the distances of the atoms from the center is calculated.
-The low value of the deviation is characteristic for the packed conformations, while larger deviations correspond to the longer conformations.
-To select $n$ conformations from all $m$ available conformations, they are sorted according to the deviation.
-Each $k-$th conformation, where $k={m\over n}$ is added to $\L$.
+The conformations are sorted according to this deviation and then the desired number of conformations are selected from this sorted list uniformly.
 

### 2017-09-10 paper.tex
set stretch to algorithm
small corrections, removed section Visualization + image

@@
 \usepackage[bookmarks=false]{hyperref}
-%\usepackage{setspace}
+\usepackage{setspace} % for setstretch in algorithm
 \pdfminorversion=4
@@
 {\small
-%\setstretch{0.88}
+\setstretch{0.88}
 \caption{\label{alg::main}Main loop of the RRT planner}
@@
     initial configuration $\qinit$
-%    goal configuration $\qgoal$,
 }
@@
 $iteration = 0$\;
+$\T$.addNode($\qinit$)\;
 \While{$iteration < \Imax$ {\bf and}  $v < n$}{
@@
 The expansion towards $\qrand$ is crucial for the growth of the tree towards unexplored areas of the configuration space.
-Traditionally, a line is constructed from $\qnear$ to $\qrand$ and the tree is expanded by a configuration located on this line.
+Traditionally, a line is constructed from $\qnear$ to $\qrand$ and the tree is expanded by a configuration located on this line in the 
+distance $\varepsilon$ from $\qnear$.
 This would however mostly lead to a collision in the case of ligands.
@@
 For each ligand conformation $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with a maximal scale.
-First, the maximal scale $\smax \in \S$ is selected and $m$ random samples are generated around $\qnear$ and tested for collisions.
+First, the maximal scale $\smax \in \S$ is selected and $m$ random samples are generated around $\qnear$ (in the distance $\varepsilon$) 
+and tested for collisions.
 The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
@@
 {\small
-%\setstretch{0.88}
-%\setstretch{\straa}
+\setstretch{0.88}
 \caption{\label{alg::expand}expand}
@@
             $\T$.addEdge($\qnear,\qnew$)\;
-            {\bf break;} // go to next conformation
+            {\bf break;} // go to the next conformation
         }
@@
 %The path $P=(q_i), q_i \in \C$ is represented as a sequence of collision-free configurations.
-The path is found in the tree even if the tree does not approach $\qgoal$ close enough.
-Considering also these non-feasible solutions is necessary to evaluate difficult areas of the tunnels, e.g. bottlenecks.
+%The path is found in the tree even if the tree does not approach $\qgoal$ close enough.
+%Considering also these non-feasible solutions is necessary to evaluate difficult areas of the tunnels, e.g. bottlenecks.
 %The utilization of all computed paths for evaluation of tunnel difficulty is described in the next section.
@@
 From motion planning point of view, both types of conformations are useful.
-The longer conformations can better fit in a narrow passage, while the packed conformations can be rotated more at a place.
+The longer conformations can better fit in a narrow passage, while the packed conformations better rotate at a place.
 In the narrow parts of the tunnels, the tree can be extended by the longer conformations, while in the more open regions, also the
@@
 To select $n$ conformations from all $m$ available conformations, they are sorted according to the deviation.
-Each $k-$th conformation, where $k={m\over n}$ is added to $\L$.
+Each $k-$th conformation from the sorted list, where $k=\lfloor{m\over n}\rfloor$, is added to $\L$.
 
@@
 %\includegraphics[width=0.25\textwidth]{fig/m040-conformations}
-\includegraphics[width=0.15\textwidth]{fig/m040-conf1} \hskip 25pt
-\includegraphics[width=0.12\textwidth]{fig/m040-conf2}
+\includegraphics[width=0.14\textwidth]{fig/m040-conf1} \hskip 25pt
+\includegraphics[width=0.11\textwidth]{fig/m040-conf2}
 \caption{\label{fig::m040c}
@@
 
-\subsection{Visualization of Results}
-
-TODO Bara: prepsat/motivace/jak se visuazlizuje
-toto je cast textu z toho posledniho paperu. Kdyztak to vyhod.
-
-%In order to fursupport the tunnel exploration process performed by biochemists.
-The above defined characteristics can be shown even as a graph, or better, presented visually by mapping them using colors to the tunnel spheres.
- Alternatively, the surface representation can be used to visualize the tunnel.
-In this case, a 3D point on the tunnel surface is colored according to the property value in its nearest tunnel part $i$,  i.e., such part $i$ whose center $c_i$ is the closest to the point among all tunnel parts.
-
-The examples of the color mapping are shown in Fig.~\ref{fig:properties}.
-The accessibility (Fig.~\ref{fig:properties}a) shows that more than the half of the tunnel is not accessible (red part of the tunnel).
-The throughput shows (Fig.~\ref{fig:properties}b) that the only difficult is the part around the bottleneck (red color in Fig.~\ref{fig:properties}b), and the second half of the tunnel is also traversable.
-
-\begin{figure*}[h]
-\centering
-\begin{tabular}{ccc}
-\includegraphics[width=0.3\textwidth]{fig/accessibility} &
-\includegraphics[width=0.3\textwidth]{fig/throughput} &
-\includegraphics[width=0.3\textwidth]{fig/ligand-scale} \\
-  (a) & (b) & (c) \\                     
-\end{tabular}
-\caption{Tunnel properties mapped onto its surface. The tunnel begins at the top left corner.
-(a) Accessibility --- green denotes the parts that were accessed by the ligand easily ($A(i) > 80$~\%), 
-    while red denotes parts that were accessible  in $<10$~\% of cases.
-(b) Throughput --- green denotes parts that are highly probable to be passed by the ligand ($T(i) > 80$~\%), 
-    while red denotes those that were hard to pass ($T(i) < 20$~\%).
-(c) Ligand scale --- the average scale of the ligand was \tylde 50\% in orange parts, \tylde 60\% in yellow parts, \tylde 80\% in yellow to green parts and \tylde 100\% in green parts.
-\label{fig:properties}
-}
-\end{figure*}
-
-\def\cstart{\overline{c_{start}}}
-\def\cend{\overline{c_{end}}}
+%\subsection{Visualization of Results}
+%
+%TODO Bara: prepsat/motivace/jak se visuazlizuje
+%toto je cast textu z toho posledniho paperu. Kdyztak to vyhod.
+%
+%%In order to fursupport the tunnel exploration process performed by biochemists.
+%The above defined characteristics can be shown even as a graph, or better, presented visually by mapping them using colors to the tunnel spheres.
+% Alternatively, the surface representation can be used to visualize the tunnel.
+%In this case, a 3D point on the tunnel surface is colored according to the property value in its nearest tunnel part $i$,  i.e., such part $i$ whose center $c_i$ is the closest to the point among all tunnel parts.
+%
+%The examples of the color mapping are shown in Fig.~\ref{fig:properties}.
+%The accessibility (Fig.~\ref{fig:properties}a) shows that more than the half of the tunnel is not accessible (red part of the tunnel).
+%The throughput shows (Fig.~\ref{fig:properties}b) that the only difficult is the part around the bottleneck (red color in Fig.~\ref{fig:properties}b), and the second half of the tunnel is also traversable.
+%
+%\begin{figure*}[h]
+%\centering
+%\begin{tabular}{ccc}
+%\includegraphics[width=0.3\textwidth]{fig/accessibility} &
+%\includegraphics[width=0.3\textwidth]{fig/throughput} &
+%\includegraphics[width=0.3\textwidth]{fig/ligand-scale} \\
+%  (a) & (b) & (c) \\                     
+%\end{tabular}
+%\caption{Tunnel properties mapped onto its surface. The tunnel begins at the top left corner.
+%(a) Accessibility --- green denotes the parts that were accessed by the ligand easily ($A(i) > 80$~\%), 
+%    while red denotes parts that were accessible  in $<10$~\% of cases.
+%(b) Throughput --- green denotes parts that are highly probable to be passed by the ligand ($T(i) > 80$~\%), 
+%    while red denotes those that were hard to pass ($T(i) < 20$~\%).
+%(c) Ligand scale --- the average scale of the ligand was \tylde 50\% in orange parts, \tylde 60\% in yellow parts, \tylde 80\% in yellow to green parts and \tylde 100\% in green parts.
+%\label{fig:properties}
+%}
+%\end{figure*}
+%
+%
+
+%\def\cstart{\overline{c_{start}}}
+%\def\cend{\overline{c_{end}}}
 
@@
 The \RA\ method was run with the following parameters:
-$\Imax=5,000$, $m=100$, $\gb=0.9$, $\dt=1.5$~\AA, $\rv=3$~\AA.
+$\Imax=5,000$, $m=100$, $\gb=0.9$, $\dt=1.5$~\AA, $\rv=2$~\AA\ and $\varepsilon=0.05$~\AA.
 Both methods were implemented in the same software framework and they utilized the same data structures and routines for collision detection and nearest-neighbor search.
 The nearest-neighbor search of both planners is based on the 6D Euclidean metric with the same weights for rotation and translation parts.
-This metric was selected based on preliminary experiments with both planners as it resulted in the higher success rates.
+This metric was selected based on preliminary experiments with both planners as it resulted in the success rates.
 
@@
 In these lines, the \RB\ method employs the retraction step in order to find a new collision-free configurations $\qnew$ that approaches $\qrand$.
-The retraction step is realized by sampling the vicinity of the configuration $\qnear$ in the distance $0.1$~\AA\ using 20 samples.
+The retraction step is realized by sampling the vicinity of the configuration $\qnear$ in the distance $0.05$~\AA\ using 20 samples.
 A collision-free sample that maximally approaches $\qrand$ is selected and added to the tree, and the retraction continues from this sample.
@@
 Other parameters of the \RB\ planner were same as for \RA.
-Planning was terminated after $\Imax$ iterations, or if the tree approached the active site to less than 0.5~\AA.
+Planning was terminated after $\Imax$ iterations, or if the tree approached the active site to the distance less than 0.5~\AA.
 
@@
 
-The results are shown in Tab.~\ref{tab::one} which shows traversability rate. 
+The computed traversability rates are shown in Tab.~\ref{tab::one}.
 The maximum percentage of traversable frames (the row Maximum in Tab.~\ref{tab::one}) is computed as the percentage of tunnels with bottleneck larger than the tested radius.
-The row Planners shows traversable rate detected by the planners.
+The row Planning shows the traversable rate computed by the planners.
 This rate decreases with the increasing radius and it is lower than the maximum traversable rate.
@@
 \caption{\label{tab::one}Traversability for one atom.}
-\renewcommand{\tabcolsep}{3.5pt}
+\renewcommand{\tabcolsep}{4.3pt}
 {\small
@@
 The traversability rates are shown in Tab.~\ref{tab::selection}, where the subscript of each ligand's 
-name shows the number of selected conformations (5 or 50) and the used method (S or P).
+name shows the number of the selected conformations (5 or 50) and the used method (S or P).
 The number after the slash in the ligand's name denotes the number of atoms of the ligand.
@@
 \small
-\renewcommand{\tabcolsep}{3pt}
+\renewcommand{\tabcolsep}{4.3pt}
 {\small
@@
 Based on the results of the previous experiment, 50 conformations using the `S' strategy were selected for the following
-ligands: m037t (1,2-Dichloroethane), m038t (1,3-Dichloropropane), m056 ( 2-Chlorobutane) and m080 (Trichloropropane), and
+ligands: m037t (1,2-Dichloroethane), m038t (1,3-Dichloropropane), m056 (2-Chlorobutane) and m080 (Trichloropropane), and
 the task was to compute trajectories for these ligands in the MD sequence.
@@
 they have same (or even larger) traversability rate than DCP.
-However, this hypothesis hasn't been tested agains MD simulations yet.
+However, this hypothesis hasn't been tested against MD simulations yet.
 
@@
 The runtimes of both algorithms increase in those frames, where the tunnel is not traversable, i.e., where the distance
-to the active site is more than 2~\AA (e.g., the frame number 60).
+to the active site is more than 2~\AA\ (for example,in the frame number 60).
 
@@
 This results in a faster exploration of the configuration space and therefore, \RA\ needs less number of iterations to reach the active site.
-The better performance of \RA\ is caused by the proposed atom-based metric, because this distance between two configurations depends on the actual shape of the ligand.
-This shape is different in each configuration.
-Contrary, the \RB\ utilizes only the single 6D Euclidean metric which is same for all conformations.
-To improve performance of \RB, the weights of the 6D Euclidean metric should differ for each conformation, but it would
-prevent the usage KD-tree for the nearest-neighbor search.
+The better performance of \RA\ is caused by the proposed atom-based metric, because this distance depends on the actual shape of the ligand and 
+it allows the ligands to retract more toward $\qrand$.
+Contrary, \RB\ utilizes the single 6D Euclidean weighted metric in the expansion, where the weights are same for all conformations.
+Depending on these weights, the metric favors samples that differ from $\qrand$ more in rotation or translation part.
+However, the weights between rotation/translation parts should depends on the shape of the ligand, which is different in each conformation.
+By using single set of weights in the 6D Euclidean metric, we implicitely favor either the packed or the longer conformations.
+%To improve performance of \RB, the weights of the 6D Euclidean metric should differ for each conformation, but it would
+%prevent the usage KD-tree for the nearest-neighbor search.
 
@@
 
-\section{Discussion - TODO prepsat}
+\section{Discussion}
+
+The experiments show the advantage of the proposed motion planner. 
+The planner outperforms the Retraction-based planner not only in the `easy' problems (with $\smin=0.5$), but also
+in the case of difficult scenarios with larger ligands ($\smin=0.8$).
+
+It is worth to mention, that the performance of both planners is influenced by their parameters, namely by
+$\Imax$.
+By increasing this parameter, we can expect better performance in finding the trajectories and consequently higher 
+traversability rates, but with the increased computational time.
+
+The internal performance of the planners is however not the main criteria.
+Fom bio-chemical point of view, the main contribution of using motion planning for detecting the ligand trajectories is the ability to
+consider
+
+The proposed planner is very simple and it can be improved in many ways: by using a less aggressive scaling-down technique (in order
+to obtain better trajectories) or by using only a subset of conformations and a better expansion (in order to speed it up).
+Having a better motion planner (in the term of the speed or even success rate) however does not automatically mean that the resulting trajectories will help the biochemists more.
+The presented work shows possible advantages of the analysis based on ligand trajectories.
+By comparing the planned trajectories with trajectories observed in MD simulations, thresholds for the success rate or minimal tunnel throughput can be set, which will help biochemists to make decisions.
 
@@
 %at the active sites.
-
-The second limitation is caused by the utilization of scaled-down ligands.
-The protein atoms fill the internal void space therefore the tunnels identified inside proteins without ligand tend
-to be very narrow.
-To enable at least some motion of the ligands inside such narrow tunnels, the scaling-down of the ligand atoms is necessary.
-It is used also in other related tool, e.g.~\cite{cortes2005path}.
-
-Due to these reasons, the computed trajectories can be considered either as too optimistic
-(e.g. because they are computed on a wide tunnel that can be in fact closed due to molecular dynamics),
-or too pessimistic (e.g. the trajectory is not found because of narrow passage around the tunnel, which can be 
-opened due to the molecular dynamics).
-Despite these limitations, testing the tunnel traversability using motion planning technique can provide chemists more information
-than the simple bottleneck radius, which is used nowadays.
-For example, possible detours from the tunnel can be identified, which indicates that even a tunnel with a small
-bottleneck can be traversed (Fig.~\ref{fig::detail}).
+%The second limitation is caused by the utilization of scaled-down ligands.
+%The protein atoms fill the internal void space therefore the tunnels identified inside proteins without ligand tend
+%to be very narrow.
+%To enable at least some motion of the ligands inside such narrow tunnels, the scaling-down of the ligand atoms is necessary.
+%It is used also in other related tool, e.g.~\cite{cortes2005path}.
+
+%Due to these reasons, the computed trajectories can be considered either as too optimistic
+%(e.g. because they are computed on a wide tunnel that can be in fact closed due to molecular dynamics),
+%or too pessimistic (e.g. the trajectory is not found because of narrow passage around the tunnel, which can be  opened due to the molecular dynamics).
+%Despite these limitations, testing the tunnel traversability using motion planning technique can provide chemists more information
+%than the simple bottleneck radius, which is used nowadays.
+%For example, possible detours from the tunnel can be identified, which indicates that even a tunnel with a small
+%bottleneck can be traversed (Fig.~\ref{fig::detail}).
 
@@
 %from proteins~\cite{cortes2010simulating,guieysse2008structure}.
-
-The proposed planner is very simple and it can be improved in many ways: by using a less aggressive scaling-down technique (in order
-to obtain better trajectories) or by using only a subset of conformations and a better expansion (in order to speed it up).
-Having a better motion planner (in the term of the speed or even success rate) however does not automatically mean that the resulting trajectories will help the biochemists more.
-The presented work shows possible advantages of the analysis based on ligand trajectories.
-By comparing the planned trajectories with trajectories observed in MD simulations, thresholds for the success rate or minimal tunnel throughput can be set, which will help biochemists to make decisions.
 

### 2017-09-10 paper.tex
discussion

@@
 
-\section{Discussion}
-
-The experiments show the advantage of the proposed motion planner. 
-The planner outperforms the Retraction-based planner not only in the `easy' problems (with $\smin=0.5$), but also
-in the case of difficult scenarios with larger ligands ($\smin=0.8$).
-
-It is worth to mention, that the performance of both planners is influenced by their parameters, namely by
-$\Imax$.
-By increasing this parameter, we can expect better performance in finding the trajectories and consequently higher 
-traversability rates, but with the increased computational time.
-
-The internal performance of the planners is however not the main criteria.
-Fom bio-chemical point of view, the main contribution of using motion planning for detecting the ligand trajectories is the ability to
-consider
-
-The proposed planner is very simple and it can be improved in many ways: by using a less aggressive scaling-down technique (in order
-to obtain better trajectories) or by using only a subset of conformations and a better expansion (in order to speed it up).
-Having a better motion planner (in the term of the speed or even success rate) however does not automatically mean that the resulting trajectories will help the biochemists more.
-The presented work shows possible advantages of the analysis based on ligand trajectories.
-By comparing the planned trajectories with trajectories observed in MD simulations, thresholds for the success rate or minimal tunnel throughput can be set, which will help biochemists to make decisions.
+\section{Discussion and utilization of the results}
+
+The work in the paper is motivated by real needs of bio-chemists who need to analyze traversability of selected ligands in the tunnels.
+To enable finding the trajectories, it is necessary to shrink the atomic radii of the ligands which has been utilized also in other tools~\cite{cortes2005path,guieysse2008structure}.
+Finding trajectories for scaled-down ligands on one hand helps to cope with the narrow passages, but the question is, how realistic
+such trajectories are.
+
+Judging the accessibility of the active sites by the ligands based only on the traversability rates may be problematic if the ligands are scaled too much.
+Instead of using a single version of the scale-down ligand, we only limit the minimal scale used by the planner (the parameter $\smin$).
+The expansion step always prefers to expand the tree with the maximal scale, that leads to a new collision-free configuration.
+The information about the scales used at the trajectory points (which can be extracted from the configuration tree of the proposed planner) brings the chemists new knowledge about the tunnels.
+It allows them to detect parts of the tunnels, where the ligand has to scale-down significantly, and parts, where the ligand can be again
+larger (Fig.~\ref{fig::dcp}).
+According to the results of the motion planinng, DCP needs to shrink down to ca. 75~\% of its size (Fig.~\ref{fig::dcp}, point B).
+However, as both the protein and the ligand adapt to each other, it is possible, that even a narrow part of the tunnel is traversable.
+This was observed in the MD simulation of the DCP ligand in the 1CQW protein, which revealed that DCP can pass the first tunnel.
 

### 2017-09-10 paper.tex
alg

@@
 
-Instead, we propose to expand the tree by considering all conformations and also by expanding the tree using the maximal possible scale that
-does not lead to a collision.
-For each ligand conformation $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with a maximal scale.
+The expansion procedure attempts to expand the tree using all the conformations $\L$ and by preferring larger scales than the smaller ones.
+For each $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with a maximal scale.
 First, the maximal scale $\smax \in \S$ is selected and $m$ random samples are generated around $\qnear$ (in the distance $\varepsilon$) 
 and tested for collisions.
+The random samples are generated similarly as in the case of $\qrand$ samples, only their translation
+part $(x,y,z)$ is generated around $\qnear$.
 The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
@@
 until a collision-free sample is found or until the minimal reduced-scale $\smin$ is reached.
-The random samples are generated similarly as in the case of $\qrand$ samples, only their translation
-part $(x,y,z)$ is generated around $\qnear$.
+By testing first the largest scale ($\smax$), the expansion procedure preffers expanding the tree by larger scales.
+This forces the planner to automatically use smaller scaled in the difficult (narrow) parts of the tunnels, while prefering
+large scales if possible.
 
@@
 
-\subsection{Influence of conformation selection}
+\subsection{Influence of the conformation selection}
 

### 2017-09-10 paper.tex
abstract 200 words

@@
 These active sites are accessible through one or more void paths, called tunnels.
-The passage of a ligand through a tunnel is directly influenced by the tunnel geometric properties, such as its length, width, and bottleneck.
-The task of the biochemists is to evaluate these properties and thus estimate the reactivity of the protein, i.e., the possibility of the ligand to reach the protein active site.
-%The knowledge about a tunnel (e.g. the length, bottleneck or physico-chemical properties) can help chemists to estimate possible interactions between the protein and the ligand.
-Nowadays, tunnels are mostly computed using Voronoi diagrams. 
-This approach helps to predict the interactions between proteins and small ligands, whose shape is approximated by a bounding sphere.
-Such approximation usually leads to highly imprecise results.
-Computing the transportation path of non-spherical ligands including their conformation changes (i.e., when the atoms of the ligand move relatively to each other) leads to the motion planning task in high-dimensional configuration space, which can be solved using sampling-based planning methods.
-This planning problem necessarily includes narrow passages, as the movements of ligands are limited by the protein atoms surrounding the ligand path.
-Considering flexible ligands increases the number of degrees of freedom, and consequently also the dimension of the configuration space.
-To address this problem, in this paper we propose a modification of the Rapidly Exploring Random Tree planner for the purpose of trajectory computations of flexible ligands, heading from the protein outside environment to its active site.
-The detected trajectory then defines the centerline of the searched tunnel.
+Nowadays, tunnels are mostly computed using Voronoi diagrams which approximate the ligand by a bounding sphere.
+This leads to highly imprecise results.
+To address this problem, we define the computation of the transportation path of non-spherical ligands including their conformation changes (i.e., when the atoms of the ligand move relatively to each other) as a motion planning task in high-dimensional configuration space.
+This can be solved using sampling-based planning methods.
+We propose a modification of the Rapidly Exploring Random Tree planner for the purpose of trajectory computations of flexible ligands.
 To cope with the narrow passage problem, the random samples are generated around a Voronoi-based tunnel, which serves us for better localization of the searched space.
@@
 \section{Introduction \& Motivation}
+
+Reactivity of proteins with other molecules plays a crucial role in many research disciplines, including protein engineering and drug design.
+Protein reactivity is tightly connected with the presence of the void space between its amino acids. 
+This space, forming a path called tunnel, can be utilized by a small molecule (ligand) which can enter the protein inner space and react with its amino acids surrounding a specific site, called the active site~\cite{gora2013gates}.
+The size and location of tunnels determine the ligand size and shape which can enter the active site.
+%The tunnel has to be wide enough and also the physico-chemical properties of the surrounding amino acids have to be compatible with the ligand.
+Biochemists are currently using computational methods for tunnel detection in order to estimate the possibility of interactions between a protein and a ligand without the need to perform laboratory experiments (Fig.~\ref{fig::motiv}).
+The most common methods detect tunnels using Voronoi diagrams~\cite{yaffe2008,caver3} assuming a spherical ligand (probe), i.e., the ligand is approximated by a bounding sphere.
+However, as the ligands are typically non-spherical and flexible, these methods can often skip possible candidate tunnels.
 
@@
 \end{figure}
-
-Reactivity of proteins with other molecules plays a crucial role in many research disciplines, including protein engineering and drug design.
-Protein reactivity is tightly connected with the presence of the void space between its amino acids. 
-This space, forming a path called tunnel, can be utilized by a small molecule (ligand) which can enter the protein inner space and react with its amino acids surrounding a specific site, called the active site~\cite{gora2013gates}.
-The size and location of tunnels determine the ligand size and shape which can enter the active site.
-%The tunnel has to be wide enough and also the physico-chemical properties of the surrounding amino acids have to be compatible with the ligand.
-Biochemists are currently using computational methods for tunnel detection in order to estimate the possibility of interactions between a protein and a ligand without the need to perform laboratory experiments (Fig.~\ref{fig::motiv}).
-The most common methods detect tunnels using Voronoi diagrams~\cite{yaffe2008,caver3} assuming a spherical ligand (probe), i.e., the ligand is approximated by a bounding sphere.
-However, as the ligands are typically non-spherical and flexible, these methods can often skip possible candidate tunnels.
 

### 2017-09-10 paper.tex
RW

@@
 
-TODO some words about MD and why it better to analyze stuff with tunnels than with MD
-
-Most of the available tools for tunnel detection are based on Voronoi diagrams~\cite{yaffe2008,caver3} or they utilize grid-based computation~\cite{sehnal2013mole}.
+%TODO some words about MD and why it better to analyze stuff with tunnels than with MD
+
+Most of the available tools for tunnel detection are based on Voronoi diagrams~\cite{yaffe2008,caver3} or they utilize a grid-based computation~\cite{sehnal2013mole}.
 %The tunnels are pathways computed for a single atom (spherical probe) using Voronio-diagrams~\cite{yaffe2008,caver3} or 
 %grid-based approaches~\cite{sehnal2013mole,petrek2006caver}.
-The resulting tunnels are represented as a sequence of spheres and characterized by their bottleneck, length, curvature, and a list of surrounding residues, which are later used to estimate the interaction possibilities.
-The main disadvantage of both grid-based and VD-based methods is that the shape of the ligand is not taken into account during the tunnel detection and it is therefore not easy to estimate if (and how) a non-spherical ligand might traverse the tunnels.
-Moreover, ligand flexibility cannot be principally considered by these approaches.
-In order to determine if a ligand can pass the tunnel, it is necessary to compute a trajectory considering the shape of the ligand and its 3D translation/rotation.
+The resulting tunnels are represented as a sequence of spheres and characterized by their bottleneck, length, curvature, and a list of surrounding residues, whose physico-chemical properties are crucial for assessing the interaction possibilities between the protein and ligand.
+More details can be found in the recently published state-of-the-art report about analysis and visualization of biomolecular cavities~\cite{Krone_2016}.
+The main disadvantage of both grid-based and VD-based methods is that the shape of the ligand is not taken into account during the tunnel detection and it is therefore not easy to estimate if (and how) a non-spherical ligand might traverse the tunnel.
+Moreover, the ligand flexibility cannot be principally considered by these approaches.
+In order to determine if the ligand can pass the tunnel, it is necessary to compute a trajectory considering the shape of the ligand and its 3D translation and rotation.
 %This can be formulated as a motion planning problem in a high-dimensional configuration space $\C$.
@@
 The well known issue of the sampling-based planners is the narrow passage problem.
-A narrow passage is a small region in the configuration space whose removal changes the connectivity of the space.
+Narrow passage is a small region in the configuration space, whose removal changes the connectivity of the space.
 Due to its low volume, the probability of sampling the narrow passage is low.
@@
 %are required in order to sample the narrow passages dense enough.
-PRM-based planners can cope with the narrow passage problem by increasing probability of sampling in given regions.
-These regions can be estimated using workspace information such as by medial axis~\cite{amatoOBRRT,amato2002using,wilmarthMAPRM} 
-or they can be identified e.g. based on number of collision-free samples around a given configurations~\cite{overmarsGauss,hsuBridge}.
-The narrow passage are however located in the configuration space and it is generally not easy to estimate their location
+PRM-based planners can cope with the narrow passage problem by increasing the probability of sampling in given regions.
+These regions can be estimated using workspace information, such as by medial axis~\cite{amatoOBRRT,amato2002using,wilmarthMAPRM}, 
+or they can be identified, e.g., based on the number of collision-free samples around a given configuration~\cite{overmarsGauss,hsuBridge}.
+The narrow passages are however located in the configuration space and it is generally not easy to estimate their location
 based only on the features of the workspace~\cite{hannaWIS}.
@@
 RRT-based planners cannot cope with the narrow passage problem simply by increasing the probability of sampling in them, as the
-growth of the configuration tree toward the random samples may be blocked by obstacles~\cite{vonasekphd}.
+growth of the configuration tree towards the random samples may be blocked by obstacles~\cite{vonasekphd}.
 To prevent this blocking, DD-RRT~\cite{yershovaDDRRT} limits the selection of nodes for expansion to a small ball. 
-The radius of the ball is set to infinity for new nodes, and decreases to a predefined radius later if the node cannot be successfully expanded.
+The radius of the ball is set to infinity for new nodes, and decreases to a predefined radius if the node cannot be successfully expanded.
 The disadvantage of DD-RRT is the sensitivity to the predefined radius.
-Therefore, authors of ADD-RRT~\cite{jailletADRRT} proposed to adapt the radius  according to the success rate of the expansion.
-To attract the tree towards a region, the random samples have to be generate there only if the tree can expand to the region.
-In~\cite{kardossRRTKK}, probability of sampling is increased in several waypoints close to narrow passages, but without specifying
-how to find these waypoints.
+Therefore, the authors of ADD-RRT~\cite{jailletADRRT} proposed to adapt the ball radius according to the success rate of the expansion.
+To attract the tree towards a given region, random samples have to be generated there only if the tree can expand to this region.
+In~\cite{kardossRRTKK}, the probability of sampling is increased in several waypoints close to narrow passages, but without specifying
+the method to find these waypoints.
 The generalization of~\cite{kardossRRTKK} is the guided sampling, where a whole path is used
 to generate samples in the configuration space.
-The path can be computed in the workspace~\cite{vonasek2009rrt}, or iteratively refined in the configuration space based on solution of a relaxed version of the problem~\cite{bayazitIRC}.
+The path can be computed in the workspace~\cite{vonasek2009rrt}, or iteratively refined in the configuration space based on the solution of a relaxed version of the problem~\cite{bayazitIRC}.
 
@@
 In Retraction-based RRT~\cite{zhangRetraction}, the tree is retracted along the boundary of the obstacles in the configuration space.
-This requires the analysis of contact space, which leads to the generalized penetration depth computation of high complexity~\cite{he2016efficient}. %\cite{he2016efficient} in RAL!! + zhang2008fast
+This requires the analysis of the contact space, which leads to the generalized penetration depth computation of high complexity~\cite{he2016efficient}.
+%\cite{he2016efficient} in RAL!!
++ zhang2008fast
 Selective Retraction-based RRT~\cite{lee2012srrrt} avoids the expensive growth in the open areas of the configuration space and focuses 
-the retractions only to the narrow passages.
-Obstacle-based RRT~\cite{amatoOBRRT} utilizes several expansion procedures based on e.g. random vectors, obstacles vectors or medial axis in order to improve growth in the narrow passages.
-In MARRT~\cite{denny2014marrt}, the newly generated samples are pushed toward the medial axis.
+the retractions only on the narrow passages.
+Obstacle-based RRT~\cite{amatoOBRRT} utilizes several expansion procedures based on, e.g., random vectors, obstacle vectors, or medial axis in order to improve the growth in the narrow passages.
+In MARRT~\cite{denny2014marrt}, the newly generated samples are pushed towards the medial axis.
 %Therefore, many algorithms uses a heuristic to compute the samples near boundary of the configuration space~\cite{amatoOBPRM,amatoOBRRT}.
 
-The probability of sampling of the narrow passages can be increased by dilating the free-space, e.g. by shrinking the geometry of 
+The probability of sampling of the narrow passages can be increased by dilating the free-space, e.g., by shrinking the geometry of 
 the robot~\cite{hsuOnProb} or the obstacles~\cite{bayazitIRC}.
-The shrinking technique has been used also in~\cite{cortes2010simulating}, where exit pathways for a small flexible molecule are
+The shrinking technique has been used also in~\cite{cortes2010simulating}, where the exit pathways for a small flexible molecule are
 computed using a RRT-based method.
 The flexible ligand is modeled as a kinematic chain, which increases the dimension of the configuration space.
-To cope with these many-DOF ligands, RRT-ML~\cite{cortes2007mlrrt} is employed as the basic planner.
-RRT-ML expands the tree primarily using those DOF that are essential for achieving motion of the ligand (i.e., rotation
-and translation) and it employs the other DOFs (i.e., that are responsible for conformation changes) if they hinder the growth of the tree.
+To cope with these many-DOF ligands, the RRT-ML~\cite{cortes2007mlrrt} approach is employed as the basic planner.
+RRT-ML expands the tree primarily using those DOF that are essential for achieving the motion of the ligand (i.e., rotation
+and translation) and it employs the other DOFs (i.e., those that are responsible for conformation changes) if they hinder the growth of the tree.
 
@@
 
-\section{Proposed method}
-
-We propose a modification of the RRT planner for computing trajectories through protein tunnels.
-The input is a set of tunnels computed using Voronoi diagrams and the task is to find trajectories along the given tunnels
+\section{Proposed Method}
+To address the problems of the existing solutions for computing the non-spherical flexible ligand trajectories through protein tunnels, we propose a modification of the RRT planner. 
+Our approach takes into account the shape and flexibility of the ligand and localizes the searched configuration space using the information about tunnels detected by the Voronoi-based approach.
+The input is therefore formed by a set of tunnels computed using Voronoi diagrams and the task is to find trajectories for a ligand along these tunnels
 leading from outside the protein to the active site.
-The tunnels are detected in a sequence of protein dynamics that is computed using MD simulations.
-
-In contrast to~\cite{cortes2010simulating}, which finds any exit pathway, we compute the trajectories only along each tunnel separately.
-This is motivated by the practical needs of chemists that are interested in the traversability of a selected tunnel.
-Computing trajectories along a single tunnel also brings advantages from the motion planning point of view.
+The tunnels are detected in a sequence time steps capturing protein dynamic behavior that is computed using molecular dynamics simulations.
+
+In contrast to~\cite{cortes2010simulating}, which searches for any exit path from the protein, we localize the searching and compute the trajectories only along each tunnel separately.
+This is motivated by the practical needs of biochemists that are interested in the actual traversability the ligand through a selected tunnel.
+Computing the trajectories along a single tunnel brings several advantages from the motion planning point of view as well.
 It decreases the volume of the configuration space to be searched and increases the relative volume of the narrow passages.
@@
 the tunnel in a user-defined distance.
-Therefore, the tunnels serve as a guiding path for the configuration tree~\cite{vonasek2009rrt}.
+Therefore, the tunnels serve as a navigation path for the configuration tree~\cite{vonasek2009rrt}.
 

### 2017-09-10 paper.tex
alg inputs updated

@@
 The flexible ligand is modeled as a kinematic chain, which increases the dimension of the configuration space.
-To cope with these many-DOF ligands, the RRT-ML~\cite{cortes2007mlrrt} approach is employed as the basic planner.
+To cope with these many-DOF ligands, the RRT-ML~\cite{cortes2007mlrrt} approach is employed as the basic planner in~\cite{cortes2010simulating}.
 RRT-ML expands the tree primarily using those DOF that are essential for achieving the motion of the ligand (i.e., rotation
@@
    ligand conformations $\L$,
-   scale limits $\smin, \smax$ and $\sdelta$
+   scale limits $\smin, \smax$ and $\sdelta$,
+   distance $\dt$ to move virtual goal
 }
@@
     $\qnear$ = nearest node in $\T$ towards $\qrand$\;
-    expand($\qnear,\qrand$)\;
+    expand($\T, \qnear,\qrand$)\;
     \For{$i= n-1,n-2,\ldots,v+1,v$}{ \nllabel{alg::main:a}
@@
 (i.e., for ligand with no translation and no rotation).
-To find the nearest atom to $\qrand$ for a ligand at the configuration $q=(x,y,z,r_x,r_y,r_z,l,s)$, the translation vector $T=(x,y,z)$ and
+To find the nearest atom to $\qrand$ for a ligand at the configuration $q=(x,y,z,r_x,r_y,r_z,l,s)$, the translation vector $V=(x,y,z)$ and
 rotation matrix $R$ given by $r_x$, $r_y$ and $r_z$ are computed.
-Then, $\qrand$ is translated to $\qrand' = R^{-1}(\qrand-T)$ and the KD-tree search is used to find the nearest point towards $\qrand'$, which
+Then, $\qrand$ is translated to $\qrand' = R^{-1}(\qrand-V)$ and the KD-tree search is used to find the nearest point towards $\qrand'$, which
 is also the nearest point of the ligand at $q$ to $\qrand$.
@@
 \KwIn{
+   tree $\T$,
    configuration $\qnear$ to be expanded,
@@
    scale limits $\smin, \smax$, and $\sdelta$,
-   tree $\T$
 }

### 2017-09-10 paper.tex
conclusion

@@
 The tunnels are pathways leading from an outer environment toward the active sites, where chemical interactions may undergo.
-So far, the tunnels have been detected using Voronoi diagrams and analyzed based on their length, bottleneck and the list of residues along them.
-Such an analysis is however useful only for small spherical ligands and it is difficult to estimate behavior of a large, non-spherical ligands.
+So far, the tunnels have been detected using Voronoi diagrams, which does not take into account shape of the ligand neither the possible 
+conformation changes.
 In the proposed work, we compute trajectories using modified RRT for whole ligands and we model the conformation changes of the
 ligands using a predefined set of conformations.
-The main idea of the proposed planner is to generate the random samples around the Voronoi-based tunnels.
-The probability of sampling is increased around so called virtual goal, which is a point moving along the tunnel. 
-The virtual goal is moved forward, if the tree approaches it close enough.
-To enable movements in narrow tunnels, the atomic radii of the ligands are scaled down.
+The atomic radii of the ligands can be scaled down up to a certain predefined factor.
+The planner generates the random samples along a tunnel computed by Voronoi diagram and it attempts to expand the configuration
+tree using all conformations. 
+The expansion prefers atoms with a higher scale, which ensures that the ligand shrinks down only if traversing narrow parts of the tunnels.
+To further boost exploration of the configuration space, a novel metric suitable for measuring distances between various shapes
+of the ligands, is used in the expansion step.
+The proposed method has been used to  evaluate trajectories of several ligands in a sequence of protein dynamics.
+Traversability of one of the tested ligands has been already approved in MD simulations.
 

### 2017-09-10 paper.tex
reference to kingsley2014

@@
 Moreover, the ligand flexibility cannot be principally considered by these approaches.
-In order to determine if the ligand can pass the tunnel, it is necessary to compute a trajectory considering the shape of the ligand and its 3D translation and rotation.
+Simulations based on molecular dynamics (MD) can directly evaluate migration of a ligand into a protein, but they are significantly more computationally demanding~\cite{kingsley2014including}. 
+Moreover, MD simulations have to be prepared for a specific ligand/protein pair, which is not practical for testing a large number of ligands.
+The geometric-based methods are therefore still preferable due to their fast speed.
+In order to determine if the ligand can pass the tunnel using the geometric-based methods, it is necessary to compute a trajectory considering the shape of the ligand and its 3D translation and rotation.
+
 %This can be formulated as a motion planning problem in a high-dimensional configuration space $\C$.

### 2017-09-10 paper.tex
method

--- 2017-09-10 paper.tex
@@
 To enable the movement of the ligand in the narrow parts of the tunnel, the free-space can be dilated by shrinking the atom radii, similarly, e.g., to~\cite{cortes2005path,hsu06multilevel}. 
-The ligand flexibility is modeled using a fixed set of predefined configurations which the ligand can occupy.
+The ligand flexibility is modeled using a fixed set of predefined conformations which the ligand can occupy.
 
@@
 The flexible ligand is modeled as a kinematic chain, which increases the dimension of the configuration space.
-To cope with these many-DOF ligands, the RRT-ML~\cite{cortes2007mlrrt} approach is employed as the basic planner in~\cite{cortes2010simulating}.
+To cope with these many-DOF ligands, the RRT-ML~\cite{cortes2007mlrrt} approach is employed as the basic planner.
 RRT-ML expands the tree primarily using those DOF that are essential for achieving the motion of the ligand (i.e., rotation
@@
 leading from outside the protein to the active site.
-The tunnels are detected in a sequence time steps capturing protein dynamic behavior that is computed using molecular dynamics simulations.
+The tunnels are detected in a sequence time steps capturing protein dynamic behavior that is computed using molecular dynamics (MD) simulations.
 
@@
 
-Due to the protein dynamics, the tunnel changes, move and merge, and their bottleneck changes in each frame.
-It can therefore be expected, that the detected tunnels are rather narrow.
-Based on the observation from MD simulations, the tunnels adapt to the ligands (and vice versa), so even narrow tunnels can serve as transportation path for thee ligands TODO CITACE??.
-To enable computing of trajectories in the narrow tunnels, we shrink the atom radii of the ligand, similarly to~\cite{cortes2010simulating,guieysse2008structure}.
-
-The ligand flexibility is modeled using a predefined set of typical conformations, therefore the number of the dimensions of the configuration space does not depend on the DOF of the ligand as in~\cite{cortes2010simulating}.
-Library of conformations are used also in different types of calculations, e.g.~\cite{kellogg}. 
-While classic 6D Euclidean metric is used in the nearest-neighbor search of RRT, we propose to use another metric during the expansion step.
-This metric measures distances between two configurations as the nearest distance of two atoms. 
-This enables the metric to consider the shape of the ligand and it supports retraction of the ligand to the narrow passages.
+Due to the protein dynamics, the tunnel changes, move and merge, and their bottleneck can change in each time step.
+Therefore, it can be expected that the detected tunnels are rather narrow.
+Based on the observations of MD simulations, tunnels adapt their shape to the passing ligands and vice versa.
+In consequence, even very narrow tunnels, being too small for the passage of a given ligand approximating sphere, can serve as the transportation path for this ligand.
+Therefore, also the existing Voronoi-based methods enable the user to decrease the size of the approximating sphere which simulates shrinking of the ligand size.
+In our solution we simulate this by shrinking the atom radii of the ligand, used also by~\cite{cortes2010simulating,guieysse2008structure}.
+
+Molecular dynamics simulations of ligands passing through proteins show that the ligands can be highly flexible.
+However, ligands cannot form arbitrary conformations. 
+Ligand can contain several torsion and dihedral angles between its atoms which determine possible ligand conformations.
+Therefore, the ligand flexibility can be modeled using only a predefined set of conformations.
+So the number of dimensions of the configuration space does not depend on the DOF of the ligand as in~\cite{cortes2010simulating}.
+The libraries containing possible ligand conformations are used also in different types of calculations, e.g., in~\cite{kellogg}.
+
+While the classic 6D Euclidean metric is used in the nearest-neighbor search of RRT, we propose to use another metric during the expansion step.
+This metric measures the distances between two configurations as the nearest distance of two atoms. 
+This enables the metric to consider the shape of the ligand and it supports the retraction of the ligand to the narrow passages.
 
@@
 
-Proteins and ligands are represented by the hard sphere model, where the radius of each sphere (atom) is given by its van der Waals radius.
+Proteins and ligands are represented by the hard sphere model, where the radius of each sphere representing an atom is given by its van der Waals radius.
 A protein tunnel is described by a sequence of collision-free spheres 
 $T=( (c_1, r_1),\ldots,(c_n,r_n) )$, where $n$ denotes the number of spheres,
-$c_i \in \R^3$ is their 3D position and $r_i > 0$ denotes the maximum collision-free radius of a sphere centered at $c_i$. 
-The tunnels can be found by the tunnel detection tools like CAVER 3.0~\cite{caver3}.
+$c_i \in \R^3$ is their 3D position, and $r_i > 0$ denotes the maximum collision-free radius of a sphere centered at $c_i$. 
+Tunnels serving as an input for our proposed method can be detected using some of the existing computational tools, such as CAVER 3.0~\cite{caver3}.
 
@@
 %Depending on the protein, the tunnel bottlenecks can be even less than $1$~\AA, which is too narrow for ligands with more than 2 atoms.
-The flexibility of a ligand is modeled using the set $\L$ of its conformations.
-We assume that a library of all possible (in the considered resolution) conformations is available (e.g., ~\cite{dunbrack}) and we later show how to select a subset of conformations suitable for motion planning.
+The flexibility of the ligand is modeled using the set $\L$ of its conformations.
+We assume that the library of all possible conformations in a considered resolution is available (e.g., ~\cite{dunbrack}) and we later show how to select a subset of conformations suitable for the motion planning.
 %The conformations $\L$ are used from a library (e.g.,~\cite{dunbrack}), or they can be prepared considering the potential energy.
 
-To enable motion of ligands in the narrow tunnels, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
-A discrete set of scales is used, i.e., $s \in \S=\{\smin, \smin+\sdelta, \ldots, \smax\}$, where 
-$\smin$ is the minimal allowed scale, $\smax=1$ is the maximal allowed scale and $\sdelta$ is the minimal difference between two scales.
+To enable the motion of ligands in the narrow tunnels, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
+In our solution we use a discrete set of scales, i.e., $s \in \S=\{\smin, \smin+\sdelta, \ldots, \smax\}$, where 
+$\smin$ is the minimal allowed scale, $\smax=1$ is the maximal allowed scale, and $\sdelta$ is the minimal difference between two scales.
 
 A configuration of the ligand $q=(x,y,z,r_x,r_y,r_z,l,s)$ is described
-by the 3D position $(x,y,z)$ of the reference point of the ligand (the geometric center), the rotation around $x$, $y$, and $z$ axes,
-the index of the conformation $l\in \L$ and the scale $s \in \S$.
+by the 3D position $(x,y,z)$ of the reference point of the ligand (its geometric center), the rotation around $x$, $y$, and $z$ axes,
+the index of the conformation $l\in \L$, and the scale $s \in \S$.
 All possible configurations form the configuration space $\C$ and the collision-free configurations
@@
 
-Depending on the chemical interactions between the ligand and the protein surface, the ligand may enter a tunnel with various rotations and in various conformations.
+Depending on the chemical interactions between the ligand and the protein surface, the ligand may enter a tunnel with various rotations and in various spatial conformations.
 The traversability of the tunnel should therefore be evaluated using multiple starting configurations $\QI$.
@@
 To find a new initial configuration, a random sample $q$ is generated around the first sphere of the tunnel 
-$c_1$ in the distance $\RI$ (the translation and rotation  parts of $q$ are generated randomly, the scale is set to $\smin$ and the conformation index is set randomly).
+$c_1$ in the distance $\RI$ (the translation and rotation parts of $q$ are generated randomly, the scale is set to $\smin$, and the conformation index is set randomly).
 If the sample $q$ is collision-free, it can be considered as a new starting configuration and it is added to $\QI$.
@@
 
-\subsection{RRT for computing ligands trajectory along a tunnel}
+\subsection{RRT for Computing Ligand's Trajectory along Tunnel}
 
@@
 both 3D rotation and 3D translation.
-The main loop of the planner (Alg.~\ref{alg::main}) is similar to original RRT, but we propose three modifications.
+The main loop of the planner (Alg.~\ref{alg::main}) is similar to the original RRT, but we propose its three modifications.
 
@@
 Setting the virtual goal to this successor allows the tree to avoid such parts of the tunnels that are not traversable or reachable by the ligand.
-This allows the tree to slightly detour from the tunnel and overcome its difficult (narrow) parts by finding an alternative trajectory nearby.
+This allows the tree to slightly detour from the tunnel and overcome its difficult (i.e., narrow) parts by finding an alternative trajectory nearby.
 The algorithm terminates after a predefined number of planning trials $\Imax$ or if the tree reaches
@@
 
-To generate the samples $\qrand$ around the virtual goal $v$, the translation part $(x,y,z)$ of $\qrand$ is generated
-from the Gaussian distribution $N(c_v,\Sigma)$, where $\Sigma$ is the diagonal matrix with diagonal entries equal to the parameter $\rv$, and the rotational
-part of $\qrand$ is generated using techniques described in~\cite{kuffnerES}.
-The other two parameters (ligand index $l$ and scale $s$) of $\qrand$ can be left zero, as these are not used in the employed
-metric for the nearest-neighbor search.
-The parameter $\rv$ influences the distribution of the random samples around the tunnel centerline. 
-By setting $\rv$ to a small value, the planner attempts to find the trajectories inside the tunnel, while higher values
-of $\rv$ cause  exploration of paths around the tunnel.
-We propose to set this parameter to the average width of the tunnel.
-
 \linesnumbered
-\begin{algorithm}[h]
+\begin{algorithm}[bh]
 {\small
@@
    ligand conformations $\L$,
-   scale limits $\smin, \smax$ and $\sdelta$,
-   distance $\dt$ to move virtual goal
+   scale limits $\smin, \smax$ and $\sdelta$
 }
@@
     $\qnear$ = nearest node in $\T$ towards $\qrand$\;
-    expand($\T, \qnear,\qrand$)\;
+    expand($\qnear,\qrand$)\;
     \For{$i= n-1,n-2,\ldots,v+1,v$}{ \nllabel{alg::main:a}
@@
 
+To generate the samples $\qrand$ around the virtual goal $v$, the translation part $(x,y,z)$ of $\qrand$ is generated
+from the Gaussian distribution $N(c_v,\Sigma)$, where $\Sigma$ is the diagonal matrix with diagonal entries equal to the parameter $\rv$, and the rotational
+part of $\qrand$ is generated using techniques described in~\cite{kuffnerES}.
+The other two parameters (ligand index $l$ and scale $s$) of $\qrand$ can be set to zero, as these are not used in the employed
+metric for the nearest-neighbor search.
+The parameter $\rv$ influences the distribution of the random samples around the tunnel centerline. 
+By setting $\rv$ to a small value, the planner attempts to find the trajectories inside the tunnel, while higher values
+of $\rv$ cause the exploration of paths around the tunnel.
+We propose to set this parameter to the average width of the tunnel.
+
 The second proposed modification is the expansion procedure (Alg.~\ref{alg::expand}), which expands the tree from $\qnear$ towards $\qrand$.
-The expansion towards $\qrand$ is crucial for the growth of the tree towards unexplored areas of the configuration space.
+The expansion towards $\qrand$ is crucial for the growth of the tree towards the unexplored areas of the configuration space.
 Traditionally, a line is constructed from $\qnear$ to $\qrand$ and the tree is expanded by a configuration located on this line in the 
 distance $\varepsilon$ from $\qnear$.
-This would however mostly lead to a collision in the case of ligands.
+However, in the case of ligands, this would mostly lead to a collision.
 
@@
 until a collision-free sample is found or until the minimal reduced-scale $\smin$ is reached.
-By testing first the largest scale ($\smax$), the expansion procedure preffers expanding the tree by larger scales.
-This forces the planner to automatically use smaller scaled in the difficult (narrow) parts of the tunnels, while prefering
+By testing first the largest scale ($\smax$), the expansion procedure prefers expanding the tree by larger scales.
+This forces the planner to automatically use smaller scales in the difficult (narrow) parts of the tunnels, while preferring
 large scales if possible.
@@
 The third modification is the utilization of another metric during the expansion procedure (line~\ref{alg::expand:a} in Alg.~\ref{alg::expand}).
-This metric is denoted $\da(\cdot,\cdot)$ and it is referred to as atom-based metric in the rest of the paper.
+This metric is denoted as $\da(\cdot,\cdot)$ and it is referred to as atom-based metric in the rest of the paper.
 To evaluate $\da(q_1,q_2)$, the absolute positions of the atoms of the ligand defined by $q_1$ and $q_2$ are computed.
@@
 is useful when multiple conformations are utilized.
-By employing this metric in the expansion procedure, the tree is expanded by such a new configuration, which atoms maximally approach
+By employing this metric in the expansion procedure, the tree is expanded by such a new configuration, whose atoms maximally approach
 the atoms of the configuration $\qrand$.
@@
 (i.e., for ligand with no translation and no rotation).
-To find the nearest atom to $\qrand$ for a ligand at the configuration $q=(x,y,z,r_x,r_y,r_z,l,s)$, the translation vector $V=(x,y,z)$ and
-rotation matrix $R$ given by $r_x$, $r_y$ and $r_z$ are computed.
-Then, $\qrand$ is translated to $\qrand' = R^{-1}(\qrand-V)$ and the KD-tree search is used to find the nearest point towards $\qrand'$, which
+To find the nearest atom to $\qrand$ for a ligand at the configuration $q=(x,y,z,r_x,r_y,r_z,l,s)$, the translation vector $T=(x,y,z)$ and
+rotation matrix $R$ given by $r_x$, $r_y$, and $r_z$ are computed.
+Then, $\qrand$ is translated to $\qrand' = R^{-1}(\qrand-T)$ and the KD-tree search is used to find the nearest point towards $\qrand'$, which
 is also the nearest point of the ligand at $q$ to $\qrand$.
 
-\begin{algorithm}[h]
+\begin{algorithm}[th]
 {\small
@@
 \KwIn{
-   tree $\T$,
    configuration $\qnear$ to be expanded,
@@
    scale limits $\smin, \smax$, and $\sdelta$,
+   tree $\T$
 }
@@
 Selection of a suitable set of conformations $\L$ is therefore crucial.
-Depending on the ligand, some conformations can be considered as more packed i.e., the atoms in such conformations are close to each other, but 
-the atoms can also form a more longer shape (Fig.~\ref{fig::m040c}).
-
-From motion planning point of view, both types of conformations are useful.
-The longer conformations can better fit in a narrow passage, while the packed conformations better rotate at a place.
-In the narrow parts of the tunnels, the tree can be extended by the longer conformations, while in the more open regions, also the
-packed configurations can be used.
-
-To select the suitable set $\L$ of conformations from a pool of available conformations, a simple technique is proposed.  
-For each conformation, the geometric center is calculated and the deviation of the distances of the atoms from the center is calculated.
-The low value of the deviation is characteristic for the packed conformations, while larger deviations correspond to the longer conformations.
-To select $n$ conformations from all $m$ available conformations, they are sorted according to the deviation.
-Each $k-$th conformation from the sorted list, where $k=\lfloor{m\over n}\rfloor$, is added to $\L$.
-
-\begin{figure}
+Depending on the ligand, some conformations can be considered as more packed i.e., the atoms in such conformations are closer to each other, but 
+the atoms can also form a more elongated shape (Fig.~\ref{fig::m040c}).
+
+From the motion planning point of view, both types of conformations are useful.
+The longer conformations can better fit in a narrow passage, while the packed conformations can be better rotated at a given position.
+In the narrow parts of the tunnel, the tree can be extended by the longer conformations, while in the wider regions, also the
+packed conformations can be used.
+
+\begin{figure}[b!]
 \centering
@@
 \end{figure}
+
+To select the suitable set $\L$ of conformations from a pool of available conformations, a simple technique is proposed.  
+For each conformation, the geometric center is calculated and the deviation of the distances of the atoms from the center is calculated.
+The low value of the deviation is characteristic for the packed conformations, while larger deviations correspond to the longer conformations.
+To select $n$ conformations from all $m$ available conformations, they are sorted according to the deviation.
+Each $k-$th conformation from the sorted list, where $k=\lfloor{m\over n}\rfloor$, is added to $\L$.
 
@@
 
-The proposed approach was used to analyze tunnel traversability towards the active site of Haloalkane dehalogenase protein (PDB ID 1CQW).
-The active site was defined using residues 38 (Asparagine), 102 (Histidine) and 103 (Aspartate).
-The molecular dynamics of the protein was computed using the Amber tool.
+The proposed approach was used to analyze tunnel traversability towards the active site of the Haloalkane dehalogenase protein (PDB ID 1CQW).
+The active site was defined using residues 38 (Asparagine), 102 (Histidine), and 103 (Aspartic Acid).
+The molecular dynamics of the protein was computed using the AMBER tool~\cite{amber}.
 In each frame, tunnels for the spherical probe 0.9~\AA\ were detected using CAVER 3.0~\cite{caver3}. 

### 2017-09-10 paper.tex

@@
 (i.e., for ligand with no translation and no rotation).
-To find the nearest atom to $\qrand$ for a ligand at the configuration $q=(x,y,z,r_x,r_y,r_z,l,s)$, the translation vector $T=(x,y,z)$ and
-rotation matrix $R$ given by $r_x$, $r_y$, and $r_z$ are computed.
-Then, $\qrand$ is translated to $\qrand' = R^{-1}(\qrand-T)$ and the KD-tree search is used to find the nearest point towards $\qrand'$, which
+To find the nearest atom to $\qrand$ for a ligand at the configuration $q=(x,y,z,r_x,r_y,r_z,l,s)$, the translation vector $V=(x,y,z)$ and
+rotation matrix $R$ given by $r_x$, $r_y$ and $r_z$ are computed.
+Then, $\qrand$ is translated to $\qrand' = R^{-1}(\qrand-V)$ and the KD-tree search is used to find the nearest point towards $\qrand'$, which
 is also the nearest point of the ligand at $q$ to $\qrand$.
@@
 
-From the motion planning point of view, both types of conformations are useful.
-The longer conformations can better fit in a narrow passage, while the packed conformations can be better rotated at a given position.
-In the narrow parts of the tunnel, the tree can be extended by the longer conformations, while in the wider regions, also the
-packed conformations can be used.
-
-\begin{figure}[b!]
+\begin{figure}[bh]
 \centering
@@
 \end{figure}
+
+From the motion planning point of view, both types of conformations are useful.
+The longer conformations can better fit in a narrow passage, while the packed conformations can be better rotated at a given position.
+In the narrow parts of the tunnel, the tree can be extended by the longer conformations, while in the wider regions, also the
+packed conformations can be used.
 

### 2017-09-10 paper.tex
first image table correected
corrections in the rest of the paper

@@
 
-The proposed approach was used to analyze tunnel traversability towards the active site of the Haloalkane dehalogenase protein (PDB ID 1CQW).
+The proposed approach was used to analyze the ligand traversability towards the active site using the main tunnel in the Haloalkane dehalogenase protein (PDB ID 1CQW).
 The active site was defined using residues 38 (Asparagine), 102 (Histidine), and 103 (Aspartic Acid).
@@
 In most of the frames, two tunnels were detected.
-The average bottleneck of the tunnels are $1.48$~\AA\ and $1.2$~\AA\ for the first and the second tunnel, resp.
-The tunnel traversability was analyzed in 100 frames.
+The average bottleneck sizes of these tunnels are $1.48$~\AA\ and $1.2$~\AA\ for the main and the side tunnel, respectively.
+The tunnel traversability was analyzed in 100 time steps (frames) of MD simulation.
 
@@
 For each scale, $|\QI|=20$ collision-free initial configurations were found in the $\RI=2$~\AA\ radius around the first sphere of the tunnel (outside the protein).
-For each frame, minimal scale $\smin$ and each initial configuration, 10 trajectories were computed, which resulted
+For each initial configuration, 10 trajectories were computed, which resulted
 in $100 \times 4 \times 20 \times 10 = 80,000$ trajectories per ligand.
 
-Two motion planners were utilized: the proposed planner (described in Alg.~\ref{alg::main} and Alg.~\ref{alg::expand}), 
-    denoted as \RA\ in the rest of the paper, and the Retraction-based RRT~\cite{zhangRetraction} denoted as \RB\ in the rest of the paper.
+Two motion planners were utilized: the proposed planner (described in Alg.~\ref{alg::main} and Alg.~\ref{alg::expand}), denoted as \RA\ in the rest of the paper, and the Retraction-based RRT~\cite{zhangRetraction}, denoted as \RB\ in the rest of the paper.
 The \RA\ method was run with the following parameters:
 $\Imax=5,000$, $m=100$, $\gb=0.9$, $\dt=1.5$~\AA, $\rv=2$~\AA\ and $\varepsilon=0.05$~\AA.
-Both methods were implemented in the same software framework and they utilized the same data structures and routines for collision detection and nearest-neighbor search.
-The nearest-neighbor search of both planners is based on the 6D Euclidean metric with the same weights for rotation and translation parts.
+Both methods were implemented in the same software framework and they utilized the same data structures and routines for the collision detection and nearest-neighbor search.
+The nearest-neighbor search of both planners was based on the 6D Euclidean metric with the same weights for rotation and translation parts.
 This metric was selected based on preliminary experiments with both planners as it resulted in the success rates.
@@
 The tested methods differ in the expansion step (lines \ref{alg::expa}--\ref{alg::expb} in Alg.~\ref{alg::expand}).
-In these lines, the \RB\ method employs the retraction step in order to find a new collision-free configurations $\qnew$ that approaches $\qrand$.
+In these lines, the \RB\ method employs the retraction step in order to find a new collision-free configuration $\qnew$ that approaches $\qrand$.
 The retraction step is realized by sampling the vicinity of the configuration $\qnear$ in the distance $0.05$~\AA\ using 20 samples.
@@
 The retraction step is repeated 5 times.
-The \RB\ method can therefore expand the tree by at most 5 new nodes per conformation, while \RA\ extends the tree only by one configuration.
-The larger number of nodes in \RB\ method is not problematic, as the KD-trees were used for nearest-neighbor search and the time to solve these queries is significantly less than computing collision detection.
-Parameters of the methods were selected in such a way that the number of collision detection queries in each expansion step is same.
-For each conformation and tested scale, the \RA\ methods attempts to find $m=100$ collision-free configurations, and \RB\ attempts to perform $5 \times 20$ collision-detection queries.
-Other parameters of the \RB\ planner were same as for \RA.
-Planning was terminated after $\Imax$ iterations, or if the tree approached the active site to the distance less than 0.5~\AA.
-
-In the following experiments, we measure performance of the methods using the traversability rate, which is the percentage of frames of the protein dynamic that are traversable by the tested ligand.
-A frame is considered traversable, if at least one trajectory out of 20 (the number of initial configurations $|\QI|$) $\times$ 10 trials approaches the active site to the distance less than 2~\AA.~\footnote{We refer to~\url{http://mrs.felk.cvut.cz/icra2018protein} for additional material}
-
-\subsection{Traversability of one atom}
-
-The aim of the first experiment is to measure performance of both planner in the simplest scenario: finding path for one sphere towards the active site.
-The radius of the sphere is $1.7$~\AA.
+The \RB\ method can therefore expand the tree by at most 5 new nodes per configuration, while \RA\ extends the tree only by one configuration.
+The larger number of nodes in \RB\ method is not problematic, as the KD-trees were used for the nearest-neighbor search and the time to solve these queries is significantly smaller than the computing collision detection.
+The parameters of the methods were selected in such a way that the number of collision detection queries in each expansion step is the same.
+For each conformation and tested scale, the \RA\ method attempts to find $m=100$ collision-free configurations, and \RB\ attempts to perform $5 \times 20$ collision-detection queries.
+Other parameters of the \RB\ planner were the same as for \RA.
+The planning was terminated after $\Imax$ iterations or if the tree approached the active site to the distance less than 0.5~\AA.
+
+In the following experiments, we measured the performance of the methods using the traversability rate, which is the percentage of frames of the protein dynamics that are traversable by the tested ligand.
+A frame is considered traversable if at least one trajectory out of 20 (the number of initial configurations $|\QI|$) $\times$ 10 trials approaches the active site to the distance less than 2~\AA.~\footnote{We refer to~\url{http://mrs.felk.cvut.cz/icra2018protein} for additional material.}
+
+\subsection{Traversability of One Atom}
+
+The aim of the first experiment is to measure the performance of both planners in the simplest scenario, i.e., finding a path for one sphere heading towards the active site.
+The radius of the sphere is set to $1.7$~\AA.
 
 The computed traversability rates are shown in Tab.~\ref{tab::one}.
-The maximum percentage of traversable frames (the row Maximum in Tab.~\ref{tab::one}) is computed as the percentage of tunnels with bottleneck larger than the tested radius.
-The row Planning shows the traversable rate computed by the planners.
+The maximum percentage of traversable frames (the row Maximum in Tab.~\ref{tab::one}) is computed as the percentage of tunnels with the bottleneck larger than the tested radius.
+The row 'Planning' shows the traversable rate computed by the planners.
 This rate decreases with the increasing radius and it is lower than the maximum traversable rate.
-The rates provided by \RA\ and \RB\ are similar which is a good indicator that the methods have been programmed correctly.
+The rates provided by \RA\ and \RB\ are similar which is a good indicator that the methods have been implemented correctly.
 Higher traversability rate can be achieved by increasing the maximum number of planning iterations $\Imax$, which would also
@@
 
-\begin{table}
+\begin{table}[ht]
 \centering
@@
 
-\subsection{Influence of the conformation selection}
-
-The aim of this experiment is to verify the procedure to select ligand conformations for the motion planning.
-The experiment was performed with the ligand DCP (Dichloropropanol), for which $1300$ conformations where available and for the ligand
+\subsection{Influence of the Conformation Selection}
+
+The aim of this experiment was to verify the procedure to select feasible ligand conformations for the motion planning task.
+The experiment was performed with the ligand DCP (Dichloropropanol), for which $1300$ conformations were available, and for the ligand
 m040 (1,5-Dichloropentane) with $7560$ available conformations.
-Two strategies were tested: the strategy described in Sec.~\ref{sec::strat} (denoted `S') and a strategy that prefers smaller conformations (denoted `P').
-The `P' method selects such conformations where the deviation of the atomic distance from the geometric center are lowest, which leads 
-to selection of the packed conformations (Fig.~\ref{fig::m040c}).
+Two strategies were tested: the strategy described in Sec.~\ref{sec::strat} (denoted `S') and the strategy that prefers smaller conformations (denoted `P').
+The `P' method selects such conformations where the deviations of the atomic distance from the geometric center are the lowest, which leads 
+to the selection of packed conformations (Fig.~\ref{fig::m040c}).
 
@@
 The number after the slash in the ligand's name denotes the number of atoms of the ligand.
-In the case of both tested ligands and 50 selected conformations, the `S' strategy leads to higher traversability rate than the `P' strategy.
-For the ligand DCP, a higher traversability rate was achieved when 50 conformations were selected.
-This experiments shows that selecting 50 conformations using the `S' strategy (Sec.~\ref{sec::strat}) leads to better results
-than using less number of conformations or when preferring the packed conformations.
-
-\begin{table}
+In the case of both tested ligands and 50 selected conformations, the `S' strategy leads to a higher traversability rate than the `P' strategy.
+For the DCP ligand, a higher traversability rate was achieved when 50 conformations were selected.
+This experiment shows that selecting 50 conformations using the `S' strategy (Sec.~\ref{sec::strat}) leads to better results
+than using a less number of conformations or when preferring the packed conformations.
+
+\begin{table}[ht]
 \centering
 \caption{\label{tab::selection}
-    Influence of the conformation selection method to the traversability rates in the first tunnel.
+    The influence of the conformation selection method to the traversability rates in the first tunnel.
     The number after~`$/$' denotes the number of atoms.
@@
 
-\subsection{Traversability of ligands}
-
-Based on the results of the previous experiment, 50 conformations using the `S' strategy were selected for the following
-ligands: m037t (1,2-Dichloroethane), m038t (1,3-Dichloropropane), m056 (2-Chlorobutane) and m080 (Trichloropropane), and
-the task was to compute trajectories for these ligands in the MD sequence.
+\subsection{Traversability of Ligands}
+
+Based on the results of the previous experiment, 50 conformations using the `S' strategy were selected for ligands m037t (1,2-Dichloroethane), m038t (1,3-Dichloropropane), m056 (2-Chlorobutane), and m080 (Trichloropropane), and
+the task was to compute the trajectories for these ligands in a given molecular dynamics simulation.
 
 The results are summarized in Tab.~\ref{tab::main}.
-The table shows the traversability rate only for the first tunnel, as no trajectory was found in the second tunnel for most of the tested ligands.
-The only exception is the ligand m037t, which was the smallest tested ligand (with 8 atoms), and therefore it can fit even inside the second tunnel.
+The table shows the traversability rate only for the main tunnel, as no trajectory was found in the side tunnel for most of the tested ligands.
+The only exception is the ligand m037t, which was the smallest tested one (with 8 atoms), and therefore it could fit even inside the side tunnel.
 The traversability rate is higher for the most scaled-down ligands ($\smin=0.5$) and it decreases with the increasing scale.
@@
 
-{\bf doufam, ze toto neni moc drze:}
-The traversability of the DCP ligand to active site was approved by the MD simulation.
-The observed trajectory of the DCP ligand in the MD simulation followed the first tunnel.
-Based on the traversability rates with $\smin=0.8$, we expect that at least ligands m037t and m038t can also reach the active site, as
-they have same (or even larger) traversability rate than DCP.
-However, this hypothesis hasn't been tested against MD simulations yet.
-
-The proposed \RA\ planner provides higher traversability rate than \RB.
-The distances of the configuration trees (measured using 3D Euclidean metric) to the active site and the runtimes in each frame
-are shown in Fig.~\ref{fig::comparison}.
-The runtimes of both algorithms increase in those frames, where the tunnel is not traversable, i.e., where the distance
-to the active site is more than 2~\AA\ (for example,in the frame number 60).
-
-The most time consuming part of the planners is the collision detection. 
-Although the both methods can utilize the same number of collision detection queries in the expansion step, \RB\ is slower than \RA.
-The reason is, that \RA\ expands the tree more closer to the random samples $\qrand$ than \RB.
-This results in a faster exploration of the configuration space and therefore, \RA\ needs less number of iterations to reach the active site.
-The better performance of \RA\ is caused by the proposed atom-based metric, because this distance depends on the actual shape of the ligand and 
-it allows the ligands to retract more toward $\qrand$.
-Contrary, \RB\ utilizes the single 6D Euclidean weighted metric in the expansion, where the weights are same for all conformations.
-Depending on these weights, the metric favors samples that differ from $\qrand$ more in rotation or translation part.
-However, the weights between rotation/translation parts should depends on the shape of the ligand, which is different in each conformation.
-By using single set of weights in the 6D Euclidean metric, we implicitely favor either the packed or the longer conformations.
-%To improve performance of \RB, the weights of the 6D Euclidean metric should differ for each conformation, but it would
-%prevent the usage KD-tree for the nearest-neighbor search.
-
-\begin{table}
+\begin{table}[ht]
 \centering
@@
 
+%{\bf doufam, ze toto neni moc drze:}
+In conclusion, the traversability of the DCP ligand to active site was approved by the MD simulation.
+The observed trajectory of the DCP ligand in the MD simulation followed the main tunnel, as expected.
+Based on the traversability rates with $\smin=0.8$, we expect that at least ligands m037t and m038t can also reach the active site, as
+they have the same (or even larger) traversability rate than the DCP.
+However, this hypothesis has not been approved yet by testing it on MD simulations.
+
+The proposed \RA\ planner provides higher traversability rate than \RB.
+The distances of the configuration trees (measured using the 3D Euclidean metric) to the active site and the runtimes in each frame
+are shown in Fig.~\ref{fig::comparison}.
+The runtimes of both algorithms increase in those frames, where the tunnel is not traversable, i.e., where the distance
+to the active site is more than 2~\AA\ (for example, in the frame no. 60).
+
 {\def\a{270}
-\begin{figure}
+\begin{figure}[hb]
 \centering
@@
 \caption{\label{fig::comparison}
-    Runtimes and distance to the active site for the m037 ligand at scale 0.6.
+    Runtimes and the distance to the active site for the m037 ligand at scale 0.6.
     The values are computed from all 200 trials (20 initial configurations $\times$ 10 trials).
@@
 }
+
+The most time-consuming part of the planners is the collision detection. 
+Although both methods can utilize the same number of collision detection queries in the expansion step, \RB\ is slower than \RA.
+The reason is that \RA\ expands the tree more closer to the random samples $\qrand$ than \RB.
+This results in a faster exploration of the configuration space and therefore, \RA\ needs less number of iterations to reach the active site.
+The better performance of \RA\ is caused by the proposed atom-based metric, because this distance depends on the actual shape of the ligand and 
+it allows the ligands to retract more towards $\qrand$.
+On the contrary, \RB\ utilizes the single 6D Euclidean weighted metric in the expansion, where the weights are the same for all conformations.
+Depending on these weights, the metric favors samples that differ from $\qrand$ more in the rotation or translation part.
+However, the weights between the rotation/translation parts should depend on the shape of the ligand, which is different in each conformation.
+By using a single set of weights in the 6D Euclidean metric, we implicitly favor either the packed or the longer conformations.
+%To improve performance of \RB, the weights of the 6D Euclidean metric should differ for each conformation, but it would
+%prevent the usage KD-tree for the nearest-neighbor search.
+
+\section{Discussion and Utilization of the Results}
+
+The work presented in the paper was motivated by the current needs of biochemists who need to analyze the traversability of selected ligands through protein tunnels.
+To find the feasible ligand trajectories, it is often required to shrink the atomic radii of the ligand, as utilized also in other existing tools.
+Finding trajectories for the scaled-down ligands enables to solve the problem with narrow passages, but it can be questionable how biochemically relevant such trajectories are.
+
+Judging the accessibility of the active site by a ligand based only on the traversability rates may be problematic if the ligand is scaled too much.
+Instead of using a single version of the scaled-down ligand, we only limit the minimal scale used by the planner (the parameter $\smin$).
+The expansion step always prefers to expand the tree with the maximal scale, that leads to a new collision-free configuration.
+The information about the scales used at the trajectory points (which can be extracted from the configuration tree of the proposed planner) brings the biochemists new knowledge about the tunnels, which was impossible to get using the previous solutions.
+Our proposed visualization allows the biochemists to easily observe those parts of the tunnel where the ligand had to be scaled-down significantly (red parts of trajectories in Fig.~\ref{fig::dcp}), and parts where the ligand could be enlarged again (green parts in Fig.~\ref{fig::dcp}).
+According to the results of the motion planing, the DCP needs to shrink down to ca. 75~\% of its size (Fig.~\ref{fig::dcp}, point B).
+However, as both the protein and ligand are mutually influence each other, it is possible that even a very narrow part of the tunnel is traversable.
+This was observed in the MD simulation of the DCP ligand in the 1CQW protein, which revealed that DCP can pass the main tunnel.
 
@@
 The color of the trajectories corresponds to the scale, that was needed to pass the given part of the tunnel.
-The ligand can enter the first tunnel with a high scale (almost 1=100~\%) (A), but then it has to scale-down (to $\sim75$~\%) to pass the
-next part of the tunnel (B). After this narrow passage, the tunnel is larger and the ligand can traverse to the narrow passage with
+The ligand can enter the first tunnel with a high scale (almost 1=100~\%) (A), but then it has to be scaled-down (to $\sim75$~\%) to pass the
+next part of the tunnel (B). After this narrow passage, the tunnel is larger again and the ligand can traverse to the narrow passage with
 scale $>75$~\% (C).
 The second tunnel can also be entered with the high scale (green), but then the narrow part is traversable only with a low scale ($\sim50$~\%) (D). 
-For the purpose of the visualization, even more scale-down to 0.3=30~\% was allowed (E) to see how the ligand pass the tunnel.
-However, the ligand can reach the active site only with very low scale ($\sim0.4=40$~\%) using the second tunnel.
+For the purpose of the visualization, even more scale-down to 0.3=30~\% was allowed (E) to see how the ligand passed the tunnel.
+However, the ligand can reach the active site only with very low scale ($\sim0.4=40$~\%) using the side tunnel.
 }
 \end{figure}
-
-\section{Discussion and utilization of the results}
-
-The work in the paper is motivated by real needs of bio-chemists who need to analyze traversability of selected ligands in the tunnels.
-To enable finding the trajectories, it is necessary to shrink the atomic radii of the ligands which has been utilized also in other tools~\cite{cortes2005path,guieysse2008structure}.
-Finding trajectories for scaled-down ligands on one hand helps to cope with the narrow passages, but the question is, how realistic
-such trajectories are.
-
-Judging the accessibility of the active sites by the ligands based only on the traversability rates may be problematic if the ligands are scaled too much.
-Instead of using a single version of the scale-down ligand, we only limit the minimal scale used by the planner (the parameter $\smin$).
-The expansion step always prefers to expand the tree with the maximal scale, that leads to a new collision-free configuration.
-The information about the scales used at the trajectory points (which can be extracted from the configuration tree of the proposed planner) brings the chemists new knowledge about the tunnels.
-It allows them to detect parts of the tunnels, where the ligand has to scale-down significantly, and parts, where the ligand can be again
-larger (Fig.~\ref{fig::dcp}).
-According to the results of the motion planinng, DCP needs to shrink down to ca. 75~\% of its size (Fig.~\ref{fig::dcp}, point B).
-However, as both the protein and the ligand adapt to each other, it is possible, that even a narrow part of the tunnel is traversable.
-This was observed in the MD simulation of the DCP ligand in the 1CQW protein, which revealed that DCP can pass the first tunnel.
 
@@
 
-In this paper, we presented a modification of the RRT planner for trajectory computation of ligands inside protein tunnels.
-The tunnels are pathways leading from an outer environment toward the active sites, where chemical interactions may undergo.
-So far, the tunnels have been detected using Voronoi diagrams, which does not take into account shape of the ligand neither the possible 
-conformation changes.
-In the proposed work, we compute trajectories using modified RRT for whole ligands and we model the conformation changes of the
-ligands using a predefined set of conformations.
-The atomic radii of the ligands can be scaled down up to a certain predefined factor.
-The planner generates the random samples along a tunnel computed by Voronoi diagram and it attempts to expand the configuration
-tree using all conformations. 
-The expansion prefers atoms with a higher scale, which ensures that the ligand shrinks down only if traversing narrow parts of the tunnels.
-To further boost exploration of the configuration space, a novel metric suitable for measuring distances between various shapes
-of the ligands, is used in the expansion step.
-The proposed method has been used to  evaluate trajectories of several ligands in a sequence of protein dynamics.
-Traversability of one of the tested ligands has been already approved in MD simulations.
+In this paper, we presented a novel modification of the RRT planner for computation of ligand trajectory inside protein tunnels.
+So far, tunnels have been mostly detected using Voronoi diagrams which do not take into account the shape and flexibility of the ligand.
+In our solution, we compute the trajectories using the modified RRT which takes into account the ligand shape and we model the ligand flexibility using a predefined set of its conformations.
+To simulate the real behavior of ligand passing through a tunnel, we enable scaling down the atomic radii of the ligands up to a certain predefined factor.
+Furthermore, our planner is localized as it generates random samples only along a tunnel computed by Voronoi diagram, and it attempts to expand the configuration
+tree using all ligand conformations. 
+The expansion prefers atoms with a higher scale, which ensures that the ligand shrinks down only if necessary, i.e., when traversing the narrow parts of the tunnel.
+To further boost the exploration of the configuration space, a novel metric suitable for measuring the distances between various ligand shapes is used in the expansion step.
+The proposed method has been used to evaluate the trajectories of several ligands in a sequence of frames capturing protein dynamics.
+The traversability of one of the tested ligands has been already approved in the MD simulations.
 

### 2017-09-10 paper.tex
ligand-approaximating changed

--- 2017-09-10 paper.tex
@@
 {\footnotesize
-\renewcommand{\arraystretch}{0.1}
-\renewcommand{\tabcolsep}{0pt}
+\renewcommand{\arraystretch}{-0.5}
+\renewcommand{\tabcolsep}{-1pt}
 \begin{tabular}{ccc}
@@
 Based on the observations of MD simulations, tunnels adapt their shape to the passing ligands and vice versa.
-In consequence, even very narrow tunnels, being too small for the passage of a given ligand approximating sphere, can serve as the transportation path for this ligand.
+In consequence, even very narrow tunnels, being too small for the passage of a given ligand-approximating sphere, can serve as the transportation path for this ligand.
 Therefore, also the existing Voronoi-based methods enable the user to decrease the size of the approximating sphere which simulates shrinking of the ligand size.
@@
 both 3D rotation and 3D translation.
-The main loop of the planner (Alg.~\ref{alg::main}) is similar to the original RRT, but we propose its three modifications.
+The main loop of the planner (Alg.~\ref{alg::main}) is similar to the original RRT, but with the following three proposed modifications.
 
 The aim of the first modification is to sample the configuration space along the tunnel being analyzed.
-To guide the growth of the tree using the tunnel, a moving virtual goal is used~\cite{vonasek2009rrt}.
+To guide the growth of the tree using the tunnel, a moving virtual goal is used. %~\cite{vonasek2009rrt}.
 The virtual goal $v, 1\le v \le n$, is the index of a sphere of the tunnel.

### 2017-09-10 paper.tex
fig 3: graphs on one line
few changes

@@
 The ligand flexibility is modeled using a fixed set of predefined conformations which the ligand can occupy.
-
 Our proposed method was tested by the biochemists who compared the results with trajectories of ligands determined experimentally in the lab.
@@
 leading from outside the protein to the active site.
-The tunnels are detected in a sequence time steps capturing protein dynamic behavior that is computed using molecular dynamics (MD) simulations.
+The tunnels are detected in a sequence time steps capturing protein dynamic behavior computed using MD simulations.
 
@@
 
-\begin{figure}[bh]
+\begin{figure}[b]
 \centering
@@
 {\small
+\renewcommand{\arraystretch}{0.7}
 \input graphProbe.tex
@@
 {\small
+\renewcommand{\arraystretch}{0.7}
 \input graphSel.tex
@@
 {\small
+\renewcommand{\arraystretch}{0.7}
 \input graph.tex
@@
 \centering
-\includegraphics[width=0.25\textwidth,angle=\a]{fig/m037tmins-03tunnel-1rrt_path_time-2}
-\includegraphics[width=0.25\textwidth,angle=\a]{fig/m037tmins-03tunnel-1trajDistToGoal-2}
+%\includegraphics[width=0.17\textwidth,angle=\a]{fig/m037tmins-03tunnel-1rrt_path_time-2}
+%\includegraphics[width=0.17\textwidth,angle=\a]{fig/m037tmins-03tunnel-1trajDistToGoal-2}
+\includegraphics[width=0.18\textwidth,angle=\a]{fig/crop1}
+\includegraphics[width=0.18\textwidth,angle=\a]{fig/crop2}
 \caption{\label{fig::comparison}

### 2017-09-10 paper.tex
joined paragraphs in the section IV-D

@@
 is useful when multiple conformations are utilized.
-By employing this metric in the expansion procedure, the tree is expanded by such a new configuration, whose atoms maximally approach
+This metric ensures that the tree is expanded by such configurations, whose atoms maximally approach
 the atoms of the configuration $\qrand$.
@@
 the atoms can also form a more elongated shape (Fig.~\ref{fig::m040c}).
+From the motion planning point of view, both types of conformations are useful.
+The longer conformations can better fit in a narrow passage, while the packed conformations can be better rotated at a given position.
+In the narrow parts of the tunnel, the tree can be extended by the longer conformations, while in the wider regions, also the
+packed conformations can be used.
 
@@
 \end{figure}
-
-From the motion planning point of view, both types of conformations are useful.
-The longer conformations can better fit in a narrow passage, while the packed conformations can be better rotated at a given position.
-In the narrow parts of the tunnel, the tree can be extended by the longer conformations, while in the wider regions, also the
-packed conformations can be used.
 

### 2017-09-10 paper.tex
title, small changes

--- 2017-09-10 paper.tex
@@
 
-\title{Motion planning for assessing accessibility of protein tunnels}
+\title{Assessing the Accessibility of Protein Tunnels using Motion Planning}
 
@@
 The ligand flexibility is modeled using a fixed set of predefined conformations which the ligand can occupy.
+
 Our proposed method was tested by the biochemists who compared the results with trajectories of ligands determined experimentally in the lab.
@@
 leading from outside the protein to the active site.
-The tunnels are detected in a sequence time steps capturing protein dynamic behavior computed using MD simulations.
+The tunnels are detected in a sequence time steps capturing protein dynamic behavior that is computed using molecular dynamics (MD) simulations.
 
@@
 
-\begin{figure}[b]
+\begin{figure}[bh]
 \centering

### 2017-09-10 paper.tex
last par. in intro to experiments shorten

--- 2017-09-10 paper.tex
@@
 The nearest-neighbor search of both planners was based on the 6D Euclidean metric with the same weights for rotation and translation parts.
-This metric was selected based on preliminary experiments with both planners as it resulted in the success rates.
+This metric was selected based on preliminary experiments with both planners as it resulted in the highest success rates.
 
@@
 In the following experiments, we measured the performance of the methods using the traversability rate, which is the percentage of frames of the protein dynamics that are traversable by the tested ligand.
-A frame is considered traversable if at least one trajectory out of 20 (the number of initial configurations $|\QI|$) $\times$ 10 trials approaches the active site to the distance less than 2~\AA.~\footnote{We refer to~\url{http://mrs.felk.cvut.cz/icra2018protein} for additional material.}
+A frame is considered traversable if at least one trajectory out of 200 ($\!|\QI|\!\times\!\!10$ trials) approaches the active site to the distance less than~2~\AA.~\footnote{We refer to~\url{http://mrs.felk.cvut.cz/icra2018protein} for additional material.}
 
@@
     The values are computed from all 200 trials (20 initial configurations $\times$ 10 trials).
-    Only each second frame is shown in the graph.
+    Only each third frame is shown in the graph.
 }

### 2017-09-10 paper.tex
intro

--- 2017-09-10 paper.tex
@@
 %The tunnel has to be wide enough and also the physico-chemical properties of the surrounding amino acids have to be compatible with the ligand.
-Biochemists are currently using computational methods for tunnel detection in order to estimate the possibility of interactions between a protein and a ligand without the need to perform laboratory experiments (Fig.~\ref{fig::motiv}).
-The most common methods detect tunnels using Voronoi diagrams~\cite{yaffe2008,caver3} assuming a spherical ligand (probe), i.e., the ligand is approximated by a bounding sphere.
-However, as the ligands are typically non-spherical and flexible, these methods can often skip possible candidate tunnels.
+Currently the biochemists are using various computational methods for tunnel detection (Fig.~\ref{fig::motiv}) in order to estimate the possibility of interactions between a protein and ligand.
+This helps them to reduce the number of necessary in-vitro lab experiments.
+The most common methods detect the tunnels using Voronoi diagrams~\cite{yaffe2008,caver3}, assuming a spherical ligand (probe), i.e., the approximation of the ligand by a bounding sphere.
+However, as the ligands are typically non-spherical, these methods can often skip possible candidate tunnels.
 
@@
 
-To provide the biochemists with a solution which detects also candidate tunnels omitted by the traditional methods, it is necessary take into account also the ligand shape and its conformation changes when the ligand moves, i.e., changes its conformation.
-Such a ligand changing the mutual position of its atoms within a molecular dynamics is denoted as a flexible ligand.
-To address this, we can formulate the task as a motion planning problem.
-It is also necessary to cope with the narrow passage problem, as the proteins are dense structures and the movements of the ligand are limited by the surrounding amino acids.
+To provide the biochemists with a solution which detects also the candidate tunnels omitted by the traditional methods, it is necessary take into account not only the ligand shape but also its conformation changes when the ligand moves.
+Such a ligand, changing the mutual position of its atoms within a molecular dynamics, is denoted as a flexible ligand.
+We can formulate this task as a motion planning problem.
+The proposed solution needs also to cope with the narrow passage problem, as proteins are dense structures limiting the ligand movements inside.
 
@@
 Therefore, they have been used for motion planning of various robotic systems.
-Besides many applications in robotics, the sampling-based planners have been already used in biochemistry as well 
+Besides many applications in robotics, the sampling-based planners have been already used in biochemistry as well, 
 %Sampling-based motion planners has been intensively studied in robotics, but they have been applied also in other research fields.

### 2017-09-10 paper.tex
discussion shorten

--- 2017-09-10 paper.tex
@@
 
-The work presented in the paper was motivated by the current needs of biochemists who need to analyze the traversability of selected ligands through protein tunnels.
+The presented work was motivated by the current needs of biochemists who need to analyze the traversability of selected ligands through protein tunnels.
 To find the feasible ligand trajectories, it is often required to shrink the atomic radii of the ligand, as utilized also in other existing tools.
-Finding trajectories for the scaled-down ligands enables to solve the problem with narrow passages, but it can be questionable how biochemically relevant such trajectories are.
+Finding trajectories for the scaled-down ligands enables to solve the problem with narrow passages, but it can be questionable how biochemically relevant such trajectories~are.
 
@@
 Instead of using a single version of the scaled-down ligand, we only limit the minimal scale used by the planner (the parameter $\smin$).
-The expansion step always prefers to expand the tree with the maximal scale, that leads to a new collision-free configuration.
+The expansion step always prefers to expand the tree by a collision-free configuration with the maximal scale.
+%that leads to a new collision-free configuration.
 The information about the scales used at the trajectory points (which can be extracted from the configuration tree of the proposed planner) brings the biochemists new knowledge about the tunnels, which was impossible to get using the previous solutions.

### 2017-09-10 paper.tex
shortening

--- 2017-09-10 paper.tex
@@
 Moreover, the ligand flexibility cannot be principally considered by these approaches.
-Simulations based on molecular dynamics (MD) can directly evaluate migration of a ligand into a protein, but they are significantly more computationally demanding~\cite{kingsley2014including}. 
+Simulations based on molecular dynamics (MD) can directly evaluate the migration of a ligand into a protein, but they are significantly more computationally demanding~\cite{kingsley2014including}. 
 Moreover, MD simulations have to be prepared for a specific ligand/protein pair, which is not practical for testing a large number of ligands.
-The geometric-based methods are therefore still preferable due to their fast speed.
-In order to determine if the ligand can pass the tunnel using the geometric-based methods, it is necessary to compute a trajectory considering the shape of the ligand and its 3D translation and rotation.
+The geometric-based methods are therefore still preferable due to their speed.
+%In order to determine if the ligand can pass the tunnel using the geometric-based methods, it is necessary to compute a trajectory considering the shape of the ligand and its 3D translation and rotation.
 
@@
 Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} is a single-query sampling-based motion planning method that 
-incrementally builds a configuration tree $\T$ rooted at the initial configuration $\qinit$.
-In each iteration of RRT, a random configuration $\qrand \in \C$ is generated and its nearest node $\qnear \in \T$ in the tree is found.
-A new configuration $\qnew$ is constructed on the line connecting $\qnear$ and $\qrand$ in the distance $\varepsilon$ from $\qnear$.
-If $\qnew$ is collision-free, it is added to the tree.
-The algorithm terminates if the tree approaches the goal configuration close enough.
+incrementally builds a configuration tree rooted at the initial configuration.
+%In each iteration of RRT, a random configuration $\qrand \in \C$ is generated and its nearest node $\qnear \in \T$ in the tree is found.
+%A new configuration $\qnew$ is constructed on the line connecting $\qnear$ and $\qrand$ in the distance $\varepsilon$ from $\qnear$.
+%If $\qnew$ is collision-free, it is added to the tree.
+%The algorithm terminates if the tree approaches the goal configuration close enough.
 
@@
 %or protein folding combined with ligand diffusion~\cite{cortes2010simulating}.
-
 The well known issue of the sampling-based planners is the narrow passage problem.
@@
 In this paper, we presented a novel modification of the RRT planner for computation of ligand trajectory inside protein tunnels.
-So far, tunnels have been mostly detected using Voronoi diagrams which do not take into account the shape and flexibility of the ligand.
+%So far, tunnels have been mostly detected using Voronoi diagrams which do not take into account the shape and flexibility of the ligand.
 In our solution, we compute the trajectories using the modified RRT which takes into account the ligand shape and we model the ligand flexibility using a predefined set of its conformations.
@@
 To further boost the exploration of the configuration space, a novel metric suitable for measuring the distances between various ligand shapes is used in the expansion step.
-The proposed method has been used to evaluate the trajectories of several ligands in a sequence of frames capturing protein dynamics.
-The traversability of one of the tested ligands has been already approved in the MD simulations.
+The proposed method has been used to test and evaluate the trajectories of several ligands in a sequence of frames capturing protein dynamics.
+%The traversability of one of the tested ligands has been already approved in the MD simulations.
 

### 2017-09-10 paper.tex
rrt description back in RW

@@
 incrementally builds a configuration tree rooted at the initial configuration.
+In each iteration of RRT, a random configuration $\qrand \in \C$ is generated and its nearest node $\qnear \in \T$ in the tree is found.
+A new configuration $\qnew$ is constructed on the line connecting $\qnear$ and $\qrand$ in the distance $\varepsilon$ from $\qnear$.
+If $\qnew$ is collision-free, it is added to the tree.
+The algorithm terminates if the tree approaches the goal configuration close enough.
+
 %In each iteration of RRT, a random configuration $\qrand \in \C$ is generated and its nearest node $\qnear \in \T$ in the tree is found.

### 2017-09-10 paper.tex
reference reorganized
shortening

@@
 %protein folding~\cite{amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},
-and for protein folding~\cite{raveh2009rapid,novinskaya2015improving}.
+and for protein folding~\cite{raveh2009rapid,novinskaya2015improving,amato2002using}.
 %or for tunnel detection~\cite{vonasek2017tunnel}.
@@
 packed conformations can be used.
-
-\begin{figure}[b]
+\vspace{-3mm}
+
+\begin{figure}[hb]
 \centering
@@
 The table shows the traversability rate only for the main tunnel, as no trajectory was found in the side tunnel for most of the tested ligands.
-The only exception is the ligand m037t, which was the smallest tested one (with 8 atoms), and therefore it could fit even inside the side tunnel.
-The traversability rate is higher for the most scaled-down ligands ($\smin=0.5$) and it decreases with the increasing scale.
-For the scale $\smin=0.8$, the most traversable frames were detected for ligands DCP (8~\%), m037t (20~\%) and m038t (16~\%).
-
-\begin{table}[ht]
+
+\begin{table}[bt]
 \centering
@@
 \end{table}
+
+The only exception is the ligand m037t, which was the smallest tested one (with 8 atoms), and therefore it could fit even inside the side tunnel.
+The traversability rate is higher for the most scaled-down ligands ($\smin=0.5$) and it decreases with the increasing scale.
+For the scale $\smin=0.8$, the most traversable frames were detected for ligands DCP (8~\%), m037t (20~\%) and m038t (16~\%).
 
@@
 
+\vspace{-3mm}
 The most time-consuming part of the planners is the collision detection. 

### 2017-09-10 paper.tex
Adam's comments

@@
 %The tunnel has to be wide enough and also the physico-chemical properties of the surrounding amino acids have to be compatible with the ligand.
-Currently the biochemists are using various computational methods for tunnel detection (Fig.~\ref{fig::motiv}) in order to estimate the possibility of interactions between a protein and ligand.
+Currently the biochemists are using various computational methods for tunnel detection (Fig.~\ref{fig::motiv}) in order to estimate the possibility of interactions between a protein and a ligand.
 This helps them to reduce the number of necessary in-vitro lab experiments.
@@
 
-To provide the biochemists with a solution which detects also the candidate tunnels omitted by the traditional methods, it is necessary take into account not only the ligand shape but also its conformation changes when the ligand moves.
+To provide the biochemists with a solution which detects also the candidate tunnels omitted by the traditional methods, it is necessary to take into account not only the ligand shape but also its conformation changes when the ligand moves.
 Such a ligand, changing the mutual position of its atoms within a molecular dynamics, is denoted as a flexible ligand.
@@
 
-Due to the protein dynamics, the tunnel changes, move and merge, and their bottleneck can change in each time step.
+Due to the protein dynamics, the tunnels move and merge, and their bottleneck can change in each time step.
 Therefore, it can be expected that the detected tunnels are rather narrow.

### 2017-12-06 paper.tex
removed long comments
state of the art updated

@@
-%\documentclass[a4paper, 10pt]{IEEEtran}      % for icra/RAL final
 \documentclass[usletter, 10pt, conference]{ieeeconf} % initial submission to RAL/ICRA
@@
 \usepackage{booktabs}
-%\usepackage{amsmath}
 \usepackage{amsfonts}
@@
 
-%D. Bedn\'{a}\v{r}:
-%Faculty of Science, Masaryk University, Kamenice 5, 625 00 Brno, Czech Republic\\
-
 \def\qrand{q_{rand}}
@@
 The size and location of tunnels determine the ligand size and shape which can enter the active site.
-%The tunnel has to be wide enough and also the physico-chemical properties of the surrounding amino acids have to be compatible with the ligand.
 Currently the biochemists are using various computational methods for tunnel detection (Fig.~\ref{fig::motiv}) in order to estimate the possibility of interactions between a protein and a ligand.
@@
 \includegraphics[width=0.14\textwidth]{fig/render37t}} \\ 
-%\hbox{
-%\vbox{
-%\hbox{\includegraphics[width=0.25\textwidth]{fig/ta-1} }
-%\hbox{\includegraphics[width=0.25\textwidth]{fig/ta-433}}
-%} 
-%}
 \end{tabular}
@@
 
-%or protein folding combined with ligand diffusion~\cite{cortes2010simulating}.
-%rrt for molecular: \cite{al2012motion}
-%survey \cite{gipson2012computational}
-%Sampling-based planners have been used in various applications in robotics~\cite{elbanhawi2014sampling} and also  %,latombe1999motion} and also
-%Solutions for tunnel detection using sampling-based methods however have not been discussed yet.
-%Understanding of interactions between proteins and other small molecules is crucial in many research fields, including drug design and protein engineering. 
-%The knowledge about the existence of the tunnels and their properties (e.g., their bottleneck and chemical properties around the tunnel),
-%is important to understand interactions between proteins and other molecules (ligands)~\cite{Koudelakova2013}.
-%As the interaction takes place in the active site, the existence of a tunnel connecting the active site with the outer environment
-%is necessary for the interaction.
-%To get the ligand to the active site, a transportation tunnel connecting the protein outer environment with the active site has to exists.
-%Biochemists decide if a tunnel can be used to transport a given ligand mainly based on the tunnel length and bottleneck (i.e., the radius of the smallest sphere that forms the tunnel).
-%This is used for example in protein engineering, where the task is to change selected properties of a protein, e.g., its stability under different outer conditions~\cite{Koudelakova2013}.
-% or its activity of the protein towards other molecules~\cite{Pavlova2009}.
-%The design of suitable ligands causing the desired changes can be speeded up by detecting and analyzing the tunnels leading to the active sites.
-%This can be achieved by detecting and studying so called tunnels in proteins which can serve as the transportation paths for the 
-%ligand from the outside environment to the active site or vice versa. 
-%Whereas the tunnel computation is already a well established research field, the simulation of ligand transportation through the detected tunnels is rather new.
-%Ligands are typically of non-spherical shape, therefore it is difficult to estimate their traversability through tunnels computed for a spherical probe.
-%The decision based only on spherical tunnels requires previous expertise in the domain and yet it may be imprecise.
-
-%Proteins are dense structures which limits movements of the ligand insider the tunnels, which leads to the narrow passage problem.
-%To cope with this problem, we propose to utilize the principle of guided sampling~\cite{vonasek2009rrt,denny2016dynamic}.
-%It helps to keep also potential solutions which do not fit to the geometric restrains but can be still feasible because of the physico-chemical properties.
-%This strategy helps to overcome the limitations of the conformational discretation introduced by the rotamer approximation.
-
-%The proposed method can be seen as an extension for the traditionally used tunnel detection tools.
-%Besides trajectory computation, visualization of the results in important. %we also propose methods for their visualization.
-%Our motivation is to help the biochemists to perform so called virtual screening, where they test the traversability of a ligand through a given tunnel.
-%In order to predict the success of ligand traversability, the virtual screening performs hundreds of thousands of tests and checks if the ligand passed through the tunnel.
-%The advantage of the proposed approach is that we can search the ligand path inside a specific tunnel which substantially decreases the computational time and resources required for the virtual screening.
-%We further propose visualization techniques for presenting the results to biochemists.
-%Therefore, methods for visualization of the results  are discussed.
-
 \section{Related Work}
-
-%The analysis of protein structure aiming to reveal the tunnels has been supported by different computational software tools which take the geometry of the protein as an input and explore the inner void space (e.g., CAVER 1.0~\cite{petrek2006caver} or MOLE~\cite{Petrek20071357}).
-%Early methods for tunnel detection utilized a discretized 3D grid, where each cell is considered as occupied or free depending
-%on the presence of atoms of the protein~\cite{petrek2006caver,Petrek20071357}.
-
-%TODO some words about MD and why it better to analyze stuff with tunnels than with MD
 
@@
 
-%This can be formulated as a motion planning problem in a high-dimensional configuration space $\C$.
-%In this case, the configuration space $\C$ has at least $6+n$ dimensions (3D rotation + 3D translation + additional degrees of freedom
-%caused by the ligand flexibility).
-
-%Tunnels can be then searched using standard graph-search methods, like Dijkstra's algorithm.
-%Besides, the grid can be used to identify other relevant properties like 
-%pockets, cavities, or channels~\cite{sehnal2013mole,petrek2006caver}.
-%One of the first grid-based approaches to the detection of tunnels in protein is the CAVER 1.0 algorithm by Pet\v{r}ek et al.~\cite{citeulike:6257975}.
-%The obvious disadvantage of the grid-based methods is the high memory demand and their dependency on the grid resolution.
-%Due to the high memory consumption, these methods are not suitable for tunnel detection in dynamic proteins and therefore they
-%are used primarily for analysis of static molecules or individual snapshots of molecular dynamics.
-%Currently the most widely used approach to tunnel detection is based on ordinary Voronoi diagrams (VD) or Weighted Voronoi Diagrams (WVD). %~\cite{caver3,yaffe2008}.
-%The ordinary VD is computed on points representing centers of all atoms, without considering the radii of atoms.
-%This may lead to detection of tunnels with incorrect bottlenecks. % i.e., incorrect radius of a smallest probe that can traverse the tunnels.
-%To consider atoms with different radii, the weights of individual points are determined by the van der Walls radii of atoms in WVD.
-%An alternative solution is to compute a non-weighted VD on an extended point set, where 
-%each atom is approximated by several spheres with a small radius~\cite{yaffe2008,caver3}.
-%VD-based methods are memory less demanding and also faster than the grid-based methods.
-%The extension of VD-based methods to dynamic molecules requires to construct VD in the frames being analyzed and 
-%finding correspondences between them.
-%The existing approaches often use hierarchical clustering to match Voronoi vertices and edges from different frames, which is computationally demanding~\cite{lindow2012dynamic,caverDetails}.
-
-%The protein tunnels computed for spherical probes contain valuable information about the protein void space.
-%Based on the biochemical properties, the chemists can decide if a given tunnel is relevant and can be possibly used by a ligand to exit the protein.
-%The biochemically relevant tunnels can be then analyzed separately.
-
-%The configuration space $\C$ is formed by all possible configurations of the ligand in the tunnel, i.e., considering its rotation, translation, and possibly also other degrees of freedom responsible for the conformation changes.
-%The dimension of the configuration space is given by the degrees of freedom (DOF) of the ligand, i.e., 6D for a rigid ligand and 6D + $n$ for a flexible
-%ligand with $n$ DOFs.
-%Sampling-based motion planning methods can be used to search this high-dimensional configuration space~\cite{Lav06}.
-
-%Sampling-based motion planning methods are suitable for computing the trajectories of the ligand as they 
-%can cope with many-DOF robots (objects) of arbitrary shapes.
-%The flexibility of ligands (or even proteins) can be modeled using a multi-link kinematic chain, where torsional angles
-%can change~\cite{songPFpath}.
-%This is useful, e.g., in the protein folding studies~\cite{al2012motion,gipson2012computational,amato2002using,raveh2009rapid} or analysis of loop motions~\cite{cortes2004geometric}.
-%This is useful, e.g., in the protein folding studies~\cite{al2012motion,gipson2012computational,amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro} or analysis of loop motions~\cite{cortes2004geometric}.
-%Sampling-based methods have also been used for 
-% tunnel detection~\cite{vonasek2016application,vonasek2017tunnel} and
-% exit pathway computation~\cite{cortes2010simulating,guieysse2008structure}.
-%none of the existing work focuses on the trajectory generation in tunnels.
-%However, not all conformations are feasible, which has to be checked using potential energy, which is time consuming.
-%The idea of sampling-based motion planning is to randomly sample the configuration space $\C$ and classify the samples as free or non-free using collision detection.
-%The free samples are stored in a roadmap (a graph structure), in which a path can be searched using standard graph-search methods.
-
 Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} is a single-query sampling-based motion planning method that 
 incrementally builds a configuration tree rooted at the initial configuration.
-In each iteration of RRT, a random configuration $\qrand \in \C$ is generated and its nearest node $\qnear \in \T$ in the tree is found.
-A new configuration $\qnew$ is constructed on the line connecting $\qnear$ and $\qrand$ in the distance $\varepsilon$ from $\qnear$.
-If $\qnew$ is collision-free, it is added to the tree.
-The algorithm terminates if the tree approaches the goal configuration close enough.
-
+Sampling-based planners can cope with many-DOF systems and they have been successfully employed also beoynd robotics, e.g. to
+identify binding sites~\cite{bayazit2001ligand} or to determine ligand exit pathway~\cite{cortes2010simulating}.
 %In each iteration of RRT, a random configuration $\qrand \in \C$ is generated and its nearest node $\qnear \in \T$ in the tree is found.
@@
 Consequently, many iterations are needed in order to put enough samples into the passages, which increases the computation time.
-%Sampling-based planners can provide poor performance in the presence of narrow passages, as many iterations
-%are required in order to sample the narrow passages dense enough.
-PRM-based planners can cope with the narrow passage problem by increasing the probability of sampling in given regions.
-These regions can be estimated using workspace information, such as by medial axis~\cite{amatoOBRRT,amato2002using,wilmarthMAPRM}, 
+PRM-based planners can cope with the narrow passage problem by increasing the probability of sampling in difficult regions, e.g.
+by sampling along medial axis~\cite{amatoOBRRT,amato2002using,wilmarthMAPRM}.
+In~\cite{bergWIG}, workspace is represented by a grid in which difficult regions are identifed using watershed labeling.
+by indetifying them in a grid-based represnetation of the workspace~
 or they can be identified, e.g., based on the number of collision-free samples around a given configuration~\cite{overmarsGauss,hsuBridge}.
@@
 RRT-based planners cannot cope with the narrow passage problem simply by increasing the probability of sampling in them, as the
-growth of the configuration tree towards the random samples may be blocked by obstacles~\cite{vonasekphd}.
+growth of the configuration tree towards the random samples may be blocked by obstacles. %~\cite{vonasekphd}.
 To prevent this blocking, DD-RRT~\cite{yershovaDDRRT} limits the selection of nodes for expansion to a small ball. 
@@
 the retractions only on the narrow passages.
-Obstacle-based RRT~\cite{amatoOBRRT} utilizes several expansion procedures based on, e.g., random vectors, obstacle vectors, or medial axis in order to improve the growth in the narrow passages.
+Obstacle-based RRT~\cite{amatoOBRRT} utilizes several expansion procedures based on, e.g., random vectors, obstacle vectors, or medial axis.
 In MARRT~\cite{denny2014marrt}, the newly generated samples are pushed towards the medial axis.
+Difficult regions can be also identified manually by a human operator~\cite{denny2018general}. 
+%The operator can also select a suitable planner for the region.<F12>
 %Therefore, many algorithms uses a heuristic to compute the samples near boundary of the configuration space~\cite{amatoOBPRM,amatoOBRRT}.
@@
 and translation) and it employs the other DOFs (i.e., those that are responsible for conformation changes) if they hinder the growth of the tree.
-
-%TODO NEKAM PRESUNU/SMAZU: 
-%A proper distance metric in the configuration space is necessary in order to determine which samples should be connected together (in the case of PRM), or which node of the tree should be expanded (in the case of RRT).
-%Generally, it is not easy to combine the rotational and transnational components of the configuration space.
-%Moreover, the selection of the metric is also influence by the shape of the objects (robots).
-%The flexible ligands can be modeled as a multi-link kinematic chain~\cite{songPFpath,cortes2005path}, but it significantly
-%increases the number of DOFs and consequently the dimension of the configuration space.
-%Moreover, by considering flexible ligands, their shape differ in each conformation, which makes the selection of a suitable metric difficult.
+In~\cite{singhLIG}, the pathways for a flexible ligand are found using Probabilistic Roadmaps.
+Each configuration in the roadmap is also evaluated using electorstatic energy.
+The high-energy configurations are discared, while low-energy conformations (and their vicinity) are sampled more dense.
+The approach~\cite{singhLIG} can be also used to predict ligand binding sites.
+
+%From reviews: vojta will solve
+%\cite{hsuBridge} uz tam je
+%\cite{bergWIG}
+%\cite{denny2018general}
+%\cite{singhLIG}
+%\cite{bayazit2001ligand} - it's in the text already.
 
@@
 This enables the metric to consider the shape of the ligand and it supports the retraction of the ligand to the narrow passages.
+
+\section{Traversability of Tunnels}
+
+\subsection{Preliminaries}
+
+Proteins and ligands are represented by the hard sphere model, where the radius of each sphere representing an atom is given by its van der Waals radius.
+A protein tunnel is described by a sequence of collision-free spheres 
+$T=( (c_1, r_1),\ldots,(c_n,r_n) )$, where $n$ denotes the number of spheres,
+$c_i \in \R^3$ is their 3D position, and $r_i > 0$ denotes the maximum collision-free radius of a sphere centered at $c_i$. 
+Tunnels serving as an input for our proposed method can be detected using some of the existing computational tools, such as CAVER 3.0~\cite{caver3}.
+
+%Conformations of the ligand may be generated in the absence of the receptor and subsequently docked[13] 
+%or conformations may be generated on-the-fly in the presence of the receptor binding cavity,[14] 
+%or with full rotational flexibility of every dihedral angle using fragment based docking.[15] 
+%Force field energy evaluation are most often used to select energetically reasonable conformations,[16] but 
+%knowledge-based methods have also been used.[17]
+%13 = https://link.springer.com/article/10.1007/BF00123666
+
+%The proteins are dense structures and the tunnels are typically narrow.
+%Depending on the protein, the tunnel bottlenecks can be even less than $1$~\AA, which is too narrow for ligands with more than 2 atoms.
+The flexibility of the ligand is modeled using the set $\L$ of its conformations.
+We assume that the library of all possible conformations in a considered resolution is available (e.g., ~\cite{dunbrack}) and we later show how to select a subset of conformations suitable for the motion planning.
+%The conformations $\L$ are used from a library (e.g.,~\cite{dunbrack}), or they can be prepared considering the potential energy.
+
+To enable the motion of ligands in the narrow tunnels, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
+In our solution we use a discrete set of scales, i.e., $s \in \S=\{\smin, \smin+\sdelta, \ldots, \smax\}$, where 
+$\smin$ is the minimal allowed scale, $\smax=1$ is the maximal allowed scale, and $\sdelta$ is the minimal difference between two scales.
+
+A configuration of the ligand $q=(x,y,z,r_x,r_y,r_z,l,s)$ is described
+by the 3D position $(x,y,z)$ of the reference point of the ligand (its geometric center), the rotation around $x$, $y$, and $z$ axes,
+the index of the conformation $l\in \L$, and the scale $s \in \S$.
+All possible configurations form the configuration space $\C$ and the collision-free configurations
+form the subspace $\CF \subseteq \C$.
+A configuration $q$ is collision-free if none of the ligand atoms scaled by $s$ and placed at the
+position defined by $q$ collides with the protein atoms.
+
+\subsection{Computing Initial Configurations}
+
+Depending on the chemical interactions between the ligand and the protein surface, the ligand may enter a tunnel with various rotations and in various spatial conformations.
+The traversability of the tunnel should therefore be evaluated using multiple starting configurations $\QI$.
+The initial configurations are searched around the beginning of the tunnel.
+To find a new initial configuration, a random sample $q$ is generated around the first sphere of the tunnel 
+$c_1$ in the distance $\RI$ (the translation and rotation parts of $q$ are generated randomly, the scale is set to $\smin$, and the conformation index is set randomly).
+If the sample $q$ is collision-free, it can be considered as a new starting configuration and it is added to $\QI$.
+%Similarly, a single goal configuration $\qgoal$ is found around the end of the tunnel. 
+For each starting configuration $\qinit \in \QI$, the trajectories are computed using a modified RRT, which is introduced in the following section.
+
+\subsection{RRT for Computing Ligand's Trajectory along Tunnel}
+
+The proposed motion planning is based on the principle of the RRT~\cite{lavalleRRT} planner.
+In each iteration, a random sample $\qrand$ is generated and its nearest node $\qnear\in\T$ in the tree is found.
+The nearest-neighbor search between $\qrand$ and the tree is performed using the weighted 6D Euclidean metric computed using
+both 3D rotation and 3D translation.
+The main loop of the planner (Alg.~\ref{alg::main}) is similar to the original RRT, but with the following three proposed modifications.
+
+The aim of the first modification is to sample the configuration space along the tunnel being analyzed.
+To guide the growth of the tree using the tunnel, a moving virtual goal is used. %~\cite{vonasek2009rrt}.
+The virtual goal $v, 1\le v \le n$, is the index of a sphere of the tunnel.
+At the beginning of the algorithm, $v=1$.
+The random samples $\qrand$ are generated around the sphere $c_v \in T$ with the probability $\gb$, and from the whole $\C$ otherwise.
+After the tree reaches the sphere $c_v$, i.e., the distance of the tree to $c_v$ is
+less than a predefined threshold $\dt$, the virtual goal is moved to the successor of the last sphere in the tunnel
+that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
+Setting the virtual goal to this successor allows the tree to avoid such parts of the tunnels that are not traversable or reachable by the ligand.
+This allows the tree to slightly detour from the tunnel and overcome its difficult (i.e., narrow) parts by finding an alternative trajectory nearby.
+The algorithm terminates after a predefined number of planning trials $\Imax$ or if the tree reaches
+the last sphere in the tunnel, i.e., when $v = n$.
+
+\linesnumbered
+\begin{algorithm}[bh]
+{\small
+\setstretch{0.88}
+\caption{\label{alg::main}Main loop of the RRT planner}
+\KwIn{
+    tunnel $T=( (c_i, r_i) )$, $i=1,\ldots,n$, with spheres centers $c_i \in \R^3$ and radii $r_i$,
+    initial configuration $\qinit$
+}
+\KwData{
+   ligand conformations $\L$,
+   scale limits $\smin, \smax$ and $\sdelta$,
+   distance $\dt$ to move virtual goal
+}
+\KwOut{
+    configuration tree $\T$\;
+}
+\hrule
+$v = 1$; // index of the virtual goal\\
+$iteration = 0$\;
+$\T$.addNode($\qinit$)\;
+\While{$iteration < \Imax$ {\bf and}  $v < n$}{
+    \eIf{$rand() < \gb$}{
+        $\qrand$ = random sample around virtual goal $c_v\!\in\!T$\;
+    }{
+        $\qrand$ = random sample from $\C$\;
+    }
+    $\qnear$ = nearest node in $\T$ towards $\qrand$\;
+    expand($\T, \qnear,\qrand$)\;
+    \For{$i= n-1,n-2,\ldots,v+1,v$}{ \nllabel{alg::main:a}
+        $d$ = nearest node in the tree towards sphere $c_i$\;
+        \If{$d < \dt $}{
+            $v = i+1$; // new virtual goal found\;
+            {\bf  break}\;
+        }
+    } \nllabel{alg::main:b}
+    $iteration = iteration+1$\;
+}
+\return $\T$\;
+}
+\end{algorithm}
+
+To generate the samples $\qrand$ around the virtual goal $v$, the translation part $(x,y,z)$ of $\qrand$ is generated
+from the Gaussian distribution $N(c_v,\Sigma)$, where $\Sigma$ is the diagonal matrix with diagonal entries equal to the parameter $\rv$, and the rotational
+part of $\qrand$ is generated using techniques described in~\cite{kuffnerES}.
+The other two parameters (ligand index $l$ and scale $s$) of $\qrand$ can be set to zero, as these are not used in the employed
+metric for the nearest-neighbor search.
+The parameter $\rv$ influences the distribution of the random samples around the tunnel centerline. 
+By setting $\rv$ to a small value, the planner attempts to find the trajectories inside the tunnel, while higher values
+of $\rv$ cause the exploration of paths around the tunnel.
+We propose to set this parameter to the average width of the tunnel.
+
+The second proposed modification is the expansion procedure (Alg.~\ref{alg::expand}), which expands the tree from $\qnear$ towards $\qrand$.
+The expansion towards $\qrand$ is crucial for the growth of the tree towards the unexplored areas of the configuration space.
+Traditionally, a line is constructed from $\qnear$ to $\qrand$ and the tree is expanded by a configuration located on this line in the 
+distance $\varepsilon$ from $\qnear$.
+However, in the case of ligands, this would mostly lead to a collision.
+
+The expansion procedure attempts to expand the tree using all the conformations $\L$ and by preferring larger scales than the smaller ones.
+For each $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with a maximal scale.
+First, the maximal scale $\smax \in \S$ is selected and $m$ random samples are generated around $\qnear$ (in the distance $\varepsilon$) 
+and tested for collisions.
+The random samples are generated similarly as in the case of $\qrand$ samples, only their translation
+part $(x,y,z)$ is generated around $\qnear$.
+The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
+If none of the tested samples is collision-free, the scale is reduced to $\smax-\sdelta$ and the search continues
+until a collision-free sample is found or until the minimal reduced-scale $\smin$ is reached.
+By testing first the largest scale ($\smax$), the expansion procedure prefers expanding the tree by larger scales.
+This forces the planner to automatically use smaller scales in the difficult (narrow) parts of the tunnels, while preferring
+large scales if possible.
+
+The third modification is the utilization of another metric during the expansion procedure (line~\ref{alg::expand:a} in Alg.~\ref{alg::expand}).
+This metric is denoted as $\da(\cdot,\cdot)$ and it is referred to as atom-based metric in the rest of the paper.
+To evaluate $\da(q_1,q_2)$, the absolute positions of the atoms of the ligand defined by $q_1$ and $q_2$ are computed.
+The minimal 3D Euclidean distance from the first set of atoms (defined by $q_1$) to the second set of atoms (defined by $q_2$) is the value of 
+$\da(q_1,q_2)$.
+By computing the atom-based metric using the absolute positions of atoms, this metric considers the shape of the ligands, which
+is useful when multiple conformations are utilized.
+This metric ensures that the tree is expanded by such configurations, whose atoms maximally approach
+the atoms of the configuration $\qrand$.
+We observed that this leads to a better construction of the configuration tree inside the proteins and it increases the probability
+of reaching the active site.
+
+The atom-based metric can be evaluated fast using KD-trees with complexity $\mathcal{O}(\log n)$, where $n$ is the number of atoms of the ligand.
+For each conformation $\l \in \L$ of the ligand, a KD-tree is built for the ligand at the configuration $q=(0,0,0,0,0,0,l,\smax)$
+(i.e., for ligand with no translation and no rotation).
+To find the nearest atom to $\qrand$ for a ligand at the configuration $q=(x,y,z,r_x,r_y,r_z,l,s)$, the translation vector $V=(x,y,z)$ and
+rotation matrix $R$ given by $r_x$, $r_y$ and $r_z$ are computed.
+Then, $\qrand$ is translated to $\qrand' = R^{-1}(\qrand-V)$ and the KD-tree search is used to find the nearest point towards $\qrand'$, which
+is also the nearest point of the ligand at $q$ to $\qrand$.
+
+\begin{algorithm}[th]
+{\small
+\setstretch{0.88}
+\caption{\label{alg::expand}expand}
+\KwIn{
+   tree $\T$,
+   configuration $\qnear$ to be expanded,
+   random configuration $\qrand$
+}
+\KwData{
+   ligand conformations $\L$,
+   scale limits $\smin, \smax$, and $\sdelta$,
+}
+\hrule
+\ForEach{$l \in \L$}{
+    \ForEach{$s \in (\smax,\smax-\sdelta, \ldots, \smin+\sdelta, \smin)$}{
+        $\qnew = \emptyset$; // empty configuration \nllabel{alg::expa} \\
+        \For{$i = 1,\ldots,m$}{
+            $q=\qnear$\;
+            $q.position$ = random 3D position around $\qnear$\;
+            $q.rotation$ = random 3D rotation\;
+            $q.l = l$\; 
+            $q.s = s$\;
+            \If{isCollisionFree($q$)}{
+                \If{$\qnew = \emptyset$ {\bf or} $\da(q, \qrand) < \da(\qnew,\qrand)$}{ \nllabel{alg::expand:a}
+                    $\qnew = q$\;
+                }
+            } 
+        }
+        \nllabel{alg::expb}
+        \If{$\qnew \ne \emptyset$} {
+            $\T$.addNode($\qnew$)\;
+            $\T$.addEdge($\qnear,\qnew$)\;
+            {\bf break;} // go to the next conformation
+        }
+    }
+}
+}
+\end{algorithm}
+
+The result of each planning trial is the tree $\T$ of collision-free configurations in which a path
+between $\qinit$ (root of the tree) and $q'$ is found, where
+ $q'$ is the nearest node to the end of the tunnel (measured using the 3D Euclidean metric). 
+%The path $P=(q_i), q_i \in \C$ is represented as a sequence of collision-free configurations.
+%The path is found in the tree even if the tree does not approach $\qgoal$ close enough.
+%Considering also these non-feasible solutions is necessary to evaluate difficult areas of the tunnels, e.g. bottlenecks.
+%The utilization of all computed paths for evaluation of tunnel difficulty is described in the next section.
+
+\subsection{Selection of Conformations for Motion Planning}
+\label{sec::strat}
+
+In the expansion step, each conformation from $\L$ is tested in order to approach $\qrand$.
+By using more conformations, there is a higher chance that at least one of them can expand the tree towards the random sample $\qrand$.
+On the other hand, using a large number of conformations also increases the time complexity of the expansion step.
+Selection of a suitable set of conformations $\L$ is therefore crucial.
+Depending on the ligand, some conformations can be considered as more packed i.e., the atoms in such conformations are closer to each other, but 
+the atoms can also form a more elongated shape (Fig.~\ref{fig::m040c}).
+From the motion planning point of view, both types of conformations are useful.
+The longer conformations can better fit in a narrow passage, while the packed conformations can be better rotated at a given position.
+In the narrow parts of the tunnel, the tree can be extended by the longer conformations, while in the wider regions, also the
+packed conformations can be used.
+\vspace{-3mm}
+
+\begin{figure}[hb]
+\centering
+\includegraphics[width=0.14\textwidth]{fig/m040-conf1} \hskip 25pt
+\includegraphics[width=0.11\textwidth]{fig/m040-conf2}
+\caption{\label{fig::m040c}
+ The examples of a `long' conformation (left) and a `packed' one for the ligand m040 (1,5-Dichloropentane).
+}
+\end{figure}
+
+To select the suitable set $\L$ of conformations from a pool of available conformations, a simple technique is proposed.  
+For each conformation, the geometric center is calculated and the deviation of the distances of the atoms from the center is calculated.
+The low value of the deviation is characteristic for the packed conformations, while larger deviations correspond to the longer conformations.
+To select $n$ conformations from all $m$ available conformations, they are sorted according to the deviation.
+Each $k-$th conformation from the sorted list, where $k=\lfloor{m\over n}\rfloor$, is added to $\L$.
+
+\section{Experimental Verification}
+
+The proposed approach was used to analyze the ligand traversability towards the active site using the main tunnel in the Haloalkane dehalogenase protein (PDB ID 1CQW).
+The active site was defined using residues 38 (Asparagine), 102 (Histidine), and 103 (Aspartic Acid).
+The molecular dynamics of the protein was computed using the AMBER tool~\cite{amber}.
+In each frame, tunnels for the spherical probe 0.9~\AA\ were detected using CAVER 3.0~\cite{caver3}. 
+In most of the frames, two tunnels were detected.
+The average bottleneck sizes of these tunnels are $1.48$~\AA\ and $1.2$~\AA\ for the main and the side tunnel, respectively.
+The tunnel traversability was analyzed in 100 time steps (frames) of MD simulation.
+
+Four different scaling-down factors were used: $\S=\{0.5,0.6,0.7,0.8\}$.
+For each scale, $|\QI|=20$ collision-free initial configurations were found in the $\RI=2$~\AA\ radius around the first sphere of the tunnel (outside the protein).
+For each initial configuration, 10 trajectories were computed, which resulted
+in $100 \times 4 \times 20 \times 10 = 80,000$ trajectories per ligand.
+
+Two motion planners were utilized: the proposed planner (described in Alg.~\ref{alg::main} and Alg.~\ref{alg::expand}), denoted as \RA\ in the rest of the paper, and the Retraction-based RRT~\cite{zhangRetraction}, denoted as \RB\ in the rest of the paper.
+The \RA\ method was run with the following parameters:
+$\Imax=5,000$, $m=100$, $\gb=0.9$, $\dt=1.5$~\AA, $\rv=2$~\AA\ and $\varepsilon=0.05$~\AA.
+Both methods were implemented in the same software framework and they utilized the same data structures and routines for the collision detection and nearest-neighbor search.
+The nearest-neighbor search of both planners was based on the 6D Euclidean metric with the same weights for rotation and translation parts.
+This metric was selected based on preliminary experiments with both planners as it resulted in the highest success rates.
+
+The tested methods differ in the expansion step (lines \ref{alg::expa}--\ref{alg::expb} in Alg.~\ref{alg::expand}).
+In these lines, the \RB\ method employs the retraction step in order to find a new collision-free configuration $\qnew$ that approaches $\qrand$.
+The retraction step is realized by sampling the vicinity of the configuration $\qnear$ in the distance $0.05$~\AA\ using 20 samples.
+A collision-free sample that maximally approaches $\qrand$ is selected and added to the tree, and the retraction continues from this sample.
+The retraction step is repeated 5 times.
+The \RB\ method can therefore expand the tree by at most 5 new nodes per configuration, while \RA\ extends the tree only by one configuration.
+The larger number of nodes in \RB\ method is not problematic, as the KD-trees were used for the nearest-neighbor search and the time to solve these queries is significantly smaller than the computing collision detection.
+The parameters of the methods were selected in such a way that the number of collision detection queries in each expansion step is the same.
+For each conformation and tested scale, the \RA\ method attempts to find $m=100$ collision-free configurations, and \RB\ attempts to perform $5 \times 20$ collision-detection queries.
+Other parameters of the \RB\ planner were the same as for \RA.
+The planning was terminated after $\Imax$ iterations or if the tree approached the active site to the distance less than 0.5~\AA.
+
+In the following experiments, we measured the performance of the methods using the traversability rate, which is the percentage of frames of the protein dynamics that are traversable by the tested ligand.
+A frame is considered traversable if at least one trajectory out of 200 ($\!|\QI|\!\times\!\!10$ trials) approaches the active site to the distance less than~2~\AA.~\footnote{We refer to~\url{http://mrs.felk.cvut.cz/icra2018protein} for additional material.}
+
+\subsection{Traversability of One Atom}
+
+The aim of the first experiment is to measure the performance of both planners in the simplest scenario, i.e., finding a path for one sphere heading towards the active site.
+The radius of the sphere is set to $1.7$~\AA.
+
+The computed traversability rates are shown in Tab.~\ref{tab::one}.
+The maximum percentage of traversable frames (the row Maximum in Tab.~\ref{tab::one}) is computed as the percentage of tunnels with the bottleneck larger than the tested radius.
+The row 'Planning' shows the traversable rate computed by the planners.
+This rate decreases with the increasing radius and it is lower than the maximum traversable rate.
+The rates provided by \RA\ and \RB\ are similar which is a good indicator that the methods have been implemented correctly.
+Higher traversability rate can be achieved by increasing the maximum number of planning iterations $\Imax$, which would also
+increase the planning time.
+For the purpose of the following experiments, we used $\Imax=5,000$ as a compromise between the performance and the runtime.
+
+\begin{table}[ht]
+\centering
+\caption{\label{tab::one}Traversability for one atom.}
+\renewcommand{\tabcolsep}{4.3pt}
+{\small
+\renewcommand{\arraystretch}{0.7}
+\input graphProbe.tex
+}
+\end{table}
+
+%\begin{figure}
+%\centering
+%\begin{tabular}{cc}
+%\includegraphics[width=0.3\textwidth]{fig/t05proteintunnels} &
+%\includegraphics[width=0.45\textwidth]{fig/ligAll}  \\
+%(a) & (b)                        
+%\end{tabular}                       
+%\caption{\label{fig::tunnel}
+%    (a) The protein 1CQW visualized using the cartoon representation (gray) with 
+%        three tunnels (red, green, blue) detected by CAVER 3.0.
+%        The first tunnel is depicted red.
+%%        its highest ranked tunnel (green) which was detected using CAVER 3.0.
+%    (b) Examples of three conformations of $\LA$ (top) and $\LB$ (bottom)
+%}
+%\end{figure}
+%
+
+\subsection{Influence of the Conformation Selection}
+
+The aim of this experiment was to verify the procedure to select feasible ligand conformations for the motion planning task.
+The experiment was performed with the ligand DCP (Dichloropropanol), for which $1300$ conformations were available, and for the ligand
+m040 (1,5-Dichloropentane) with $7560$ available conformations.
+Two strategies were tested: the strategy described in Sec.~\ref{sec::strat} (denoted `S') and the strategy that prefers smaller conformations (denoted `P').
+The `P' method selects such conformations where the deviations of the atomic distance from the geometric center are the lowest, which leads 
+to the selection of packed conformations (Fig.~\ref{fig::m040c}).
+
+The traversability rates are shown in Tab.~\ref{tab::selection}, where the subscript of each ligand's 
+name shows the number of the selected conformations (5 or 50) and the used method (S or P).
+The number after the slash in the ligand's name denotes the number of atoms of the ligand.
+In the case of both tested ligands and 50 selected conformations, the `S' strategy leads to a higher traversability rate than the `P' strategy.
+For the DCP ligand, a higher traversability rate was achieved when 50 conformations were selected.
+This experiment shows that selecting 50 conformations using the `S' strategy (Sec.~\ref{sec::strat}) leads to better results
+than using a less number of conformations or when preferring the packed conformations.
+
+\begin{table}[ht]
+\centering
+\caption{\label{tab::selection}
+    The influence of the conformation selection method to the traversability rates in the first tunnel.
+    The number after~`$/$' denotes the number of atoms.
+}
+\small
+\renewcommand{\tabcolsep}{4.3pt}
+{\small
+\renewcommand{\arraystretch}{0.7}
+\input graphSel.tex
+}
+\end{table}
+
+\subsection{Traversability of Ligands}
+
+Based on the results of the previous experiment, 50 conformations using the `S' strategy were selected for ligands m037t (1,2-Dichloroethane), m038t (1,3-Dichloropropane), m056 (2-Chlorobutane), and m080 (Trichloropropane), and
+the task was to compute the trajectories for these ligands in a given molecular dynamics simulation.
+
+The results are summarized in Tab.~\ref{tab::main}.
+The table shows the traversability rate only for the main tunnel, as no trajectory was found in the side tunnel for most of the tested ligands.
+
+\begin{table}[bt]
+\centering
+\caption{\label{tab::main}
+    Traversability of the tunnels for ligands with 50 conformations. 
+    The number after '$/$' denotes the number of atoms.
+}
+\small
+\renewcommand{\tabcolsep}{4.3pt}
+{\small
+\renewcommand{\arraystretch}{0.7}
+\input graph.tex
+}
+\end{table}
+
+The only exception is the ligand m037t, which was the smallest tested one (with 8 atoms), and therefore it could fit even inside the side tunnel.
+The traversability rate is higher for the most scaled-down ligands ($\smin=0.5$) and it decreases with the increasing scale.
+For the scale $\smin=0.8$, the most traversable frames were detected for ligands DCP (8~\%), m037t (20~\%) and m038t (16~\%).
+
+In conclusion, the traversability of the DCP ligand to active site was approved by the MD simulation.
+The observed trajectory of the DCP ligand in the MD simulation followed the main tunnel, as expected.
+Based on the traversability rates with $\smin=0.8$, we expect that at least ligands m037t and m038t can also reach the active site, as
+they have the same (or even larger) traversability rate than the DCP.
+However, this hypothesis has not been approved yet by testing it on MD simulations.
+
+The proposed \RA\ planner provides higher traversability rate than \RB.
+The distances of the configuration trees (measured using the 3D Euclidean metric) to the active site and the runtimes in each frame
+are shown in Fig.~\ref{fig::comparison}.
+The runtimes of both algorithms increase in those frames, where the tunnel is not traversable, i.e., where the distance
+to the active site is more than 2~\AA\ (for example, in the frame no. 60).
+
+{\def\a{270}
+\begin{figure}[hb]
+\centering
+%\includegraphics[width=0.17\textwidth,angle=\a]{fig/m037tmins-03tunnel-1rrt_path_time-2}
+%\includegraphics[width=0.17\textwidth,angle=\a]{fig/m037tmins-03tunnel-1trajDistToGoal-2}
+\includegraphics[width=0.18\textwidth,angle=\a]{fig/crop1}
+\includegraphics[width=0.18\textwidth,angle=\a]{fig/crop2}
+\caption{\label{fig::comparison}
+    Runtimes and the distance to the active site for the m037 ligand at scale 0.6.
+    The values are computed from all 200 trials (20 initial configurations $\times$ 10 trials).
+    Only each third frame is shown in the graph.
+}
+\end{figure}
+}
+
+\vspace{-3mm}
+The most time-consuming part of the planners is the collision detection. 
+Although both methods can utilize the same number of collision detection queries in the expansion step, \RB\ is slower than \RA.
+The reason is that \RA\ expands the tree more closer to the random samples $\qrand$ than \RB.
+This results in a faster exploration of the configuration space and therefore, \RA\ needs less number of iterations to reach the active site.
+The better performance of \RA\ is caused by the proposed atom-based metric, because this distance depends on the actual shape of the ligand and 
+it allows the ligands to retract more towards $\qrand$.
+On the contrary, \RB\ utilizes the single 6D Euclidean weighted metric in the expansion, where the weights are the same for all conformations.
+Depending on these weights, the metric favors samples that differ from $\qrand$ more in the rotation or translation part.
+However, the weights between the rotation/translation parts should depend on the shape of the ligand, which is different in each conformation.
+By using a single set of weights in the 6D Euclidean metric, we implicitly favor either the packed or the longer conformations.
+
+\section{Discussion and Utilization of the Results}
+
+The presented work was motivated by the current needs of biochemists who need to analyze the traversability of selected ligands through protein tunnels.
+To find the feasible ligand trajectories, it is often required to shrink the atomic radii of the ligand, as utilized also in other existing tools.
+Finding trajectories for the scaled-down ligands enables to solve the problem with narrow passages, but it can be questionable how biochemically relevant such trajectories~are.
+
+Judging the accessibility of the active site by a ligand based only on the traversability rates may be problematic if the ligand is scaled too much.
+Instead of using a single version of the scaled-down ligand, we only limit the minimal scale used by the planner (the parameter $\smin$).
+The expansion step always prefers to expand the tree by a collision-free configuration with the maximal scale.
+%that leads to a new collision-free configuration.
+The information about the scales used at the trajectory points (which can be extracted from the configuration tree of the proposed planner) brings the biochemists new knowledge about the tunnels, which was impossible to get using the previous solutions.
+Our proposed visualization allows the biochemists to easily observe those parts of the tunnel where the ligand had to be scaled-down significantly (red parts of trajectories in Fig.~\ref{fig::dcp}), and parts where the ligand could be enlarged again (green parts in Fig.~\ref{fig::dcp}).
+According to the results of the motion planing, the DCP needs to shrink down to ca. 75~\% of its size (Fig.~\ref{fig::dcp}, point B).
+However, as both the protein and ligand are mutually influence each other, it is possible that even a very narrow part of the tunnel is traversable.
+This was observed in the MD simulation of the DCP ligand in the 1CQW protein, which revealed that DCP can pass the main tunnel.
+
+\begin{figure}
+\centering
+\includegraphics[width=0.4\textwidth]{fig/dcp-image1Label}\\
+\includegraphics[width=0.2\textwidth]{fig/colormap}
+\caption{\label{fig::dcp}
+Visualization of trajectories for the ligand DCP in the two examined tunnels.  
+The color of the trajectories corresponds to the scale, that was needed to pass the given part of the tunnel.
+The ligand can enter the first tunnel with a high scale (almost 1=100~\%) (A), but then it has to be scaled-down (to $\sim75$~\%) to pass the
+next part of the tunnel (B). After this narrow passage, the tunnel is larger again and the ligand can traverse to the narrow passage with
+scale $>75$~\% (C).
+The second tunnel can also be entered with the high scale (green), but then the narrow part is traversable only with a low scale ($\sim50$~\%) (D). 
+For the purpose of the visualization, even more scale-down to 0.3=30~\% was allowed (E) to see how the ligand passed the tunnel.
+However, the ligand can reach the active site only with very low scale ($\sim0.4=40$~\%) using the side tunnel.
+}
+\end{figure}
+
+\section{Conclusion}
+
+In this paper, we presented a novel modification of the RRT planner for computation of ligand trajectory inside protein tunnels.
+%So far, tunnels have been mostly detected using Voronoi diagrams which do not take into account the shape and flexibility of the ligand.
+In our solution, we compute the trajectories using the modified RRT which takes into account the ligand shape and we model the ligand flexibility using a predefined set of its conformations.
+To simulate the real behavior of ligand passing through a tunnel, we enable scaling down the atomic radii of the ligands up to a certain predefined factor.
+Furthermore, our planner is localized as it generates random samples only along a tunnel computed by Voronoi diagram, and it attempts to expand the configuration
+tree using all ligand conformations. 
+The expansion prefers atoms with a higher scale, which ensures that the ligand shrinks down only if necessary, i.e., when traversing the narrow parts of the tunnel.
+To further boost the exploration of the configuration space, a novel metric suitable for measuring the distances between various ligand shapes is used in the expansion step.
+The proposed method has been used to test and evaluate the trajectories of several ligands in a sequence of frames capturing protein dynamics.
+
+\bibliographystyle{plain}
+\bibliography{paper}
+
+\end{document}
+
+% ====================================================
+% ====================================================
+% ====================================================
+% ====================================================
+% ====================================================
+% ====================================================
+
+% ============  unused ideas =========================
+
+%neni treba \cite{cheng01reducing}
+%Using a library of predefined conformations has been used also in other tasks, e.g.~\cite{kellogg}.
+%This shrinking technique is also used in the MoMa-LigPath tool~\cite{cortes2005path}, which utilizes RRT to find pathways for unbinding a ligand  from a protein.
+%It is necessary to use an apporpriated distance metric in the configuration space.
+%Generally, it is hard to combine the translational and rotational components in the configuration space.
+%In~\cite{zhangRetraction}, the DISP metric~\cite{zhang2008fast}.
+%energetic constraints are translated into geometric ones by considering a steric model of the molecules and applying collision detection
+%\cite{deAngulo2005biocd}
+%First the biredge test is applied to identify regions aeround narrow passages, optimization-based retraction operatoin sleectively only at those regions selective rrt~\cite{lee2012srrrt}
+%Free-space dilation is an effective aproach for narrow passage sampling~\cite{hsu06multilevel}. 
+%It is necessary to define proper method for dilating and determin the amount of dilation.
+%\cite{saha2005finding}
+%DISP metric \cite{zhang2008fast}
+
+\subsection{Traversability Characteristics}
+
+For each initial configuration $\qinit \in \QI$, $K$ trajectories are computed, which results in the set of $K |\QI|$ trajectories.
+All these trajectories are used to compute the following properties of the tunnel $T$.
+A trajectory $P$ reaches the tunnel sphere $c_i \in T$, if the 3D Euclidean distance of the 
+nearest configuration $q \in P$ towards $c_i$ is less than $r_i$ (radius of the sphere $c_i$).
+Let $N(i)$ denote the number of trajectories that reached $i-$th sphere of the tunnel, $i=1,\ldots,n$.
+Three basic characteristics of the tunnel are computed from the trajectories: accessibility, throughput, and the scaling factor.
+
+The accessibility $A(i)=N(i)/N$ of the sphere $i$ is the probability of reaching the sphere $i$, where $N=K|\QI|$ is the total number of trajectories.
+The accessibility shows how probable it is to pass the tunnel up to the sphere $i$.
+Obviously, the most important is $A(n)$ of the last sphere of the tunnel, which can be considered as the overall difficulty of the tunnel.
+The  ligand passage may however be strongly affected by the first bottleneck, so the parts of the tunnel located behind
+the bottleneck have low accessibility.
+
+The throughput $T(i)$ is the ratio of trajectories that passed sphere $i$ (i.e., visited sphere $i+1$) and reached the sphere $i$,
+i.e., $T(i) = N(i+1) / N(i)$. 
+The throughput is not computed for the last sphere ($i=N$).
+The throughput shows a local accessibility of tunnel parts and it can be used to detect places where most of the trajectories ends.
+
+The proposed planner is allowed to scale down the radii of ligand spheres up to the allowed scale $\smin$. 
+It can be expected that narrower parts of the tunnel are more often passed with a more scaled-down ligand than
+the wider parts.
+Ligand scale $L(i)$ at the tunnel sphere $i$ is the average scale of ligand that reaches the sphere $i$.
+
+\begin{figure}
+\centering
+{
+\renewcommand{\tabcolsep}{-5pt}
+\begin{tabular}{ccc}
+\includegraphics[width=0.38\textwidth]{fig/t05proteinBottle2T} & 
+\includegraphics[width=0.38\textwidth]{fig/t04goodT} & 
+\includegraphics[width=0.38\textwidth]{fig/t05goodT} \\
+\includegraphics[width=0.38\textwidth]{fig/t05thpT} &
+\includegraphics[width=0.38\textwidth]{fig/t04badT} &
+\includegraphics[width=0.38\textwidth]{fig/t05badT} \\ 
+(a) & (b) & (c)                       
+\end{tabular}                       
+\caption{\label{fig::tunnel2}
+    (a) Classic bottleneck for spherical probe (top) and visualization of throughput (bottom) for the ligand $\LA$.
+    (b) Successful (green, top) trajectories that reached end points of the tunnels,
+        and unsuccessful ones (red, bottom) for $\smin=0.4$.
+    (c) Successful and unsuccessful trajectories for $\smin=0.5$.
+}
+}
+\end{figure}
+
+\begin{figure}
+\centering
+{
+\renewcommand{\tabcolsep}{0pt}
+\begin{tabular}{ccc}
+\includegraphics[width=0.325\textwidth]{fig/tunne4ac}& 
+\includegraphics[width=0.325\textwidth]{fig/tunne4bc} & 
+\includegraphics[width=0.325\textwidth]{fig/tunne4cc} \\
+(a) & (b) & (c)                        
+\end{tabular}
+}
+\caption{\label{fig::detail}
+    Detailed view to alternative pathway around the third tunnel in 1CQW.
+    The successful trajectories are in green (a) the unsuccessful in red (b).
+    (c) Shows visualization of protein atoms around the tunnel.
+}
+\end{figure}
+
+Besides the color mapping, it is also necessary to show the computed trajectories.
+%This is useful, for example, to find out in which part of the tunnel the trajectories detour.
+Simple visualization of all trajectories could however be too slow for an interactive work.
+Therefore, the trajectories are first clustered and then only the clusters are visualized.
+Due to different lengths of the trajectories, they are first converted to a normalized form.
+Let $\cstart$ represent the average starting position of all trajectories and let $d_{max}$ be the 3D Euclidean distance
+of the most distant configuration from $\cstart$ among all trajectories.
+A set of $M$ spheres centered at $\cstart$ are created with radii $r'_i=i {d_{max} \over M}$, where $i=0,\ldots, M-1$.
+The trajectory $P=(q_1,\ldots,q_n)$ of length $n$ is represented by the normalized vector $v=(x_1,\ldots,x_M)$ of length $M$,
+where $x_i$ is the 3D position of the nearest configuration $q \in P$ to the surface of the $i$-th sphere with the radius $r'_i$.
+The distance between two normalized trajectories $v_i$ and $v_j$ is defined as
+$d(v_i, v_j) = \frac{1}{M} \sum_{1 \leq k \leq N} |x_k^i x_k^j|$.
+This distance is used in the UPGMA clustering technique~\cite{sokal1958statistical}.
+The trajectories can be visualized using a representative of each cluster.
+The number of trajectories in each cluster is represented by the width of the polyline (Fig.~\ref{fig:trajectories}).
+%In this manner, we are able to convey the information about all different trajectories together (see Fig.~\ref{fig:trajectories} right).
+
+\begin{figure}
+\centering
+\begin{tabular}{cc}
+\includegraphics[width=0.4\textwidth]{fig/trajectories-all} &
+\includegraphics[width=0.4\textwidth]{fig/trajectories-clustered-21} \\
+(a) & (b) \\                       
+\end{tabular}
+\caption{Visualization of the trajectories. The tunnel begins in the top left corner.
+(a) All trajectories ($\sim5000$) colored according to whether they reached the end of the tunnel (green) or not (red). 
+(b) Visualization using clusters of trajectories. 
+%Trajectories clustered according to their proximity colored by distinct colors.
+%The different line widths represent the number of trajectories in a cluster.
+\label{fig:trajectories}
+}
+\end{figure}
+
+\begin{table}
+\centering
+\caption{\label{tab::rrt}
+    Runtime and success ratio for ligand $\LA$ (left) and $\LB$ (right).
+}
+{
+\small
+\renewcommand{\tabcolsep}{1pt}
+\input table.003.tex
+\hskip 4pt   
+\input table.004.tex
+}
+\end{table}
+
+Trajectories for $\LA$ in all detected tunnels were classified as successful if they reached the last sphere of the tunnel
+to distance $\dt=2$~\AA and they were considered unsuccessful otherwise.
+The throughput computed from the trajectories shows that the tunnels are difficult not only around bottlenecks, but
+also in other places.
+The comparison of the classic bottleneck (i.e., measured by the radius of spherical probe) and throughput is depicted in Fig.~\ref{fig::tunnel2}a.
+The trajectories for $\smin=0.4$ are depicted in Fig.~\ref{fig::tunnel2}b and for $\smin=0.5$ in Fig.~\ref{fig::tunnel2}c.
+The successful trajectories for $\smin=0.4$ reveal that the end of the tunnel No. 3 can be approached by two different pathways (one in the tunnel and another one outside the tunnel).
+The detail is depicted in Fig.~\ref{fig::detail}.
+Despite the low bottleneck of this tunnel, the ligand may reach its end using the alternative pathway.
 
@@
 
-\section{Traversability of Tunnels}
-
-\subsection{Preliminaries}
-
-Proteins and ligands are represented by the hard sphere model, where the radius of each sphere representing an atom is given by its van der Waals radius.
-A protein tunnel is described by a sequence of collision-free spheres 
-$T=( (c_1, r_1),\ldots,(c_n,r_n) )$, where $n$ denotes the number of spheres,
-$c_i \in \R^3$ is their 3D position, and $r_i > 0$ denotes the maximum collision-free radius of a sphere centered at $c_i$. 
-Tunnels serving as an input for our proposed method can be detected using some of the existing computational tools, such as CAVER 3.0~\cite{caver3}.
-
-%from wiki:
 %Multiple static structures experimentally determined for the same protein in different conformations are often used to emulate receptor flexibility.[19] Alternatively rotamer libraries of amino acid side chains that surround the binding cavity may be searched to generate alternate but energetically reasonable protein conformations.[20][21]
@@
 % 21 = \cite{taylor2003fds}
-
-%Conformations of the ligand may be generated in the absence of the receptor and subsequently docked[13] 
-%or conformations may be generated on-the-fly in the presence of the receptor binding cavity,[14] 
-%or with full rotational flexibility of every dihedral angle using fragment based docking.[15] 
-%Force field energy evaluation are most often used to select energetically reasonable conformations,[16] but 
-%knowledge-based methods have also been used.[17]
-%13 = https://link.springer.com/article/10.1007/BF00123666
-
-%The proteins are dense structures and the tunnels are typically narrow.
-%Depending on the protein, the tunnel bottlenecks can be even less than $1$~\AA, which is too narrow for ligands with more than 2 atoms.
-The flexibility of the ligand is modeled using the set $\L$ of its conformations.
-We assume that the library of all possible conformations in a considered resolution is available (e.g., ~\cite{dunbrack}) and we later show how to select a subset of conformations suitable for the motion planning.
-%The conformations $\L$ are used from a library (e.g.,~\cite{dunbrack}), or they can be prepared considering the potential energy.
-
-To enable the motion of ligands in the narrow tunnels, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
-In our solution we use a discrete set of scales, i.e., $s \in \S=\{\smin, \smin+\sdelta, \ldots, \smax\}$, where 
-$\smin$ is the minimal allowed scale, $\smax=1$ is the maximal allowed scale, and $\sdelta$ is the minimal difference between two scales.
-
-A configuration of the ligand $q=(x,y,z,r_x,r_y,r_z,l,s)$ is described
-by the 3D position $(x,y,z)$ of the reference point of the ligand (its geometric center), the rotation around $x$, $y$, and $z$ axes,
-the index of the conformation $l\in \L$, and the scale $s \in \S$.
-All possible configurations form the configuration space $\C$ and the collision-free configurations
-form the subspace $\CF \subseteq \C$.
-A configuration $q$ is collision-free if none of the ligand atoms scaled by $s$ and placed at the
-position defined by $q$ collides with the protein atoms.
-%The set of all feasible configurations at scale $s$ is denoted $\CFD \subseteq \C$.
-
-\subsection{Computing Initial Configurations}
-
-Depending on the chemical interactions between the ligand and the protein surface, the ligand may enter a tunnel with various rotations and in various spatial conformations.
-The traversability of the tunnel should therefore be evaluated using multiple starting configurations $\QI$.
-The initial configurations are searched around the beginning of the tunnel.
-To find a new initial configuration, a random sample $q$ is generated around the first sphere of the tunnel 
-$c_1$ in the distance $\RI$ (the translation and rotation parts of $q$ are generated randomly, the scale is set to $\smin$, and the conformation index is set randomly).
-If the sample $q$ is collision-free, it can be considered as a new starting configuration and it is added to $\QI$.
-%Similarly, a single goal configuration $\qgoal$ is found around the end of the tunnel. 
-For each starting configuration $\qinit \in \QI$, the trajectories are computed using a modified RRT, which is introduced in the following section.
-
-\subsection{RRT for Computing Ligand's Trajectory along Tunnel}
-
-The proposed motion planning is based on the principle of the RRT~\cite{lavalleRRT} planner.
-In each iteration, a random sample $\qrand$ is generated and its nearest node $\qnear\in\T$ in the tree is found.
-The nearest-neighbor search between $\qrand$ and the tree is performed using the weighted 6D Euclidean metric computed using
-both 3D rotation and 3D translation.
-The main loop of the planner (Alg.~\ref{alg::main}) is similar to the original RRT, but with the following three proposed modifications.
-
-The aim of the first modification is to sample the configuration space along the tunnel being analyzed.
-To guide the growth of the tree using the tunnel, a moving virtual goal is used. %~\cite{vonasek2009rrt}.
-The virtual goal $v, 1\le v \le n$, is the index of a sphere of the tunnel.
-At the beginning of the algorithm, $v=1$.
-The random samples $\qrand$ are generated around the sphere $c_v \in T$ with the probability $\gb$, and from the whole $\C$ otherwise.
-After the tree reaches the sphere $c_v$, i.e., the distance of the tree to $c_v$ is
-less than a predefined threshold $\dt$, the virtual goal is moved to the successor of the last sphere in the tunnel
-that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
-Setting the virtual goal to this successor allows the tree to avoid such parts of the tunnels that are not traversable or reachable by the ligand.
-This allows the tree to slightly detour from the tunnel and overcome its difficult (i.e., narrow) parts by finding an alternative trajectory nearby.
-The algorithm terminates after a predefined number of planning trials $\Imax$ or if the tree reaches
-the last sphere in the tunnel, i.e., when $v = n$.
-
-\linesnumbered
-\begin{algorithm}[bh]
-{\small
-\setstretch{0.88}
-\caption{\label{alg::main}Main loop of the RRT planner}
-\KwIn{
-    tunnel $T=( (c_i, r_i) )$, $i=1,\ldots,n$, with spheres centers $c_i \in \R^3$ and radii $r_i$,
-    initial configuration $\qinit$
-}
-\KwData{
-   ligand conformations $\L$,
-   scale limits $\smin, \smax$ and $\sdelta$,
-   distance $\dt$ to move virtual goal
-}
-\KwOut{
-    configuration tree $\T$\;
-}
-\hrule
-$v = 1$; // index of the virtual goal\\
-$iteration = 0$\;
-$\T$.addNode($\qinit$)\;
-\While{$iteration < \Imax$ {\bf and}  $v < n$}{
-    \eIf{$rand() < \gb$}{
-        $\qrand$ = random sample around virtual goal $c_v\!\in\!T$\;
-    }{
-        $\qrand$ = random sample from $\C$\;
-    }
-    $\qnear$ = nearest node in $\T$ towards $\qrand$\;
-    expand($\T, \qnear,\qrand$)\;
-    \For{$i= n-1,n-2,\ldots,v+1,v$}{ \nllabel{alg::main:a}
-        $d$ = nearest node in the tree towards sphere $c_i$\;
-        \If{$d < \dt $}{
-            $v = i+1$; // new virtual goal found\;
-            {\bf  break}\;
-        }
-    } \nllabel{alg::main:b}
-    $iteration = iteration+1$\;
-}
-\return $\T$\;
-}
-\end{algorithm}
-
-To generate the samples $\qrand$ around the virtual goal $v$, the translation part $(x,y,z)$ of $\qrand$ is generated
-from the Gaussian distribution $N(c_v,\Sigma)$, where $\Sigma$ is the diagonal matrix with diagonal entries equal to the parameter $\rv$, and the rotational
-part of $\qrand$ is generated using techniques described in~\cite{kuffnerES}.
-The other two parameters (ligand index $l$ and scale $s$) of $\qrand$ can be set to zero, as these are not used in the employed
-metric for the nearest-neighbor search.
-The parameter $\rv$ influences the distribution of the random samples around the tunnel centerline. 
-By setting $\rv$ to a small value, the planner attempts to find the trajectories inside the tunnel, while higher values
-of $\rv$ cause the exploration of paths around the tunnel.
-We propose to set this parameter to the average width of the tunnel.
-
-The second proposed modification is the expansion procedure (Alg.~\ref{alg::expand}), which expands the tree from $\qnear$ towards $\qrand$.
-The expansion towards $\qrand$ is crucial for the growth of the tree towards the unexplored areas of the configuration space.
-Traditionally, a line is constructed from $\qnear$ to $\qrand$ and the tree is expanded by a configuration located on this line in the 
-distance $\varepsilon$ from $\qnear$.
-However, in the case of ligands, this would mostly lead to a collision.
-
-The expansion procedure attempts to expand the tree using all the conformations $\L$ and by preferring larger scales than the smaller ones.
-For each $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with a maximal scale.
-First, the maximal scale $\smax \in \S$ is selected and $m$ random samples are generated around $\qnear$ (in the distance $\varepsilon$) 
-and tested for collisions.
-The random samples are generated similarly as in the case of $\qrand$ samples, only their translation
-part $(x,y,z)$ is generated around $\qnear$.
-The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
-If none of the tested samples is collision-free, the scale is reduced to $\smax-\sdelta$ and the search continues
-until a collision-free sample is found or until the minimal reduced-scale $\smin$ is reached.
-By testing first the largest scale ($\smax$), the expansion procedure prefers expanding the tree by larger scales.
-This forces the planner to automatically use smaller scales in the difficult (narrow) parts of the tunnels, while preferring
-large scales if possible.
-
-The third modification is the utilization of another metric during the expansion procedure (line~\ref{alg::expand:a} in Alg.~\ref{alg::expand}).
-This metric is denoted as $\da(\cdot,\cdot)$ and it is referred to as atom-based metric in the rest of the paper.
-To evaluate $\da(q_1,q_2)$, the absolute positions of the atoms of the ligand defined by $q_1$ and $q_2$ are computed.
-The minimal 3D Euclidean distance from the first set of atoms (defined by $q_1$) to the second set of atoms (defined by $q_2$) is the value of 
-$\da(q_1,q_2)$.
-By computing the atom-based metric using the absolute positions of atoms, this metric considers the shape of the ligands, which
-is useful when multiple conformations are utilized.
-This metric ensures that the tree is expanded by such configurations, whose atoms maximally approach
-the atoms of the configuration $\qrand$.
-We observed that this leads to a better construction of the configuration tree inside the proteins and it increases the probability
-of reaching the active site.
-
-%While the nearest neighbors are searched using 6D Euclidean metric, which can be efficiently realized using KD-trees, the expansion
-%procedure utilizes a metric that is sensitive to the shape of the ligands.
-%This is necessary as the ligands adopt various shapes according to the actual conformations.
-%To minimize the distance between $\qrand$ and the ligand, the shape of the ligand should be considered.
-
-%Therefore, we propose to measure the distance as the smallest 3D Euclidean distance between atoms placed at $\qrand$ and a configuration $q$ being tested.
-The atom-based metric can be evaluated fast using KD-trees with complexity $\mathcal{O}(\log n)$, where $n$ is the number of atoms of the ligand.
-For each conformation $\l \in \L$ of the ligand, a KD-tree is built for the ligand at the configuration $q=(0,0,0,0,0,0,l,\smax)$
-(i.e., for ligand with no translation and no rotation).
-To find the nearest atom to $\qrand$ for a ligand at the configuration $q=(x,y,z,r_x,r_y,r_z,l,s)$, the translation vector $V=(x,y,z)$ and
-rotation matrix $R$ given by $r_x$, $r_y$ and $r_z$ are computed.
-Then, $\qrand$ is translated to $\qrand' = R^{-1}(\qrand-V)$ and the KD-tree search is used to find the nearest point towards $\qrand'$, which
-is also the nearest point of the ligand at $q$ to $\qrand$.
-
-\begin{algorithm}[th]
-{\small
-\setstretch{0.88}
-\caption{\label{alg::expand}expand}
-\KwIn{
-   tree $\T$,
-   configuration $\qnear$ to be expanded,
-   random configuration $\qrand$
-}
-\KwData{
-   ligand conformations $\L$,
-   scale limits $\smin, \smax$, and $\sdelta$,
-}
-%\KwOut{
-%    set $S$ of collision-free configurations reachable from $\qnear$\;
-%}
-\hrule
-\ForEach{$l \in \L$}{
-    \ForEach{$s \in (\smax,\smax-\sdelta, \ldots, \smin+\sdelta, \smin)$}{
-        $\qnew = \emptyset$; // empty configuration \nllabel{alg::expa} \\
-        \For{$i = 1,\ldots,m$}{
-            $q=\qnear$\;
-            $q.position$ = random 3D position around $\qnear$\;
-            $q.rotation$ = random 3D rotation\;
-            $q.l = l$\; 
-            $q.s = s$\;
-            \If{isCollisionFree($q$)}{
-                \If{$\qnew = \emptyset$ {\bf or} $\da(q, \qrand) < \da(\qnew,\qrand)$}{ \nllabel{alg::expand:a}
-                    $\qnew = q$\;
-                }
-            } 
-        }
-        \nllabel{alg::expb}
-        \If{$\qnew \ne \emptyset$} {
-            $\T$.addNode($\qnew$)\;
-            $\T$.addEdge($\qnear,\qnew$)\;
-            {\bf break;} // go to the next conformation
-        }
-    }
-}
-}
-\end{algorithm}
-
-The result of each planning trial is the tree $\T$ of collision-free configurations in which a path
-between $\qinit$ (root of the tree) and $q'$ is found, where
- $q'$ is the nearest node to the end of the tunnel (measured using the 3D Euclidean metric). 
-%The path $P=(q_i), q_i \in \C$ is represented as a sequence of collision-free configurations.
-%The path is found in the tree even if the tree does not approach $\qgoal$ close enough.
-%Considering also these non-feasible solutions is necessary to evaluate difficult areas of the tunnels, e.g. bottlenecks.
-%The utilization of all computed paths for evaluation of tunnel difficulty is described in the next section.
-
-\subsection{Selection of Conformations for Motion Planning}
-\label{sec::strat}
-
-In the expansion step, each conformation from $\L$ is tested in order to approach $\qrand$.
-By using more conformations, there is a higher chance that at least one of them can expand the tree towards the random sample $\qrand$.
-On the other hand, using a large number of conformations also increases the time complexity of the expansion step.
-Selection of a suitable set of conformations $\L$ is therefore crucial.
-Depending on the ligand, some conformations can be considered as more packed i.e., the atoms in such conformations are closer to each other, but 
-the atoms can also form a more elongated shape (Fig.~\ref{fig::m040c}).
-From the motion planning point of view, both types of conformations are useful.
-The longer conformations can better fit in a narrow passage, while the packed conformations can be better rotated at a given position.
-In the narrow parts of the tunnel, the tree can be extended by the longer conformations, while in the wider regions, also the
-packed conformations can be used.
-\vspace{-3mm}
-
-\begin{figure}[hb]
-\centering
-%\includegraphics[width=0.25\textwidth]{fig/m040-conformations}
-\includegraphics[width=0.14\textwidth]{fig/m040-conf1} \hskip 25pt
-\includegraphics[width=0.11\textwidth]{fig/m040-conf2}
-\caption{\label{fig::m040c}
- The examples of a `long' conformation (left) and a `packed' one for the ligand m040 (1,5-Dichloropentane).
-}
-\end{figure}
-
-To select the suitable set $\L$ of conformations from a pool of available conformations, a simple technique is proposed.  
-For each conformation, the geometric center is calculated and the deviation of the distances of the atoms from the center is calculated.
-The low value of the deviation is characteristic for the packed conformations, while larger deviations correspond to the longer conformations.
-To select $n$ conformations from all $m$ available conformations, they are sorted according to the deviation.
-Each $k-$th conformation from the sorted list, where $k=\lfloor{m\over n}\rfloor$, is added to $\L$.
-
-%\subsection{Visualization of Results}
-%
-%TODO Bara: prepsat/motivace/jak se visuazlizuje
-%toto je cast textu z toho posledniho paperu. Kdyztak to vyhod.
-%
-%%In order to fursupport the tunnel exploration process performed by biochemists.
-%The above defined characteristics can be shown even as a graph, or better, presented visually by mapping them using colors to the tunnel spheres.
-% Alternatively, the surface representation can be used to visualize the tunnel.
-%In this case, a 3D point on the tunnel surface is colored according to the property value in its nearest tunnel part $i$,  i.e., such part $i$ whose center $c_i$ is the closest to the point among all tunnel parts.
-%
-%The examples of the color mapping are shown in Fig.~\ref{fig:properties}.
-%The accessibility (Fig.~\ref{fig:properties}a) shows that more than the half of the tunnel is not accessible (red part of the tunnel).
-%The throughput shows (Fig.~\ref{fig:properties}b) that the only difficult is the part around the bottleneck (red color in Fig.~\ref{fig:properties}b), and the second half of the tunnel is also traversable.
-%
-%\begin{figure*}[h]
-%\centering
-%\begin{tabular}{ccc}
-%\includegraphics[width=0.3\textwidth]{fig/accessibility} &
-%\includegraphics[width=0.3\textwidth]{fig/throughput} &
-%\includegraphics[width=0.3\textwidth]{fig/ligand-scale} \\
-%  (a) & (b) & (c) \\                     
-%\end{tabular}
-%\caption{Tunnel properties mapped onto its surface. The tunnel begins at the top left corner.
-%(a) Accessibility --- green denotes the parts that were accessed by the ligand easily ($A(i) > 80$~\%), 
-%    while red denotes parts that were accessible  in $<10$~\% of cases.
-%(b) Throughput --- green denotes parts that are highly probable to be passed by the ligand ($T(i) > 80$~\%), 
-%    while red denotes those that were hard to pass ($T(i) < 20$~\%).
-%(c) Ligand scale --- the average scale of the ligand was \tylde 50\% in orange parts, \tylde 60\% in yellow parts, \tylde 80\% in yellow to green parts and \tylde 100\% in green parts.
-%\label{fig:properties}
-%}
-%\end{figure*}
-%
-%
-
-%\def\cstart{\overline{c_{start}}}
-%\def\cend{\overline{c_{end}}}
-
-\section{Experimental Verification}
-
-The proposed approach was used to analyze the ligand traversability towards the active site using the main tunnel in the Haloalkane dehalogenase protein (PDB ID 1CQW).
-The active site was defined using residues 38 (Asparagine), 102 (Histidine), and 103 (Aspartic Acid).
-The molecular dynamics of the protein was computed using the AMBER tool~\cite{amber}.
-In each frame, tunnels for the spherical probe 0.9~\AA\ were detected using CAVER 3.0~\cite{caver3}. 
-In most of the frames, two tunnels were detected.
-The average bottleneck sizes of these tunnels are $1.48$~\AA\ and $1.2$~\AA\ for the main and the side tunnel, respectively.
-The tunnel traversability was analyzed in 100 time steps (frames) of MD simulation.
-
-Four different scaling-down factors were used: $\S=\{0.5,0.6,0.7,0.8\}$.
-For each scale, $|\QI|=20$ collision-free initial configurations were found in the $\RI=2$~\AA\ radius around the first sphere of the tunnel (outside the protein).
-For each initial configuration, 10 trajectories were computed, which resulted
-in $100 \times 4 \times 20 \times 10 = 80,000$ trajectories per ligand.
-
-Two motion planners were utilized: the proposed planner (described in Alg.~\ref{alg::main} and Alg.~\ref{alg::expand}), denoted as \RA\ in the rest of the paper, and the Retraction-based RRT~\cite{zhangRetraction}, denoted as \RB\ in the rest of the paper.
-The \RA\ method was run with the following parameters:
-$\Imax=5,000$, $m=100$, $\gb=0.9$, $\dt=1.5$~\AA, $\rv=2$~\AA\ and $\varepsilon=0.05$~\AA.
-Both methods were implemented in the same software framework and they utilized the same data structures and routines for the collision detection and nearest-neighbor search.
-The nearest-neighbor search of both planners was based on the 6D Euclidean metric with the same weights for rotation and translation parts.
-This metric was selected based on preliminary experiments with both planners as it resulted in the highest success rates.
-
-The tested methods differ in the expansion step (lines \ref{alg::expa}--\ref{alg::expb} in Alg.~\ref{alg::expand}).
-In these lines, the \RB\ method employs the retraction step in order to find a new collision-free configuration $\qnew$ that approaches $\qrand$.
-The retraction step is realized by sampling the vicinity of the configuration $\qnear$ in the distance $0.05$~\AA\ using 20 samples.
-A collision-free sample that maximally approaches $\qrand$ is selected and added to the tree, and the retraction continues from this sample.
-The retraction step is repeated 5 times.
-The \RB\ method can therefore expand the tree by at most 5 new nodes per configuration, while \RA\ extends the tree only by one configuration.
-The larger number of nodes in \RB\ method is not problematic, as the KD-trees were used for the nearest-neighbor search and the time to solve these queries is significantly smaller than the computing collision detection.
-The parameters of the methods were selected in such a way that the number of collision detection queries in each expansion step is the same.
-For each conformation and tested scale, the \RA\ method attempts to find $m=100$ collision-free configurations, and \RB\ attempts to perform $5 \times 20$ collision-detection queries.
-Other parameters of the \RB\ planner were the same as for \RA.
-The planning was terminated after $\Imax$ iterations or if the tree approached the active site to the distance less than 0.5~\AA.
-
-In the following experiments, we measured the performance of the methods using the traversability rate, which is the percentage of frames of the protein dynamics that are traversable by the tested ligand.
-A frame is considered traversable if at least one trajectory out of 200 ($\!|\QI|\!\times\!\!10$ trials) approaches the active site to the distance less than~2~\AA.~\footnote{We refer to~\url{http://mrs.felk.cvut.cz/icra2018protein} for additional material.}
-
-\subsection{Traversability of One Atom}
-
-The aim of the first experiment is to measure the performance of both planners in the simplest scenario, i.e., finding a path for one sphere heading towards the active site.
-The radius of the sphere is set to $1.7$~\AA.
-
-The computed traversability rates are shown in Tab.~\ref{tab::one}.
-The maximum percentage of traversable frames (the row Maximum in Tab.~\ref{tab::one}) is computed as the percentage of tunnels with the bottleneck larger than the tested radius.
-The row 'Planning' shows the traversable rate computed by the planners.
-This rate decreases with the increasing radius and it is lower than the maximum traversable rate.
-The rates provided by \RA\ and \RB\ are similar which is a good indicator that the methods have been implemented correctly.
-Higher traversability rate can be achieved by increasing the maximum number of planning iterations $\Imax$, which would also
-increase the planning time.
-For the purpose of the following experiments, we used $\Imax=5,000$ as a compromise between the performance and the runtime.
-
-\begin{table}[ht]
-\centering
-\caption{\label{tab::one}Traversability for one atom.}
-\renewcommand{\tabcolsep}{4.3pt}
-{\small
-\renewcommand{\arraystretch}{0.7}
-\input graphProbe.tex
-}
-\end{table}
-
-%
-%{\def\a{270}
-%\begin{figure}
-%\centering
-%\renewcommand{\tabcolsep}{0pt}
-%\begin{tabular}{cccc}
-%\includegraphics[width=0.15\textwidth,angle=\a]{fig/m001-oneAtom-cdCount} &
-%\includegraphics[width=0.15\textwidth,angle=\a]{fig/m001-oneAtom-rrt_path_time} &
-%\includegraphics[width=0.15\textwidth,angle=\a]{fig/m037t-cdCount} & 
-%\includegraphics[width=0.15\textwidth,angle=\a]{fig/m037t-rrt_path_time} \\
-%\multicolumn{2}{c}{One-atom} &
-%\multicolumn{2}{c}{m037}  \\
-%\multicolumn{4}{c}{%
-%\includegraphics[width=0.15\textwidth,angle=\a]{fig/m040-10-cdCount} 
-%\includegraphics[width=0.15\textwidth,angle=\a]{fig/m040-10-rrt_path_time} 
-%\includegraphics[width=0.15\textwidth,angle=\a]{fig/m040-10-trajDistToGoal}} \\
-%\multicolumn{4}{c}{m040} 
-%\end{tabular}
-%\end{figure}
-%}
-%
-
-%\begin{figure}
-%\centering
-%\begin{tabular}{cc}
-%\includegraphics[width=0.3\textwidth]{fig/t05proteintunnels} &
-%\includegraphics[width=0.45\textwidth]{fig/ligAll}  \\
-%(a) & (b)                        
-%\end{tabular}                       
-%\caption{\label{fig::tunnel}
-%    (a) The protein 1CQW visualized using the cartoon representation (gray) with 
-%        three tunnels (red, green, blue) detected by CAVER 3.0.
-%        The first tunnel is depicted red.
-%%        its highest ranked tunnel (green) which was detected using CAVER 3.0.
-%    (b) Examples of three conformations of $\LA$ (top) and $\LB$ (bottom)
-%}
-%\end{figure}
-%
-
-\subsection{Influence of the Conformation Selection}
-
-The aim of this experiment was to verify the procedure to select feasible ligand conformations for the motion planning task.
-The experiment was performed with the ligand DCP (Dichloropropanol), for which $1300$ conformations were available, and for the ligand
-m040 (1,5-Dichloropentane) with $7560$ available conformations.
-Two strategies were tested: the strategy described in Sec.~\ref{sec::strat} (denoted `S') and the strategy that prefers smaller conformations (denoted `P').
-The `P' method selects such conformations where the deviations of the atomic distance from the geometric center are the lowest, which leads 
-to the selection of packed conformations (Fig.~\ref{fig::m040c}).
-
-The traversability rates are shown in Tab.~\ref{tab::selection}, where the subscript of each ligand's 
-name shows the number of the selected conformations (5 or 50) and the used method (S or P).
-The number after the slash in the ligand's name denotes the number of atoms of the ligand.
-In the case of both tested ligands and 50 selected conformations, the `S' strategy leads to a higher traversability rate than the `P' strategy.
-For the DCP ligand, a higher traversability rate was achieved when 50 conformations were selected.
-This experiment shows that selecting 50 conformations using the `S' strategy (Sec.~\ref{sec::strat}) leads to better results
-than using a less number of conformations or when preferring the packed conformations.
-
-\begin{table}[ht]
-\centering
-\caption{\label{tab::selection}
-    The influence of the conformation selection method to the traversability rates in the first tunnel.
-    The number after~`$/$' denotes the number of atoms.
-}
-\small
-\renewcommand{\tabcolsep}{4.3pt}
-{\small
-\renewcommand{\arraystretch}{0.7}
-\input graphSel.tex
-}
-\end{table}
-
-\subsection{Traversability of Ligands}
-
-Based on the results of the previous experiment, 50 conformations using the `S' strategy were selected for ligands m037t (1,2-Dichloroethane), m038t (1,3-Dichloropropane), m056 (2-Chlorobutane), and m080 (Trichloropropane), and
-the task was to compute the trajectories for these ligands in a given molecular dynamics simulation.
-
-The results are summarized in Tab.~\ref{tab::main}.
-The table shows the traversability rate only for the main tunnel, as no trajectory was found in the side tunnel for most of the tested ligands.
-
-\begin{table}[bt]
-\centering
-\caption{\label{tab::main}
-    Traversability of the tunnels for ligands with 50 conformations. 
-    The number after '$/$' denotes the number of atoms.
-}
-\small
-\renewcommand{\tabcolsep}{4.3pt}
-{\small
-\renewcommand{\arraystretch}{0.7}
-\input graph.tex
-}
-\end{table}
-
-The only exception is the ligand m037t, which was the smallest tested one (with 8 atoms), and therefore it could fit even inside the side tunnel.
-The traversability rate is higher for the most scaled-down ligands ($\smin=0.5$) and it decreases with the increasing scale.
-For the scale $\smin=0.8$, the most traversable frames were detected for ligands DCP (8~\%), m037t (20~\%) and m038t (16~\%).
-
-%{\bf doufam, ze toto neni moc drze:}
-In conclusion, the traversability of the DCP ligand to active site was approved by the MD simulation.
-The observed trajectory of the DCP ligand in the MD simulation followed the main tunnel, as expected.
-Based on the traversability rates with $\smin=0.8$, we expect that at least ligands m037t and m038t can also reach the active site, as
-they have the same (or even larger) traversability rate than the DCP.
-However, this hypothesis has not been approved yet by testing it on MD simulations.
-
-The proposed \RA\ planner provides higher traversability rate than \RB.
-The distances of the configuration trees (measured using the 3D Euclidean metric) to the active site and the runtimes in each frame
-are shown in Fig.~\ref{fig::comparison}.
-The runtimes of both algorithms increase in those frames, where the tunnel is not traversable, i.e., where the distance
-to the active site is more than 2~\AA\ (for example, in the frame no. 60).
-
-{\def\a{270}
-\begin{figure}[hb]
-\centering
-%\includegraphics[width=0.17\textwidth,angle=\a]{fig/m037tmins-03tunnel-1rrt_path_time-2}
-%\includegraphics[width=0.17\textwidth,angle=\a]{fig/m037tmins-03tunnel-1trajDistToGoal-2}
-\includegraphics[width=0.18\textwidth,angle=\a]{fig/crop1}
-\includegraphics[width=0.18\textwidth,angle=\a]{fig/crop2}
-\caption{\label{fig::comparison}
-    Runtimes and the distance to the active site for the m037 ligand at scale 0.6.
-    The values are computed from all 200 trials (20 initial configurations $\times$ 10 trials).
-    Only each third frame is shown in the graph.
-}
-\end{figure}
-}
-
-\vspace{-3mm}
-The most time-consuming part of the planners is the collision detection. 
-Although both methods can utilize the same number of collision detection queries in the expansion step, \RB\ is slower than \RA.
-The reason is that \RA\ expands the tree more closer to the random samples $\qrand$ than \RB.
-This results in a faster exploration of the configuration space and therefore, \RA\ needs less number of iterations to reach the active site.
-The better performance of \RA\ is caused by the proposed atom-based metric, because this distance depends on the actual shape of the ligand and 
-it allows the ligands to retract more towards $\qrand$.
-On the contrary, \RB\ utilizes the single 6D Euclidean weighted metric in the expansion, where the weights are the same for all conformations.
-Depending on these weights, the metric favors samples that differ from $\qrand$ more in the rotation or translation part.
-However, the weights between the rotation/translation parts should depend on the shape of the ligand, which is different in each conformation.
-By using a single set of weights in the 6D Euclidean metric, we implicitly favor either the packed or the longer conformations.
-%To improve performance of \RB, the weights of the 6D Euclidean metric should differ for each conformation, but it would
-%prevent the usage KD-tree for the nearest-neighbor search.
-
-\section{Discussion and Utilization of the Results}
-
-The presented work was motivated by the current needs of biochemists who need to analyze the traversability of selected ligands through protein tunnels.
-To find the feasible ligand trajectories, it is often required to shrink the atomic radii of the ligand, as utilized also in other existing tools.
-Finding trajectories for the scaled-down ligands enables to solve the problem with narrow passages, but it can be questionable how biochemically relevant such trajectories~are.
-
-Judging the accessibility of the active site by a ligand based only on the traversability rates may be problematic if the ligand is scaled too much.
-Instead of using a single version of the scaled-down ligand, we only limit the minimal scale used by the planner (the parameter $\smin$).
-The expansion step always prefers to expand the tree by a collision-free configuration with the maximal scale.
-%that leads to a new collision-free configuration.
-The information about the scales used at the trajectory points (which can be extracted from the configuration tree of the proposed planner) brings the biochemists new knowledge about the tunnels, which was impossible to get using the previous solutions.
-Our proposed visualization allows the biochemists to easily observe those parts of the tunnel where the ligand had to be scaled-down significantly (red parts of trajectories in Fig.~\ref{fig::dcp}), and parts where the ligand could be enlarged again (green parts in Fig.~\ref{fig::dcp}).
-According to the results of the motion planing, the DCP needs to shrink down to ca. 75~\% of its size (Fig.~\ref{fig::dcp}, point B).
-However, as both the protein and ligand are mutually influence each other, it is possible that even a very narrow part of the tunnel is traversable.
-This was observed in the MD simulation of the DCP ligand in the 1CQW protein, which revealed that DCP can pass the main tunnel.
-
-\begin{figure}
-\centering
-\includegraphics[width=0.4\textwidth]{fig/dcp-image1Label}\\
-\includegraphics[width=0.2\textwidth]{fig/colormap}
-\caption{\label{fig::dcp}
-Visualization of trajectories for the ligand DCP in the two examined tunnels.  
-The color of the trajectories corresponds to the scale, that was needed to pass the given part of the tunnel.
-The ligand can enter the first tunnel with a high scale (almost 1=100~\%) (A), but then it has to be scaled-down (to $\sim75$~\%) to pass the
-next part of the tunnel (B). After this narrow passage, the tunnel is larger again and the ligand can traverse to the narrow passage with
-scale $>75$~\% (C).
-The second tunnel can also be entered with the high scale (green), but then the narrow part is traversable only with a low scale ($\sim50$~\%) (D). 
-For the purpose of the visualization, even more scale-down to 0.3=30~\% was allowed (E) to see how the ligand passed the tunnel.
-However, the ligand can reach the active site only with very low scale ($\sim0.4=40$~\%) using the side tunnel.
-}
-\end{figure}
-
-%The computed trajectories cannot be considered as `real', they should rather be considered as a hint for the biochemists.
-%One of the reason is that the trajectories are computed inside a static protein and static tunnels.
-%Real proteins are dynamic structures and consequently also the tunnels are dynamic: they move, merge or even disappear due to motions of protein atoms.
-%Contrary, usually static tunnels are used in protein engineering.
-%Despite such a simplification, researches often used information about the static tunnels  to estimate the possibility of chemical reactions
-%at the active sites.
-%The second limitation is caused by the utilization of scaled-down ligands.
-%The protein atoms fill the internal void space therefore the tunnels identified inside proteins without ligand tend
-%to be very narrow.
-%To enable at least some motion of the ligands inside such narrow tunnels, the scaling-down of the ligand atoms is necessary.
-%It is used also in other related tool, e.g.~\cite{cortes2005path}.
-
-%Due to these reasons, the computed trajectories can be considered either as too optimistic
-%(e.g. because they are computed on a wide tunnel that can be in fact closed due to molecular dynamics),
-%or too pessimistic (e.g. the trajectory is not found because of narrow passage around the tunnel, which can be  opened due to the molecular dynamics).
-%Despite these limitations, testing the tunnel traversability using motion planning technique can provide chemists more information
-%than the simple bottleneck radius, which is used nowadays.
-%For example, possible detours from the tunnel can be identified, which indicates that even a tunnel with a small
-%bottleneck can be traversed (Fig.~\ref{fig::detail}).
-
-%Most related works, e.g. those based on motion planning, are usually focused on protein folding, or finding arbitrary pathways
-%from proteins~\cite{cortes2010simulating,guieysse2008structure}.
-
-%The runtime of the planner is currently given mainly by the collision-detection.
-%As the ligand is assumed to travel only along the tunnel centerline (in the distance defined by $\rv$),
-%only atoms located in the vicinity of the tunnel can be considered for collision detection.
-%Fast collision-detection can be achieved either using fast hierarchical data structures as OBB (implemented, e.g., in~\cite{ozcollide})
-%    or using collision-detection system developed directly for proteins~\cite{ruiz2005biocd}.
-%We assume that a conformation $l \in \L$ can be switched to all other conformations in $\L$.
-
-%parmas:
-%$\RI$
-%$\rv$
-%$\dt$
-%The presented expansion procedure favors the samples with maximal scale (they are tested first), the trajectory contains samples with higher scales at the easy part of the tunnels (e.i., wide areas), and the smaller scales are used only if the ligand has to fit into a narrow space.
-
-\section{Conclusion}
-
-In this paper, we presented a novel modification of the RRT planner for computation of ligand trajectory inside protein tunnels.
-%So far, tunnels have been mostly detected using Voronoi diagrams which do not take into account the shape and flexibility of the ligand.
-In our solution, we compute the trajectories using the modified RRT which takes into account the ligand shape and we model the ligand flexibility using a predefined set of its conformations.
-To simulate the real behavior of ligand passing through a tunnel, we enable scaling down the atomic radii of the ligands up to a certain predefined factor.
-Furthermore, our planner is localized as it generates random samples only along a tunnel computed by Voronoi diagram, and it attempts to expand the configuration
-tree using all ligand conformations. 
-The expansion prefers atoms with a higher scale, which ensures that the ligand shrinks down only if necessary, i.e., when traversing the narrow parts of the tunnel.
-To further boost the exploration of the configuration space, a novel metric suitable for measuring the distances between various ligand shapes is used in the expansion step.
-The proposed method has been used to test and evaluate the trajectories of several ligands in a sequence of frames capturing protein dynamics.
-%The traversability of one of the tested ligands has been already approved in the MD simulations.
-
-\bibliographystyle{plain}
-\bibliography{paper}
-
-\end{document}
-
-% ============  unused ideas =========================
-
-%neni treba \cite{cheng01reducing}
-%Using a library of predefined conformations has been used also in other tasks, e.g.~\cite{kellogg}.
-%This shrinking technique is also used in the MoMa-LigPath tool~\cite{cortes2005path}, which utilizes RRT to find pathways for unbinding a ligand  from a protein.
-%It is necessary to use an apporpriated distance metric in the configuration space.
-%Generally, it is hard to combine the translational and rotational components in the configuration space.
-%In~\cite{zhangRetraction}, the DISP metric~\cite{zhang2008fast}.
-%energetic constraints are translated into geometric ones by considering a steric model of the molecules and applying collision detection
-%\cite{deAngulo2005biocd}
-%First the biredge test is applied to identify regions aeround narrow passages, optimization-based retraction operatoin sleectively only at those regions selective rrt~\cite{lee2012srrrt}
-%Free-space dilation is an effective aproach for narrow passage sampling~\cite{hsu06multilevel}. 
-%It is necessary to define proper method for dilating and determin the amount of dilation.
-%\cite{saha2005finding}
-%DISP metric \cite{zhang2008fast}
-
-\subsection{Traversability Characteristics}
-
-For each initial configuration $\qinit \in \QI$, $K$ trajectories are computed, which results in the set of $K |\QI|$ trajectories.
-All these trajectories are used to compute the following properties of the tunnel $T$.
-A trajectory $P$ reaches the tunnel sphere $c_i \in T$, if the 3D Euclidean distance of the 
-nearest configuration $q \in P$ towards $c_i$ is less than $r_i$ (radius of the sphere $c_i$).
-Let $N(i)$ denote the number of trajectories that reached $i-$th sphere of the tunnel, $i=1,\ldots,n$.
-Three basic characteristics of the tunnel are computed from the trajectories: accessibility, throughput, and the scaling factor.
-
-The accessibility $A(i)=N(i)/N$ of the sphere $i$ is the probability of reaching the sphere $i$, where $N=K|\QI|$ is the total number of trajectories.
-The accessibility shows how probable it is to pass the tunnel up to the sphere $i$.
-Obviously, the most important is $A(n)$ of the last sphere of the tunnel, which can be considered as the overall difficulty of the tunnel.
-The  ligand passage may however be strongly affected by the first bottleneck, so the parts of the tunnel located behind
-the bottleneck have low accessibility.
-
-The throughput $T(i)$ is the ratio of trajectories that passed sphere $i$ (i.e., visited sphere $i+1$) and reached the sphere $i$,
-i.e., $T(i) = N(i+1) / N(i)$. 
-The throughput is not computed for the last sphere ($i=N$).
-The throughput shows a local accessibility of tunnel parts and it can be used to detect places where most of the trajectories ends.
-
-The proposed planner is allowed to scale down the radii of ligand spheres up to the allowed scale $\smin$. 
-It can be expected that narrower parts of the tunnel are more often passed with a more scaled-down ligand than
-the wider parts.
-Ligand scale $L(i)$ at the tunnel sphere $i$ is the average scale of ligand that reaches the sphere $i$.
-
-\begin{figure}
-\centering
-{
-\renewcommand{\tabcolsep}{-5pt}
-\begin{tabular}{ccc}
-\includegraphics[width=0.38\textwidth]{fig/t05proteinBottle2T} & 
-\includegraphics[width=0.38\textwidth]{fig/t04goodT} & 
-\includegraphics[width=0.38\textwidth]{fig/t05goodT} \\
-\includegraphics[width=0.38\textwidth]{fig/t05thpT} &
-\includegraphics[width=0.38\textwidth]{fig/t04badT} &
-\includegraphics[width=0.38\textwidth]{fig/t05badT} \\ 
-(a) & (b) & (c)                       
-\end{tabular}                       
-\caption{\label{fig::tunnel2}
-    (a) Classic bottleneck for spherical probe (top) and visualization of throughput (bottom) for the ligand $\LA$.
-    (b) Successful (green, top) trajectories that reached end points of the tunnels,
-        and unsuccessful ones (red, bottom) for $\smin=0.4$.
-    (c) Successful and unsuccessful trajectories for $\smin=0.5$.
-}
-}
-\end{figure}
-
-\begin{figure}
-\centering
-{
-\renewcommand{\tabcolsep}{0pt}
-\begin{tabular}{ccc}
-\includegraphics[width=0.325\textwidth]{fig/tunne4ac}& 
-\includegraphics[width=0.325\textwidth]{fig/tunne4bc} & 
-\includegraphics[width=0.325\textwidth]{fig/tunne4cc} \\
-(a) & (b) & (c)                        
-\end{tabular}
-}
-\caption{\label{fig::detail}
-    Detailed view to alternative pathway around the third tunnel in 1CQW.
-    The successful trajectories are in green (a) the unsuccessful in red (b).
-    (c) Shows visualization of protein atoms around the tunnel.
-}
-\end{figure}
-
-Besides the color mapping, it is also necessary to show the computed trajectories.
-%This is useful, for example, to find out in which part of the tunnel the trajectories detour.
-Simple visualization of all trajectories could however be too slow for an interactive work.
-Therefore, the trajectories are first clustered and then only the clusters are visualized.
-Due to different lengths of the trajectories, they are first converted to a normalized form.
-Let $\cstart$ represent the average starting position of all trajectories and let $d_{max}$ be the 3D Euclidean distance
-of the most distant configuration from $\cstart$ among all trajectories.
-A set of $M$ spheres centered at $\cstart$ are created with radii $r'_i=i {d_{max} \over M}$, where $i=0,\ldots, M-1$.
-The trajectory $P=(q_1,\ldots,q_n)$ of length $n$ is represented by the normalized vector $v=(x_1,\ldots,x_M)$ of length $M$,
-where $x_i$ is the 3D position of the nearest configuration $q \in P$ to the surface of the $i$-th sphere with the radius $r'_i$.
-The distance between two normalized trajectories $v_i$ and $v_j$ is defined as
-$d(v_i, v_j) = \frac{1}{M} \sum_{1 \leq k \leq N} |x_k^i x_k^j|$.
-This distance is used in the UPGMA clustering technique~\cite{sokal1958statistical}.
-The trajectories can be visualized using a representative of each cluster.
-The number of trajectories in each cluster is represented by the width of the polyline (Fig.~\ref{fig:trajectories}).
-%In this manner, we are able to convey the information about all different trajectories together (see Fig.~\ref{fig:trajectories} right).
-
-\begin{figure}
-\centering
-\begin{tabular}{cc}
-\includegraphics[width=0.4\textwidth]{fig/trajectories-all} &
-\includegraphics[width=0.4\textwidth]{fig/trajectories-clustered-21} \\
-(a) & (b) \\                       
-\end{tabular}
-\caption{Visualization of the trajectories. The tunnel begins in the top left corner.
-(a) All trajectories ($\sim5000$) colored according to whether they reached the end of the tunnel (green) or not (red). 
-(b) Visualization using clusters of trajectories. 
-%Trajectories clustered according to their proximity colored by distinct colors.
-%The different line widths represent the number of trajectories in a cluster.
-\label{fig:trajectories}
-}
-\end{figure}
-
-\begin{table}
-\centering
-\caption{\label{tab::rrt}
-    Runtime and success ratio for ligand $\LA$ (left) and $\LB$ (right).
-}
-{
-\small
-\renewcommand{\tabcolsep}{1pt}
-\input table.003.tex
-\hskip 4pt   
-\input table.004.tex
-}
-\end{table}
-
-Trajectories for $\LA$ in all detected tunnels were classified as successful if they reached the last sphere of the tunnel
-to distance $\dt=2$~\AA and they were considered unsuccessful otherwise.
-The throughput computed from the trajectories shows that the tunnels are difficult not only around bottlenecks, but
-also in other places.
-The comparison of the classic bottleneck (i.e., measured by the radius of spherical probe) and throughput is depicted in Fig.~\ref{fig::tunnel2}a.
-The trajectories for $\smin=0.4$ are depicted in Fig.~\ref{fig::tunnel2}b and for $\smin=0.5$ in Fig.~\ref{fig::tunnel2}c.
-The successful trajectories for $\smin=0.4$ reveal that the end of the tunnel No. 3 can be approached by two different pathways (one in the tunnel and another one outside the tunnel).
-The detail is depicted in Fig.~\ref{fig::detail}.
-Despite the low bottleneck of this tunnel, the ligand may reach its end using the alternative pathway.

### 2017-12-06 paper.tex
color map included in last image
added Ondra

@@
     Adam Jur\v{c}\'\i k$^{2}$,
+    Ond\v{r}ej V\'{a}vra$^{3}$,
     Martin Saska$^{1}$
@@
 \thanks{$^{2}$        
-Faculty of Informatics, Masaryk University, Botanick\'a 68a, 602~00 Brno, Czech Republic,
+Faculty of Informatics, Masaryk University, Botanick\'a 68a, 602~00 Brno, Czech Republic}
+\thanks{$^{3}$        
+Faculty of Informatics, Masaryk University, Botanick\'a 68a, 602~00 Brno, Czech Republic}
 %The presented work has been supported by the Czech Science Foundation (GA{\v C}R) under research project No. 17-07690S.
@@
 %}%
-}
+%}
 }
@@
 \includegraphics[width=0.4\textwidth]{fig/dcp-image1Label}\\
-\includegraphics[width=0.2\textwidth]{fig/colormap}
+%\includegraphics[width=0.2\textwidth]{fig/colormap}
 \caption{\label{fig::dcp}

### 2017-12-07 paper.tex
Intro and some comments in the cover letter

@@
 \usepackage{multirow}
+\usepackage{color}
 \usepackage{flushend}
@@
 The most common methods detect the tunnels using Voronoi diagrams~\cite{yaffe2008,caver3}, assuming a spherical ligand (probe), i.e., the approximation of the ligand by a bounding sphere.
-However, as the ligands are typically non-spherical, these methods can often skip possible candidate tunnels.
+\textcolor{blue}{However, the ligands are typically of a non-spherical shape which is completely ignored by these methods. 
+Additionally, when studying the tunnel evolution and behavior over time, the traditional methods detect tunnels in individual time frames and perform their clustering to reveal their temporal correspondence.
+In other words, they do not take into account the ligand position in the previous step, which leads to situations, when the information about a given tunnel is completely missing in few or many frames.}
 
@@
 
-To provide the biochemists with a solution which detects also the candidate tunnels omitted by the traditional methods, it is necessary to take into account not only the ligand shape but also its conformation changes when the ligand moves.
+\textcolor{blue}{To provide the biochemists with a solution which detects tunnels also in these frames and understand the ligand passage, it is necessary to take into account not only the ligand shape but also its conformation changes when the ligand moves.}
 Such a ligand, changing the mutual position of its atoms within a molecular dynamics, is denoted as a flexible ligand.
@@
 The ligand flexibility is modeled using a fixed set of predefined conformations which the ligand can occupy.
+
+\textcolor{blue}{Another advantage of our method in comparison with the existing approaches is that we enable the users to trace the ligand path leading from the outside solvent towards the protein active site.}
+
 Our proposed method was tested by the biochemists who compared the results with trajectories of ligands determined experimentally in the lab.
+\textcolor{blue}{In the following we first mention several approaches and methods related to our solution and then describe our proposed method in detail.}
 

### 2017-12-07 paper.tex
cutting of text, algorithm updated

@@
 Technick\'a 2, 166 27 Prague, Czech Republic,
-{\tt vonasek@labe.felk.cvut.cz}}
-\thanks{$^{2}$        
-Faculty of Informatics, Masaryk University, Botanick\'a 68a, 602~00 Brno, Czech Republic}
-\thanks{$^{3}$        
-Faculty of Informatics, Masaryk University, Botanick\'a 68a, 602~00 Brno, Czech Republic}
+{\tt vonasek@labe.felk.cvut.cz} $^{2}$        
+Faculty of Informatics, Masaryk University, Botanick\'a 68a, 602~00 Brno, Czech Republic 
+$^{3}$        
+Loschmidt Laboratories, Department of Experimental Biology RECETOX,Faculty of Science, Masaryk University
+    Kamenice 5, 625 00 Brno, Czech Republic}
+    }
 %The presented work has been supported by the Czech Science Foundation (GA{\v C}R) under research project No. 17-07690S.
@@
 %}
-}
+%}
 
@@
 \begin{tabular}{ccc}
-\includegraphics[width=0.15\textwidth]{fig/motiv1} &
-\includegraphics[width=0.17\textwidth]{fig/motiv2lab} &
-\includegraphics[width=0.16\textwidth]{fig/motiv3}  \\
+\includegraphics[width=0.14\textwidth]{fig/motiv1} &
+\includegraphics[width=0.15\textwidth]{fig/motiv2lab} &
+\includegraphics[width=0.14\textwidth]{fig/motiv3}  \\
 Protein 1CQW & Active site & Detected tunnels \\ %& Example of a  \\
@@
 \caption{\label{fig::motiv}
+    \small
     Tunnels in the DhaA haloalkane dehalogenase protein (PDB ID 1CQW) with possible trajectories of 1-Chlorpropan ligand (bottom left) and 1,2-Dichloroethane (bottom right).
@@
 Such a ligand, changing the mutual position of its atoms within a molecular dynamics, is denoted as a flexible ligand.
-We can formulate this task as a motion planning problem.
-The proposed solution needs also to cope with the narrow passage problem, as proteins are dense structures limiting the ligand movements inside.
+We can formulate this task as a motion planning problem and employ sampling-based methods to solve it.
+%The proposed solution needs also to cope with the narrow passage problem, as proteins are dense structures limiting the ligand movements inside.
 
-The idea of the sampling-based motion planning is to randomly sample the configuration space of the robot and to connect the collision-free samples
+Sampling-based planners randomly sample the configuration space of the robot and to connect the collision-free samples
 into a graph structure (roadmap).
 A path in the roadmap then corresponds to a motion in the workspace.
+The sampling-based planners can cope with many-DOF (Degrees of Freedom) robots of an arbitrary shape.
+Besides many applications in robotics, the sampling-based planners have been already used in biochemistry as well, 
+to study loop motions~\cite{cortes2004geometric}
+and for protein folding~\cite{raveh2009rapid,amato2002using}. %novinskaya2015improving
 %The random samples are classified as free or non-free using black-box collision detection which allows the sampling-based methods to find paths for robots of various shapes. 
 %The sampling-based planners can cope with robots with many degrees of freedom (DOF).
-Two most commonly used sampling-based techniques are Probabilistic Roadmaps (PRM)~\cite{kavrakiForPP} and Rapidly Exploring Random Trees (RRT)~\cite{lavalleRRT}.
-The sampling-based planners can cope with many-DOF (Degrees of Freedom) robots of an arbitrary shape.
-Therefore, they have been used for motion planning of various robotic systems.
-Besides many applications in robotics, the sampling-based planners have been already used in biochemistry as well, 
+%Two most commonly used sampling-based techniques are Probabilistic Roadmaps (PRM)~\cite{kavrakiForPP} and Rapidly Exploring Random Trees (RRT)~\cite{lavalleRRT}.
+%Therefore, they have been used for motion planning of various robotic systems.
 %Sampling-based motion planners has been intensively studied in robotics, but they have been applied also in other research fields.
 %In bio-chemistry, the planners have been used to study
-to study loop motions~\cite{cortes2004geometric}
 %protein folding~\cite{amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},
-and for protein folding~\cite{raveh2009rapid,novinskaya2015improving,amato2002using}.
 %or for tunnel detection~\cite{vonasek2017tunnel}.
 
-In this paper, we propose a modification to RRT to address the problem of planning a non-spherical flexible ligand passage through a given protein.
+In this paper, we propose a modification to Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} planner  to address the problem of planning a non-spherical flexible ligand passage through a given protein.
 We localize the generation of the random samples, i.e., instead of sampling the whole configuration space we focus only on the space surrounding a potential tunnel detected by the Voronoi-based method (using a smaller ligand probe in order to avoid omitting possibly feasible tunnel candidates).
@@
 %grid-based approaches~\cite{sehnal2013mole,petrek2006caver}.
-The resulting tunnels are represented as a sequence of spheres and characterized by their bottleneck, length, curvature, and a list of surrounding residues, whose physico-chemical properties are crucial for assessing the interaction possibilities between the protein and ligand.
+The tunnels are represented as a sequence of spheres and characterized by their bottleneck, length, curvature, and a list of surrounding residues, whose physico-chemical properties are crucial for assessing the interaction possibilities between the protein and ligand.
 More details can be found in the recently published state-of-the-art report about analysis and visualization of biomolecular cavities~\cite{Krone_2016}.
@@
 \section{Proposed Method}
-To address the problems of the existing solutions for computing the non-spherical flexible ligand trajectories through protein tunnels, we propose a modification of the RRT planner. 
-Our approach takes into account the shape and flexibility of the ligand and localizes the searched configuration space using the information about tunnels detected by the Voronoi-based approach.
-The input is therefore formed by a set of tunnels computed using Voronoi diagrams and the task is to find trajectories for a ligand along these tunnels
+%To address the problems of the existing solutions for computing the non-spherical flexible ligand trajectories through protein tunnels, we propose a modification of the RRT planner. 
+We propose a modification of RRT to compute trajectories for non-spherical flexible ligands through protein tunnels.
+%Our approach takes into account the shape and flexibility of the ligand and localizes the search around given tunnels. configuration space to 
+%using the information about tunnels detected by the Voronoi-based approach.
+The input is formed by a set of tunnels computed using Voronoi diagrams and the task is to find trajectories for a ligand along these tunnels
 leading from outside the protein to the active site.
@@
 
-In contrast to~\cite{cortes2010simulating}, which searches for any exit path from the protein, we localize the searching and compute the trajectories only along each tunnel separately.
-This is motivated by the practical needs of biochemists that are interested in the actual traversability the ligand through a selected tunnel.
+In contrast to~\cite{cortes2010simulating}, which searches for any exit path from the protein, we localize the searching and compute the trajectories only along each tunnel separately which allows biochemists to analyze selected tunnels.
+%This is motivated by the practical needs of biochemists that are interested in the actual traversability the ligand through a selected tunnel.
 Computing the trajectories along a single tunnel brings several advantages from the motion planning point of view as well.
 It decreases the volume of the configuration space to be searched and increases the relative volume of the narrow passages.
-To further cope with the narrow passage problem, the random samples are not generated in the whole configuration space, but only (mainly) around
-the tunnel in a user-defined distance.
-Therefore, the tunnels serve as a navigation path for the configuration tree~\cite{vonasek2009rrt}.
+The random samples are not generated in the whole configuration space, but only around the tunnel in a user-defined distance which helps to cope with the narrow passage problem.
+%The tunnels serve as a navigation path for the configuration tree~\cite{vonasek2009rrt}.
 
@@
 Tunnels serving as an input for our proposed method can be detected using some of the existing computational tools, such as CAVER 3.0~\cite{caver3}.
-
 %Conformations of the ligand may be generated in the absence of the receptor and subsequently docked[13] 
@@
 %13 = https://link.springer.com/article/10.1007/BF00123666
-
 %The proteins are dense structures and the tunnels are typically narrow.
 %Depending on the protein, the tunnel bottlenecks can be even less than $1$~\AA, which is too narrow for ligands with more than 2 atoms.
-The flexibility of the ligand is modeled using the set $\L$ of its conformations.
-We assume that the library of all possible conformations in a considered resolution is available (e.g., ~\cite{dunbrack}) and we later show how to select a subset of conformations suitable for the motion planning.
-%The conformations $\L$ are used from a library (e.g.,~\cite{dunbrack}), or they can be prepared considering the potential energy.
-
+The ligand is modeled using the set $\L$ of its conformations.
+The conformations are prepared using the Rosetta tool and ranked according their potential energy. Only conformations
+with energy lower than a user-defined threshold are used.
+%The flexibility of the ligand is modeled using the set $\L$ of its conformations.
+%We assume that the library of all possible conformations in a considered resolution is available
+%(e.g., ~\cite{dunbrack}) 
+%and we later show how to select a subset of conformations suitable for the motion planning.
 To enable the motion of ligands in the narrow tunnels, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
@@
 position defined by $q$ collides with the protein atoms.
+We assume that the ligand can change from conformation $l_1$ at configuration $q_1$ to a conformation $l_2$ at configuration $q_2$
+if both $q_1$ and $q_2$ are collision-free and if the distance from $q_1$ to $q_2$ is less than the resolution of the planner ($\varepsilon=0.05$~\AA).
 
@@
 The initial configurations are searched around the beginning of the tunnel.
-To find a new initial configuration, a random sample $q$ is generated around the first sphere of the tunnel 
+A random sample $q$ is generated around the first sphere of the tunnel 
 $c_1$ in the distance $\RI$ (the translation and rotation parts of $q$ are generated randomly, the scale is set to $\smin$, and the conformation index is set randomly).
@@
 
-The second proposed modification is the expansion procedure (Alg.~\ref{alg::expand}), which expands the tree from $\qnear$ towards $\qrand$.
-The expansion towards $\qrand$ is crucial for the growth of the tree towards the unexplored areas of the configuration space.
-Traditionally, a line is constructed from $\qnear$ to $\qrand$ and the tree is expanded by a configuration located on this line in the 
-distance $\varepsilon$ from $\qnear$.
-However, in the case of ligands, this would mostly lead to a collision.
+The second modification is the expansion procedure (Alg.~\ref{alg::expand}). % which expands the tree from $\qnear$ towards $\qrand$.
+%The expansion towards $\qrand$ is crucial for the growth of the tree towards the unexplored areas of the configuration space.
+The traditional straight-line expansion (i.e., new configuration is constructed on line from $\qnear$ to $\qrand$ in the distance
+        $\varepsilon$ from $\qnear$) more likely lead to collisions, especially in narrow void space.
+%Traditionally, a line is constructed from $\qnear$ to $\qrand$ and the tree is expanded by a configuration located on this line in the 
+%distance $\varepsilon$ from $\qnear$.
+%However, in the case of ligands, this would mostly lead to a collision.
 
-The expansion procedure attempts to expand the tree using all the conformations $\L$ and by preferring larger scales than the smaller ones.
+The proposed expansion attempts to expand the tree using all the conformations $\L$ and by preferring larger scales than the smaller ones.
 For each $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with a maximal scale.
@@
 
-In the expansion step, each conformation from $\L$ is tested in order to approach $\qrand$.
+Each conformation from $\L$ is tested during the tree expansion. %in order to approach $\qrand$.
 By using more conformations, there is a higher chance that at least one of them can expand the tree towards the random sample $\qrand$.
-On the other hand, using a large number of conformations also increases the time complexity of the expansion step.
+Using a large number of conformations also increases the time complexity of the expansion step.
 Selection of a suitable set of conformations $\L$ is therefore crucial.
@@
 packed conformations can be used.
-\vspace{-3mm}
+%\vspace{-3mm}
 
@@
 \caption{\label{fig::m040c}
+    \small
  The examples of a `long' conformation (left) and a `packed' one for the ligand m040 (1,5-Dichloropentane).
@@
 
-The proposed approach was used to analyze the ligand traversability towards the active site using the main tunnel in the Haloalkane dehalogenase protein (PDB ID 1CQW).
+The proposed method was used to compute ligand traversability towards the active site using two main tunnels in the Haloalkane dehalogenase protein (PDB ID 1CQW) which consists of 4650 atoms.
 The active site was defined using residues 38 (Asparagine), 102 (Histidine), and 103 (Aspartic Acid).
-The molecular dynamics of the protein was computed using the AMBER tool~\cite{amber}.
+MD of the protein was computed using the AMBER tool (details about the MD simulations are described in~\cite{marques2017catalytic}).
 In each frame, tunnels for the spherical probe 0.9~\AA\ were detected using CAVER 3.0~\cite{caver3}. 
@@
 The average bottleneck sizes of these tunnels are $1.48$~\AA\ and $1.2$~\AA\ for the main and the side tunnel, respectively.
-The tunnel traversability was analyzed in 100 time steps (frames) of MD simulation.
+The tunnel traversability was analyzed in 100 consecutive frames of MD simulation in part where the protein was stabilized.
 
@@
 
-Two motion planners were utilized: the proposed planner (described in Alg.~\ref{alg::main} and Alg.~\ref{alg::expand}), denoted as \RA\ in the rest of the paper, and the Retraction-based RRT~\cite{zhangRetraction}, denoted as \RB\ in the rest of the paper.
+Two motion planners were compared: the proposed planner (described in Alg.~\ref{alg::main} and Alg.~\ref{alg::expand}), denoted as \RA\ in the rest of the paper, and the Retraction-based RRT~\cite{zhangRetraction}, denoted as \RB\ in the rest of the paper.
 The \RA\ method was run with the following parameters:
 $\Imax=5,000$, $m=100$, $\gb=0.9$, $\dt=1.5$~\AA, $\rv=2$~\AA\ and $\varepsilon=0.05$~\AA.
-Both methods were implemented in the same software framework and they utilized the same data structures and routines for the collision detection and nearest-neighbor search.
+Both methods utilize the same data structures and low-level routines (e.g., collision-detection) to enable fair comparison.
+%were implemented in the same software framework and they utilized the same data structures and routines for the collision detection and nearest-neighbor search.
 The nearest-neighbor search of both planners was based on the 6D Euclidean metric with the same weights for rotation and translation parts.
@@
 \centering
-\caption{\label{tab::one}Traversability for one atom.}
+\caption{\label{tab::one}
+    \small
+    Traversability for one atom.
+}
 \renewcommand{\tabcolsep}{4.3pt}
@@
 \caption{\label{tab::selection}
+    \small
     The influence of the conformation selection method to the traversability rates in the first tunnel.
@@
 the task was to compute the trajectories for these ligands in a given molecular dynamics simulation.
+Models of the ligands in Mol2 in Mol2 formats were converted to param files~\cite{meiler2006rosettaligand} containing the topology, rotatable bonds, atom types and partial charges. 
+Rotamer library was created by rotation of all rotatable bonds with 30$^\circ$ step. 
+The energy cut-off for individual conformations was set to 0.075 %(-rot_ensemble_ecutoff 0.075) 
+and Rosetta database 3.2.1 was used during the conformers generation.
+
+%\cite{meiler2006rosettaligand}
+%All substrates were prepared by XXXXXX program and saved in mol2 file.
+Mol2 formats were converted to param files [1] containing the topology, rotatable bonds, atom types and partial charges by molfile_to_params.py script supplemented in Rosetta Scripts [2].
+
+%Rotamer library was created by rotation of all rotatable bonds with 30$^{\circ}$ step. 
+%The energy cut-off for individual conformations was set to 0.075. % (-rot_ensemble_ecutoff 0.075) 
+%and Rosetta database 3.2.1 was used during the conformers generation.
+%Conformations passing the energy criteria were output as multi-pdb file.
 
@@
 \caption{\label{tab::main}
+    \small
     Traversability of the tunnels for ligands with 50 conformations. 
@@
 \caption{\label{fig::comparison}
+    \small
     Runtimes and the distance to the active site for the m037 ligand at scale 0.6.
@@
 \caption{\label{fig::dcp}
+    \small
 Visualization of trajectories for the ligand DCP in the two examined tunnels.  

### 2017-12-07 paper.tex
sota intro to planning moved to indtorduction

@@
 
-Sampling-based planners randomly sample the configuration space of the robot and to connect the collision-free samples
+Sampling-based planners randomly sample the configuration space of the robot and connect the collision-free samples
 into a graph structure (roadmap).
-A path in the roadmap then corresponds to a motion in the workspace.
-The sampling-based planners can cope with many-DOF (Degrees of Freedom) robots of an arbitrary shape.
+%A path in the roadmap then corresponds to a motion in the workspace.
+Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} is one of the most used sampling-based planners.
+%The sampling-based planners can cope with many-DOF (Degrees of Freedom) robots of an arbitrary shape.
+%Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} is a single-query sampling-based motion planning method that 
+%It incrementally builds a configuration tree rooted at the initial configuration.
 Besides many applications in robotics, the sampling-based planners have been already used in biochemistry as well, 
-to study loop motions~\cite{cortes2004geometric}
-and for protein folding~\cite{raveh2009rapid,amato2002using}. %novinskaya2015improving
+to study loop motions~\cite{cortes2004geometric}, to identify binding sites~\cite{bayazit2001ligand}, for exit
+pathway detection~\cite{cortes2010simulating} and especially for protein folding~\cite{raveh2009rapid,amato2002using}. %novinskaya2015improving
+
+%Sampling-based planners can cope with many-DOF systems and they have been successfully employed also beoynd robotics, e.g. to
+%identify binding sites~\cite{bayazit2001ligand} or to determine ligand exit pathway~\cite{cortes2010simulating}.
 %The random samples are classified as free or non-free using black-box collision detection which allows the sampling-based methods to find paths for robots of various shapes. 
@@
 
-In this paper, we propose a modification to Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} planner  to address the problem of planning a non-spherical flexible ligand passage through a given protein.
+In this paper, we propose a modification of the RRT planner to address the problem of planning a non-spherical flexible ligand passage through a given protein.
 We localize the generation of the random samples, i.e., instead of sampling the whole configuration space we focus only on the space surrounding a potential tunnel detected by the Voronoi-based method (using a smaller ligand probe in order to avoid omitting possibly feasible tunnel candidates).
@@
 The geometric-based methods are therefore still preferable due to their speed.
+
+The task of finding the trajectories for ligand inside the proteins can be formulated as a motion planning problem.
+Sampling-based planners like RRT can be used to find the trajectories even for flexible ligands, as they can cope with many-DOF systems and they can consider arbitrary shape.
+%The computation of the ligand trajectories can be solved also using motion 
 %In order to determine if the ligand can pass the tunnel using the geometric-based methods, it is necessary to compute a trajectory considering the shape of the ligand and its 3D translation and rotation.
-
-Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} is a single-query sampling-based motion planning method that 
-incrementally builds a configuration tree rooted at the initial configuration.
-Sampling-based planners can cope with many-DOF systems and they have been successfully employed also beoynd robotics, e.g. to
-identify binding sites~\cite{bayazit2001ligand} or to determine ligand exit pathway~\cite{cortes2010simulating}.
 %In each iteration of RRT, a random configuration $\qrand \in \C$ is generated and its nearest node $\qnear \in \T$ in the tree is found.
@@
 %The algorithm terminates if the tree approaches the goal configuration close enough.
-
 %rrt for molecular: \cite{al2012motion}
@@
 or they can be identified, e.g., based on the number of collision-free samples around a given configuration~\cite{overmarsGauss,hsuBridge}.
-The narrow passages are however located in the configuration space and it is generally not easy to estimate their location
-based only on the features of the workspace~\cite{hannaWIS}.
+%The narrow passages are however located in the configuration space and it is generally not easy to estimate their location
+%based only on the features of the workspace~\cite{hannaWIS}.
 %.g. based on amount of collision-free samples around a given sample~\cite{overmarsGauss} or e.g. based
@@
 \renewcommand{\tabcolsep}{4.3pt}
-{\small
+{\footnotesize
 \renewcommand{\arraystretch}{0.7}

### 2017-12-07 paper.tex
last two paragraphs of experiments moved to discussion

@@
 
-In conclusion, the traversability of the DCP ligand to active site was approved by the MD simulation.
+In conclusion, the traversability of the DCP ligand to active site was approved by the MD simulation~\cite{marques2017catalytic}.
 The observed trajectory of the DCP ligand in the MD simulation followed the main tunnel, as expected.
@@
 However, this hypothesis has not been approved yet by testing it on MD simulations.
-
-The proposed \RA\ planner provides higher traversability rate than \RB.
-The distances of the configuration trees (measured using the 3D Euclidean metric) to the active site and the runtimes in each frame
-are shown in Fig.~\ref{fig::comparison}.
-The runtimes of both algorithms increase in those frames, where the tunnel is not traversable, i.e., where the distance
-to the active site is more than 2~\AA\ (for example, in the frame no. 60).
 
@@
 
-\vspace{-3mm}
+\section{Discussion and Utilization of the Results}
+
+The presented work was motivated by the current needs of biochemists who need to analyze the traversability of selected ligands through protein tunnels.
+To find the feasible ligand trajectories, it is often required to shrink the atomic radii of the ligand, as utilized also in other existing tools.
+Finding trajectories for the scaled-down ligands enables to solve the problem with narrow passages, but it can be questionable how biochemically relevant such trajectories~are.
+
+Judging the accessibility of the active site by a ligand based only on the traversability rates may be problematic if the ligand is scaled too much.
+Instead of using a single version of the scaled-down ligand, we only limit the minimal scale used by the planner (the parameter $\smin$).
+The expansion step always prefers to expand the tree by a collision-free configuration with the maximal scale.
+%that leads to a new collision-free configuration.
+The information about the scales used at the trajectory points (which can be extracted from the configuration tree of the proposed planner) brings the biochemists new knowledge about the tunnels, which was impossible to get using the previous solutions.
+Our proposed visualization allows the biochemists to easily observe those parts of the tunnel where the ligand had to be scaled-down significantly (red parts of trajectories in Fig.~\ref{fig::dcp}), and parts where the ligand could be enlarged again (green parts in Fig.~\ref{fig::dcp}).
+According to the results of the motion planing, the DCP needs to shrink down to ca. 75~\% of its size (Fig.~\ref{fig::dcp}, point B).
+However, as both the protein and ligand are mutually influence each other, it is possible that even a very narrow part of the tunnel is traversable.
+This was observed in the MD simulation of the DCP ligand in the 1CQW protein, which revealed that DCP can pass the main tunnel~\cite{marques2017catalytic}.
+
+%originally last two paragraphs in the experiment section, for resubmition, they are moved here to discussion
+
+The proposed \RA\ planner provides higher traversability rate than \RB.
+The distances of the configuration trees (measured using the 3D Euclidean metric) to the active site and the runtimes in each frame
+are shown in Fig.~\ref{fig::comparison}.
+The runtimes of both algorithms increase in those frames, where the tunnel is not traversable, i.e., where the distance
+to the active site is more than 2~\AA\ (for example, in the frame no. 60).
+
 The most time-consuming part of the planners is the collision detection. 
@@
 By using a single set of weights in the 6D Euclidean metric, we implicitly favor either the packed or the longer conformations.
-
-\section{Discussion and Utilization of the Results}
-
-The presented work was motivated by the current needs of biochemists who need to analyze the traversability of selected ligands through protein tunnels.
-To find the feasible ligand trajectories, it is often required to shrink the atomic radii of the ligand, as utilized also in other existing tools.
-Finding trajectories for the scaled-down ligands enables to solve the problem with narrow passages, but it can be questionable how biochemically relevant such trajectories~are.
-
-Judging the accessibility of the active site by a ligand based only on the traversability rates may be problematic if the ligand is scaled too much.
-Instead of using a single version of the scaled-down ligand, we only limit the minimal scale used by the planner (the parameter $\smin$).
-The expansion step always prefers to expand the tree by a collision-free configuration with the maximal scale.
-%that leads to a new collision-free configuration.
-The information about the scales used at the trajectory points (which can be extracted from the configuration tree of the proposed planner) brings the biochemists new knowledge about the tunnels, which was impossible to get using the previous solutions.
-Our proposed visualization allows the biochemists to easily observe those parts of the tunnel where the ligand had to be scaled-down significantly (red parts of trajectories in Fig.~\ref{fig::dcp}), and parts where the ligand could be enlarged again (green parts in Fig.~\ref{fig::dcp}).
-According to the results of the motion planing, the DCP needs to shrink down to ca. 75~\% of its size (Fig.~\ref{fig::dcp}, point B).
-However, as both the protein and ligand are mutually influence each other, it is possible that even a very narrow part of the tunnel is traversable.
-This was observed in the MD simulation of the DCP ligand in the 1CQW protein, which revealed that DCP can pass the main tunnel.
 

### 2017-12-08 paper.tex
sota updated

@@
 \thanks{$^{1}$ Faculty of Electrical Engineering,  Czech Technical University in Prague, 
-Technick\'a 2, 166 27 Prague, Czech Republic,
-{\tt vonasek@labe.felk.cvut.cz} $^{2}$        
-Faculty of Informatics, Masaryk University, Botanick\'a 68a, 602~00 Brno, Czech Republic 
-$^{3}$        
-Loschmidt Laboratories, Department of Experimental Biology RECETOX,Faculty of Science, Masaryk University
+Technick\'a 2, 166 27 Prague, Czech Republic, {\tt vonasek@labe.felk.cvut.cz};
+$^{2}$ Faculty of Informatics, Masaryk University, Botanick\'a 68a, 602~00 Brno, Czech Republic;
+$^{3}$ Loschmidt Laboratories, Department of Experimental Biology RECETOX, Faculty of Science, Masaryk University
     Kamenice 5, 625 00 Brno, Czech Republic}
@@
 Additionally, when studying the tunnel evolution and behavior over time, the traditional methods detect tunnels in individual time frames and perform their clustering to reveal their temporal correspondence.
-In other words, they do not take into account the ligand position in the previous step, which leads to situations, when the information about a given tunnel is completely missing in few or many frames.}
+Therefore, they do not take into account the ligand position in the previous step, which leads to situations, when the information about a given tunnel is completely missing in few or many frames.}
 
@@
 
-Sampling-based planners randomly sample the configuration space of the robot and connect the collision-free samples
+Sampling-based planners randomly sample the configuration space of the robot (ligand) and connect the collision-free samples
 into a graph structure (roadmap).
-%A path in the roadmap then corresponds to a motion in the workspace.
 Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} is one of the most used sampling-based planners.
-%The sampling-based planners can cope with many-DOF (Degrees of Freedom) robots of an arbitrary shape.
-%Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} is a single-query sampling-based motion planning method that 
-%It incrementally builds a configuration tree rooted at the initial configuration.
 Besides many applications in robotics, the sampling-based planners have been already used in biochemistry as well, 
-to study loop motions~\cite{cortes2004geometric}, to identify binding sites~\cite{bayazit2001ligand}, for exit
-pathway detection~\cite{cortes2010simulating} and especially for protein folding~\cite{raveh2009rapid,amato2002using}. %novinskaya2015improving
+e.g., to study loop motions~\cite{cortes2004geometric}, to identify binding sites~\cite{bayazit2001ligand}, for exit
+pathway detection~\cite{cortes2010simulating,cortes2005path} and especially for protein folding~\cite{raveh2009rapid,amato2002using}. %novinskaya2015improving
 
@@
 In this paper, we propose a modification of the RRT planner to address the problem of planning a non-spherical flexible ligand passage through a given protein.
-We localize the generation of the random samples, i.e., instead of sampling the whole configuration space we focus only on the space surrounding a potential tunnel detected by the Voronoi-based method (using a smaller ligand probe in order to avoid omitting possibly feasible tunnel candidates).
+The random samples are generated around a given tunnel instead in the whole configuration space.
+%We localize the generation of the random samples, i.e., instead of sampling the whole configuration space we focus only on the space surrounding a potential tunnel detected by the Voronoi-based method (using a smaller ligand probe in order to avoid omitting possibly feasible tunnel candidates).
 The potential tunnel is represented by a set of consecutive spheres surrounding the tunnel centerline.
-This optimization allows us to find trajectories along the given tunnel, which is faster than searching for any trajectory in the whole protein.
-To enable the movement of the ligand in the narrow parts of the tunnel, the free-space can be dilated by shrinking the atom radii, similarly, e.g., to~\cite{cortes2005path,hsu06multilevel}. 
+To enable the movement of the ligand in the narrow parts of the tunnel, the free-space is dilated by shrinking the atom radii, similarly, e.g., to~\cite{hsu06multilevel}.
+%cortes2005path removed as it is not clear it they really rescale
 The ligand flexibility is modeled using a fixed set of predefined conformations which the ligand can occupy.
 
+Finding trajectories only along a given tunnel tunnel is faster than computing any exit-pathway from the protein due to decreased
+volume of the configuration space to be searched.
 \textcolor{blue}{Another advantage of our method in comparison with the existing approaches is that we enable the users to trace the ligand path leading from the outside solvent towards the protein active site.}
@@
 The main disadvantage of both grid-based and VD-based methods is that the shape of the ligand is not taken into account during the tunnel detection and it is therefore not easy to estimate if (and how) a non-spherical ligand might traverse the tunnel.
-Moreover, the ligand flexibility cannot be principally considered by these approaches.
+The ligand flexibility cannot be principally considered by these approaches.
 Simulations based on molecular dynamics (MD) can directly evaluate the migration of a ligand into a protein, but they are significantly more computationally demanding~\cite{kingsley2014including}. 
@@
 
-The task of finding the trajectories for ligand inside the proteins can be formulated as a motion planning problem.
-Sampling-based planners like RRT can be used to find the trajectories even for flexible ligands, as they can cope with many-DOF systems and they can consider arbitrary shape.
+{\textcolor{blue}{The traversability of a tunnel by a flexible ligand can be formulated as a motion planning problem and solved using
+    sampling-based motion planners.
+These planners can cope with objects (robots) with many Degrees of Freedom (DOF)  and they can also consider arbitrary geometry.}
+Two widely used planners are Probabilistic Roadmaps (PRM)~\cite{kavrakiForPP} and Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT}.
+%Sampling-based planners like RRT can be used to find the trajectories even for flexible ligands, as they can cope with many-DOF systems and they can consider arbitrary shape.
 %The computation of the ligand trajectories can be solved also using motion 
@@
 In~\cite{bergWIG}, workspace is represented by a grid in which difficult regions are identifed using watershed labeling.
-by indetifying them in a grid-based represnetation of the workspace~
-or they can be identified, e.g., based on the number of collision-free samples around a given configuration~\cite{overmarsGauss,hsuBridge}.
+The regions can be estimated online, e.g., based on the number of collision-free samples around a given configuration~\cite{overmarsGauss,hsuBridge}.
 %The narrow passages are however located in the configuration space and it is generally not easy to estimate their location
@@
 The radius of the ball is set to infinity for new nodes, and decreases to a predefined radius if the node cannot be successfully expanded.
-The disadvantage of DD-RRT is the sensitivity to the predefined radius.
-Therefore, the authors of ADD-RRT~\cite{jailletADRRT} proposed to adapt the ball radius according to the success rate of the expansion.
+Automatic adaptation of this radius was proposed in ADD-RRT~\cite{jailletADRRT}.
+%The disadvantage of DD-RRT is the sensitivity to the predefined radius.
+%Therefore, the authors of ADD-RRT~\cite{jailletADRRT} proposed to adapt the ball radius according to the success rate of the expansion.
 To attract the tree towards a given region, random samples have to be generated there only if the tree can expand to this region.

### 2017-12-08 paper.tex
algorithm cut

@@
 To generate the samples $\qrand$ around the virtual goal $v$, the translation part $(x,y,z)$ of $\qrand$ is generated
-from the Gaussian distribution $N(c_v,\Sigma)$, where $\Sigma$ is the diagonal matrix with diagonal entries equal to the parameter $\rv$, and the rotational
-part of $\qrand$ is generated using techniques described in~\cite{kuffnerES}.
-The other two parameters (ligand index $l$ and scale $s$) of $\qrand$ can be set to zero, as these are not used in the employed
-metric for the nearest-neighbor search.
+from the Gaussian distribution $N(c_v,\Sigma)$, where $\Sigma$ is the diagonal matrix with diagonal entries equal to the parameter $\rv$, 
+and the rotational part of $\qrand$ is generated using techniques described in~\cite{kuffnerES}.
+The ligand index $l$ and scale $s$ can be set to zero, as these are not used in the employed metric for the nearest-neighbor search.
 The parameter $\rv$ influences the distribution of the random samples around the tunnel centerline. 
-By setting $\rv$ to a small value, the planner attempts to find the trajectories inside the tunnel, while higher values
-of $\rv$ cause the exploration of paths around the tunnel.
 We propose to set this parameter to the average width of the tunnel.
+%By setting $\rv$ to a small value, the planner attempts to find the trajectories inside the tunnel, while higher values
+%of $\rv$ cause the exploration of paths around the tunnel.
 
@@
 The traditional straight-line expansion (i.e., new configuration is constructed on line from $\qnear$ to $\qrand$ in the distance
-        $\varepsilon$ from $\qnear$) more likely lead to collisions, especially in narrow void space.
-%Traditionally, a line is constructed from $\qnear$ to $\qrand$ and the tree is expanded by a configuration located on this line in the 
-%distance $\varepsilon$ from $\qnear$.
-%However, in the case of ligands, this would mostly lead to a collision.
-
-The proposed expansion attempts to expand the tree using all the conformations $\L$ and by preferring larger scales than the smaller ones.
-For each $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with a maximal scale.
-First, the maximal scale $\smax \in \S$ is selected and $m$ random samples are generated around $\qnear$ (in the distance $\varepsilon$) 
+$\varepsilon$ from $\qnear$) more likely leads to collisions, especially in narrow void space.
+We expand the tree considering all $\L$ conformations.
+%The proposed expansion attempts to expand the tree using all the conformations $\L$ and by preferring larger scales than the smaller ones.
+For each $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ which the highest scale.
+First, the maximal scale $\smax \in \S$ is considered and $m$ random samples are generated around $\qnear$ (in the distance $\varepsilon$) 
 and tested for collisions.
-The random samples are generated similarly as in the case of $\qrand$ samples, only their translation
-part $(x,y,z)$ is generated around $\qnear$.
+The random samples are generated similarly as in the case of $\qrand$ samples, the translation part $(x,y,z)$ is generated around $\qnear$.
 The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
@@
 until a collision-free sample is found or until the minimal reduced-scale $\smin$ is reached.
-By testing first the largest scale ($\smax$), the expansion procedure prefers expanding the tree by larger scales.
-This forces the planner to automatically use smaller scales in the difficult (narrow) parts of the tunnels, while preferring
-large scales if possible.
+By testing first the largest scale ($\smax$), the expansion procedure prefers expanding the tree by larger scales, and the
+smaller scales are used only in difficult (narrow) regions.
 
@@
 From the motion planning point of view, both types of conformations are useful.
-The longer conformations can better fit in a narrow passage, while the packed conformations can be better rotated at a given position.
-In the narrow parts of the tunnel, the tree can be extended by the longer conformations, while in the wider regions, also the
-packed conformations can be used.
+The longer conformations can better fit in a narrow passage, while the packed conformations can easily rotate on the spot.
+%In the narrow parts of the tunnel, the tree can be extended by the longer conformations, while in the wider regions, also the
+%packed conformations can be used.
 %\vspace{-3mm}
@@
 
-To select the suitable set $\L$ of conformations from a pool of available conformations, a simple technique is proposed.  
-For each conformation, the geometric center is calculated and the deviation of the distances of the atoms from the center is calculated.
-The low value of the deviation is characteristic for the packed conformations, while larger deviations correspond to the longer conformations.
-To select $n$ conformations from all $m$ available conformations, they are sorted according to the deviation.
-Each $k-$th conformation from the sorted list, where $k=\lfloor{m\over n}\rfloor$, is added to $\L$.
+The pool of available conformations $\L$ consists all conformations (upto some resolution) with energy lower than a user speficied threshold, which can still contain too many conformations (thousands or even more).
+To select a subset of conformation, the distance of the ligand atoms to the geometric center of the ligand is computed and 
+deviation of these distances is calculated.
+Low deviation is characteristic for the packed conformations, while larger deviations correspond to the longer conformations.
+To select $n$ conformations from all $|\L|$ available conformations, they are sorted according to the deviation.
+Each $k-$th conformation from the sorted list, where $k=\lfloor{|\L| / n}\rfloor$, is used.
 
@@
 The active site was defined using residues 38 (Asparagine), 102 (Histidine), and 103 (Aspartic Acid).
-MD of the protein was computed using the AMBER tool (details about the MD simulations are described in~\cite{marques2017catalytic}).
+MD of the protein was computed using the AMBER 12 tool (details about the MD simulations are described in~\cite{marques2017catalytic}).
 In each frame, tunnels for the spherical probe 0.9~\AA\ were detected using CAVER 3.0~\cite{caver3}. 

### 2017-12-08 paper.tex
updated main graph, comments to text a

@@
 \def\RB{A$_{2}$}
+\def\RC{A$_{1}^{*}$}
+\def\RD{A$_1^P$}
 
@@
 This metric was selected based on preliminary experiments with both planners as it resulted in the highest success rates.
+The \RA\ method was also tested without the atom-based metric.
+In this case (denoted \RC), the distance in the expansion step is measured using 6D Euclidean metric.
 
@@
 In the following experiments, we measured the performance of the methods using the traversability rate, which is the percentage of frames of the protein dynamics that are traversable by the tested ligand.
-A frame is considered traversable if at least one trajectory out of 200 ($\!|\QI|\!\times\!\!10$ trials) approaches the active site to the distance less than~2~\AA.~\footnote{We refer to~\url{http://mrs.felk.cvut.cz/icra2018protein} for additional material.}
+A frame is considered traversable if at least one trajectory out of 200 ($\!|\QI|\!\times\!\!10$ trials) approaches the active site to the distance less than~2~\AA. 
+%~\footnote{We refer to~\url{http://mrs.felk.cvut.cz/icra2018protein} for additional material.}
 
@@
 
-The aim of the first experiment is to measure the performance of both planners in the simplest scenario, i.e., finding a path for one sphere heading towards the active site.
-The radius of the sphere is set to $1.7$~\AA.
-
+The performance of \RA\ and \RB\ planners is first measured by finding trajecotries for only one atom of VDW radii $1.7$~\AA.
+The \RC\ variant is not tested here as it behaves same as \RA\ in the case of one-atom ligand.
 The computed traversability rates are shown in Tab.~\ref{tab::one}.
@@
 
-Based on the results of the previous experiment, 50 conformations using the `S' strategy were selected for ligands m037t (1,2-Dichloroethane), m038t (1,3-Dichloropropane), m056 (2-Chlorobutane), and m080 (Trichloropropane), and
-the task was to compute the trajectories for these ligands in a given molecular dynamics simulation.
-Models of the ligands in Mol2 in Mol2 formats were converted to param files~\cite{meiler2006rosettaligand} containing the topology, rotatable bonds, atom types and partial charges. 
-Rotamer library was created by rotation of all rotatable bonds with 30$^\circ$ step. 
-The energy cut-off for individual conformations was set to 0.075 %(-rot_ensemble_ecutoff 0.075) 
-and Rosetta database 3.2.1 was used during the conformers generation.
+Based on the results of the previous experiment, 50 conformations using the `S' strategy were selected for ligands m037t (1,2-Dichloroethane), m038t (1,3-Dichloropropane), m056 (2-Chlorobutane), and m080 (Trichloropropane).
+Models of the ligands in Mol2 formats were converted to param files~\cite{meiler2006rosettaligand} containing the topology, rotatable bonds, atom types and partial charges. 
+Rotamer library was created by rotation of all rotatable bonds with 30$^\circ$ step using Rosetta3.
+The energy cut-off for individual conformations was set to 0.075. %(-rot_ensemble_ecutoff 0.075) 
+%and Rosetta database 3.2.1 was used during the conformers generation.
 
@@
 
-The results are summarized in Tab.~\ref{tab::main}.
-The table shows the traversability rate only for the main tunnel, as no trajectory was found in the side tunnel for most of the tested ligands.
-
 \begin{table}[bt]
@@
     \small
-    Traversability of the tunnels for ligands with 50 conformations. 
+    Traversability rate [\%] over 100 frames for ligands with 50 conformations. 
     The number after '$/$' denotes the number of atoms.
+    Each frame of protein dynamics consists of 4650 atoms.
 }
 \small
-\renewcommand{\tabcolsep}{4.3pt}
-{\footnotesize
+\renewcommand{\tabcolsep}{2pt}
+{
+\scriptsize
 \renewcommand{\arraystretch}{0.7}
-\input graph.tex
+\input graphNew.tex
 }
@@
 
+%{
+%\footnotesize
+%\begin{table}
+%\caption{New}
+%\renewcommand{\tabcolsep}{2pt}
+%\scriptsize
+%\input g2.tex
+%\end{table}
+%}
+
+The results are summarized in Tab.~\ref{tab::main}.
+The table shows the traversability rate only for the main tunnel, as no trajectory was found in the side tunnel for most of the tested ligands.
 The only exception is the ligand m037t, which was the smallest tested one (with 8 atoms), and therefore it could fit even inside the side tunnel.
 The traversability rate is higher for the most scaled-down ligands ($\smin=0.5$) and it decreases with the increasing scale.
-For the scale $\smin=0.8$, the most traversable frames were detected for ligands DCP (8~\%), m037t (20~\%) and m038t (16~\%).
-
-In conclusion, the traversability of the DCP ligand to active site was approved by the MD simulation~\cite{marques2017catalytic}.
-The observed trajectory of the DCP ligand in the MD simulation followed the main tunnel, as expected.
-Based on the traversability rates with $\smin=0.8$, we expect that at least ligands m037t and m038t can also reach the active site, as
-they have the same (or even larger) traversability rate than the DCP.
-However, this hypothesis has not been approved yet by testing it on MD simulations.
+For the ligand DCP, trajectories were detected in 10~\% of frames for scale $\smin=0.8$ and the traversability of this ligand
+to the active site was approved by a separate MD simulation~\cite{marques2017catalytic}.
+%For the scale $\smin=0.8$, the most traversable frames were detected for ligands DCP (18~\%), m037t (45~\%) and m038t (30~\%).
+%The traversability of the DCP ligand to active site was approved by a separate MD simulation~\cite{marques2017catalytic}.
+Based on the number of success rate, also other ligands (e.g., m037t with trav. rate 45~\% at $\smin$=0.8) seem to be promissing.
+However, their binding was not approved yet by MD dedicated simulations.
+
+%The observed trajectory of the DCP ligand in the MD simulation followed the main tunnel, as expected.
+%Based on the traversability rates with $\smin=0.8$, we expect that at least ligands m037t and m038t can also reach the active site, as
+%they have the same (or even larger) traversability rate than the DCP.
+%However, this hypothesis has not been approved yet by testing it on MD simulations.
 
@@
 }
+
+Beside scaling-down the ligand, also the atoms of the protein can be scaled-down, which is used e.g. in~\cite{cortes2005path} which
+scales  atoms of both the ligand and the protein to 80~\% of their size.
+We tested also this case (column \RD\ in Tab.~\ref{tab::main}).
+The results are same as for ligand scale $\smin=0.6$ (except DCP and m040 which can be results of the randomization).
 

### 2017-12-08 paper.tex
discussion

@@
 
-The proposed motion planning is based on the principle of the RRT~\cite{lavalleRRT} planner.
+The proposed algorithm is based on the principle of the RRT~\cite{lavalleRRT} planner.
 In each iteration, a random sample $\qrand$ is generated and its nearest node $\qnear\in\T$ in the tree is found.
@@
 $\da(q_1,q_2)$.
+This metric is sensitive to the shape of the ligand (that is given by it's conformation) as it computes the distance using the two nearest atoms of the two configurations.
+
 By computing the atom-based metric using the absolute positions of atoms, this metric considers the shape of the ligands, which
@@
 Depending on the ligand, some conformations can be considered as more packed i.e., the atoms in such conformations are closer to each other, but 
-the atoms can also form a more elongated shape (Fig.~\ref{fig::m040c}).
+the atoms can also form a more elongated shape (Tab.~\ref{fig::m040c}).
 From the motion planning point of view, both types of conformations are useful.
@@
 %\vspace{-3mm}
-
-\begin{figure}[hb]
-\centering
-\includegraphics[width=0.14\textwidth]{fig/m040-conf1} \hskip 25pt
-\includegraphics[width=0.11\textwidth]{fig/m040-conf2}
-\caption{\label{fig::m040c}
-    \small
- The examples of a `long' conformation (left) and a `packed' one for the ligand m040 (1,5-Dichloropentane).
-}
-\end{figure}
 
@@
 
-\begin{table}[ht]
+\begin{table}[t]
 \centering
@@
 \renewcommand{\tabcolsep}{4.3pt}
-{\footnotesize
+{\scriptsize
 \renewcommand{\arraystretch}{0.7}
@@
 
-The results are summarized in Tab.~\ref{tab::main}.
+The traversability rate is shown in Tab.~\ref{tab::main}. 
 The table shows the traversability rate only for the main tunnel, as no trajectory was found in the side tunnel for most of the tested ligands.
@@
 
+Beside scaling-down the ligand, also the atoms of the protein can be scaled-down, which is used e.g. in~\cite{cortes2005path} which
+scales  atoms of both the ligand and the protein to 80~\% of their size.
+We tested also this case (column \RD\ in Tab.~\ref{tab::main}).
+The results are same as for ligand scale $\smin=0.6$ (except DCP and m040 which can be result of the randomization).
+
+%The number of conformations (out of 50 available) used in the successfull trajectories are shown in Tab.~\ref{tab::m040c}.
 %The observed trajectory of the DCP ligand in the MD simulation followed the main tunnel, as expected.
@@
 
+\begin{table}[t]
+\caption{\label{fig::m040c}
+\small
+Examples of elongated/packed conformation for m040 (1,5-Dichloropentane).
+The table shows the average number of conformations (out of 50) used in trajectories reaching the active site.
+}
+\centering
+{\footnotesize
+\def\arraystretch{0.9}
+\begin{tabular}{cc}
+\rotatebox{0}{\hskip 5pt Long} \\ \includegraphics[width=0.1\textwidth]{fig/m040-conf1} \\
+\rotatebox{0}{Packed} \\ \includegraphics[width=0.078\textwidth]{fig/m040-conf2}
+\end{tabular}
+}
+{\scriptsize
+\def\arraystretch{0.9}
+\input usedc.tex
+}
+\end{table}
+
+\section{Discussion and Utilization of the Results}
+
+%originally last two paragraphs in the experiment section, for resubmition, they are moved here to discussion
+The proposed \RA\ planner provides higher traversability rate than RRT-Retraction (\RB).
+The distances of the configuration trees (measured using the 3D Euclidean metric) to the active site and the runtimes in each frame
+are shown in Fig.~\ref{fig::comparison}.
+The runtimes of both algorithms increase in those frames, where the tunnel is not traversable, i.e., where the distance
+to the active site is more than 2~\AA\ (for example, in the frame no. 60).
+The comparison between \RA\ and \RC\ shows that utilization of the  atom-based metric (used in \RA) leads to higher traversability rate
+in the case of large scales and larger ligands (e.g., m056 with $\smin=0.8$: 15~\% with the metric, but to only 6~\% without it).
+On the other hand, there is not practical difference between \RA\ and \RC\ in the case of more scaled-down ligands.
+
+%The most time-consuming part of the planners is the collision detection. 
+%Although both methods can utilize the same number of collision detection queries in the expansion step, 
+%\RB\ is slower than \RA, because  \RA\ expands the tree more closer to the random samples $\qrand$ than \RB.
+%This results in a faster exploration of the configuration space and therefore, \RA\ needs less number of iterations to reach the active site.
+
+%The better performance of \RA\ is caused by the proposed atom-based metric, because this distance depends on the actual shape of the ligand and 
+%it allows the ligands to retract more towards $\qrand$.
+%On the contrary, \RB\ utilizes the single 6D Euclidean weighted metric in the expansion, where the weights are the same for all conformations.
+%Depending on these weights, the metric favors samples that differ from $\qrand$ more in the rotation or translation part.
+%However, the weights between the rotation/translation parts should depend on the shape of the ligand, which is different in each conformation.
+%By using a single set of weights in the 6D Euclidean metric, we implicitly favor either the packed or the longer conformations.
+
+The presented work was motivated by the current needs of biochemists who need to analyze the traversability of selected ligands through selected protein tunnels.
+Due to low bottleneck of the tunnels, it is often required to shrink the atomic radii of the ligand, as utilized also in other existing tools.
+Finding trajectories for the scaled-down ligands enables to solve the problem with narrow passages, but it can be questionable how biochemically relevant such trajectories~are.
+
+Judging the accessibility of the active site by a ligand based only on the traversability rates may be problematic if the ligand is scaled too much.
+Instead of using a single version of the scaled-down ligand, we only limit the minimal scale used by the planner (the parameter $\smin$) and
+the planner prefers to apply the large scales in possible.
+%The expansion step always prefers to expand the tree by a collision-free configuration with the maximal scale.
+%that leads to a new collision-free configuration.
+The information about the scales used at the trajectory points (which can be extracted from the configuration tree of the proposed planner) brings the biochemists new knowledge about the tunnels, which was impossible to get using the previous solutions.
+Our proposed visualization allows the biochemists to observe those parts of the tunnel where the ligand had to be scaled-down significantly (red parts of trajectories in Fig.~\ref{fig::dcp}), and parts where the ligand could be enlarged again (green parts in Fig.~\ref{fig::dcp}).
+According to the results of the motion planing, the DCP needs to shrink down to ca. 75~\% of its size (Fig.~\ref{fig::dcp}, point B).
+However, as both the protein and ligand are mutually influence each other, it is possible that even a very narrow part of the tunnel is traversable.
+This was observed in the MD simulation of the DCP ligand in the 1CQW protein, which revealed that DCP can pass the main tunnel~\cite{marques2017catalytic}.
+
 {\def\a{270}
-\begin{figure}[hb]
+\begin{figure}[t]
 \centering
@@
 }
-
-Beside scaling-down the ligand, also the atoms of the protein can be scaled-down, which is used e.g. in~\cite{cortes2005path} which
-scales  atoms of both the ligand and the protein to 80~\% of their size.
-We tested also this case (column \RD\ in Tab.~\ref{tab::main}).
-The results are same as for ligand scale $\smin=0.6$ (except DCP and m040 which can be results of the randomization).
-
-\section{Discussion and Utilization of the Results}
-
-The presented work was motivated by the current needs of biochemists who need to analyze the traversability of selected ligands through protein tunnels.
-To find the feasible ligand trajectories, it is often required to shrink the atomic radii of the ligand, as utilized also in other existing tools.
-Finding trajectories for the scaled-down ligands enables to solve the problem with narrow passages, but it can be questionable how biochemically relevant such trajectories~are.
-
-Judging the accessibility of the active site by a ligand based only on the traversability rates may be problematic if the ligand is scaled too much.
-Instead of using a single version of the scaled-down ligand, we only limit the minimal scale used by the planner (the parameter $\smin$).
-The expansion step always prefers to expand the tree by a collision-free configuration with the maximal scale.
-%that leads to a new collision-free configuration.
-The information about the scales used at the trajectory points (which can be extracted from the configuration tree of the proposed planner) brings the biochemists new knowledge about the tunnels, which was impossible to get using the previous solutions.
-Our proposed visualization allows the biochemists to easily observe those parts of the tunnel where the ligand had to be scaled-down significantly (red parts of trajectories in Fig.~\ref{fig::dcp}), and parts where the ligand could be enlarged again (green parts in Fig.~\ref{fig::dcp}).
-According to the results of the motion planing, the DCP needs to shrink down to ca. 75~\% of its size (Fig.~\ref{fig::dcp}, point B).
-However, as both the protein and ligand are mutually influence each other, it is possible that even a very narrow part of the tunnel is traversable.
-This was observed in the MD simulation of the DCP ligand in the 1CQW protein, which revealed that DCP can pass the main tunnel~\cite{marques2017catalytic}.
-
-%originally last two paragraphs in the experiment section, for resubmition, they are moved here to discussion
-
-The proposed \RA\ planner provides higher traversability rate than \RB.
-The distances of the configuration trees (measured using the 3D Euclidean metric) to the active site and the runtimes in each frame
-are shown in Fig.~\ref{fig::comparison}.
-The runtimes of both algorithms increase in those frames, where the tunnel is not traversable, i.e., where the distance
-to the active site is more than 2~\AA\ (for example, in the frame no. 60).
-
-The most time-consuming part of the planners is the collision detection. 
-Although both methods can utilize the same number of collision detection queries in the expansion step, \RB\ is slower than \RA.
-The reason is that \RA\ expands the tree more closer to the random samples $\qrand$ than \RB.
-This results in a faster exploration of the configuration space and therefore, \RA\ needs less number of iterations to reach the active site.
-The better performance of \RA\ is caused by the proposed atom-based metric, because this distance depends on the actual shape of the ligand and 
-it allows the ligands to retract more towards $\qrand$.
-On the contrary, \RB\ utilizes the single 6D Euclidean weighted metric in the expansion, where the weights are the same for all conformations.
-Depending on these weights, the metric favors samples that differ from $\qrand$ more in the rotation or translation part.
-However, the weights between the rotation/translation parts should depend on the shape of the ligand, which is different in each conformation.
-By using a single set of weights in the 6D Euclidean metric, we implicitly favor either the packed or the longer conformations.
 
@@
 For the purpose of the visualization, even more scale-down to 0.3=30~\% was allowed (E) to see how the ligand passed the tunnel.
-However, the ligand can reach the active site only with very low scale ($\sim0.4=40$~\%) using the side tunnel.
+The ligand can reach the active site only at a very low scale ($\sim0.4=40$~\%) using the side tunnel.
 }

### 2017-12-08 paper.tex
missing files

@@
 $^{3}$ Loschmidt Laboratories, Department of Experimental Biology RECETOX, Faculty of Science, Masaryk University
-    Kamenice 5, 625 00 Brno, Czech Republic}
-    }
+    Kamenice 5, 625 00 Brno, Czech Republic. 
+The presented work has been supported by the Czech Science Foundation (GA{\v C}R) under research project No. 17-07690S.
+}}
 %The presented work has been supported by the Czech Science Foundation (GA{\v C}R) under research project No. 17-07690S.
@@
 %The traversability of the DCP ligand to active site was approved by a separate MD simulation~\cite{marques2017catalytic}.
-Based on the number of success rate, also other ligands (e.g., m037t with trav. rate 45~\% at $\smin$=0.8) seem to be promissing.
+Based on the number of success rate, also other ligands (e.g., m037t with traversability rate 45~\% at $\smin$=0.8) seem to be promissing.
 However, their binding was not approved yet by MD dedicated simulations.
@@
 
-Judging the accessibility of the active site by a ligand based only on the traversability rates may be problematic if the ligand is scaled too much.
+Judging the accessibility of the active site by a ligand based only on the traversability rates may be problematic if the ligand is scaled down too much.
 Instead of using a single version of the scaled-down ligand, we only limit the minimal scale used by the planner (the parameter $\smin$) and
-the planner prefers to apply the large scales in possible.
+the planner prefers to apply the higher scales if possible.
 %The expansion step always prefers to expand the tree by a collision-free configuration with the maximal scale.
 %that leads to a new collision-free configuration.
-The information about the scales used at the trajectory points (which can be extracted from the configuration tree of the proposed planner) brings the biochemists new knowledge about the tunnels, which was impossible to get using the previous solutions.
-Our proposed visualization allows the biochemists to observe those parts of the tunnel where the ligand had to be scaled-down significantly (red parts of trajectories in Fig.~\ref{fig::dcp}), and parts where the ligand could be enlarged again (green parts in Fig.~\ref{fig::dcp}).
+The information about the scales used at the trajectory points % (which can be extracted from the configuration tree of the proposed planner) 
+brings the biochemists new knowledge about the tunnels, which was impossible to get using the VD-based computations.
+The  visualization allows the biochemists to observe those parts of the tunnel where the ligand had to be scaled-down significantly (red parts of trajectories in Fig.~\ref{fig::dcp}), and parts where the ligand could be enlarged again (green parts in Fig.~\ref{fig::dcp}).
 According to the results of the motion planing, the DCP needs to shrink down to ca. 75~\% of its size (Fig.~\ref{fig::dcp}, point B).

### 2017-12-08 paper.tex
time/distance graphs images updated by A_1*

@@
     \small
-    The influence of the conformation selection method to the traversability rates in the first tunnel.
+    The influence of the conformation selection method on the traversability rates in the first (main) tunnel.
     The number after~`$/$' denotes the number of atoms.
@@
 in the case of large scales and larger ligands (e.g., m056 with $\smin=0.8$: 15~\% with the metric, but to only 6~\% without it).
-On the other hand, there is not practical difference between \RA\ and \RC\ in the case of more scaled-down ligands.
+Both \RA\ and \RC\ provided trajectories with the same traversability rate in the case of lower scales ($\smin<0.6$).
 
@@
 
-{\def\a{270}
+{\def\a{0}
 \begin{figure}[t]
@@
 %\includegraphics[width=0.17\textwidth,angle=\a]{fig/m037tmins-03tunnel-1trajDistToGoal-2}
-\includegraphics[width=0.18\textwidth,angle=\a]{fig/crop1}
-\includegraphics[width=0.18\textwidth,angle=\a]{fig/crop2}
+\includegraphics[width=0.2\textwidth,angle=\a]{fig/crop1}
+\includegraphics[width=0.2\textwidth,angle=\a]{fig/crop2}
 \caption{\label{fig::comparison}
     \small
-    Runtimes and the distance to the active site for the m037 ligand at scale 0.6.
+    Runtimes and the distance to the active site for the m037 ligand at scale 0.7.
     The values are computed from all 200 trials (20 initial configurations $\times$ 10 trials).

### 2017-12-08 paper.tex
changed ordering of title in the graphs 1cqw -> 4e46

@@
 
-The proposed method was used to compute ligand traversability towards the active site using two main tunnels in the Haloalkane dehalogenase protein (PDB ID 1CQW) which consists of 4650 atoms.
+The proposed method was used to compute ligand traversability towards the active site using two main tunnels in the Haloalkane dehalogenase protein (PDB ID 4E46) which consists of 4650 atoms.
 The active site was defined using residues 38 (Asparagine), 102 (Histidine), and 103 (Aspartic Acid).
-MD of the protein was computed using the AMBER 12 tool (details about the MD simulations are described in~\cite{marques2017catalytic}).
+MD of the protein was computed using the AMBER~12 (details about the MD simulations of DhaAwt/4E46 are described in~\cite{marques2017catalytic}).
 In each frame, tunnels for the spherical probe 0.9~\AA\ were detected using CAVER 3.0~\cite{caver3}. 
@@
 %\includegraphics[width=0.17\textwidth,angle=\a]{fig/m037tmins-03tunnel-1trajDistToGoal-2}
-\includegraphics[width=0.2\textwidth,angle=\a]{fig/crop1}
-\includegraphics[width=0.2\textwidth,angle=\a]{fig/crop2}
+\includegraphics[width=0.22\textwidth,angle=\a]{fig/crop1}
+\includegraphics[width=0.22\textwidth,angle=\a]{fig/crop2}
 \caption{\label{fig::comparison}

### 2017-12-08 paper.tex
intro correction

@@
 \textcolor{blue}{However, the ligands are typically of a non-spherical shape which is completely ignored by these methods. 
-Additionally, when studying the tunnel evolution and behavior over time, the traditional methods detect tunnels in individual time frames and perform their clustering to reveal their temporal correspondence.
-Therefore, they do not take into account the ligand position in the previous step, which leads to situations, when the information about a given tunnel is completely missing in few or many frames.}
+Our approach aims to overcome this problem by taking into account the actual shape of the ligand and additionally also its evolution over time, i.e. its flexibility.}
 
@@
 
-Our proposed method was tested by the biochemists who compared the results with trajectories of ligands determined experimentally in the lab.
+Our proposed method was tested by the biochemists who compared the results with already known trajectories of ligands, determined by their traditional approaches.
 \textcolor{blue}{In the following we first mention several approaches and methods related to our solution and then describe our proposed method in detail.}

### 2017-12-09 paper.tex
text all updated

@@
 The most common methods detect the tunnels using Voronoi diagrams~\cite{yaffe2008,caver3}, assuming a spherical ligand (probe), i.e., the approximation of the ligand by a bounding sphere.
-\textcolor{blue}{However, the ligands are typically of a non-spherical shape which is completely ignored by these methods. 
-Our approach aims to overcome this problem by taking into account the actual shape of the ligand and additionally also its evolution over time, i.e. its flexibility.}
+However, the ligands are typically of a non-spherical shape which is completely ignored by these methods.
 
@@
 \includegraphics[width=0.14\textwidth]{fig/motiv3}  \\
-Protein 1CQW & Active site & Detected tunnels \\ %& Example of a  \\
+Protein  & Active site & Detected tunnels \\ %& Example of a  \\
              &            & (orange)         \\  %& trajectory
@@
     \small
-    Tunnels in the DhaA haloalkane dehalogenase protein (PDB ID 1CQW) with possible trajectories of 1-Chlorpropan ligand (bottom left) and 1,2-Dichloroethane (bottom right).
+    Tunnels in the Haloalkane dehalogenase protein with possible trajectories of 1-Chlorpropan ligand (bottom left) and 1,2-Dichloroethane (bottom right).
 }
@@
 
-\textcolor{blue}{To provide the biochemists with a solution which detects tunnels also in these frames and understand the ligand passage, it is necessary to take into account not only the ligand shape but also its conformation changes when the ligand moves.}
+Our approach aims to overcome this problem by taking into account the actual shape of the ligand.
+This necessary requires to consider also conformational changes.
+%Our approach aims to overcome this problem by taking into account the actual shape of the ligand and also its conformations.
+%\textcolor{blue}{To provide the biochemists with a solution which detects tunnels also in these frames and understand the ligand passage, it is necessary to take into account not only the ligand shape but also its conformation changes when the ligand moves.}
+%The proposed solution needs also to cope with the narrow passage problem, as proteins are dense structures limiting the ligand movements inside.
 Such a ligand, changing the mutual position of its atoms within a molecular dynamics, is denoted as a flexible ligand.
-We can formulate this task as a motion planning problem and employ sampling-based methods to solve it.
-%The proposed solution needs also to cope with the narrow passage problem, as proteins are dense structures limiting the ligand movements inside.
-
+
+The analysis of traversability of the tunnels by flexible ligands can be formulated as a motion planning problem and
+solve by sampling-based planners.
 Sampling-based planners randomly sample the configuration space of the robot (ligand) and connect the collision-free samples
@@
 volume of the configuration space to be searched.
-\textcolor{blue}{Another advantage of our method in comparison with the existing approaches is that we enable the users to trace the ligand path leading from the outside solvent towards the protein active site.}
+Another advantage of our method in comparison with the existing approaches is that we enable the users to trace the ligand path leading from the outside solvent towards the protein active site.
 
 Our proposed method was tested by the biochemists who compared the results with already known trajectories of ligands, determined by their traditional approaches.
-\textcolor{blue}{In the following we first mention several approaches and methods related to our solution and then describe our proposed method in detail.}
+In the following we first mention several approaches and methods related to our solution and then describe our proposed method in detail.
 
@@
 
-{\textcolor{blue}{The traversability of a tunnel by a flexible ligand can be formulated as a motion planning problem and solved using
-    sampling-based motion planners.
-These planners can cope with objects (robots) with many Degrees of Freedom (DOF)  and they can also consider arbitrary geometry.}
+The traversability of a tunnel by a flexible ligand can be formulated as a motion planning problem and solved using sampling-based motion planners.
+These planners can cope with objects (robots) with many Degrees of Freedom (DOF)  and they can also consider arbitrary geometry.
 Two widely used planners are Probabilistic Roadmaps (PRM)~\cite{kavrakiForPP} and Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT}.
@@
 by sampling along medial axis~\cite{amatoOBRRT,amato2002using,wilmarthMAPRM}.
-In~\cite{bergWIG}, workspace is represented by a grid in which difficult regions are identifed using watershed labeling.
+In~\cite{bergWIG}, workspace is represented by a grid in which difficult regions are identified using watershed labeling.
 The regions can be estimated online, e.g., based on the number of collision-free samples around a given configuration~\cite{overmarsGauss,hsuBridge}.
@@
 In~\cite{singhLIG}, the pathways for a flexible ligand are found using Probabilistic Roadmaps.
-Each configuration in the roadmap is also evaluated using electorstatic energy.
-The high-energy configurations are discared, while low-energy conformations (and their vicinity) are sampled more dense.
+Each configuration in the roadmap is also evaluated using electrostatic energy.
+The high-energy configurations are discarded, while low-energy conformations (and their vicinity) are sampled more dense.
 The approach~\cite{singhLIG} can be also used to predict ligand binding sites.
@@
 $c_1$ in the distance $\RI$ (the translation and rotation parts of $q$ are generated randomly, the scale is set to $\smin$, and the conformation index is set randomly).
-If the sample $q$ is collision-free, it can be considered as a new starting configuration and it is added to $\QI$.
+If the sample $q$ is collision-free, it is added to $\QI$ as a new starting configuration.
 %Similarly, a single goal configuration $\qgoal$ is found around the end of the tunnel. 
+The distance $\RI$ influences how far are the initial configurations from the start of the tunnel.
+By placing them too far, the planner would need more iterations to enter the tunnel, which would increase the runtime.
+Contrary, too low $\RI$ may disable to find any collision-free configurations, especially for large ligands.
+We suggest to set $\RI$ to the radius of the first sphere in the tunnel and increase it only if no collision-free configuration
+can be found around.
 For each starting configuration $\qinit \in \QI$, the trajectories are computed using a modified RRT, which is introduced in the following section.
@@
 
+Movements of the ligands inside tunnels are blocked by many atoms forming the tunnel.
+It is not always possible for the ligand to strictly follow the centerline of the tunnel and slight detours
+should be allowed, which is controlled by the parameter $\dt$.
+This parameter should be set considering existence of cavities or other tunnels around the tunnel being investigated.
+In the case of alone tunnels (which is very common), we recommend to use double of the tunnel average width.
+
 \linesnumbered
@@
 We propose to set this parameter to the average width of the tunnel.
+
 %By setting $\rv$ to a small value, the planner attempts to find the trajectories inside the tunnel, while higher values
@@
 $\da(q_1,q_2)$.
-This metric is sensitive to the shape of the ligand (that is given by it's conformation) as it computes the distance using the two nearest atoms of the two configurations.
-
-By computing the atom-based metric using the absolute positions of atoms, this metric considers the shape of the ligands, which
-is useful when multiple conformations are utilized.
-This metric ensures that the tree is expanded by such configurations, whose atoms maximally approach
-the atoms of the configuration $\qrand$.
-We observed that this leads to a better construction of the configuration tree inside the proteins and it increases the probability
-of reaching the active site.
+This metric is sensitive to the shape of the ligand (that depends on its conformation) as it computes the distance using the two nearest atoms of the two configurations.
+It allows us to expand the tree by a configuration whose nearest atom approaches its nearest atom in $\qrand$, which supports
+retraction of the ligands towards $\qrand$.
+
+%By computing the atom-based metric using the absolute positions of atoms, this metric considers the shape of the ligands, which
+%is useful when multiple conformations are utilized.
+%This metric ensures that the tree is expanded by such configurations, whose atoms maximally approach
+%the atoms of the configuration $\qrand$.
+%We observed that this leads to a better construction of the configuration tree inside the proteins and it increases the probability of reaching the active site.
 
@@
 
-The pool of available conformations $\L$ consists all conformations (upto some resolution) with energy lower than a user speficied threshold, which can still contain too many conformations (thousands or even more).
-To select a subset of conformation, the distance of the ligand atoms to the geometric center of the ligand is computed and 
-deviation of these distances is calculated.
+The pool of available conformations $\L$ consists all conformations (upto some resolution) with energy lower than a user specified threshold, which can still contain too many conformations (thousands or even more).
+We select a subset by computing how the atoms of the ligand are spread around its geometric mean.
+The distance of centers of ligand atoms to the geometric center of the ligand is computed, and the deviation of these
+distances then characterize the spread.
 Low deviation is characteristic for the packed conformations, while larger deviations correspond to the longer conformations.
@@
 Other parameters of the \RB\ planner were the same as for \RA.
-The planning was terminated after $\Imax$ iterations or if the tree approached the active site to the distance less than 0.5~\AA.
-
-In the following experiments, we measured the performance of the methods using the traversability rate, which is the percentage of frames of the protein dynamics that are traversable by the tested ligand.
+The planning was terminated after $\Imax$ iterations or if the tree approached the active site to the distance less than 2~\AA\ (i.e., the 
+geometric center of the ligand was closer than 2~\AA\ from the active site).
+
+The performance of the methods is evaluated using the traversability rate, which is the percentage of frames of the protein dynamics that are traversable by the tested ligand.
 A frame is considered traversable if at least one trajectory out of 200 ($\!|\QI|\!\times\!\!10$ trials) approaches the active site to the distance less than~2~\AA. 
+This distance is measured as 3D distance between geometric center of the ligand and the position of the active site.
 %~\footnote{We refer to~\url{http://mrs.felk.cvut.cz/icra2018protein} for additional material.}
@@
 
-The performance of \RA\ and \RB\ planners is first measured by finding trajecotries for only one atom of VDW radii $1.7$~\AA.
+The performance of \RA\ and \RB\ planners is first measured by finding trajectories for only one atom of VDW radii $1.7$~\AA.
 The \RC\ variant is not tested here as it behaves same as \RA\ in the case of one-atom ligand.
@@
     The number after~`$/$' denotes the number of atoms.
+    The protein has 4650 atoms.
 }
@@
     The number after '$/$' denotes the number of atoms.
-    Each frame of protein dynamics consists of 4650 atoms.
+    The protein has 4650 atoms.
 }
@@
 %The traversability of the DCP ligand to active site was approved by a separate MD simulation~\cite{marques2017catalytic}.
-Based on the number of success rate, also other ligands (e.g., m037t with traversability rate 45~\% at $\smin$=0.8) seem to be promissing.
+Based on the number of success rate, also other ligands (e.g., m037t with traversability rate 45~\% at $\smin$=0.8) seem to be promising.
 However, their binding was not approved yet by MD dedicated simulations.
@@
 Trajectories for $\LA$ in all detected tunnels were classified as successful if they reached the last sphere of the tunnel
-to distance $\dt=2$~\AA and they were considered unsuccessful otherwise.
+to distance $2$~\AA or less and they were considered unsuccessful otherwise.
 The throughput computed from the trajectories shows that the tunnels are difficult not only around bottlenecks, but

### 2017-12-09 paper.tex
corrections upto V-C

@@
 We propose a modification of the Rapidly Exploring Random Tree planner for the purpose of trajectory computations of flexible ligands.
-To cope with the narrow passage problem, the random samples are generated around a Voronoi-based tunnel, which serves us for better localization of the searched space.
+To cope with the narrow passage problem, the random samples are generated around a Voronoi-based tunnel.
+%which serves us for better localization of the searched space.
 The flexibility of the ligand is modeled using a fixed set of known conformations, which does not increase the number of dimensions of the configuration space.
@@
 This space, forming a path called tunnel, can be utilized by a small molecule (ligand) which can enter the protein inner space and react with its amino acids surrounding a specific site, called the active site~\cite{gora2013gates}.
-The size and location of tunnels determine the ligand size and shape which can enter the active site.
+\textcolor{blue}{The size and location of tunnels determine the ligand size and shape which can enter the active site.}
 Currently the biochemists are using various computational methods for tunnel detection (Fig.~\ref{fig::motiv}) in order to estimate the possibility of interactions between a protein and a ligand.
 This helps them to reduce the number of necessary in-vitro lab experiments.
-The most common methods detect the tunnels using Voronoi diagrams~\cite{yaffe2008,caver3}, assuming a spherical ligand (probe), i.e., the approximation of the ligand by a bounding sphere.
+The most common tools detect the tunnels using Voronoi diagrams~\cite{yaffe2008,caver3}, assuming a spherical ligand (probe), 
+\textcolor{blue}{i.e., they approximate the ligand by a bounding sphere.}
 However, the ligands are typically of a non-spherical shape which is completely ignored by these methods.
@@
 %The proposed solution needs also to cope with the narrow passage problem, as proteins are dense structures limiting the ligand movements inside.
-Such a ligand, changing the mutual position of its atoms within a molecular dynamics, is denoted as a flexible ligand.
+Such a ligand, changing the mutual position of its atoms, is denoted as a flexible ligand.
 
 The analysis of traversability of the tunnels by flexible ligands can be formulated as a motion planning problem and
-solve by sampling-based planners.
+solved by sampling-based planners.
 Sampling-based planners randomly sample the configuration space of the robot (ligand) and connect the collision-free samples
@@
 
-In this paper, we propose a modification of the RRT planner to address the problem of planning a non-spherical flexible ligand passage through a given protein.
+In this paper, we propose a modification of the RRT planner to find trajectories for non-spherical flexible ligands inside protein tunnels.
 The random samples are generated around a given tunnel instead in the whole configuration space.
@@
 volume of the configuration space to be searched.
-Another advantage of our method in comparison with the existing approaches is that we enable the users to trace the ligand path leading from the outside solvent towards the protein active site.
-
-Our proposed method was tested by the biochemists who compared the results with already known trajectories of ligands, determined by their traditional approaches.
-In the following we first mention several approaches and methods related to our solution and then describe our proposed method in detail.
+Another advantage of our method in comparison with the existing approaches is that we enable the users to trace the ligand path leading from the
+\textcolor{blue}{outside/outer?} solvent towards the protein active site.
+
+The proposed method was tested by the biochemists who compared the results with already known trajectories of ligands, determined by their traditional approaches.
+In the following section, we first mention several approaches and methods related to our solution and then we describe the proposed method in detail.
 
@@
 In~\cite{bergWIG}, workspace is represented by a grid in which difficult regions are identified using watershed labeling.
-The regions can be estimated online, e.g., based on the number of collision-free samples around a given configuration~\cite{overmarsGauss,hsuBridge}.
+The regions can be estimated online, e.g., based on the number of collision-free samples around a given configuration~\cite{overmarsGauss,hsuBridge} or even identified by a human operator~\cite{denny2018general}.
 %The narrow passages are however located in the configuration space and it is generally not easy to estimate their location
@@
 
-RRT-based planners cannot cope with the narrow passage problem simply by increasing the probability of sampling in them, as the
+RRT-based planners cannot cope with the narrow passage problem simply by increasing the probability of sampling in the difficult regions, as the
 growth of the configuration tree towards the random samples may be blocked by obstacles. %~\cite{vonasekphd}.
@@
 In~\cite{kardossRRTKK}, the probability of sampling is increased in several waypoints close to narrow passages, but without specifying
-the method to find these waypoints.
+how to find these waypoints.
 The generalization of~\cite{kardossRRTKK} is the guided sampling, where a whole path is used
@@
 In MARRT~\cite{denny2014marrt}, the newly generated samples are pushed towards the medial axis.
-Difficult regions can be also identified manually by a human operator~\cite{denny2018general}. 
 %The operator can also select a suitable planner for the region.<F12>
@@
 RRT-ML expands the tree primarily using those DOF that are essential for achieving the motion of the ligand (i.e., rotation
-and translation) and it employs the other DOFs (i.e., those that are responsible for conformation changes) if they hinder the growth of the tree.
-In~\cite{singhLIG}, the pathways for a flexible ligand are found using Probabilistic Roadmaps.
+and translation) and it employs the other DOFs (i.e., those that are responsible for \textcolor{blue}{conformation/al?} changes) if they hinder the growth of the tree.
+In~\cite{singhLIG}, the pathways for a flexible ligand are found using PRM.
 Each configuration in the roadmap is also evaluated using electrostatic energy.
@@
 leading from outside the protein to the active site.
-The tunnels are detected in a sequence time steps capturing protein dynamic behavior computed using MD simulations.
+The tunnels are detected in a sequence of time frames of MD simulations.
+%capturing protein dynamic behavior. 
+%The frames are computed using MD simulations.
 
@@
 Computing the trajectories along a single tunnel brings several advantages from the motion planning point of view as well.
-It decreases the volume of the configuration space to be searched and increases the relative volume of the narrow passages.
+By selecting a tunnel to be analyzed, the users help the planner to focus only on a subset of the whole configuration
+space, similarly as in~\cite{denny2018general}.
+Consequently, the volume of the configuration space to be searched is decreased, while the relative volume of the narrow passages is increased.
 The random samples are not generated in the whole configuration space, but only around the tunnel in a user-defined distance which helps to cope with the narrow passage problem.
@@
 Therefore, the ligand flexibility can be modeled using only a predefined set of conformations.
-So the number of dimensions of the configuration space does not depend on the DOF of the ligand as in~\cite{cortes2010simulating}.
+By using the set of predefined conformations, it is not necessary to model the additional degrees of freedom, so 
+the dimension of the conifguration space is not increased as in~\cite{cortes2010simulating}.
+%So the number of dimensions of the configuration space does not depend on the DOF of the ligand as in~\cite{cortes2010simulating}.
 The libraries containing possible ligand conformations are used also in different types of calculations, e.g., in~\cite{kellogg}.
@@
 The ligand is modeled using the set $\L$ of its conformations.
-The conformations are prepared using the Rosetta tool and ranked according their potential energy. Only conformations
-with energy lower than a user-defined threshold are used.
+Candidate conformations are prepared by discrete sampling of degrees of freedom of the ligand.
+Each candidate conformation is ranked according its energy and only low-energy conformation are used in $\L$.
+%The conformations are prepared using the Rosetta tool and ranked according their potential energy. Only conformations
+%with energy lower than a user-defined threshold are used.
 %The flexibility of the ligand is modeled using the set $\L$ of its conformations.
@@
 
-Movements of the ligands inside tunnels are blocked by many atoms forming the tunnel.
+Movements of the ligands inside a tunnel are blocked by many atoms forming the tunnel.
 It is not always possible for the ligand to strictly follow the centerline of the tunnel and slight detours
@@
 %The expansion towards $\qrand$ is crucial for the growth of the tree towards the unexplored areas of the configuration space.
-The traditional straight-line expansion (i.e., new configuration is constructed on line from $\qnear$ to $\qrand$ in the distance
+The traditional straight-line expansion (i.e., new configuration is constructed on the line from $\qnear$ to $\qrand$ in the distance
 $\varepsilon$ from $\qnear$) more likely leads to collisions, especially in narrow void space.
@@
 %The proposed expansion attempts to expand the tree using all the conformations $\L$ and by preferring larger scales than the smaller ones.
-For each $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ which the highest scale.
+For each $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with the highest scale.
 First, the maximal scale $\smax \in \S$ is considered and $m$ random samples are generated around $\qnear$ (in the distance $\varepsilon$) 
@@
 The distance of centers of ligand atoms to the geometric center of the ligand is computed, and the deviation of these
-distances then characterize the spread.
+distances then characterizes the spread.
 Low deviation is characteristic for the packed conformations, while larger deviations correspond to the longer conformations.
@@
 The active site was defined using residues 38 (Asparagine), 102 (Histidine), and 103 (Aspartic Acid).
-MD of the protein was computed using the AMBER~12 (details about the MD simulations of DhaAwt/4E46 are described in~\cite{marques2017catalytic}).
+MD of the protein was computed using AMBER~12 (details about the used MD simulations of DhaAwt/4E46 are described in~\cite{marques2017catalytic}).
 In each frame, tunnels for the spherical probe 0.9~\AA\ were detected using CAVER 3.0~\cite{caver3}. 
@@
 The \RA\ method was also tested without the atom-based metric.
-In this case (denoted \RC), the distance in the expansion step is measured using 6D Euclidean metric.
-
-The tested methods differ in the expansion step (lines \ref{alg::expa}--\ref{alg::expb} in Alg.~\ref{alg::expand}).
+In this case (denoted \RC), the distance in the expansion step is measured using the 6D Euclidean metric.
+
+The methods \RA\ and \RB\ differ in the expansion step (lines \ref{alg::expa}--\ref{alg::expb} in Alg.~\ref{alg::expand}).
 In these lines, the \RB\ method employs the retraction step in order to find a new collision-free configuration $\qnew$ that approaches $\qrand$.
@@
 
-The performance of \RA\ and \RB\ planners is first measured by finding trajectories for only one atom of VDW radii $1.7$~\AA.
-The \RC\ variant is not tested here as it behaves same as \RA\ in the case of one-atom ligand.
+The performance of \RA\ and \RB\ planners was first measured by finding trajectories for only one atom of VDW radii $1.7$~\AA.
+The \RC\ variant was not tested here as it behaves same as \RA\ in the case of one-atom ligand.
 The computed traversability rates are shown in Tab.~\ref{tab::one}.
@@
     \small
-    Traversability for one atom.
+    Traversability rate [\%] of one atom.
 }
@@
 The `P' method selects such conformations where the deviations of the atomic distance from the geometric center are the lowest, which leads 
-to the selection of packed conformations (Fig.~\ref{fig::m040c}).
+to the selection of packed conformations (examples are shown in Tab.~\ref{fig::m040c}).
 

### 2017-12-09 paper.tex
removed text beyond end of document

@@
 
-%{
-%\footnotesize
-%\begin{table}
-%\caption{New}
-%\renewcommand{\tabcolsep}{2pt}
-%\scriptsize
-%\input g2.tex
-%\end{table}
-%}
-
 The traversability rate is shown in Tab.~\ref{tab::main}. 
@@
 %DISP metric \cite{zhang2008fast}
-
-\subsection{Traversability Characteristics}
-
-For each initial configuration $\qinit \in \QI$, $K$ trajectories are computed, which results in the set of $K |\QI|$ trajectories.
-All these trajectories are used to compute the following properties of the tunnel $T$.
-A trajectory $P$ reaches the tunnel sphere $c_i \in T$, if the 3D Euclidean distance of the 
-nearest configuration $q \in P$ towards $c_i$ is less than $r_i$ (radius of the sphere $c_i$).
-Let $N(i)$ denote the number of trajectories that reached $i-$th sphere of the tunnel, $i=1,\ldots,n$.
-Three basic characteristics of the tunnel are computed from the trajectories: accessibility, throughput, and the scaling factor.
-
-The accessibility $A(i)=N(i)/N$ of the sphere $i$ is the probability of reaching the sphere $i$, where $N=K|\QI|$ is the total number of trajectories.
-The accessibility shows how probable it is to pass the tunnel up to the sphere $i$.
-Obviously, the most important is $A(n)$ of the last sphere of the tunnel, which can be considered as the overall difficulty of the tunnel.
-The  ligand passage may however be strongly affected by the first bottleneck, so the parts of the tunnel located behind
-the bottleneck have low accessibility.
-
-The throughput $T(i)$ is the ratio of trajectories that passed sphere $i$ (i.e., visited sphere $i+1$) and reached the sphere $i$,
-i.e., $T(i) = N(i+1) / N(i)$. 
-The throughput is not computed for the last sphere ($i=N$).
-The throughput shows a local accessibility of tunnel parts and it can be used to detect places where most of the trajectories ends.
-
-The proposed planner is allowed to scale down the radii of ligand spheres up to the allowed scale $\smin$. 
-It can be expected that narrower parts of the tunnel are more often passed with a more scaled-down ligand than
-the wider parts.
-Ligand scale $L(i)$ at the tunnel sphere $i$ is the average scale of ligand that reaches the sphere $i$.
-
-\begin{figure}
-\centering
-{
-\renewcommand{\tabcolsep}{-5pt}
-\begin{tabular}{ccc}
-\includegraphics[width=0.38\textwidth]{fig/t05proteinBottle2T} & 
-\includegraphics[width=0.38\textwidth]{fig/t04goodT} & 
-\includegraphics[width=0.38\textwidth]{fig/t05goodT} \\
-\includegraphics[width=0.38\textwidth]{fig/t05thpT} &
-\includegraphics[width=0.38\textwidth]{fig/t04badT} &
-\includegraphics[width=0.38\textwidth]{fig/t05badT} \\ 
-(a) & (b) & (c)                       
-\end{tabular}                       
-\caption{\label{fig::tunnel2}
-    (a) Classic bottleneck for spherical probe (top) and visualization of throughput (bottom) for the ligand $\LA$.
-    (b) Successful (green, top) trajectories that reached end points of the tunnels,
-        and unsuccessful ones (red, bottom) for $\smin=0.4$.
-    (c) Successful and unsuccessful trajectories for $\smin=0.5$.
-}
-}
-\end{figure}
-
-\begin{figure}
-\centering
-{
-\renewcommand{\tabcolsep}{0pt}
-\begin{tabular}{ccc}
-\includegraphics[width=0.325\textwidth]{fig/tunne4ac}& 
-\includegraphics[width=0.325\textwidth]{fig/tunne4bc} & 
-\includegraphics[width=0.325\textwidth]{fig/tunne4cc} \\
-(a) & (b) & (c)                        
-\end{tabular}
-}
-\caption{\label{fig::detail}
-    Detailed view to alternative pathway around the third tunnel in 1CQW.
-    The successful trajectories are in green (a) the unsuccessful in red (b).
-    (c) Shows visualization of protein atoms around the tunnel.
-}
-\end{figure}
-
-Besides the color mapping, it is also necessary to show the computed trajectories.
-%This is useful, for example, to find out in which part of the tunnel the trajectories detour.
-Simple visualization of all trajectories could however be too slow for an interactive work.
-Therefore, the trajectories are first clustered and then only the clusters are visualized.
-Due to different lengths of the trajectories, they are first converted to a normalized form.
-Let $\cstart$ represent the average starting position of all trajectories and let $d_{max}$ be the 3D Euclidean distance
-of the most distant configuration from $\cstart$ among all trajectories.
-A set of $M$ spheres centered at $\cstart$ are created with radii $r'_i=i {d_{max} \over M}$, where $i=0,\ldots, M-1$.
-The trajectory $P=(q_1,\ldots,q_n)$ of length $n$ is represented by the normalized vector $v=(x_1,\ldots,x_M)$ of length $M$,
-where $x_i$ is the 3D position of the nearest configuration $q \in P$ to the surface of the $i$-th sphere with the radius $r'_i$.
-The distance between two normalized trajectories $v_i$ and $v_j$ is defined as
-$d(v_i, v_j) = \frac{1}{M} \sum_{1 \leq k \leq N} |x_k^i x_k^j|$.
-This distance is used in the UPGMA clustering technique~\cite{sokal1958statistical}.
-The trajectories can be visualized using a representative of each cluster.
-The number of trajectories in each cluster is represented by the width of the polyline (Fig.~\ref{fig:trajectories}).
-%In this manner, we are able to convey the information about all different trajectories together (see Fig.~\ref{fig:trajectories} right).
-
-\begin{figure}
-\centering
-\begin{tabular}{cc}
-\includegraphics[width=0.4\textwidth]{fig/trajectories-all} &
-\includegraphics[width=0.4\textwidth]{fig/trajectories-clustered-21} \\
-(a) & (b) \\                       
-\end{tabular}
-\caption{Visualization of the trajectories. The tunnel begins in the top left corner.
-(a) All trajectories ($\sim5000$) colored according to whether they reached the end of the tunnel (green) or not (red). 
-(b) Visualization using clusters of trajectories. 
-%Trajectories clustered according to their proximity colored by distinct colors.
-%The different line widths represent the number of trajectories in a cluster.
-\label{fig:trajectories}
-}
-\end{figure}
-
-\begin{table}
-\centering
-\caption{\label{tab::rrt}
-    Runtime and success ratio for ligand $\LA$ (left) and $\LB$ (right).
-}
-{
-\small
-\renewcommand{\tabcolsep}{1pt}
-\input table.003.tex
-\hskip 4pt   
-\input table.004.tex
-}
-\end{table}
-
-Trajectories for $\LA$ in all detected tunnels were classified as successful if they reached the last sphere of the tunnel
-to distance $2$~\AA or less and they were considered unsuccessful otherwise.
-The throughput computed from the trajectories shows that the tunnels are difficult not only around bottlenecks, but
-also in other places.
-The comparison of the classic bottleneck (i.e., measured by the radius of spherical probe) and throughput is depicted in Fig.~\ref{fig::tunnel2}a.
-The trajectories for $\smin=0.4$ are depicted in Fig.~\ref{fig::tunnel2}b and for $\smin=0.5$ in Fig.~\ref{fig::tunnel2}c.
-The successful trajectories for $\smin=0.4$ reveal that the end of the tunnel No. 3 can be approached by two different pathways (one in the tunnel and another one outside the tunnel).
-The detail is depicted in Fig.~\ref{fig::detail}.
-Despite the low bottleneck of this tunnel, the ligand may reach its end using the alternative pathway.
-
-%approximate the distance metric using nonlinear parametric regression model 
-%\cite{palmieri2015Distance}
-%true cost-to-go function implies solving two-point boundary problem which can be as expensive as solving the planning query itself.
-
-%Motion planning for flexible ligands in protein tunnels brings two main issues: 
-%the ligand flexibility increases the dimension of the configuration space, and the necessity to plan in protein tunnels
-%leads to the narrow passage problem.
-%A narrow passage is a region in the configuration space whose removal changes the connectivity of the free space~\cite{hannaWIS}.
-%Narrow passages have smaller volume than other regions and it is therefore difficult to sample them dense enough using uniform distribution, that is used in the basic sampling-based planners.
-%The presence of narrow passages in the configuration space, especially if they contain part of the solution, requires many iterations
-%in order to put enough samples there and it consequently increases the planning time.
-%Typical tunnels have bottlenecks smaller than $1.0$~\AA, so they can already be considered as narrow passages for ligands with more than two atoms.
-%%The protein tunnels have the properties of the narrow passages, as they are very narrow
-%To cope with the narrow passages, they have to be sampled more dense.
-%For the family of RRT planners, it is useful to change the distribution of random samples according to the growth of the tree~\cite{vonasek2009rrt,vonasekphd,denny2016dynamic}.
-%%about their 3D position, which is represented by the tunnel centerline.
-%%The centerline can be used to guide the growth of the RRT tree through the configuration space~\cite{vonasek2009rrt,denny2016dynamic}.
-%
-%%The additional degrees of freedom required to model the flexibility increase the dimension of the configuration space.
-%The kinematic chain representation used to model the ligand flexibility can generate all possible conformations but it also increases the dimension of the configuration space.
-%Not all of conformations are however feasible and it is therefore necessary to verify the feasibility of a given conformation based on energy, which is time consuming.
-%%RRT-based planners are sensitive to the employed metric, that should consider also the flexibility-related DOFs.
-%An alternative solution is to employ a library of known conformations.
-%The conformation changes of protein amino acids and ligand are represented by so called rotamers, stored in different rotamer libraries (e.g., Dunbrack library~\cite{dunbrack}).
-%%Rotamers differ in the angles between their atoms.
-%Possible rotamer conformations correspond to energetically and geometrically favorable positions.
-%%This solution is common also in different types of calculations, e.g.~\cite{kellogg}.
-
-%is used also in other related tools like like MoMa-LigPath~\cite{cortes2005path}
-%Such techniques have been used for motion planning of deformable objects, 
-%e.g., in~\cite{frank2008efficient,bayazit2001ligand,alterovitz2008motion,lamiraux2001flexible,kavraki1998towards,gayle2005path}, 
-%e.g., in~\cite{frank2008efficient,bayazit2001ligand,alterovitz2008motion,lamiraux2001flexible,gayle2005path}, 
-%and for motion planning of rigid objects among 
-%flexible obstacles~\cite{rodriguez2006planning,frank2008efficient,phillips2014representation}.
-
-%Due to the interaction, the relative positions of the ligand's atoms can change.
-% in cortes2005path:
-%In this paper two kinds of large-amplitude motion are treated: protein loop conformational changes (involving pro-
-%tein backbone flexibility) and ligand trajectories to deep active sites in proteins (involving ligand and protein side-chain flex-
-%ibility). First studies performed using our two-stage approach (geometric search followed by energy refinements) show that,
-%    compared to classical molecular modeling methods, quite   similar results can be obtained with a performance gain of
-% several orders of magnitude. Furthermore, our results also  indicate that the geometric stage can provide highly valuable information to biologists.
-%This technique is similar to generation of samples near the medial exist of the environment
-%medial axis~\cite{wilmarthMAPRM,foskey01hybrid,guibas1999probabilistic,hoff2000interactive,yang2004adapting,amatoOBRRT}.
-%or by guiding the growth of the tree around a general path in the workspace~\cite{vonasek2009rrt,denny2014marrt}.
-
-%Multiple static structures experimentally determined for the same protein in different conformations are often used to emulate receptor flexibility.[19] Alternatively rotamer libraries of amino acid side chains that surround the binding cavity may be searched to generate alternate but energetically reasonable protein conformations.[20][21]
-% 19 = \cite{totrov2008flexible}
-% 20 = \cite{hartmann2009docking}
-% 21 = \cite{taylor2003fds}

### 2017-12-09 paper.tex
table 2 regeneratd, new type of graph in Fig, 2

@@
 
-The aim of this experiment was to verify the procedure to select feasible ligand conformations for the motion planning task.
-The experiment was performed with the ligand DCP (Dichloropropanol), for which $1300$ conformations were available, and for the ligand
-m040 (1,5-Dichloropentane) with $7560$ available conformations.
-Two strategies were tested: the strategy described in Sec.~\ref{sec::strat} (denoted `S') and the strategy that prefers smaller conformations (denoted `P').
+The aim of this experiment was to verify the procedure to select feasible ligand conformations from an available pool of conformations.
+Models  of the ligands in Mol2 formats were converted to param files~\cite{meiler2006rosettaligand} containing the topology, rotatable bonds, atom types and partial charges. 
+The pool of conformations $\L$ was created by rotation of all rotatable bonds with 30$^\circ$ step using Rosetta3.
+The energy cut-off for individual conformations was set to 0.075. %(-rot_ensemble_ecutoff 0.075) 
+Two strategies to select a subset from the $\L$ conformations were tested: 
+the strategy described in Sec.~\ref{sec::strat} (denoted `S') and the strategy that prefers smaller conformations (denoted `P').
 The `P' method selects such conformations where the deviations of the atomic distance from the geometric center are the lowest, which leads 
-to the selection of packed conformations (examples are shown in Tab.~\ref{fig::m040c}).
+to the selection of packed conformations.
+The strategies were tested using the ligand DCP (Dichloropropanol), for which $1300$ conformations were created, and for the ligand
+m040 (1,5-Dichloropentane) with $7560$ conformations.
 
@@
 In the case of both tested ligands and 50 selected conformations, the `S' strategy leads to a higher traversability rate than the `P' strategy.
-For the DCP ligand, a higher traversability rate was achieved when 50 conformations were selected.
+For the DCP ligand, higher traversability rate was achieved when 50 conformations were selected.
 This experiment shows that selecting 50 conformations using the `S' strategy (Sec.~\ref{sec::strat}) leads to better results
@@
 \renewcommand{\arraystretch}{0.7}
-\input graphSel.tex
+\input graphSelNew.tex
 }
@@
 
-Based on the results of the previous experiment, 50 conformations using the `S' strategy were selected for ligands m037t (1,2-Dichloroethane), m038t (1,3-Dichloropropane), m056 (2-Chlorobutane), and m080 (Trichloropropane).
-Models of the ligands in Mol2 formats were converted to param files~\cite{meiler2006rosettaligand} containing the topology, rotatable bonds, atom types and partial charges. 
-Rotamer library was created by rotation of all rotatable bonds with 30$^\circ$ step using Rosetta3.
-The energy cut-off for individual conformations was set to 0.075. %(-rot_ensemble_ecutoff 0.075) 
+Based on the results of the previous experiment, 50 conformations using the `S' strategy were selected for ligands 
+m037t (1,2-Dichloroethane), m038t (1,3-Dichloropropane), m056 (2-Chlorobutane), and m080 (Trichloropropane).
+
     %and Rosetta database 3.2.1 was used during the conformers generation. 
-
 %\cite{meiler2006rosettaligand}
@@
 Mol2 formats were converted to param files [1] containing the topology, rotatable bonds, atom types and partial charges by molfile_to_params.py script supplemented in Rosetta Scripts [2].
-
 %Rotamer library was created by rotation of all rotatable bonds with 30$^{\circ}$ step. 
@@
 The traversability rate is higher for the most scaled-down ligands ($\smin=0.5$) and it decreases with the increasing scale.
-For the ligand DCP, trajectories were detected in 10~\% of frames for scale $\smin=0.8$ and the traversability of this ligand
+For the ligand DCP, trajectories were detected in 18~\% of frames for scale $\smin=0.8$ and the traversability of this ligand
 to the active site was approved by a separate MD simulation~\cite{marques2017catalytic}.
@@
 
-Beside scaling-down the ligand, also the atoms of the protein can be scaled-down, which is used e.g. in~\cite{cortes2005path} which
-scales  atoms of both the ligand and the protein to 80~\% of their size.
-We tested also this case (column \RD\ in Tab.~\ref{tab::main}).
-The results are same as for ligand scale $\smin=0.6$ (except DCP and m040 which can be result of the randomization).
+Beside scaling-down the ligand, also the atoms of the protein can be scaled, which is used e.g. in~\cite{cortes2005path}.
+The tool~\cite{cortes2005path} uses the scale 0.8 for both the ligand and the protein.
+We tested also this case (the column \RD\ in Tab.~\ref{tab::main}).
+The results are same as for ligand scale $\smin=0.6$ (except DCP and m040 which can be a result of the randomization).
 
@@
 are shown in Fig.~\ref{fig::comparison}.
-The runtimes of both algorithms increase in those frames, where the tunnel is not traversable, i.e., where the distance
-to the active site is more than 2~\AA\ (for example, in the frame no. 60).
-The comparison between \RA\ and \RC\ shows that utilization of the  atom-based metric (used in \RA) leads to higher traversability rate
+The runtimes of \RB\ are higher than runtimes of \RA and the std. deviations as well.
+The variant \RC\ is faster than \RA\ due to simple evaluation of the metric in the expansion step.
+%The runtimes of both algorithms increase in those frames, where the tunnel is not traversable, i.e., where the distance
+%to the active site is more than 2~\AA\ (for example, in the frame no. 60).
+The utilization of the  atom-based metric (used in \RA) leads to higher traversability rate
 in the case of large scales and larger ligands (e.g., m056 with $\smin=0.8$: 15~\% with the metric, but to only 6~\% without it).

### 2017-12-09 paper.tex
corections

@@
 \textcolor{blue}{The size and location of tunnels determine the ligand size and shape which can enter the active site.}
-Currently the biochemists are using various computational methods for tunnel detection (Fig.~\ref{fig::motiv}) in order to estimate the possibility of interactions between a protein and a ligand.
+Currently, the biochemists use various computational methods for tunnel detection (Fig.~\ref{fig::motiv}) in order to estimate the possibility of interactions between a protein and a ligand.
 This helps them to reduce the number of necessary in-vitro lab experiments.
-The most common tools detect the tunnels using Voronoi diagrams~\cite{yaffe2008,caver3}, assuming a spherical ligand (probe), 
-\textcolor{blue}{i.e., they approximate the ligand by a bounding sphere.}
+The most common tools detect the tunnels using Voronoi diagrams~\cite{yaffe2008,caver3} assuming a spherical ligand (probe), 
+i.e., they approximate the ligand by a bounding sphere.
 However, the ligands are typically of a non-spherical shape which is completely ignored by these methods.
@@
 Our approach aims to overcome this problem by taking into account the actual shape of the ligand.
-This necessary requires to consider also conformational changes.
+This necessarily requires to consider also conformational changes, i.e., the mutual position of the ligand atoms can change.
+Such a ligand is denoted as flexible ligand.
 %Our approach aims to overcome this problem by taking into account the actual shape of the ligand and also its conformations.
@@
 %The proposed solution needs also to cope with the narrow passage problem, as proteins are dense structures limiting the ligand movements inside.
-Such a ligand, changing the mutual position of its atoms, is denoted as a flexible ligand.
+%Such a ligand, for which the mutual position of its atoms can change, is denoted as a flexible ligand.
 
@@
 
-Finding trajectories only along a given tunnel tunnel is faster than computing any exit-pathway from the protein due to decreased
+Finding trajectories only along a given tunnel tunnel is faster than computing any exit pathway from the protein due to decreased
 volume of the configuration space to be searched.
@@
 More details can be found in the recently published state-of-the-art report about analysis and visualization of biomolecular cavities~\cite{Krone_2016}.
-The main disadvantage of both grid-based and VD-based methods is that the shape of the ligand is not taken into account during the tunnel detection and it is therefore not easy to estimate if (and how) a non-spherical ligand might traverse the tunnel.
+The main disadvantage of both grid-based and VD-based methods is that the shape of the ligand is not taken into account during the tunnel detection and it is therefore not easy to estimate if (and how) a non-spherical ligand can traverse the tunnel.
 The ligand flexibility cannot be principally considered by these approaches.
@@
 The traversability of a tunnel by a flexible ligand can be formulated as a motion planning problem and solved using sampling-based motion planners.
-These planners can cope with objects (robots) with many Degrees of Freedom (DOF)  and they can also consider arbitrary geometry.
+These planners can cope with objects (robots) with many Degrees of Freedom (DOF)  and they can also consider objects of arbitrary shape.
 Two widely used planners are Probabilistic Roadmaps (PRM)~\cite{kavrakiForPP} and Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT}.
@@
 RRT-ML expands the tree primarily using those DOF that are essential for achieving the motion of the ligand (i.e., rotation
-and translation) and it employs the other DOFs (i.e., those that are responsible for \textcolor{blue}{conformation/al?} changes) if they hinder the growth of the tree.
+and translation) and it employs the other DOFs (i.e., those that are responsible for conformational changes) if they hinder the growth of the tree.
 In~\cite{singhLIG}, the pathways for a flexible ligand are found using PRM.
@@
 
-In contrast to~\cite{cortes2010simulating}, which searches for any exit path from the protein, we localize the searching and compute the trajectories only along each tunnel separately which allows biochemists to analyze selected tunnels.
+In contrast to~\cite{cortes2010simulating}, which searches for any exit pathway from the protein, we localize the searching and compute the trajectories only along each tunnel separately which allows biochemists to analyze selected tunnels.
 %This is motivated by the practical needs of biochemists that are interested in the actual traversability the ligand through a selected tunnel.
 Computing the trajectories along a single tunnel brings several advantages from the motion planning point of view as well.
+The random samples are not generated in the whole configuration space, but only around the tunnel in a user-defined distance which helps to cope with the narrow passage problem.
+%The tunnels serve as a navigation path for the configuration tree~\cite{vonasek2009rrt}.
 By selecting a tunnel to be analyzed, the users help the planner to focus only on a subset of the whole configuration
@@
 Consequently, the volume of the configuration space to be searched is decreased, while the relative volume of the narrow passages is increased.
-The random samples are not generated in the whole configuration space, but only around the tunnel in a user-defined distance which helps to cope with the narrow passage problem.
-%The tunnels serve as a navigation path for the configuration tree~\cite{vonasek2009rrt}.
 
 Due to the protein dynamics, the tunnels move and merge, and their bottleneck can change in each time step.
-Therefore, it can be expected that the detected tunnels are rather narrow.
-Based on the observations of MD simulations, tunnels adapt their shape to the passing ligands and vice versa.
-In consequence, even very narrow tunnels, being too small for the passage of a given ligand-approximating sphere, can serve as the transportation path for this ligand.
+It can be expected that the tunnels in MD simulations without ligands are rather narrow.
+However, the tunnels adapt their shape to the passing ligands and vice versa.
+%Therefore, it can be expected that the detected tunnels are rather narrow.
+In consequence, even very narrow tunnels, being too small for the passage of a given ligand-approximating sphere, can serve as the transportation path for the ligand.
 Therefore, also the existing Voronoi-based methods enable the user to decrease the size of the approximating sphere which simulates shrinking of the ligand size.
@@
 $c_i \in \R^3$ is their 3D position, and $r_i > 0$ denotes the maximum collision-free radius of a sphere centered at $c_i$. 
-Tunnels serving as an input for our proposed method can be detected using some of the existing computational tools, such as CAVER 3.0~\cite{caver3}.
+Tunnels serving as an input for the proposed method can be detected using some of the existing computational tools, such as CAVER 3.0~\cite{caver3}.
 %Conformations of the ligand may be generated in the absence of the receptor and subsequently docked[13] 
@@
 Candidate conformations are prepared by discrete sampling of degrees of freedom of the ligand.
-Each candidate conformation is ranked according its energy and only low-energy conformation are used in $\L$.
+Each candidate conformation is ranked according its energy and only low-energy conformations are used in $\L$.
 %The conformations are prepared using the Rosetta tool and ranked according their potential energy. Only conformations
@@
 \linesnumbered
-\begin{algorithm}[bh]
+\begin{algorithm}[hb]
 {\small
@@
 rotation matrix $R$ given by $r_x$, $r_y$ and $r_z$ are computed.
-Then, $\qrand$ is translated to $\qrand' = R^{-1}(\qrand-V)$ and the KD-tree search is used to find the nearest point towards $\qrand'$, which
-is also the nearest point of the ligand at $q$ to $\qrand$.
+Then, $\qrand$ is translated to $\qrand' = R^{-1}(\qrand-V)$ and the KD-tree search is used to find the nearest atom from $q$ towards atoms of $\qrand'$.
+%is also the nearest point of the ligand at $q$ to $\qrand$.
 
@@
     The protein has 4650 atoms. 
+    The rate is shown in [\%].
 }
@@
     The protein has 4650 atoms.
+    The rate is shown in [\%].
 }
@@
 Based on the number of success rate, also other ligands (e.g., m037t with traversability rate 45~\% at $\smin$=0.8) seem to be promising.
-However, their binding was not approved yet by MD dedicated simulations.
+However, their binding was not approved yet by dedicated MD simulations.
 

### 2017-12-09 paper.tex
corrections

@@
 The tool~\cite{cortes2005path} uses the scale 0.8 for both the ligand and the protein.
-We tested also this case (the column \RD\ in Tab.~\ref{tab::main}).
+We tested also this case using \RA\ (the column \RD\ in Tab.~\ref{tab::main}).
 The results are same as for ligand scale $\smin=0.6$  (except DCP and m040 which can be a result of the randomization).
@@
 in the case of large scales and larger ligands (e.g., m056 with $\smin=0.8$: 15~\% with the metric, but to only 6~\% without it).
-Both \RA\ and \RC\ provided trajectories with the same traversability rate in the case of lower scales ($\smin<0.6$).
+%Both \RA\ and \RC\ provided trajectories with the same traversability rate in the case of lower scales ($\smin<0.6$).
 
@@
 According to the results of the motion planing, the DCP needs to shrink down to ca. 75~\% of its size (Fig.~\ref{fig::dcp}, point B).
-However, as both the protein and ligand are mutually influence each other, it is possible that even a very narrow part of the tunnel is traversable.
-This was observed in the MD simulation of the DCP ligand in the 1CQW protein, which revealed that DCP can pass the main tunnel~\cite{marques2017catalytic}.
+As both the protein and the ligand mutually influence each other, it is possible that even a narrow part of the tunnel is traversable.
+This was observed in the MD simulation of the DCP ligand in the 4E46 protein, which revealed that DCP can pass the main tunnel~\cite{marques2017catalytic}.
 
@@
 
-In this paper, we presented a novel modification of the RRT planner for computation of ligand trajectory inside protein tunnels.
-%So far, tunnels have been mostly detected using Voronoi diagrams which do not take into account the shape and flexibility of the ligand.
-In our solution, we compute the trajectories using the modified RRT which takes into account the ligand shape and we model the ligand flexibility using a predefined set of its conformations.
+We presented a novel modification of the RRT planner to compute trajectories of ligands inside selected protein tunnels.
+In our solution, we take into account the ligand shape and we model the ligand flexibility using a predefined set of its conformations.
 To simulate the real behavior of ligand passing through a tunnel, we enable scaling down the atomic radii of the ligands up to a certain predefined factor.
-Furthermore, our planner is localized as it generates random samples only along a tunnel computed by Voronoi diagram, and it attempts to expand the configuration
+The planner generates the random samples only along a tunnel computed by Voronoi diagram, and it attempts to expand the configuration
 tree using all ligand conformations. 

### 2017-12-09 paper.tex
corrections

@@
 We tested also this case using \RA\ (the column \RD\ in Tab.~\ref{tab::main}).
-The results are same as for ligand scale $\smin=0.6$  (except DCP and m040 which can be a result of the randomization).
+The results are close the results of \RA\ and $\smin=0.6$  (except DCP and m040 which can be a result of the randomization).
 
@@
 %that leads to a new collision-free configuration.
-The information about the scales used at the trajectory points % (which can be extracted from the configuration tree of the proposed planner) 
+The information about the scales used at the trajectory points 
 brings the biochemists new knowledge about the tunnels, which was impossible to get using the VD-based computations.
-The  visualization allows the biochemists to observe those parts of the tunnel where the ligand had to be scaled-down significantly (red parts of trajectories in Fig.~\ref{fig::dcp}), and parts where the ligand could be enlarged again (green parts in Fig.~\ref{fig::dcp}).
-According to the results of the motion planing, the DCP needs to shrink down to ca. 75~\% of its size (Fig.~\ref{fig::dcp}, point B).
-As both the protein and the ligand mutually influence each other, it is possible that even a narrow part of the tunnel is traversable.
-This was observed in the MD simulation of the DCP ligand in the 4E46 protein, which revealed that DCP can pass the main tunnel~\cite{marques2017catalytic}.
+By visualizing all trajectories computed for different scales, we can observe those parts of the tunnel where the ligand had to be scaled-down significantly (red parts of trajectories in Fig.~\ref{fig::dcp}), and parts where the ligand could be enlarged again (green parts in Fig.~\ref{fig::dcp}).
+According to the results of the motion planing, the DCP ligand needs to shrink down to ca. 75~\% of its size, but only at one part of the tunnel (Fig.~\ref{fig::dcp}, point B). 
+Beyond this point, the ligand can move at larger scales again.
+As both the protein and the ligand mutually influence each other, it is possible that even such a narrow part of the tunnel is traversable.
+This was observed in a separate MD simulation of the DCP ligand in the 4E46 protein, which revealed that DCP can pass the main tunnel~\cite{marques2017catalytic}.
 

### 2017-12-09 paper.tex
first 2 pages checked

@@
 Nowadays, tunnels are mostly computed using Voronoi diagrams which approximate the ligand by a bounding sphere.
-This leads to highly imprecise results.
-To address this problem, we define the computation of the transportation path of non-spherical ligands including their conformation changes (i.e., when the atoms of the ligand move relatively to each other) as a motion planning task in high-dimensional configuration space.
+This can lead to highly imprecise results.
+To address this problem, we define the computation of the transportation path of non-spherical ligands including their conformation changes (i.e., when the atoms of the ligand move relatively to each other) as a motion planning task in the high-dimensional configuration space.
 This can be solved using sampling-based planning methods.
-We propose a modification of the Rapidly Exploring Random Tree planner for the purpose of trajectory computations of flexible ligands.
-To cope with the narrow passage problem, the random samples are generated around a Voronoi-based tunnel.
+We propose a modification of the Rapidly Exploring Random Tree planner for the purpose of trajectory computations for flexible ligands.
+To cope with the narrow passage problem, the random samples are generated around a tunnel detected by a Voronoi-based approach.
 %which serves us for better localization of the searched space.
@@
 
-Reactivity of proteins with other molecules plays a crucial role in many research disciplines, including protein engineering and drug design.
+The reactivity of proteins with other molecules plays a crucial role in many research disciplines, including protein engineering and drug design.
 Protein reactivity is tightly connected with the presence of the void space between its amino acids. 
-This space, forming a path called tunnel, can be utilized by a small molecule (ligand) which can enter the protein inner space and react with its amino acids surrounding a specific site, called the active site~\cite{gora2013gates}.
-\textcolor{blue}{The size and location of tunnels determine the ligand size and shape which can enter the active site.}
-Currently, the biochemists use various computational methods for tunnel detection (Fig.~\ref{fig::motiv}) in order to estimate the possibility of interactions between a protein and a ligand.
-This helps them to reduce the number of necessary in-vitro lab experiments.
+This space, forming a path called a tunnel, can be utilized by a small molecule (ligand) which can enter the protein inner space and react with the amino acids surrounding the protein active site~\cite{gora2013gates} (see Fig.~\ref{fig::motiv}).
+The location and size of a tunnel determine the size and shape of a ligand passing through this tunnel to the active site.
+Currently, the biochemists can use various computational methods for tunnel detection in order to estimate the possibility of interactions between a protein and a ligand.
+This can help them to reduce the number of necessary in-vitro experiments.
 The most common tools detect the tunnels using Voronoi diagrams~\cite{yaffe2008,caver3} assuming a spherical ligand (probe), 
 i.e., they approximate the ligand by a bounding sphere.
-However, the ligands are typically of a non-spherical shape which is completely ignored by these methods.
+However, ligands are typically of a non-spherical shape which is completely ignored by these methods.
 
@@
 Our approach aims to overcome this problem by taking into account the actual shape of the ligand.
-This necessarily requires to consider also conformational changes, i.e., the mutual position of the ligand atoms can change.
+This necessarily requires to consider also conformational changes of the ligand, i.e., the mutual position of the ligand atoms can change.
 Such a ligand is denoted as flexible ligand.
@@
 
-The analysis of traversability of the tunnels by flexible ligands can be formulated as a motion planning problem and
-solved by sampling-based planners.
-Sampling-based planners randomly sample the configuration space of the robot (ligand) and connect the collision-free samples
-into a graph structure (roadmap).
+The analysis of the traversability of tunnels by flexible ligands can be formulated as a motion planning problem and solved by sampling-based planners.
+The sampling-based planners randomly sample the configuration space of the robot (ligand) and connect the collision-free samples into a graph structure (roadmap).
 Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} is one of the most used sampling-based planners.
@@
 e.g., to study loop motions~\cite{cortes2004geometric}, to identify binding sites~\cite{bayazit2001ligand}, for exit
-pathway detection~\cite{cortes2010simulating,cortes2005path} and especially for protein folding~\cite{raveh2009rapid,amato2002using}. %novinskaya2015improving
+pathway detection~\cite{cortes2010simulating,cortes2005path}, and especially for protein folding~\cite{raveh2009rapid,amato2002using}. %novinskaya2015improving
 
@@
 In this paper, we propose a modification of the RRT planner to find trajectories for non-spherical flexible ligands inside protein tunnels.
-The random samples are generated around a given tunnel instead in the whole configuration space.
+Random samples are generated only around a given tunnel instead of the whole configuration space.
 %We localize the generation of the random samples, i.e., instead of sampling the whole configuration space we focus only on the space surrounding a potential tunnel detected by the Voronoi-based method (using a smaller ligand probe in order to avoid omitting possibly feasible tunnel candidates).
@@
 %cortes2005path removed as it is not clear it they really rescale
-The ligand flexibility is modeled using a fixed set of predefined conformations which the ligand can occupy.
+The ligand flexibility is modeled using a fixed set of predefined conformations which can be taken by the ligand.
 
-Finding trajectories only along a given tunnel tunnel is faster than computing any exit pathway from the protein due to decreased
-volume of the configuration space to be searched.
-Another advantage of our method in comparison with the existing approaches is that we enable the users to trace the ligand path leading from the
-\textcolor{blue}{outside/outer?} solvent towards the protein active site.
+Finding trajectories only along a given tunnel is faster than computing all possible exit path, due to the decreased volume of the searched configuration space.
+Another advantage of our method, in comparison with the existing approaches, is that we enable the users to trace the ligand path leading from the
+outer solvent towards the protein active site.
 
 The proposed method was tested by the biochemists who compared the results with already known trajectories of ligands, determined by their traditional approaches.
-In the following section, we first mention several approaches and methods related to our solution and then we describe the proposed method in detail.
+In the following section, we first mention several approaches and methods related to our solution and then describe the proposed method in detail.
 
@@
 %grid-based approaches~\cite{sehnal2013mole,petrek2006caver}.
-The tunnels are represented as a sequence of spheres and characterized by their bottleneck, length, curvature, and a list of surrounding residues, whose physico-chemical properties are crucial for assessing the interaction possibilities between the protein and ligand.
-More details can be found in the recently published state-of-the-art report about analysis and visualization of biomolecular cavities~\cite{Krone_2016}.
+Tunnels are represented as a sequence of spheres and characterized by their bottleneck, length, curvature, and the list of surrounding residues, whose physico-chemical properties are crucial for assessing the protein-ligand interaction possibilities.
+More details can be found in the recently published state-of-the-art report about the analysis and visualization of biomolecular cavities~\cite{Krone_2016}.
 The main disadvantage of both grid-based and VD-based methods is that the shape of the ligand is not taken into account during the tunnel detection and it is therefore not easy to estimate if (and how) a non-spherical ligand can traverse the tunnel.
 The ligand flexibility cannot be principally considered by these approaches.
-Simulations based on molecular dynamics (MD) can directly evaluate the migration of a ligand into a protein, but they are significantly more computationally demanding~\cite{kingsley2014including}. 
-Moreover, MD simulations have to be prepared for a specific ligand/protein pair, which is not practical for testing a large number of ligands.
-The geometric-based methods are therefore still preferable due to their speed.
+Simulations based on molecular dynamics (MD) can directly evaluate the migration of the ligand into the protein, but they are significantly more computationally demanding~\cite{kingsley2014including}.
+Moreover, MD simulations have to be prepared for a specific ligand/protein pair, which is not practical for testing a large number of different ligands.
+The geometry-based methods are therefore still preferable due to their speed.
 
-The traversability of a tunnel by a flexible ligand can be formulated as a motion planning problem and solved using sampling-based motion planners.
-These planners can cope with objects (robots) with many Degrees of Freedom (DOF)  and they can also consider objects of arbitrary shape.
-Two widely used planners are Probabilistic Roadmaps (PRM)~\cite{kavrakiForPP} and Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT}.
+The traversability of a tunnel by a flexible ligand can be formulated as a motion planning problem and solved using the sampling-based motion planners.
+These planners can cope with objects (robots) with many Degrees of Freedom (DOF) and they can also consider objects of an arbitrary shape.
+Two widely used planners are the Probabilistic Roadmaps (PRM)~\cite{kavrakiForPP} and the Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT}.
 %Sampling-based planners like RRT can be used to find the trajectories even for flexible ligands, as they can cope with many-DOF systems and they can consider arbitrary shape.
@@
 The well known issue of the sampling-based planners is the narrow passage problem.
-Narrow passage is a small region in the configuration space, whose removal changes the connectivity of the space.
+The narrow passage is a small region in the configuration space, whose removal changes the connectivity of the space.
 Due to its low volume, the probability of sampling the narrow passage is low.
 Consequently, many iterations are needed in order to put enough samples into the passages, which increases the computation time.
-PRM-based planners can cope with the narrow passage problem by increasing the probability of sampling in difficult regions, e.g.
-by sampling along medial axis~\cite{amatoOBRRT,amato2002using,wilmarthMAPRM}.
-In~\cite{bergWIG}, workspace is represented by a grid in which difficult regions are identified using watershed labeling.
+The PRM-based planners can cope with the narrow passage problem by increasing the probability of sampling in difficult regions, e.g., by sampling along the medial axis~\cite{amatoOBRRT,amato2002using,wilmarthMAPRM}.
+In~\cite{bergWIG}, the workspace is represented by a grid in which the difficult regions are identified using watershed labeling.
 The regions can be estimated online, e.g., based on the number of collision-free samples around a given configuration~\cite{overmarsGauss,hsuBridge} or even identified by a human operator~\cite{denny2018general}.
@@
 
-RRT-based planners cannot cope with the narrow passage problem simply by increasing the probability of sampling in the difficult regions, as the
-growth of the configuration tree towards the random samples may be blocked by obstacles. %~\cite{vonasekphd}.
-To prevent this blocking, DD-RRT~\cite{yershovaDDRRT} limits the selection of nodes for expansion to a small ball. 
-The radius of the ball is set to infinity for new nodes, and decreases to a predefined radius if the node cannot be successfully expanded.
-Automatic adaptation of this radius was proposed in ADD-RRT~\cite{jailletADRRT}.
+The RRT-based planners cannot cope with the narrow passage problem simply by increasing the probability of sampling in the difficult regions, as the growth of the configuration tree towards the random samples may be blocked by obstacles.
+%~\cite{vonasekphd}.
+To prevent this blocking, the DD-RRT~\cite{yershovaDDRRT} limits the selection of nodes for the expansion to a small ball. 
+The radius of this ball is set to infinity for new nodes, and decreases to a predefined radius if the node cannot be successfully expanded.
+An automatic adaptation of this radius was proposed in the ADD-RRT~\cite{jailletADRRT}.
 %The disadvantage of DD-RRT is the sensitivity to the predefined radius.
@@
 To attract the tree towards a given region, random samples have to be generated there only if the tree can expand to this region.
-In~\cite{kardossRRTKK}, the probability of sampling is increased in several waypoints close to narrow passages, but without specifying
-how to find these waypoints.
-The generalization of~\cite{kardossRRTKK} is the guided sampling, where a whole path is used
-to generate samples in the configuration space.
+In~\cite{kardossRRTKK}, the probability of sampling is increased in several waypoints close to narrow passages, but without specifying how to find these waypoints.
+The generalization of~\cite{kardossRRTKK} is the guided sampling, where the whole path is used to generate samples in the configuration space.
 The path can be computed in the workspace~\cite{vonasek2009rrt}, or iteratively refined in the configuration space based on the solution of a relaxed version of the problem~\cite{bayazitIRC}.
@@
 %The task of the retraction step is to find a contact configuration around $\qnear$ that minimizes the distance to $\qrand$.
-In Retraction-based RRT~\cite{zhangRetraction}, the tree is retracted along the boundary of the obstacles in the configuration space.
+In the Retraction-based RRT~\cite{zhangRetraction}, the tree is retracted along the boundary of the obstacles in the configuration space.
 This requires the analysis of the contact space, which leads to the generalized penetration depth computation of high complexity~\cite{he2016efficient}.
@@
 + zhang2008fast
-Selective Retraction-based RRT~\cite{lee2012srrrt} avoids the expensive growth in the open areas of the configuration space and focuses 
-the retractions only on the narrow passages.
-Obstacle-based RRT~\cite{amatoOBRRT} utilizes several expansion procedures based on, e.g., random vectors, obstacle vectors, or medial axis.
-In MARRT~\cite{denny2014marrt}, the newly generated samples are pushed towards the medial axis.
+The Selective Retraction-based RRT~\cite{lee2012srrrt} avoids the expensive growth in the open areas of the configuration space and focuses the retractions only on the narrow passages.
+The Obstacle-based RRT~\cite{amatoOBRRT} utilizes several expansion procedures based on, e.g., random vectors, obstacle vectors, or medial axis.
+In MARRT~\cite{denny2014marrt}, newly generated samples are pushed towards the medial axis.
 %The operator can also select a suitable planner for the region.<F12>
@@
 
-The probability of sampling of the narrow passages can be increased by dilating the free-space, e.g., by shrinking the geometry of 
-the robot~\cite{hsuOnProb} or the obstacles~\cite{bayazitIRC}.
-The shrinking technique has been used also in~\cite{cortes2010simulating}, where the exit pathways for a small flexible molecule are
-computed using a RRT-based method.
+The probability of sampling of the narrow passages can be increased by dilating the free-space, e.g., by shrinking the geometry of the robot~\cite{hsuOnProb} or the obstacles~\cite{bayazitIRC}.
+The shrinking technique has been used also in~\cite{cortes2010simulating}, where the exit pathways for a small flexible molecule are computed using a RRT-based method.
 The flexible ligand is modeled as a kinematic chain, which increases the dimension of the configuration space.
 To cope with these many-DOF ligands, the RRT-ML~\cite{cortes2007mlrrt} approach is employed as the basic planner in~\cite{cortes2010simulating}.
-RRT-ML expands the tree primarily using those DOF that are essential for achieving the motion of the ligand (i.e., rotation
-and translation) and it employs the other DOFs (i.e., those that are responsible for conformational changes) if they hinder the growth of the tree.
+RRT-ML expands the tree primarily using those DOF, that are essential for achieving the motion of the ligand (i.e., rotation
+and translation) and it employs other DOFs (i.e., those that are responsible for conformational changes) if they hinder the growth of the tree.
 In~\cite{singhLIG}, the pathways for a flexible ligand are found using PRM.
-Each configuration in the roadmap is also evaluated using electrostatic energy.
-The high-energy configurations are discarded, while low-energy conformations (and their vicinity) are sampled more dense.
-The approach~\cite{singhLIG} can be also used to predict ligand binding sites.
+Each configuration in the roadmap is also evaluated using the electrostatic energy.
+The high-energy configurations are discarded, while the low-energy conformations (and their vicinity) are sampled densely.
+The approach~\cite{singhLIG} can be also used to predict the ligand binding sites.
 

### 2017-12-09 paper.tex
CL

@@
 The ligand is modeled using the set $\L$ of its conformations.
-Candidate conformations are prepared by discrete sampling of degrees of freedom of the ligand.
-Each candidate conformation is ranked according its energy and only low-energy conformations are used in $\L$.
+The candidate conformations are prepared by discrete sampling of degrees of freedom of the ligand.
+Each candidate conformation is ranked according to its energy and only low-energy conformations are used in $\L$.
 %The conformations are prepared using the Rosetta tool and ranked according their potential energy. Only conformations
@@
 %Similarly, a single goal configuration $\qgoal$ is found around the end of the tunnel. 
-The distance $\RI$ influences how far are the initial configurations from the start of the tunnel.
+The distance $\RI$ influences how far the initial configurations are from the start of the tunnel.
 By placing them too far, the planner would need more iterations to enter the tunnel, which would increase the runtime.
@@
 
-Movements of the ligands inside a tunnel are blocked by many atoms forming the tunnel.
+The movements of the ligands inside a tunnel are blocked by many atoms forming the tunnel.
 It is not always possible for the ligand to strictly follow the centerline of the tunnel and slight detours
-should be allowed, which is controlled by the parameter $\dt$.
-This parameter should be set considering existence of cavities or other tunnels around the tunnel being investigated.
-In the case of alone tunnels (which is very common), we recommend to use double of the tunnel average width.
+should be allowed, which is controlled by the $\dt$ parameter.
+This parameter should be set considering the existence of cavities or other tunnels around the tunnel being investigated.
+In the case of alone tunnels (which is very common), we recommend to use double value of the tunnel average width.
 
@@
 It allows us to expand the tree by a configuration whose nearest atom approaches its nearest atom in $\qrand$, which supports
-retraction of the ligands towards $\qrand$.
+the retraction of the ligands towards $\qrand$.
 
@@
 
-The pool of available conformations $\L$ consists all conformations (upto some resolution) with energy lower than a user specified threshold, which can still contain too many conformations (thousands or even more).
+The pool of available conformations $\L$ consists all conformations (up to some resolution) with energy lower than a user specified threshold, which can still contain too many conformations (thousands or even more).
 We select a subset by computing how the atoms of the ligand are spread around its geometric mean.
@@
 
-The proposed method was used to compute ligand traversability towards the active site using two main tunnels in the Haloalkane dehalogenase protein (PDB ID 4E46) which consists of 4650 atoms.
+The proposed method was used to compute the ligand traversability towards the active site using two main tunnels in the Haloalkane dehalogenase protein (PDB ID 4E46) which consists of 4650 atoms.
 The active site was defined using residues 38 (Asparagine), 102 (Histidine), and 103 (Aspartic Acid).
@@
 A frame is considered traversable if at least one trajectory out of 200 ($\!|\QI|\!\times\!\!10$ trials) approaches the active site to the distance less than~2~\AA. 
-This distance is measured as 3D distance between geometric center of the ligand and the position of the active site.
+This distance is measured as 3D distance between the geometric center of the ligand and the position of the active site.
 %~\footnote{We refer to~\url{http://mrs.felk.cvut.cz/icra2018protein} for additional material.}
@@
 The aim of this experiment was to verify the procedure to select feasible ligand conformations from an available pool of conformations.
-Models  of the ligands in Mol2 formats were converted to param files~\cite{meiler2006rosettaligand} containing the topology, rotatable bonds, atom types and partial charges. 
-The pool of conformations $\L$ was created by rotation of all rotatable bonds with 30$^\circ$ step using Rosetta3.
+Models  of the ligands in Mol2 formats were converted to param files~\cite{meiler2006rosettaligand}, containing the topology, rotatable bonds, atom types, and partial charges. 
+The pool of conformations $\L$ was created by the rotation of all rotatable bonds with 30$^\circ$ step using Rosetta3.
 The energy cut-off for individual conformations was set to 0.075. %(-rot_ensemble_ecutoff 0.075) 

### 2017-12-09 paper.tex
final check

@@
 
-In contrast to~\cite{cortes2010simulating}, which searches for any exit pathway from the protein, we localize the searching and compute the trajectories only along each tunnel separately which allows biochemists to analyze selected tunnels.
+In contrast to~\cite{cortes2010simulating}, which searches for any exit pathway from the protein, we localize the searching and compute the trajectories only along each tunnel separately which allows the biochemists to analyze selected tunnels.
 %This is motivated by the practical needs of biochemists that are interested in the actual traversability the ligand through a selected tunnel.
@@
 In consequence, even very narrow tunnels, being too small for the passage of a given ligand-approximating sphere, can serve as the transportation path for the ligand.
-Therefore, also the existing Voronoi-based methods enable the user to decrease the size of the approximating sphere which simulates shrinking of the ligand size.
+Therefore, also the existing Voronoi-based methods enable the user to decrease the size of the approximating sphere which simulates the shrinking of the ligand size.
 In our solution we simulate this by shrinking the atom radii of the ligand, used also by~\cite{cortes2010simulating,guieysse2008structure}.
@@
 However, ligands cannot form arbitrary conformations. 
-Ligand can contain several torsion and dihedral angles between its atoms which determine possible ligand conformations.
+The ligand can contain several torsion and dihedral angles between its atoms which determine possible ligand conformations.
 Therefore, the ligand flexibility can be modeled using only a predefined set of conformations.
 By using the set of predefined conformations, it is not necessary to model the additional degrees of freedom, so 
-the dimension of the conifguration space is not increased as in~\cite{cortes2010simulating}.
+the dimension of the configuration space is not increased as in~\cite{cortes2010simulating}.
 %So the number of dimensions of the configuration space does not depend on the DOF of the ligand as in~\cite{cortes2010simulating}.
@@
 Selection of a suitable set of conformations $\L$ is therefore crucial.
-Depending on the ligand, some conformations can be considered as more packed i.e., the atoms in such conformations are closer to each other, but 
+Depending on the ligand, some conformations can be considered as more packed, i.e., the atoms in such conformations are closer to each other, but 
 the atoms can also form a more elongated shape (Tab.~\ref{fig::m040c}).
@@
 To simulate the real behavior of ligand passing through a tunnel, we enable scaling down the atomic radii of the ligands up to a certain predefined factor.
-The planner generates the random samples only along a tunnel computed by Voronoi diagram, and it attempts to expand the configuration
+The planner generates the random samples only along a tunnel computed by the Voronoi diagram, and it attempts to expand the configuration
 tree using all ligand conformations. 

### 2018-02-22 paper.tex
FINAL VERSION
related work

@@
     Adam Jur\v{c}\'\i k$^{2}$,
-    Ond\v{r}ej V\'{a}vra$^{3}$,
-    Martin Saska$^{1}$
+    Ond\v{r}ej V\'{a}vra$^{3}$
+%    Martin Saska$^{1}$
 \thanks{$^{1}$ Faculty of Electrical Engineering,  Czech Technical University in Prague, 
@@
 \def\da{d_{atom}}
+\def\dist{d}
 \def\R{\mathbb{R}}
@@
 These active sites are accessible through one or more void paths, called tunnels.
-Nowadays, tunnels are mostly computed using Voronoi diagrams which approximate the ligand by a bounding sphere.
+Nowadays, tunnels are mostly computed using Voronoi diagrams which approximate the ligand by a bounding sphere and 
+the traversability of the tunnels is determined based only on the radial bottlenecks.
 This can lead to highly imprecise results.
 To address this problem, we define the computation of the transportation path of non-spherical ligands including their conformation changes (i.e., when the atoms of the ligand move relatively to each other) as a motion planning task in the high-dimensional configuration space.
-This can be solved using sampling-based planning methods.
 We propose a modification of the Rapidly Exploring Random Tree planner for the purpose of trajectory computations for flexible ligands.
@@
 \section{Related Work}
+{\color{red}
+Migration of the ligands into a protein can be directly evaluated using Molecular dynamics (MD) simulations, that consider both 
+the protein and the ligand flexible.
+However, MD simulations are considerably computational demanding~\cite{kingsley2014including} and they have to be prepared
+for particular ligand/protein pair, which is not practical for testing a large number of different ligands.
+Therefore, computational approaches that consider only geometry of the protein/ligand are still preferable due to their high speed.
 
+The widely used approaches to evaluate possiblity of ligand migration into the protein rely on the tunnel detection.
 Most of the available tools for tunnel detection are based on Voronoi diagrams~\cite{yaffe2008,caver3} or they utilize a grid-based computation~\cite{sehnal2013mole}.
-%The tunnels are pathways computed for a single atom (spherical probe) using Voronio-diagrams~\cite{yaffe2008,caver3} or 
-%grid-based approaches~\cite{sehnal2013mole,petrek2006caver}.
 Tunnels are represented as a sequence of spheres and characterized by their bottleneck, length, curvature, and the list of surrounding residues, whose physico-chemical properties are crucial for assessing the protein-ligand interaction possibilities.
-More details can be found in the recently published state-of-the-art report about the analysis and visualization of biomolecular cavities~\cite{Krone_2016}.
-The main disadvantage of both grid-based and VD-based methods is that the shape of the ligand is not taken into account during the tunnel detection and it is therefore not easy to estimate if (and how) a non-spherical ligand can traverse the tunnel.
-The ligand flexibility cannot be principally considered by these approaches.
-Simulations based on molecular dynamics (MD) can directly evaluate the migration of the ligand into the protein, but they are significantly more computationally demanding~\cite{kingsley2014including}.
-Moreover, MD simulations have to be prepared for a specific ligand/protein pair, which is not practical for testing a large number of different ligands.
-The geometry-based methods are therefore still preferable due to their speed.
+The tunnels are detected either on a single conformation of the protein, or, on a set of protein conformations.
+The protein conformations can be prepared again using MD simulations.
+Computationally less costly approaches for generating protein conformations are based on constrained geometric simulations like
+ROCK~\cite{lei2004sampling}, FRODA~\cite{wells2005constrained} and tCONCOORD~\cite{seeliger2007geometry}.
+Examples of utilization of protein tunnels can be found in the recently published state-of-the-art report about the analysis and visualization of biomolecular cavities~\cite{Krone_2016}.
+
+The main disadvantage of both grid-based and VD-based tunnel detection methods is that the shape of the ligand is not taken into account during the tunnel detection and it is therefore not easy to estimate if (and how) a non-spherical ligand can traverse the tunnel.
+The ligand flexibility cannot be principally considered by Voronoi-based approaches.
+}
+
+%Simulations based on molecular dynamics (MD) can directly evaluate the migration of the ligand into the protein, but they are significantly more computationally demanding~\cite{kingsley2014including}.
+%Moreover, MD simulations have to be prepared for a specific ligand/protein pair, which is not practical for testing a large number of different ligands.
+%The geometry-based methods are therefore still preferable due to their speed.
 
@@
 
-The probability of sampling of the narrow passages can be increased by dilating the free-space, e.g., by shrinking the geometry of the robot~\cite{hsuOnProb} or the obstacles~\cite{bayazitIRC}.
+%The probability of sampling of the narrow passages can be increased by dilating the free-space, e.g., by shrinking the geometry of the robot~\cite{hsuOnProb} or the obstacles~\cite{bayazitIRC}.
+The probability of sampling of the narrow passages can be increased by dilating the free-space, e.g., by shrinking the geometry of the robot or the obstacles~\cite{bayazitIRC}.
+
 The shrinking technique has been used also in~\cite{cortes2010simulating}, where the exit pathways for a small flexible molecule are computed using a RRT-based method.
@@
 The high-energy configurations are discarded, while the low-energy conformations (and their vicinity) are sampled densely.
-The approach~\cite{singhLIG} can be also used to predict the ligand binding sites.
+The approach~\cite{singhLIG} can also predict the ligand binding sites.
 
@@
 
-While the classic 6D Euclidean metric is used in the nearest-neighbor search of RRT, we propose to use another metric during the expansion step.
-This metric measures the distances between two configurations as the nearest distance of two atoms. 
-This enables the metric to consider the shape of the ligand and it supports the retraction of the ligand to the narrow passages.
+%While the classic 6D Euclidean metric is used in the nearest-neighbor search of RRT, we propose to use another metric during the expansion step.
+%This metric measures the distances between two configurations as the nearest distance of two atoms. 
+%This enables the metric to consider the shape of the ligand and it supports the retraction of the ligand to the narrow passages.
 
@@
 both 3D rotation and 3D translation.
-The main loop of the planner (Alg.~\ref{alg::main}) is similar to the original RRT, but with the following three proposed modifications.
+The main loop of the planner (Alg.~\ref{alg::main}) is similar to the original RRT, but with the following two proposed modifications.
 
@@
 \linesnumbered
-\begin{algorithm}[hb]
+\begin{algorithm}[b]
 {\small
@@
 
-The third modification is the utilization of another metric during the expansion procedure (line~\ref{alg::expand:a} in Alg.~\ref{alg::expand}).
-This metric is denoted as $\da(\cdot,\cdot)$ and it is referred to as atom-based metric in the rest of the paper.
-To evaluate $\da(q_1,q_2)$, the absolute positions of the atoms of the ligand defined by $q_1$ and $q_2$ are computed.
-The minimal 3D Euclidean distance from the first set of atoms (defined by $q_1$) to the second set of atoms (defined by $q_2$) is the value of 
-$\da(q_1,q_2)$.
-This metric is sensitive to the shape of the ligand (that depends on its conformation) as it computes the distance using the two nearest atoms of the two configurations.
-It allows us to expand the tree by a configuration whose nearest atom approaches its nearest atom in $\qrand$, which supports
-the retraction of the ligands towards $\qrand$.
+%The third modification is the utilization of another metric during the expansion procedure (line~\ref{alg::expand:a} in Alg.~\ref{alg::expand}).
+%This metric is denoted as $\da(\cdot,\cdot)$ and it is referred to as atom-based metric in the rest of the paper.
+%To evaluate $\da(q_1,q_2)$, the absolute positions of the atoms of the ligand defined by $q_1$ and $q_2$ are computed.
+%The minimal 3D Euclidean distance from the first set of atoms (defined by $q_1$) to the second set of atoms (defined by $q_2$) is the value of 
+%$\da(q_1,q_2)$.
+%This metric is sensitive to the shape of the ligand (that depends on its conformation) as it computes the distance using the two nearest atoms of the two configurations.
+%It allows us to expand the tree by a configuration whose nearest atom approaches its nearest atom in $\qrand$, which supports
+%the retraction of the ligands towards $\qrand$.
 
@@
 
-The atom-based metric can be evaluated fast using KD-trees with complexity $\mathcal{O}(\log n)$, where $n$ is the number of atoms of the ligand.
-For each conformation $\l \in \L$ of the ligand, a KD-tree is built for the ligand at the configuration $q=(0,0,0,0,0,0,l,\smax)$
-(i.e., for ligand with no translation and no rotation).
-To find the nearest atom to $\qrand$ for a ligand at the configuration $q=(x,y,z,r_x,r_y,r_z,l,s)$, the translation vector $V=(x,y,z)$ and
-rotation matrix $R$ given by $r_x$, $r_y$ and $r_z$ are computed.
-Then, $\qrand$ is translated to $\qrand' = R^{-1}(\qrand-V)$ and the KD-tree search is used to find the nearest atom from $q$ towards atoms of $\qrand'$.
-%is also the nearest point of the ligand at $q$ to $\qrand$.
+%The atom-based metric can be evaluated fast using KD-trees with complexity $\mathcal{O}(\log n)$, where $n$ is the number of atoms of the ligand.
+%For each conformation $\l \in \L$ of the ligand, a KD-tree is built for the ligand at the configuration $q=(0,0,0,0,0,0,l,\smax)$
+%(i.e., for ligand with no translation and no rotation).
+%To find the nearest atom to $\qrand$ for a ligand at the configuration $q=(x,y,z,r_x,r_y,r_z,l,s)$, the translation vector $V=(x,y,z)$ and
+%rotation matrix $R$ given by $r_x$, $r_y$ and $r_z$ are computed.
+%Then, $\qrand$ is translated to $\qrand' = R^{-1}(\qrand-V)$ and the KD-tree search is used to find the nearest atom from $q$ towards atoms of $\qrand'$.
+%
 
-\begin{algorithm}[th]
+\begin{algorithm}[b]
 {\small
@@
             \If{isCollisionFree($q$)}{
-                \If{$\qnew = \emptyset$ {\bf or} $\da(q, \qrand) < \da(\qnew,\qrand)$}{ \nllabel{alg::expand:a}
+                \If{$\qnew = \emptyset$ {\bf or} $\dist(q, \qrand) < \dist(\qnew,\qrand)$}{ \nllabel{alg::expand:a}
                     $\qnew = q$\;
@@
 This metric was selected based on preliminary experiments with both planners as it resulted in the highest success rates.
-The \RA\ method was also tested without the atom-based metric.
-In this case (denoted \RC), the distance in the expansion step is measured using the 6D Euclidean metric.
+%The \RA\ method was also tested without the atom-based metric.
+%In this case (denoted \RC), the distance in the expansion step is measured using the 6D Euclidean metric.
 
@@
 %that leads to a new collision-free configuration.
+
 The information about the scales used at the trajectory points 
@@
 \centering
-\includegraphics[width=0.4\textwidth]{fig/dcp-image1Label}\\
+\includegraphics[width=0.35\textwidth]{fig/dcp-image1Label}\\
 %\includegraphics[width=0.2\textwidth]{fig/colormap}
+\vspace{-3mm}
 \caption{\label{fig::dcp}

### 2018-02-23 paper.tex
tables updated, experiments reorganized

@@
     Ond\v{r}ej V\'{a}vra$^{3}$
-%    Martin Saska$^{1}$
+    Martin Saska$^{1}$
 \thanks{$^{1}$ Faculty of Electrical Engineering,  Czech Technical University in Prague, 
@@
 %which serves us for better localization of the searched space.
-The flexibility of the ligand is modeled using a fixed set of known conformations, which does not increase the number of dimensions of the configuration space.
+The flexibility of the ligand is modeled using a fixed set of known conformations. 
+%which does not increase the number of dimensions of the configuration space.
 The proposed method has been evaluated in a virtual screening scenario to estimate the traversability of tunnels for several ligands.
@@
 \section{Proposed Method}
-%To address the problems of the existing solutions for computing the non-spherical flexible ligand trajectories through protein tunnels, we propose a modification of the RRT planner. 
 We propose a modification of RRT to compute trajectories for non-spherical flexible ligands through protein tunnels.
-%Our approach takes into account the shape and flexibility of the ligand and localizes the search around given tunnels. configuration space to 
-%using the information about tunnels detected by the Voronoi-based approach.
-The input is formed by a set of tunnels computed using Voronoi diagrams and the task is to find trajectories for a ligand along these tunnels
-leading from outside the protein to the active site.
-The tunnels are detected in a sequence of time frames of MD simulations.
+The input of the proposed method is a set of tunnels computed using Voronoi diagrams, and the task is to find trajectories for a ligand along these tunnels leading from outside the protein to the active site.
+The tunnels are detected on selected protein conformations resulting e.g. from MD simulations or from tool providing conformation sampling
+~\cite{lei2004sampling,wells2005constrained,seeliger2007geometry}.
+%The tunnels are detected in a sequence of time frames of MD simulations.
 %capturing protein dynamic behavior. 
@@
 %So the number of dimensions of the configuration space does not depend on the DOF of the ligand as in~\cite{cortes2010simulating}.
-The libraries containing possible ligand conformations are used also in different types of calculations, e.g., in~\cite{kellogg}.
+%The libraries containing possible ligand conformations are used also in different types of calculations, e.g., in~\cite{kellogg}.
 
@@
 A configuration of the ligand $q=(x,y,z,r_x,r_y,r_z,l,s)$ is described
-by the 3D position $(x,y,z)$ of the reference point of the ligand (its geometric center), the rotation around $x$, $y$, and $z$ axes,
+by the 3D position $(x,y,z)$ of the reference point of the ligand (e.g. the geometric center or a selected atom), the rotation around $x$, $y$, and $z$ axes,
 the index of the conformation $l\in \L$, and the scale $s \in \S$.
@@
 Setting the virtual goal to this successor allows the tree to avoid such parts of the tunnels that are not traversable or reachable by the ligand.
-This allows the tree to slightly detour from the tunnel and overcome its difficult (i.e., narrow) parts by finding an alternative trajectory nearby.
+The tree can slightly detour from the tunnel and overcome difficult parts (e.g., the tunnel bottleneck) by finding an alternative trajectory nearby.
+%It is not always possible for the ligand to strictly follow the centerline of the tunnel and slight detours
+%should be allowed, which is controlled by the $\dt$ parameter.
+%parameter should be set considering the existence of cavities or other tunnels around the tunnel being investigated.
+The parameter $\dt$ determines how far can the tree detour from the tunnel centerline.
+In the case of alone tunnels (which is very common), we recommend to use double value of the tunnel average width.
 The algorithm terminates after a predefined number of planning trials $\Imax$ or if the tree reaches
@@
 
-The movements of the ligands inside a tunnel are blocked by many atoms forming the tunnel.
-It is not always possible for the ligand to strictly follow the centerline of the tunnel and slight detours
-should be allowed, which is controlled by the $\dt$ parameter.
-This parameter should be set considering the existence of cavities or other tunnels around the tunnel being investigated.
-In the case of alone tunnels (which is very common), we recommend to use double value of the tunnel average width.
+%The movements of the ligands inside a tunnel are blocked by many atoms forming the tunnel.
 
@@
 
-The proposed method was used to compute the ligand traversability towards the active site using two main tunnels in the Haloalkane dehalogenase protein (PDB ID 4E46) which consists of 4650 atoms.
-The active site was defined using residues 38 (Asparagine), 102 (Histidine), and 103 (Aspartic Acid).
+The proposed algorithm was verified in two scenarios: on a sequence frames of MD simulation and on a set of protein conformations provided by the tCONCOORD tool~\cite{seeliger2007geometry}.
+
+\subsection{Analysis of MD simulation}
+
+\subsubsection{Algorithm and data setup}
+
+The task is to compute the ligand traversability towards the active site using two main tunnels in the Haloalkane dehalogenase protein (DhaAwt, PDB ID 4E46) which consists of 4650 atoms.
 MD of the protein was computed using AMBER~12 (details about the used MD simulations of DhaAwt/4E46 are described in~\cite{marques2017catalytic}).
-In each frame, tunnels for the spherical probe 0.9~\AA\ were detected using CAVER 3.0~\cite{caver3}. 
+The tunnels were detected in 100 consecutive frames of MD simulation in part where the protein was stabilized.
+The tunnels for probe 0.9~\AA\ towards the active site defined by residues 38 (Asparagine), 102 (Histidine), and 103 (Aspartic Acid) were
+computed using the CAVER 3.0~\cite{caver3} tool.
 In most of the frames, two tunnels were detected.
 The average bottleneck sizes of these tunnels are $1.48$~\AA\ and $1.2$~\AA\ for the main and the side tunnel, respectively.
-The tunnel traversability was analyzed in 100 consecutive frames of MD simulation in part where the protein was stabilized.
+
+Models  of the ligands in Mol2 formats were converted to param files~\cite{meiler2006rosettaligand}, containing the topology, rotatable bonds, atom types, and partial charges. 
+The pool of conformations $\L$ was created by the rotation of all rotatable bonds with 30$^\circ$ step using Rosetta3.
+The energy cut-off for individual conformations was set to 0.075. %(-rot_ensemble_ecutoff 0.075)
 
@@
 
-\subsection{Traversability of One Atom}
-
-The performance of \RA\ and \RB\ planners was first measured by finding trajectories for only one atom of VDW radii $1.7$~\AA.
-The \RC\ variant was not tested here as it behaves same as \RA\ in the case of one-atom ligand.
-The computed traversability rates are shown in Tab.~\ref{tab::one}.
-The maximum percentage of traversable frames (the row Maximum in Tab.~\ref{tab::one}) is computed as the percentage of tunnels with the bottleneck larger than the tested radius.
-The row 'Planning' shows the traversable rate computed by the planners.
-This rate decreases with the increasing radius and it is lower than the maximum traversable rate.
-The rates provided by \RA\ and \RB\ are similar which is a good indicator that the methods have been implemented correctly.
-Higher traversability rate can be achieved by increasing the maximum number of planning iterations $\Imax$, which would also
-increase the planning time.
-For the purpose of the following experiments, we used $\Imax=5,000$ as a compromise between the performance and the runtime.
-
-\begin{table}[t]
-\centering
-\caption{\label{tab::one}
-    \small
-    Traversability rate [\%] of one atom.
-}
-\renewcommand{\tabcolsep}{4.3pt}
-{\scriptsize
-\renewcommand{\arraystretch}{0.7}
-\input graphProbe.tex
-}
-\end{table}
-
-%\begin{figure}
+%\subsection{Traversability of One Atom}
+%
+%The performance of \RA\ and \RB\ planners was first measured by finding trajectories for only one atom of VDW radii $1.7$~\AA.
+%The \RC\ variant was not tested here as it behaves same as \RA\ in the case of one-atom ligand.
+%The computed traversability rates are shown in Tab.~\ref{tab::one}.
+%The maximum percentage of traversable frames (the row Maximum in Tab.~\ref{tab::one}) is computed as the percentage of tunnels with the bottleneck larger than the tested radius.
+%The row 'Planning' shows the traversable rate computed by the planners.
+%This rate decreases with the increasing radius and it is lower than the maximum traversable rate.
+%The rates provided by \RA\ and \RB\ are similar which is a good indicator that the methods have been implemented correctly.
+%Higher traversability rate can be achieved by increasing the maximum number of planning iterations $\Imax$, which would also
+%increase the planning time.
+%For the purpose of the following experiments, we used $\Imax=5,000$ as a compromise between the performance and the runtime.
+%
+%\begin{table}[t]
 %\centering
-%\begin{tabular}{cc}
-%\includegraphics[width=0.3\textwidth]{fig/t05proteintunnels} &
-%\includegraphics[width=0.45\textwidth]{fig/ligAll}  \\
-%(a) & (b)                        
-%\end{tabular}                       
-%\caption{\label{fig::tunnel}
-%    (a) The protein 1CQW visualized using the cartoon representation (gray) with 
-%        three tunnels (red, green, blue) detected by CAVER 3.0.
-%        The first tunnel is depicted red.
-%%        its highest ranked tunnel (green) which was detected using CAVER 3.0.
-%    (b) Examples of three conformations of $\LA$ (top) and $\LB$ (bottom)
+%\caption{\label{tab::one}
+%    \small
+%    Traversability rate [\%] of one atom.
 %}
-%\end{figure}
+%\renewcommand{\tabcolsep}{4.3pt}
+%{\scriptsize
+%\renewcommand{\arraystretch}{0.7}
+%\input graphProbe.tex
+%}
+%\end{table}
+%
+%
+%
+%%\begin{figure}
+%%\centering
+%%\begin{tabular}{cc}
+%%\includegraphics[width=0.3\textwidth]{fig/t05proteintunnels} &
+%%\includegraphics[width=0.45\textwidth]{fig/ligAll}  \\
+%%(a) & (b)                        
+%%\end{tabular}                       
+%%\caption{\label{fig::tunnel}
+%%    (a) The protein 1CQW visualized using the cartoon representation (gray) with 
+%%        three tunnels (red, green, blue) detected by CAVER 3.0.
+%%        The first tunnel is depicted red.
+%%%        its highest ranked tunnel (green) which was detected using CAVER 3.0.
+%%    (b) Examples of three conformations of $\LA$ (top) and $\LB$ (bottom)
+%%}
+%%\end{figure}
+%%
 %
 
-\subsection{Influence of the Conformation Selection}
+\subsubsection{Influence of the Conformation Selection}
 
-The aim of this experiment was to verify the procedure to select feasible ligand conformations from an available pool of conformations.
-Models  of the ligands in Mol2 formats were converted to param files~\cite{meiler2006rosettaligand}, containing the topology, rotatable bonds, atom types, and partial charges. 
-The pool of conformations $\L$ was created by the rotation of all rotatable bonds with 30$^\circ$ step using Rosetta3.
-The energy cut-off for individual conformations was set to 0.075. %(-rot_ensemble_ecutoff 0.075) 
+The number of generated ligand conformations $|\L|$ depends on the number of rotatable bonds of the ligands.
+Although only the low-energy conformations are generated and saved to $\L$, the number of all conformations can still be large for
+ligands with many rotatable bonds.
+The aim of this experiment was to verify the procedure to select a subset of ligand conformations from $\L$.
+
 Two strategies to select a subset from the $\L$ conformations were tested: 
-the strategy described in Sec.~\ref{sec::strat} (denoted `S') and the strategy that prefers smaller conformations (denoted `P').
+the strategy described in Sec.~\ref{sec::strat} (denoted `S') and the strategy that prefers smaller (packed) conformations (denoted `P').
 The `P' method selects such conformations where the deviations of the atomic distance from the geometric center are the lowest, which leads 
 to the selection of packed conformations.
-The strategies were tested using the ligand DCP (Dichloropropanol), for which $1300$ conformations were created, and for the ligand
-m040 (1,5-Dichloropentane) with $7560$ conformations.
+The strategies were tested using the ligand DCP (Dichloropropanol), for which  $|\L|=1300$ conformations were created, and for the ligand
+m040 (1,5-Dichloropentane) with $|\L|=7560$ conformations.
 
@@
 {\scriptsize
-\renewcommand{\arraystretch}{0.7}
+%\renewcommand{\arraystretch}{0.7}
 \input graphSelNew.tex
@@
 
-\subsection{Traversability of Ligands}
+\subsubsection{Traversability of Ligands}
 
@@
 \scriptsize
-\renewcommand{\arraystretch}{0.7}
-\input graphNew.tex
+%\renewcommand{\arraystretch}{0.7}
+%\input graphNew.tex %with \rc (a1*) column
+\input graphNewNoA3.tex %without \rc (a1*) column
 }
@@
 {\footnotesize
-\def\arraystretch{0.9}
+%\def\arraystretch{0.9}
 \begin{tabular}{cc}
@@
 
+\subsection{Analysis of protein conformations}
+
+The next studied case was the analysis of Leukotriene A4 hydrolase (PDB ID 4L2L).
+The task was to find trajectory for the ligand 4-(4-benzylphenyl)-1,3-thiazol-2-amine, which is a known inhibitor of the protein 4L2L.
+We therefore assume that the ligand can reach the active site.
+%which was in the protein crystal structure (PDB ID 4L2L). 
+%Based on the location of inhibitor inside the protein we assume that the inhibitor is able 
+The tCONCOORD tool~\cite{seeliger2007geometry} was used to generate 1000 conformations of the protein 4L2L (without the inhibitor).
+The initial pdb structure was minimized by GROMACS 4.0.7~\cite{hess2008gromacs} using the input parameters recommended on the example page for tCONCOORD. 
+The protein conformations were generated using default settings for tCONCOORD in the PERT mode. % which only perturbs the starting configuration of atoms instead of complete randomization.
+The tunnels towards the active site (defined by residues 134 (Alanine) and 311 (Phenylalanine) for probe 0.9~\AA\ were calculated
+using CAVER 3.0.
+All 1000 conformations contained at least 5 tunnels, more than 900 conformations contained 9 tunnels.
+The ordering of the tunnels can slightly change due to conformational changes, so the traversability
+to the active site was computed for the first three tunnels (with average bottlenecks 1.22~\AA, 1.20~\AA and 1.19~\AA).
+Due to size of the ligand (which has 33 atoms) and the low bottlenecks of the tunnels, we
+
+%R1: https://www.ncbi.nlm.nih.gov/pubmed/17997973
+%R2: https://www.ncbi.nlm.nih.gov/pubmed/26620784
+
+\begin{figure}
+\centering
+\caption{\label{fig::inhibitor}Example of inhibitor in 4L2L ... }
+\end{figure}
+
 \section{Discussion and Utilization of the Results}
@@
 are shown in Fig.~\ref{fig::comparison}.
-The runtimes of \RB\ are higher than runtimes of \RA and the std. deviations as well.
-The variant \RC\ is faster than \RA\ due to simple evaluation of the metric in the expansion step.
+The runtimes of \RB\ are higher than runtimes of \RA\ and the std. deviations as well.
+%The variant \RC\ is faster than \RA\ due to simple evaluation of the metric in the expansion step.
 %The runtimes of both algorithms increase in those frames, where the tunnel is not traversable, i.e., where the distance
 %to the active site is more than 2~\AA\ (for example, in the frame no. 60).
-The utilization of the  atom-based metric (used in \RA) leads to higher traversability rate
-in the case of large scales and larger ligands (e.g., m056 with $\smin=0.8$: 15~\% with the metric, but to only 6~\% without it).
+%The utilization of the  atom-based metric (used in \RA) leads to higher traversability rate
+%in the case of large scales and larger ligands (e.g., m056 with $\smin=0.8$: 15~\% with the metric, but to only 6~\% without it).
 %Both \RA\ and \RC\ provided trajectories with the same traversability rate in the case of lower scales ($\smin<0.6$).
@@
 The expansion prefers atoms with a higher scale, which ensures that the ligand shrinks down only if necessary, i.e., when traversing the narrow parts of the tunnel.
-To further boost the exploration of the configuration space, a novel metric suitable for measuring the distances between various ligand shapes is used in the expansion step.
-The proposed method has been used to test and evaluate the trajectories of several ligands in a sequence of frames capturing protein dynamics.
+%To further boost the exploration of the configuration space, a novel metric suitable for measuring the distances between various ligand shapes is used in the expansion step.
+%The proposed method has been used to test and evaluate the trajectories of several ligands in a sequence of frames capturing protein dynamics.
 

### 2018-02-23 paper.tex
experiments, added new image

@@
 All 1000 conformations contained at least 5 tunnels, more than 900 conformations contained 9 tunnels.
-The ordering of the tunnels can slightly change due to conformational changes, so the traversability
-to the active site was computed for the first three tunnels (with average bottlenecks 1.22~\AA, 1.20~\AA and 1.19~\AA).
-Due to size of the ligand (which has 33 atoms) and the low bottlenecks of the tunnels, we
+The detected the tunnels slightly differ due to conformational changes.
+Therefore, the traversability towards the active site was computed for the first three tunnels (with average bottlenecks 1.22~\AA, 1.20~\AA and 1.19~\AA with the average lengths 22.3~\AA, 25.3~\AA and 28.4~\AA respectively).
+The conformations of the ligand were prepared in the same way as in the previous section.
+From the available $|\L|=736$ conformations, 100 were selected using the `S' strategy.
 
@@
 \centering
-\caption{\label{fig::inhibitor}Example of inhibitor in 4L2L ... }
+\includegraphics[width=0.4\textwidth]{fig/4l2l-detail}
+\caption{\label{fig::inhibitor}Example of inhibitor in 4L2L ... TOOO some better image }
 \end{figure}

### 2018-02-23 paper.tex
experiments

@@
 e.g., to study loop motions~\cite{cortes2004geometric}, to identify binding sites~\cite{bayazit2001ligand}, for exit
-pathway detection~\cite{cortes2010simulating,cortes2005path}, and especially for protein folding~\cite{raveh2009rapid,amato2002using}. %novinskaya2015improving
+pathway detection~\cite{cortes2010simulating,cortes2005path}, for tunnel detection~\cite{vonasek2017tunnel,vonasek2016application}, and especially for protein folding~\cite{raveh2009rapid,amato2002using}.
+%novinskaya2015improving
 
@@
 
-The number of generated ligand conformations $|\L|$ depends on the number of rotatable bonds of the ligands.
-Although only the low-energy conformations are generated and saved to $\L$, the number of all conformations can still be large for
-ligands with many rotatable bonds.
-The aim of this experiment was to verify the procedure to select a subset of ligand conformations from $\L$.
+The number of generated ligand conformations $|\L|$ depends on the number of rotatable bonds of the ligands and eventhough only
+low-enery conformations are stored in $\L$, the set can still contain thousands of conformations.
+For a fast motion planning, we utilize a subset of the original set $\L$.
+The aim of this experiment was to verify the procedure to select the subset.
 
-Two strategies to select a subset from the $\L$ conformations were tested: 
+Two selection strategies were tested:
 the strategy described in Sec.~\ref{sec::strat} (denoted `S') and the strategy that prefers smaller (packed) conformations (denoted `P').
@@
 
+The proposed \RA\ planner provides higher traversability rate than RRT-Retraction (\RB).
+The distances of the configuration trees (measured using the 3D Euclidean metric) to the active site and the runtimes in each frame
+are shown in Fig.~\ref{fig::comparison}.
+The runtimes of \RB\ are higher than runtimes of \RA\ and the std. deviations as well.
+
 %The number of conformations (out of 50 available) used in the successfull trajectories are shown in Tab.~\ref{tab::m040c}.
@@
 
-The next studied case was the analysis of Leukotriene A4 hydrolase (PDB ID 4L2L).
-The task was to find trajectory for the ligand 4-(4-benzylphenyl)-1,3-thiazol-2-amine, which is a known inhibitor of the protein 4L2L.
-We therefore assume that the ligand can reach the active site.
-%which was in the protein crystal structure (PDB ID 4L2L). 
+%The case was the analysis of Leukotriene A4 hydrolase (PDB ID 4L2L).
+The task of the second analysis was to find trajectory for the ligand 4-(4-benzylphenyl)-1,3-thiazol-2-amine into the active
+site of A4 hydrolase (PDB ID 4L2L).
+%This ligand is a known inhibitor of the protein 4L2L.
 %Based on the location of inhibitor inside the protein we assume that the inhibitor is able 
@@
 The protein conformations were generated using default settings for tCONCOORD in the PERT mode. % which only perturbs the starting configuration of atoms instead of complete randomization.
-The tunnels towards the active site (defined by residues 134 (Alanine) and 311 (Phenylalanine) for probe 0.9~\AA\ were calculated
-using CAVER 3.0.
+The tunnels to the active site defined by residues 134 (Alanine) and 311 (Phenylalanine) were computed by CAVER 3.0 for the probe 0.9~\AA.
 All 1000 conformations contained at least 5 tunnels, more than 900 conformations contained 9 tunnels.
 The detected the tunnels slightly differ due to conformational changes.
-Therefore, the traversability towards the active site was computed for the first three tunnels (with average bottlenecks 1.22~\AA, 1.20~\AA and 1.19~\AA with the average lengths 22.3~\AA, 25.3~\AA and 28.4~\AA respectively).
+Therefore, the traversability towards the active site was computed for the first three tunnels (with average bottlenecks 1.22~\AA, 1.20~\AA\ and 1.19~\AA\ with the average lengths 22.3~\AA, 25.3~\AA\ and 28.4~\AA, respectively).
 The conformations of the ligand were prepared in the same way as in the previous section.
 From the available $|\L|=736$ conformations, 100 were selected using the `S' strategy.
+
+As the tested ligand is a known inhibitor of the protein 4L2L, we assume that the ligand can reach the active site.
+We tested several scaling-down factors from $\smin=0.4$ to $\smin=0.6$.
+
+Example of a found trajectory with the depicted final position of the ligand is depicted in Fig.~\ref{fig::inhibitor}.
+TODO: report num of travewrsable frames
 
@@
 %originally last two paragraphs in the experiment section, for resubmition, they are moved here to discussion
-The proposed \RA\ planner provides higher traversability rate than RRT-Retraction (\RB).
-The distances of the configuration trees (measured using the 3D Euclidean metric) to the active site and the runtimes in each frame
-are shown in Fig.~\ref{fig::comparison}.
-The runtimes of \RB\ are higher than runtimes of \RA\ and the std. deviations as well.
 %The variant \RC\ is faster than \RA\ due to simple evaluation of the metric in the expansion step.
@@
 %Both \RA\ and \RC\ provided trajectories with the same traversability rate in the case of lower scales ($\smin<0.6$).
-
 %The most time-consuming part of the planners is the collision detection. 
@@
 %This results in a faster exploration of the configuration space and therefore, \RA\ needs less number of iterations to reach the active site.
-
 %The better performance of \RA\ is caused by the proposed atom-based metric, because this distance depends on the actual shape of the ligand and 
@@
 
-The presented work was motivated by the current needs of biochemists who need to analyze the traversability of selected ligands through selected protein tunnels.
+The presented work is motivated by the current needs of biochemists who need to analyze the traversability of selected ligands through selected protein tunnels.
 Due to low bottleneck of the tunnels, it is often required to shrink the atomic radii of the ligand, as utilized also in other existing tools.
@@
 the planner prefers to apply the higher scales if possible.
-%The expansion step always prefers to expand the tree by a collision-free configuration with the maximal scale.
-%that leads to a new collision-free configuration.
 
-The information about the scales used at the trajectory points 
-brings the biochemists new knowledge about the tunnels, which was impossible to get using the VD-based computations.
+{\color{red}
+    Neni to blbost? (ta prvni veta)
+Determination of the proper scaling-down factor is beyond the scope of this paper, where we focus on the algorithmic part of the problem.
+The computed trajectories are further visualized and analyzed.
+The scales used at the trajectory points brings the biochemists new knowledge about the tunnels, which was impossible to get using the VD-based computations.
 By visualizing all trajectories computed for different scales, we can observe those parts of the tunnel where the ligand had to be scaled-down significantly (red parts of trajectories in Fig.~\ref{fig::dcp}), and parts where the ligand could be enlarged again (green parts in Fig.~\ref{fig::dcp}).
@@
 Beyond this point, the ligand can move at larger scales again.
+}
+
 As both the protein and the ligand mutually influence each other, it is possible that even such a narrow part of the tunnel is traversable.
 This was observed in a separate MD simulation of the DCP ligand in the 4E46 protein, which revealed that DCP can pass the main tunnel~\cite{marques2017catalytic}. 
+The necessity to scaled-down even more than 0.8, which is a value used by other tools~\cite{cortes2010simulating}, was
+shown also in the second experiment, where the trajectory for inhibitor of the 4L2L protein required scaled-down to 0.5 or 0.6.
 
@@
 We presented a novel modification of the RRT planner to compute trajectories of ligands inside selected protein tunnels.
-In our solution, we take into account the ligand shape and we model the ligand flexibility using a predefined set of its conformations.
+The ligand shape is taken into account and the ligand flexibility is modeled using a predefined set of its conformations.
+The ligand conformations are prepared using the Rosetta tool, ranked by their energy and only low-energy conformations are used in the planning.
 To simulate the real behavior of ligand passing through a tunnel, we enable scaling down the atomic radii of the ligands up to a certain predefined factor.

### 2018-02-23 paper.tex
update from Bara

@@
 The location and size of a tunnel determine the size and shape of a ligand passing through this tunnel to the active site.
-Currently, the biochemists can use various computational methods for tunnel detection in order to estimate the possibility of interactions between a protein and a ligand.
+Currently, the biochemists can use various computational methods for tunnel detection in order to estimate the possibility of interactions between the protein and ligand.
 This can help them to reduce the number of necessary in-vitro experiments.
-The most common tools detect the tunnels using Voronoi diagrams~\cite{yaffe2008,caver3} assuming a spherical ligand (probe), 
+The most common tools detect the tunnels using Voronoi diagrams~\cite{yaffe2008,caver3}, assuming a spherical ligand (probe), 
 i.e., they approximate the ligand by a bounding sphere.
-However, ligands are typically of a non-spherical shape which is completely ignored by these methods.
+However, ligands are typically of a non-spherical shape, which is completely ignored by these methods.
 
@@
 Our approach aims to overcome this problem by taking into account the actual shape of the ligand.
-This necessarily requires to consider also conformational changes of the ligand, i.e., the mutual position of the ligand atoms can change.
-Such a ligand is denoted as flexible ligand.
+This necessarily requires to consider also conformational changes of the ligand, i.e., changes of the mutual position of the ligand atoms.
+Such ligand is denoted as a flexible ligand.
 %Our approach aims to overcome this problem by taking into account the actual shape of the ligand and also its conformations.
@@
 
-Finding trajectories only along a given tunnel is faster than computing all possible exit path, due to the decreased volume of the searched configuration space.
+Finding trajectories only along a given tunnel is faster than computing all possible exit paths, due to the decreased volume of the searched configuration space.
 Another advantage of our method, in comparison with the existing approaches, is that we enable the users to trace the ligand path leading from the
@@
 \section{Related Work}
-{\color{red}
-Migration of the ligands into a protein can be directly evaluated using Molecular dynamics (MD) simulations, that consider both 
-the protein and the ligand flexible.
+Transportation of ligands into the protein active site can be directly evaluated using molecular dynamics (MD) simulations, where both the protein and ligand are flexible.
 However, MD simulations are considerably computational demanding~\cite{kingsley2014including} and they have to be prepared
-for particular ligand/protein pair, which is not practical for testing a large number of different ligands.
-Therefore, computational approaches that consider only geometry of the protein/ligand are still preferable due to their high speed.
+for a particular ligand/protein pair, which is not practical for testing a large number of different ligands interacting with the protein.
+Therefore, computational approaches that consider only the static geometry of the ligand are still preferable due to their high speed.
 
-The widely used approaches to evaluate possiblity of ligand migration into the protein rely on the tunnel detection.
-Most of the available tools for tunnel detection are based on Voronoi diagrams~\cite{yaffe2008,caver3} or they utilize a grid-based computation~\cite{sehnal2013mole}.
-Tunnels are represented as a sequence of spheres and characterized by their bottleneck, length, curvature, and the list of surrounding residues, whose physico-chemical properties are crucial for assessing the protein-ligand interaction possibilities.
-The tunnels are detected either on a single conformation of the protein, or, on a set of protein conformations.
-The protein conformations can be prepared again using MD simulations.
-Computationally less costly approaches for generating protein conformations are based on constrained geometric simulations like
-ROCK~\cite{lei2004sampling}, FRODA~\cite{wells2005constrained} and tCONCOORD~\cite{seeliger2007geometry}.
+The widely used approaches to evaluate the ligand traversability rely on the tunnel detection.
+Most of the available tools for tunnel detection are based on Voronoi diagrams~\cite{yaffe2008,caver3,sehnal2013mole}.
+Tunnels are represented as a sequence of spheres and characterized by their bottleneck, length, curvature, and the list of surrounding residues, whose physico-chemical properties are crucial for assessing of the feasibility of the protein-ligand interactions.
+The tunnels are detected either on a single conformation of the protein, or, on a set of protein conformations, generated by MD simulations.
+Computationally less demanding approaches for generating protein conformations are based on constrained geometric simulations, such as
+ROCK~\cite{lei2004sampling}, FRODA~\cite{wells2005constrained}, and tCONCOORD~\cite{seeliger2007geometry}.
 Examples of utilization of protein tunnels can be found in the recently published state-of-the-art report about the analysis and visualization of biomolecular cavities~\cite{Krone_2016}.
 
-The main disadvantage of both grid-based and VD-based tunnel detection methods is that the shape of the ligand is not taken into account during the tunnel detection and it is therefore not easy to estimate if (and how) a non-spherical ligand can traverse the tunnel.
-The ligand flexibility cannot be principally considered by Voronoi-based approaches.
-}
+The main disadvantage of Voronoi diagram based tunnel detection methods is that the shape of the ligand is not taken into account during the tunnel detection.
+Therefore, it is not easy to estimate if (and how) a non-spherical ligand can traverse the tunnel.
+The ligand flexibility cannot be principally considered by Voronoi-based approaches and other techniques have to be adopted to solve this problem.
 
@@
 To cope with these many-DOF ligands, the RRT-ML~\cite{cortes2007mlrrt} approach is employed as the basic planner in~\cite{cortes2010simulating}.
-RRT-ML expands the tree primarily using those DOF, that are essential for achieving the motion of the ligand (i.e., rotation
+RRT-ML expands the tree primarily using those DOFs, that are essential for achieving the motion of the ligand (i.e., rotation
 and translation) and it employs other DOFs (i.e., those that are responsible for conformational changes) if they hinder the growth of the tree.
@@
 Each configuration in the roadmap is also evaluated using the electrostatic energy.
-The high-energy configurations are discarded, while the low-energy conformations (and their vicinity) are sampled densely.
+The high-energy configurations are discarded, while the low-energy configurations (and their vicinity) are sampled densely.
 The approach~\cite{singhLIG} can also predict the ligand binding sites.
@@
 We propose a modification of RRT to compute trajectories for non-spherical flexible ligands through protein tunnels.
-The input of the proposed method is a set of tunnels computed using Voronoi diagrams, and the task is to find trajectories for a ligand along these tunnels leading from outside the protein to the active site.
-The tunnels are detected on selected protein conformations resulting e.g. from MD simulations or from tool providing conformation sampling
+The input of the proposed method is a set of tunnels computed using Voronoi diagrams, and the task is to find trajectories for a ligand along these tunnels leading from the protein surface to the inner buried active site.
+The tunnels are detected on selected protein conformations, resulting, e.g., from MD simulations or from a tool providing the conformation sampling
 ~\cite{lei2004sampling,wells2005constrained,seeliger2007geometry}.
@@
 
-In contrast to~\cite{cortes2010simulating}, which searches for any exit pathway from the protein, we localize the searching and compute the trajectories only along each tunnel separately which allows the biochemists to analyze selected tunnels.
+In contrast to~\cite{cortes2010simulating}, which searches for any exit pathway from the protein, we localize the searching and compute the trajectories only along each tunnel separately, which allows the biochemists to analyze selected tunnels.
 %This is motivated by the practical needs of biochemists that are interested in the actual traversability the ligand through a selected tunnel.
 Computing the trajectories along a single tunnel brings several advantages from the motion planning point of view as well.
-The random samples are not generated in the whole configuration space, but only around the tunnel in a user-defined distance which helps to cope with the narrow passage problem.
+The random samples are not generated in the whole configuration space, but only around the tunnel in a user-defined distance. 
+This helps to cope with the narrow passage problem.
 %The tunnels serve as a navigation path for the configuration tree~\cite{vonasek2009rrt}.
@@
 It can be expected that the tunnels in MD simulations without ligands are rather narrow.
-However, the tunnels adapt their shape to the passing ligands and vice versa.
+However, the tunnels can adapt their shape to the passing ligands and vice versa.
 %Therefore, it can be expected that the detected tunnels are rather narrow.
-In consequence, even very narrow tunnels, being too small for the passage of a given ligand-approximating sphere, can serve as the transportation path for the ligand.
+In consequence, even very narrow tunnels, being too small for the passage of a given ligand-approximating sphere, can potentially serve as the transportation path for the ligand.
 Therefore, also the existing Voronoi-based methods enable the user to decrease the size of the approximating sphere which simulates the shrinking of the ligand size.
-In our solution we simulate this by shrinking the atom radii of the ligand, used also by~\cite{cortes2010simulating,guieysse2008structure}.
+In our solution, we simulate this by shrinking the atom radii of the ligand, used also by~\cite{cortes2010simulating,guieysse2008structure}.
 
@@
 Proteins and ligands are represented by the hard sphere model, where the radius of each sphere representing an atom is given by its van der Waals radius.
-A protein tunnel is described by a sequence of collision-free spheres 
+Protein tunnel is described by a sequence of collision-free spheres 
 $T=( (c_1, r_1),\ldots,(c_n,r_n) )$, where $n$ denotes the number of spheres,
-$c_i \in \R^3$ is their 3D position, and $r_i > 0$ denotes the maximum collision-free radius of a sphere centered at $c_i$. 
+$c_i \in \R^3$ their 3D position, and $r_i > 0$ the maximum collision-free radius of a sphere centered at $c_i$. 
 Tunnels serving as an input for the proposed method can be detected using some of the existing computational tools, such as CAVER 3.0~\cite{caver3}.
@@
 To enable the motion of ligands in the narrow tunnels, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
-In our solution we use a discrete set of scales, i.e., $s \in \S=\{\smin, \smin+\sdelta, \ldots, \smax\}$, where 
+In our solution, we use the discrete set of scales, i.e., $s \in \S=\{\smin, \smin+\sdelta, \ldots, \smax\}$, where 
 $\smin$ is the minimal allowed scale, $\smax=1$ is the maximal allowed scale, and $\sdelta$ is the minimal difference between two scales.
@@
 A configuration of the ligand $q=(x,y,z,r_x,r_y,r_z,l,s)$ is described
-by the 3D position $(x,y,z)$ of the reference point of the ligand (e.g. the geometric center or a selected atom), the rotation around $x$, $y$, and $z$ axes,
-the index of the conformation $l\in \L$, and the scale $s \in \S$.
+by the 3D position $(x,y,z)$ of the reference point of the ligand (e.g., the geometric center or a selected atom), rotation around $x$, $y$, and $z$ axes,
+index of the conformation $l\in \L$, and scale $s \in \S$.
 All possible configurations form the configuration space $\C$ and the collision-free configurations
@@
 A configuration $q$ is collision-free if none of the ligand atoms scaled by $s$ and placed at the
-position defined by $q$ collides with the protein atoms.
-We assume that the ligand can change from conformation $l_1$ at configuration $q_1$ to a conformation $l_2$ at configuration $q_2$
-if both $q_1$ and $q_2$ are collision-free and if the distance from $q_1$ to $q_2$ is less than the resolution of the planner ($\varepsilon=0.05$~\AA).
+position defined by $q$ collides with protein atoms.
+We assume that the ligand can change from conformation $l_1$ in configuration $q_1$ to conformation $l_2$ in configuration $q_2$
+if both $q_1$ and $q_2$ are collision-free and if the distance from $q_1$ to $q_2$ is smaller than the resolution of the planner ($\varepsilon=0.05$~\AA).
 
@@
 
-Depending on the chemical interactions between the ligand and the protein surface, the ligand may enter a tunnel with various rotations and in various spatial conformations.
-The traversability of the tunnel should therefore be evaluated using multiple starting configurations $\QI$.
+Depending on the chemical interactions between the ligand and the protein surface, the ligand may enter the tunnel in various rotations and spatial conformations.
+Therefore, the traversability of the tunnel should be evaluated using multiple starting configurations $\QI$.
 The initial configurations are searched around the beginning of the tunnel.
@@
 $c_1$ in the distance $\RI$ (the translation and rotation parts of $q$ are generated randomly, the scale is set to $\smin$, and the conformation index is set randomly).
-If the sample $q$ is collision-free, it is added to $\QI$ as a new starting configuration.
+If the sample $q$ is collision-free, it is added to $\QI$ as new starting configuration.
 %Similarly, a single goal configuration $\qgoal$ is found around the end of the tunnel. 
@@
 By placing them too far, the planner would need more iterations to enter the tunnel, which would increase the runtime.
-Contrary, too low $\RI$ may disable to find any collision-free configurations, especially for large ligands.
+Contrary, too low $\RI$ may prevent the algorithm to find any collision-free configurations, especially for large ligands.
 We suggest to set $\RI$ to the radius of the first sphere in the tunnel and increase it only if no collision-free configuration
@@
 The parameter $\dt$ determines how far can the tree detour from the tunnel centerline.
-In the case of alone tunnels (which is very common), we recommend to use double value of the tunnel average width.
+In case of tunnels not having other tunnels in the close vicinity (which is very common), we recommend to use the double value of the tunnel average width.
 The algorithm terminates after a predefined number of planning trials $\Imax$ or if the tree reaches
@@
 The traditional straight-line expansion (i.e., new configuration is constructed on the line from $\qnear$ to $\qrand$ in the distance
-$\varepsilon$ from $\qnear$) more likely leads to collisions, especially in narrow void space.
+$\varepsilon$ from $\qnear$) more likely leads to collisions, especially in the narrow void space.
 We expand the tree considering all $\L$ conformations.
@@
 and tested for collisions.
-The random samples are generated similarly as in the case of $\qrand$ samples, the translation part $(x,y,z)$ is generated around $\qnear$.
+The random samples are generated similarly as in case of $\qrand$ samples, the translation part $(x,y,z)$ is generated around $\qnear$.
 The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
@@
 From the motion planning point of view, both types of conformations are useful.
-The longer conformations can better fit in a narrow passage, while the packed conformations can easily rotate on the spot.
+The longer conformations can better fit in the narrow passage, while the packed conformations can easily rotate on the spot.
 %In the narrow parts of the tunnel, the tree can be extended by the longer conformations, while in the wider regions, also the
@@
 
-The pool of available conformations $\L$ consists all conformations (up to some resolution) with energy lower than a user specified threshold, which can still contain too many conformations (thousands or even more).
-We select a subset by computing how the atoms of the ligand are spread around its geometric mean.
+The pool of available conformations $\L$ consists of all conformations (up to some resolution) with the energy lower than a user-specified threshold, which can still contain too many conformations (thousands or even more).
+We further select a subset by computing how the atoms of the ligand are spread around its geometric center.
 The distance of centers of ligand atoms to the geometric center of the ligand is computed, and the deviation of these
 distances then characterizes the spread.
-Low deviation is characteristic for the packed conformations, while larger deviations correspond to the longer conformations.
+Low deviation is characteristic for the packed conformations, while larger deviations correspond to the elongated conformations.
 To select $n$ conformations from all $|\L|$ available conformations, they are sorted according to the deviation.
@@
 
-The task is to compute the ligand traversability towards the active site using two main tunnels in the Haloalkane dehalogenase protein (DhaAwt, PDB ID 4E46) which consists of 4650 atoms.
-MD of the protein was computed using AMBER~12 (details about the used MD simulations of DhaAwt/4E46 are described in~\cite{marques2017catalytic}).
-The tunnels were detected in 100 consecutive frames of MD simulation in part where the protein was stabilized.
+The task is to compute the ligand traversability towards the active site using two main tunnels in the Haloalkane dehalogenase protein (DhaAwt, PDB ID 4E46), which consists of 4650 atoms.
+The MD simulation was computed using AMBER~12 (details about the used MD simulations of DhaAwt/4E46 are described in~\cite{marques2017catalytic}).
+The tunnels were detected in 100 consecutive frames of MD simulation in the part where the protein was already stabilized.
 The tunnels for probe 0.9~\AA\ towards the active site defined by residues 38 (Asparagine), 102 (Histidine), and 103 (Aspartic Acid) were
@@
 In most of the frames, two tunnels were detected.
-The average bottleneck sizes of these tunnels are $1.48$~\AA\ and $1.2$~\AA\ for the main and the side tunnel, respectively.
+The average bottleneck sizes of these tunnels were $1.48$~\AA\ and $1.2$~\AA\ for the main and side tunnel, respectively.
 
-Models  of the ligands in Mol2 formats were converted to param files~\cite{meiler2006rosettaligand}, containing the topology, rotatable bonds, atom types, and partial charges. 
+Models  of the ligands in Mol2 formats were converted to the param files~\cite{meiler2006rosettaligand}, containing the topology, rotatable bonds, atom types, and partial charges. 
 The pool of conformations $\L$ was created by the rotation of all rotatable bonds with 30$^\circ$ step using Rosetta3.
 The energy cut-off for individual conformations was set to 0.075. %(-rot_ensemble_ecutoff 0.075) 
-
 Four different scaling-down factors were used: $\S=\{0.5,0.6,0.7,0.8\}$.
@@
 Two motion planners were compared: the proposed planner (described in Alg.~\ref{alg::main} and Alg.~\ref{alg::expand}), denoted as \RA\ in the rest of the paper, and the Retraction-based RRT~\cite{zhangRetraction}, denoted as \RB\ in the rest of the paper.
-The \RA\ method was run with the following parameters:
+The \RA\ method was launched with the following parameters:
 $\Imax=5,000$, $m=100$, $\gb=0.9$, $\dt=1.5$~\AA, $\rv=2$~\AA\ and $\varepsilon=0.05$~\AA.
@@
 The nearest-neighbor search of both planners was based on the 6D Euclidean metric with the same weights for rotation and translation parts.
-This metric was selected based on preliminary experiments with both planners as it resulted in the highest success rates.
+This metric was selected based on preliminary experiments with both planners, as it resulted in the highest success rates.
 %The \RA\ method was also tested without the atom-based metric.
@@
 A frame is considered traversable if at least one trajectory out of 200 ($\!|\QI|\!\times\!\!10$ trials) approaches the active site to the distance less than~2~\AA. 
-This distance is measured as 3D distance between the geometric center of the ligand and the position of the active site.
+This distance is measured as the 3D distance between the geometric center of the ligand and the position of the active site.
 %~\footnote{We refer to~\url{http://mrs.felk.cvut.cz/icra2018protein} for additional material.}
@@
 
-The number of generated ligand conformations $|\L|$ depends on the number of rotatable bonds of the ligands and eventhough only
-low-enery conformations are stored in $\L$, the set can still contain thousands of conformations.
-For a fast motion planning, we utilize a subset of the original set $\L$.
-The aim of this experiment was to verify the procedure to select the subset.
+The number of generated ligand conformations $|\L|$ depends on the number of rotatable bonds of the ligands and even though only
+the low-energy conformations are stored in $\L$, the set can still contain thousands of conformations.
+For fast motion planning we utilize a subset of the original set $\L$.
+The aim of this experiment was to verify the procedure for selection of the subset.
 
@@
 The number after the slash in the ligand's name denotes the number of atoms of the ligand.
-In the case of both tested ligands and 50 selected conformations, the `S' strategy leads to a higher traversability rate than the `P' strategy.
-For the DCP ligand, higher traversability rate was achieved when 50 conformations were selected.
+In case of both tested ligands and 50 selected conformations, the `S' strategy leads to a higher traversability rate than the `P' strategy.
+For the DCP ligand, the higher traversability rate was achieved when 50 conformations were selected.
 This experiment shows that selecting 50 conformations using the `S' strategy (Sec.~\ref{sec::strat}) leads to better results
-than using a less number of conformations or when preferring the packed conformations.
+than using the smaller number of conformations or when preferring packed conformations.
 
@@
     The number after~`$/$' denotes the number of atoms.
-    The protein has 4650 atoms. 
+    The tested protein consists of 4650 atoms. 
     The rate is shown in [\%].
@@
     The number after '$/$' denotes the number of atoms.
-    The protein has 4650 atoms.
+    The tested protein consists of 4650 atoms.
     The rate is shown in [\%].
@@
 The table shows the traversability rate only for the main tunnel, as no trajectory was found in the side tunnel for most of the tested ligands.
-The only exception is the ligand m037t, which was the smallest tested one (with 8 atoms), and therefore it could fit even inside the side tunnel.
+The only exception is the ligand m037t, which was the smallest tested one (consisting of 8 atoms), and therefore it could fit even inside the side tunnel.
 The traversability rate is higher for the most scaled-down ligands ($\smin=0.5$) and it decreases with the increasing scale.
-For the ligand DCP, trajectories were detected in 18~\% of frames for scale $\smin=0.8$ and the traversability of this ligand
+For the DCP ligand, the trajectories were detected in 18~\% of frames for scale $\smin=0.8$ and the traversability of this ligand
 to the active site was approved by a separate MD simulation~\cite{marques2017catalytic}.
@@
 %The traversability of the DCP ligand to active site was approved by a separate MD simulation~\cite{marques2017catalytic}.
-Based on the number of success rate, also other ligands (e.g., m037t with traversability rate 45~\% at $\smin$=0.8) seem to be promising.
-However, their binding was not approved yet by dedicated MD simulations.
+Based on the success rate, also other ligands (e.g., m037t with traversability rate 45~\% at $\smin$=0.8) seem to be promising.
+However, their binding was not yet approved by dedicated MD simulations.
 
-Beside scaling-down the ligand, also the atoms of the protein can be scaled, which is used e.g. in~\cite{cortes2005path}.
-The tool~\cite{cortes2005path} uses the scale 0.8 for both the ligand and the protein.
-We tested also this case using \RA\ (the column \RD\ in Tab.~\ref{tab::main}).
-The results are close the results of \RA\ and $\smin=0.6$  (except DCP and m040 which can be a result of the randomization).
+Besides scaling-down the ligand, also the atoms of the protein can be scaled, which is used, e.g., in~\cite{cortes2005path}.
+The tool~\cite{cortes2005path} uses the scale 0.8 for both the ligand and protein.
+We tested also this option using \RA\ (the column \RD\ in Tab.~\ref{tab::main}).
+The results are similar to the results of \RA\ and $\smin=0.6$ (except for DCP and m040, which can be a result of randomization).
 
-The proposed \RA\ planner provides higher traversability rate than RRT-Retraction (\RB).
+The proposed \RA\ planner provides higher traversability rate than the RRT-Retraction (\RB).
 The distances of the configuration trees (measured using the 3D Euclidean metric) to the active site and the runtimes in each frame
 are shown in Fig.~\ref{fig::comparison}.
-The runtimes of \RB\ are higher than runtimes of \RA\ and the std. deviations as well.
+The runtimes of \RB\ are higher than the runtimes of \RA\ and the standard deviations as well.
 
@@
 
-\subsection{Analysis of protein conformations}
+\subsection{Analysis of Protein Conformations}
 
 %The case was the analysis of Leukotriene A4 hydrolase (PDB ID 4L2L).
-The task of the second analysis was to find trajectory for the ligand 4-(4-benzylphenyl)-1,3-thiazol-2-amine into the active
+The task of the second analysis was to find the trajectory for the ligand 4-(4-benzylphenyl)-1,3-thiazol-2-amine into the active
 site of A4 hydrolase (PDB ID 4L2L).
@@
 The tCONCOORD tool~\cite{seeliger2007geometry} was used to generate 1000 conformations of the protein 4L2L (without the inhibitor).
-The initial pdb structure was minimized by GROMACS 4.0.7~\cite{hess2008gromacs} using the input parameters recommended on the example page for tCONCOORD. 
-The protein conformations were generated using default settings for tCONCOORD in the PERT mode. % which only perturbs the starting configuration of atoms instead of complete randomization.
-The tunnels to the active site defined by residues 134 (Alanine) and 311 (Phenylalanine) were computed by CAVER 3.0 for the probe 0.9~\AA.
+The initial PDB structure was minimized by GROMACS 4.0.7~\cite{hess2008gromacs} using the input parameters recommended on the example page for tCONCOORD. 
+The protein conformations were generated using the default settings for tCONCOORD in the PERT mode. % which only perturbs the starting configuration of atoms instead of complete randomization.
+The tunnels leading to the active site defined by residues 134 (Alanine) and 311 (Phenylalanine) were computed by CAVER 3.0, using the probe 0.9~\AA.
 All 1000 conformations contained at least 5 tunnels, more than 900 conformations contained 9 tunnels.
-The detected the tunnels slightly differ due to conformational changes.
-Therefore, the traversability towards the active site was computed for the first three tunnels (with average bottlenecks 1.22~\AA, 1.20~\AA\ and 1.19~\AA\ with the average lengths 22.3~\AA, 25.3~\AA\ and 28.4~\AA, respectively).
+The detected the tunnels slightly differ due to the conformational changes.
+Therefore, the ligand traversability towards the active site was computed for the first three tunnels (with average bottlenecks 1.22~\AA, 1.20~\AA\, and 1.19~\AA\ with the average lengths 22.3~\AA, 25.3~\AA\, and 28.4~\AA, respectively).
 The conformations of the ligand were prepared in the same way as in the previous section.
@@
 
-Example of a found trajectory with the depicted final position of the ligand is depicted in Fig.~\ref{fig::inhibitor}.
-TODO: report num of travewrsable frames
+The example of the found trajectory showing the final position of the ligand is depicted in Fig.~\ref{fig::inhibitor}.
+TODO: report num of traversable frames
 
@@
 \includegraphics[width=0.4\textwidth]{fig/4l2l-detail}
-\caption{\label{fig::inhibitor}Example of inhibitor in 4L2L ... TOOO some better image }
+\caption{\label{fig::inhibitor}Example of inhibitor in 4L2L ... TODO some better image }
 \end{figure}
@@
 The presented work is motivated by the current needs of biochemists who need to analyze the traversability of selected ligands through selected protein tunnels.
-Due to low bottleneck of the tunnels, it is often required to shrink the atomic radii of the ligand, as utilized also in other existing tools.
+Due to the low bottleneck of tunnels, it is often required to shrink the atomic radii of the ligand, as utilized also in other existing tools.
 Finding trajectories for the scaled-down ligands enables to solve the problem with narrow passages, but it can be questionable how biochemically relevant such trajectories~are.
@@
 
-{\color{red}
-    Neni to blbost? (ta prvni veta)
-Determination of the proper scaling-down factor is beyond the scope of this paper, where we focus on the algorithmic part of the problem.
-The computed trajectories are further visualized and analyzed.
-The scales used at the trajectory points brings the biochemists new knowledge about the tunnels, which was impossible to get using the VD-based computations.
+Determination of the proper scaling-down factor requires long-term testing of many exemplary simulations and needs to be performed by the biochemists. 
+Therefore, it is beyond the scope of this paper, which focuses namely on the algorithmic part of the problem.
+The scales used at the trajectory points can provide the biochemists with new knowledge about the ligand behavior, which was impossible to get using the Voronoi-based approaches.
+The computed trajectories can further visualized and analyzed.
 By visualizing all trajectories computed for different scales, we can observe those parts of the tunnel where the ligand had to be scaled-down significantly (red parts of trajectories in Fig.~\ref{fig::dcp}), and parts where the ligand could be enlarged again (green parts in Fig.~\ref{fig::dcp}).
-According to the results of the motion planing, the DCP ligand needs to shrink down to ca. 75~\% of its size, but only at one part of the tunnel (Fig.~\ref{fig::dcp}, point B). 
+According to the results of the motion planing, the DCP ligand needs to shrink down to approximately 75~\% of its size, but only at one part of the tunnel (Fig.~\ref{fig::dcp}, point B). 
 Beyond this point, the ligand can move at larger scales again.
-}
 
-As both the protein and the ligand mutually influence each other, it is possible that even such a narrow part of the tunnel is traversable.
+As both the protein and ligand mutually influence each other, it is possible that even such a narrow part of the tunnel is traversable.
 This was observed in a separate MD simulation of the DCP ligand in the 4E46 protein, which revealed that DCP can pass the main tunnel~\cite{marques2017catalytic}. 
 The necessity to scaled-down even more than 0.8, which is a value used by other tools~\cite{cortes2010simulating}, was
-shown also in the second experiment, where the trajectory for inhibitor of the 4L2L protein required scaled-down to 0.5 or 0.6.
+shown also in the second experiment, where the trajectory for the inhibitor of the 4L2L protein required scaling down to 0.5 or 0.6.
 
@@
 The second tunnel can also be entered with the high scale (green), but then the narrow part is traversable only with a low scale ($\sim50$~\%) (D). 
-For the purpose of the visualization, even more scale-down to 0.3=30~\% was allowed (E) to see how the ligand passed the tunnel.
+For the visualization purposes, even more scale-down to 0.3=30~\% was allowed (E) to see how the ligand passed the tunnel.
 The ligand can reach the active site only at a very low scale ($\sim0.4=40$~\%) using the side tunnel.
@@
 The ligand conformations are prepared using the Rosetta tool, ranked by their energy and only low-energy conformations are used in the planning.
-To simulate the real behavior of ligand passing through a tunnel, we enable scaling down the atomic radii of the ligands up to a certain predefined factor.
+To simulate the real behavior of ligand passing through the tunnel, we enable scaling down the atomic radii of the ligand up to a certain predefined factor.
 The planner generates the random samples only along a tunnel computed by the Voronoi diagram, and it attempts to expand the configuration

### 2018-02-24 paper.tex
conformation selection rewritten

@@
 the traversability of the tunnels is determined based only on the radial bottlenecks.
-This can lead to highly imprecise results.
+This can lead to imprecise results as the shape of the ligand is not considered.
 To address this problem, we define the computation of the transportation path of non-spherical ligands including their conformation changes (i.e., when the atoms of the ligand move relatively to each other) as a motion planning task in the high-dimensional configuration space.
@@
 The tunnels are detected either on a single conformation of the protein, or, on a set of protein conformations, generated by MD simulations.
-Computationally less demanding approaches for generating protein conformations are based on constrained geometric simulations, such as
+Computationally less demanding approaches for generating possible protein conformations are based on constrained geometric simulations, such as
 ROCK~\cite{lei2004sampling}, FRODA~\cite{wells2005constrained}, and tCONCOORD~\cite{seeliger2007geometry}.
@@
 The generalization of~\cite{kardossRRTKK} is the guided sampling, where the whole path is used to generate samples in the configuration space.
-The path can be computed in the workspace~\cite{vonasek2009rrt}, or iteratively refined in the configuration space based on the solution of a relaxed version of the problem~\cite{bayazitIRC}.
+The guiding path can be computed in the workspace~\cite{vonasek2009rrt}, or iteratively refined in the configuration space based on the solution of a relaxed version of the problem~\cite{bayazitIRC}.
 
@@
 In the Retraction-based RRT~\cite{zhangRetraction}, the tree is retracted along the boundary of the obstacles in the configuration space.
-This requires the analysis of the contact space, which leads to the generalized penetration depth computation of high complexity~\cite{he2016efficient}.
+%This requires the analysis of the contact space, which leads to the generalized penetration depth computation of high complexity~\cite{he2016efficient}.
 %\cite{he2016efficient} in RAL!!
@@
 The input of the proposed method is a set of tunnels computed using Voronoi diagrams, and the task is to find trajectories for a ligand along these tunnels leading from the protein surface to the inner buried active site.
-The tunnels are detected on selected protein conformations, resulting, e.g., from MD simulations or from a tool providing the conformation sampling
+The tunnels are detected on selected protein conformations, 
+    resulting, e.g., from MD simulations or from a tool providing the protein conformation sampling
 ~\cite{lei2004sampling,wells2005constrained,seeliger2007geometry}.
@@
 Therefore, also the existing Voronoi-based methods enable the user to decrease the size of the approximating sphere which simulates the shrinking of the ligand size.
-In our solution, we simulate this by shrinking the atom radii of the ligand, used also by~\cite{cortes2010simulating,guieysse2008structure}.
+In our solution, we simulate this by shrinking the atom radii of the ligand, which is used also by~\cite{cortes2010simulating,guieysse2008structure}.
 
@@
 The ligand is modeled using the set $\L$ of its conformations.
-The candidate conformations are prepared by discrete sampling of degrees of freedom of the ligand.
+The conformations are prepared by discrete sampling of degrees of freedom of the ligand. % (e.g., using the Rosetta3 tool).
 Each candidate conformation is ranked according to its energy and only low-energy conformations are used in $\L$.
@@
 %and we later show how to select a subset of conformations suitable for the motion planning.
+
 To enable the motion of ligands in the narrow tunnels, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
@@
 We assume that the ligand can change from conformation $l_1$ in configuration $q_1$ to conformation $l_2$ in configuration $q_2$
-if both $q_1$ and $q_2$ are collision-free and if the distance from $q_1$ to $q_2$ is smaller than the resolution of the planner ($\varepsilon=0.05$~\AA).
+if both $q_1$ and $q_2$ are collision-free and if the distance from $q_1$ to $q_2$ is smaller than the resolution $\varepsilon$ of the planner ($\varepsilon=0.05$~\AA).
 
@@
 %The expansion towards $\qrand$ is crucial for the growth of the tree towards the unexplored areas of the configuration space.
-The traditional straight-line expansion (i.e., new configuration is constructed on the line from $\qnear$ to $\qrand$ in the distance
-$\varepsilon$ from $\qnear$) more likely leads to collisions, especially in the narrow void space.
+%The traditional straight-line expansion (i.e., new configuration is constructed on the line from $\qnear$ to $\qrand$ in the distance
+%$\varepsilon$ from $\qnear$) more likely leads to collisions, especially in the narrow void space.
 We expand the tree considering all $\L$ conformations.
@@
 Using a large number of conformations also increases the time complexity of the expansion step.
-Selection of a suitable set of conformations $\L$ is therefore crucial.
-Depending on the ligand, some conformations can be considered as more packed, i.e., the atoms in such conformations are closer to each other, but 
-the atoms can also form a more elongated shape (Tab.~\ref{fig::m040c}).
-From the motion planning point of view, both types of conformations are useful.
-The longer conformations can better fit in the narrow passage, while the packed conformations can easily rotate on the spot.
+The pool of available conformations $\L$ consists of all conformations (up to some resolution) with the energy lower than a user-specified threshold, which can still contain too many conformations (thousands or even more).
+
+In the case of large sets $\L$, only a subset of the available ligand conformations can be used in the planning.
+To select the subset, the conformations are ranked using the spread of the atoms around the geometric center of the ligand.
+The distance of centers of ligand atoms to the geometric center of the ligand is computed, and the deviation of these
+distances then characterizes the spread.
+Low deviation is characteristic for `packed' conformations, while larger deviations correspond to the elongated conformations (Tab.~\ref{fig::m040c}).
+To select the subset of $n$ representative conformations from all $|\L|$ available conformations, they are sorted according to the deviation.
+Each $k-$th conformation from the sorted list, where $k=\lfloor{|\L| / n}\rfloor$, is used.
+The selected conformations are then used in the planning process.
+
+%Selection of a suitable set of conformations $\L$ is therefore crucial.
+%Depending on the ligand, some conformations can be considered as more packed, i.e., the atoms in such conformations are closer to each other, but%the atoms can also form a more elongated shape (Tab.~\ref{fig::m040c}).
+%From the motion planning point of view, both types of conformations are useful.
+%The longer conformations can better fit in the narrow passage, while the packed conformations can easily rotate on the spot.
 %In the narrow parts of the tunnel, the tree can be extended by the longer conformations, while in the wider regions, also the
@@
 %\vspace{-3mm}
-
-The pool of available conformations $\L$ consists of all conformations (up to some resolution) with the energy lower than a user-specified threshold, which can still contain too many conformations (thousands or even more).
-We further select a subset by computing how the atoms of the ligand are spread around its geometric center.
-The distance of centers of ligand atoms to the geometric center of the ligand is computed, and the deviation of these
-distances then characterizes the spread.
-Low deviation is characteristic for the packed conformations, while larger deviations correspond to the elongated conformations.
-To select $n$ conformations from all $|\L|$ available conformations, they are sorted according to the deviation.
-Each $k-$th conformation from the sorted list, where $k=\lfloor{|\L| / n}\rfloor$, is used.
+%We further select a subset by computing how the atoms of the ligand are spread around its geometric center.
 
@@
 The retraction step is repeated 5 times.
-The \RB\ method can therefore expand the tree by at most 5 new nodes per configuration, while \RA\ extends the tree only by one configuration.
+The \RB\ method can therefore expand the tree by at most 5 new nodes per ligand conformation, while \RA\ extends the tree only by one new node.
 The larger number of nodes in \RB\ method is not problematic, as the KD-trees were used for the nearest-neighbor search and the time to solve these queries is significantly smaller than the computing collision detection.
@@
 The task of the second analysis was to find the trajectory for the ligand 4-(4-benzylphenyl)-1,3-thiazol-2-amine into the active
-site of A4 hydrolase (PDB ID 4L2L).
+site of A4 hydrolase (PDB ID 4L2L, 9666 atoms).
 %This ligand is a known inhibitor of the protein 4L2L.
@@
 All 1000 conformations contained at least 5 tunnels, more than 900 conformations contained 9 tunnels.
-The detected the tunnels slightly differ due to the conformational changes.
-Therefore, the ligand traversability towards the active site was computed for the first three tunnels (with average bottlenecks 1.22~\AA, 1.20~\AA\, and 1.19~\AA\ with the average lengths 22.3~\AA, 25.3~\AA\, and 28.4~\AA, respectively).
+The ordering of the detected tunnels slightly differ in different conformations.
+Therefore, the ligand traversability towards the active site was computed for the first three widest tunnels (with average bottlenecks 1.22~\AA, 1.20~\AA\, and 1.19~\AA\ with the average lengths 22.3~\AA, 25.3~\AA\, and 28.4~\AA, respectively).
 The conformations of the ligand were prepared in the same way as in the previous section.

### 2018-02-24 paper.tex
new images
images of 4l2l solution + new  table

@@
 Protein reactivity is tightly connected with the presence of the void space between its amino acids. 
-This space, forming a path called a tunnel, can be utilized by a small molecule (ligand) which can enter the protein inner space and react with the amino acids surrounding the protein active site~\cite{gora2013gates} (see Fig.~\ref{fig::motiv}).
+This space, forming a path called a tunnel, can be utilized by a small molecule (ligand) which can enter the protein inner space and react with the amino acids surrounding the protein active site~\cite{gora2013gates,marques2017enzyme} (see Fig.~\ref{fig::motiv}).
 The location and size of a tunnel determine the size and shape of a ligand passing through this tunnel to the active site.
@@
 In this paper, we propose a modification of the RRT planner to find trajectories for non-spherical flexible ligands inside protein tunnels.
+A tunnel is represented by a set of consecutive spheres surrounding the tunnel centerline.
 Random samples are generated only around a given tunnel instead of the whole configuration space.
 %We localize the generation of the random samples, i.e., instead of sampling the whole configuration space we focus only on the space surrounding a potential tunnel detected by the Voronoi-based method (using a smaller ligand probe in order to avoid omitting possibly feasible tunnel candidates).
-The potential tunnel is represented by a set of consecutive spheres surrounding the tunnel centerline.
 To enable the movement of the ligand in the narrow parts of the tunnel, the free-space is dilated by shrinking the atom radii, similarly, e.g., to~\cite{hsu06multilevel}.
@@
 for a particular ligand/protein pair, which is not practical for testing a large number of different ligands interacting with the protein.
-Therefore, computational approaches that consider only the static geometry of the ligand are still preferable due to their high speed.
+Therefore, computational approaches considering only geometry of ligand and proteins are still preferable due to their high speed.
 
@@
 Consequently, many iterations are needed in order to put enough samples into the passages, which increases the computation time.
-The PRM-based planners can cope with the narrow passage problem by increasing the probability of sampling in difficult regions, e.g., by sampling along the medial axis~\cite{amatoOBRRT,amato2002using,wilmarthMAPRM}.
+The PRM-based planners can cope with the narrow passage problem by increasing the probability of sampling in difficult regions, e.g., by sampling along the medial axis~\cite{wilmarthMAPRM}.
 In~\cite{bergWIG}, the workspace is represented by a grid in which the difficult regions are identified using watershed labeling.
@@
 %~\cite{vonasekphd}.
-To prevent this blocking, the DD-RRT~\cite{yershovaDDRRT} limits the selection of nodes for the expansion to a small ball. 
+To prevent this blocking, DD-RRT~\cite{yershovaDDRRT} limits the selection of nodes for the expansion to a small ball. 
 The radius of this ball is set to infinity for new nodes, and decreases to a predefined radius if the node cannot be successfully expanded.
@@
 
-Due to the protein dynamics, the tunnels move and merge, and their bottleneck can change in each time step.
-It can be expected that the tunnels in MD simulations without ligands are rather narrow.
-However, the tunnels can adapt their shape to the passing ligands and vice versa.
+It has been observed, that the tunnels detected in protein conformations without ligands (provided by MD simulations without ligands or by
+protein conformation sampling) are rather narrow. 
+However, in protein dynamics with ligand, the tunnels can move, merge and adapt to the shape of the passing ligand and vice versa.
 %Therefore, it can be expected that the detected tunnels are rather narrow.
+%Due to the protein dynamics, the tunnels move and merge, and their bottleneck can change in each time step.
+%The tunnels can adapt their shape to the passing ligands and vice versa.
 In consequence, even very narrow tunnels, being too small for the passage of a given ligand-approximating sphere, can potentially serve as the transportation path for the ligand.
@@
 
-Molecular dynamics simulations of ligands passing through proteins show that the ligands can be highly flexible.
-However, ligands cannot form arbitrary conformations. 
-The ligand can contain several torsion and dihedral angles between its atoms which determine possible ligand conformations.
-Therefore, the ligand flexibility can be modeled using only a predefined set of conformations.
+MD simulations of ligands passing through proteins further show that the ligands can be highly flexible.
+The possible conformations of a ligand are determined by the torsion and dihedral angles between its atoms. 
+The low-energy conformations are preferred.
+%the ligand can contain several torsion and dihedral angles between its atoms which determine possible ligand conformations.
+Therefore, the ligand flexibility can be modeled using only a predefined set of low-energy conformations.
 By using the set of predefined conformations, it is not necessary to model the additional degrees of freedom, so 
@@
 and tested for collisions.
-The random samples are generated similarly as in case of $\qrand$ samples, the translation part $(x,y,z)$ is generated around $\qnear$.
+The random samples are generated similarly as in the case of $\qrand$ samples, the translation part $(x,y,z)$ is generated around $\qnear$.
 The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
@@
 distances then characterizes the spread.
-Low deviation is characteristic for `packed' conformations, while larger deviations correspond to the elongated conformations (Tab.~\ref{fig::m040c}).
+Low deviation is characteristic for `packed' conformations, while larger deviations correspond to the elongated conformations (Tab.~\ref{tab::m040c}).
 To select the subset of $n$ representative conformations from all $|\L|$ available conformations, they are sorted according to the deviation.
@@
 
-The proposed algorithm was verified in two scenarios: on a sequence frames of MD simulation and on a set of protein conformations provided by the tCONCOORD tool~\cite{seeliger2007geometry}.
+The proposed algorithm was verified in two scenarios: on a sequence of frames of MD simulation and on a set of protein conformations provided 
+the conformation sampling approach.
+%by the tCONCOORD tool~\cite{seeliger2007geometry}.
 
@@
 Four different scaling-down factors were used: $\S=\{0.5,0.6,0.7,0.8\}$.
+
 For each scale, $|\QI|=20$ collision-free initial configurations were found in the $\RI=2$~\AA\ radius around the first sphere of the tunnel (outside the protein).
@@
 The parameters of the methods were selected in such a way that the number of collision detection queries in each expansion step is the same.
-For each conformation and tested scale, the \RA\ method attempts to find $m=100$ collision-free configurations, and \RB\ attempts to perform $5 \times 20$ collision-detection queries.
+For each conformation and tested scale, the \RA\ method checks $m=100$ candidate configurations for collision, and \RB\ checks $5 \times 20$ configurations for collision.
+%collision-detection queries.
 Other parameters of the \RB\ planner were the same as for \RA.
@@
 This experiment shows that selecting 50 conformations using the `S' strategy (Sec.~\ref{sec::strat}) leads to better results
-than using the smaller number of conformations or when preferring packed conformations.
+than using the smaller number of conformations or when preferring the packed conformations.
 
-\begin{table}[ht]
+\begin{table}[bt]
 \centering
@@
     The number after~`$/$' denotes the number of atoms.
-    The tested protein consists of 4650 atoms. 
     The rate is shown in [\%].
@@
 
-Based on the results of the previous experiment, 50 conformations using the `S' strategy were selected for ligands 
+Based on the results of the previous experiment, 50 conformations using the `S' strategy (Sec.~\ref{sec::strat})  were selected for ligands 
 m037t (1,2-Dichloroethane), m038t (1,3-Dichloropropane), m056 (2-Chlorobutane), and m080 (Trichloropropane).
@@
     The number after '$/$' denotes the number of atoms.
-    The tested protein consists of 4650 atoms.
-    The rate is shown in [\%].
 }
@@
 \input graphNewNoA3.tex %without \rc (a1*) column
+%\input traversability.tex
 }
@@
 The traversability rate is shown in Tab.~\ref{tab::main}. 
-The table shows the traversability rate only for the main tunnel, as no trajectory was found in the side tunnel for most of the tested ligands.
+Only the traversability of the main tunnel is shown, as no trajectory was found in the side tunnel for most of the tested ligands.
 The only exception is the ligand m037t, which was the smallest tested one (consisting of 8 atoms), and therefore it could fit even inside the side tunnel.
@@
 However, their binding was not yet approved by dedicated MD simulations.
+Most of the used 50 ligand conformations where used in the planning, which is shown in Tab.~\ref{tab::m040c}.
 
-Besides scaling-down the ligand, also the atoms of the protein can be scaled, which is used, e.g., in~\cite{cortes2005path}.
-The tool~\cite{cortes2005path} uses the scale 0.8 for both the ligand and protein.
+Besides scaling-down the ligand, also the atoms of the protein can be scaled, which is used, e.g., in~\cite{cortes2005path}, where
+%The tool~\cite{cortes2005path} uses 
+the scale 0.8 is used for both the ligand and protein.
 We tested also this option using \RA\ (the column \RD\ in Tab.~\ref{tab::main}).
@@
 \begin{table}[t]
-\caption{\label{fig::m040c}
+\caption{\label{tab::m040c}
 \small
@@
 \begin{tabular}{cc}
-\rotatebox{0}{\hskip 5pt Long} \\ \includegraphics[width=0.1\textwidth]{fig/m040-conf1} \\
-\rotatebox{0}{Packed} \\ \includegraphics[width=0.078\textwidth]{fig/m040-conf2}
+\rotatebox{0}{\hskip 5pt Long} \\ \includegraphics[width=0.07\textwidth]{fig/m040-conf1} \\
+\rotatebox{0}{Packed} \\ \includegraphics[width=0.0375\textwidth]{fig/m040-conf2}
 \end{tabular}
@@
 %Based on the location of inhibitor inside the protein we assume that the inhibitor is able 
-The tCONCOORD tool~\cite{seeliger2007geometry} was used to generate 1000 conformations of the protein 4L2L (without the inhibitor).
+The tCONCOORD tool~\cite{seeliger2007geometry} was used to generate 1000 conformations of the protein 4L2L.
 The initial PDB structure was minimized by GROMACS 4.0.7~\cite{hess2008gromacs} using the input parameters recommended on the example page for tCONCOORD. 
@@
 The conformations of the ligand were prepared in the same way as in the previous section.
-From the available $|\L|=736$ conformations, 100 were selected using the `S' strategy.
+From the available $|\L|=736$ low-energy conformations, 100 were selected using the `S' strategy.
 
-As the tested ligand is a known inhibitor of the protein 4L2L, we assume that the ligand can reach the active site.
-We tested several scaling-down factors from $\smin=0.4$ to $\smin=0.6$.
+\begin{table}[bt]
+\centering
+\caption{\label{tab::4l2l}
+    \small
+    Traversability rate [\%] in 1000 conformations of the protein 4L2L in the first three tunnels.
+}
+\small
+\renewcommand{\tabcolsep}{2pt}
+{
+\scriptsize
+\input traversability.tex
+}
+\end{table}
 
-The example of the found trajectory showing the final position of the ligand is depicted in Fig.~\ref{fig::inhibitor}.
-TODO: report num of traversable frames
+As the tested ligand is a known inhibitor of the protein 4L2L~\cite{marques2017enzyme} we assume that the ligand can reach the active site.
+Due to large size of the ligand (33 atoms), low scale-down factors $\smin\in\{0.4,0.5,0.6\}$ were used.
+The traversability detected by the \RA\ method is show in Tab.~\ref{tab::4l2l} and the example
+of a final trajectory to the active site is depicted in Fig.~\ref{fig::inhibitor}.
 
@@
 \centering
-\includegraphics[width=0.4\textwidth]{fig/4l2l-detail}
-\caption{\label{fig::inhibitor}Example of inhibitor in 4L2L ... TODO some better image }
+\vskip -5pt
+\includegraphics[width=0.4\textwidth]{fig/4l2l-solution}
+\caption{\small
+    \label{fig::inhibitor}Example of a trajectory inside the tunnel No. 1 of the 4L2L protein with the final position of the ligand.
+}
 \end{figure}
@@
 
-Judging the accessibility of the active site by a ligand based only on the traversability rates may be problematic if the ligand is scaled down too much.
-Instead of using a single version of the scaled-down ligand, we only limit the minimal scale used by the planner (the parameter $\smin$) and
-the planner prefers to apply the higher scales if possible.
-
 Determination of the proper scaling-down factor requires long-term testing of many exemplary simulations and needs to be performed by the biochemists. 
@@
 The scales used at the trajectory points can provide the biochemists with new knowledge about the ligand behavior, which was impossible to get using the Voronoi-based approaches.
-The computed trajectories can further visualized and analyzed.
-By visualizing all trajectories computed for different scales, we can observe those parts of the tunnel where the ligand had to be scaled-down significantly (red parts of trajectories in Fig.~\ref{fig::dcp}), and parts where the ligand could be enlarged again (green parts in Fig.~\ref{fig::dcp}).
+%The computed trajectories can further visualized and analyzed.
+Judging the accessibility of the active site by a ligand based only on the traversability rates may be problematic if the ligand is scaled down too much.
+The visualization of the computed trajectories can show 
+%By visualizing all trajectories computed for different scales, we can observe 
+those parts of the tunnel where the ligand had to be scaled-down significantly (red parts of trajectories in Fig.~\ref{fig::dcp}), and parts where the ligand could be enlarged again (green parts in Fig.~\ref{fig::dcp}).
+
 According to the results of the motion planing, the DCP ligand needs to shrink down to approximately 75~\% of its size, but only at one part of the tunnel (Fig.~\ref{fig::dcp}, point B). 
 Beyond this point, the ligand can move at larger scales again.
-
 As both the protein and ligand mutually influence each other, it is possible that even such a narrow part of the tunnel is traversable.
@@
 shown also in the second experiment, where the trajectory for the inhibitor of the 4L2L protein required scaling down to 0.5 or 0.6.
+
+%Instead of using a single version of the scaled-down ligand, we only limit the minimal scale used by the planner (the parameter $\smin$) and
+%the planner prefers to apply the higher scales if possible.
 
@@
 %\includegraphics[width=0.17\textwidth,angle=\a]{fig/m037tmins-03tunnel-1trajDistToGoal-2}
-\includegraphics[width=0.22\textwidth,angle=\a]{fig/crop1}
-\includegraphics[width=0.22\textwidth,angle=\a]{fig/crop2}
+\includegraphics[width=0.22\textwidth,angle=\a]{fig/crop3}
+\includegraphics[width=0.22\textwidth,angle=\a]{fig/crop4}
 \caption{\label{fig::comparison}
@@
 \centering
+\vskip -5pt
 \includegraphics[width=0.35\textwidth]{fig/dcp-image1Label}\\
 %\includegraphics[width=0.2\textwidth]{fig/colormap}
-\vspace{-3mm}
+\vspace{-5pt}
 \caption{\label{fig::dcp}

### 2018-02-24 paper.tex
added percents to all tables

@@
 The tunnels were detected in 100 consecutive frames of MD simulation in the part where the protein was already stabilized.
-The tunnels for probe 0.9~\AA\ towards the active site defined by residues 38 (Asparagine), 102 (Histidine), and 103 (Aspartic Acid) were
+The tunnels for the probe 0.9~\AA\ towards the active site defined by residues 38 (Asparagine), 102 (Histidine), and 103 (Aspartic Acid) were
 computed using the CAVER 3.0~\cite{caver3} tool.
@@
 The energy cut-off for individual conformations was set to 0.075. %(-rot_ensemble_ecutoff 0.075)
+
 Four different scaling-down factors were used: $\S=\{0.5,0.6,0.7,0.8\}$.
-
 For each scale, $|\QI|=20$ collision-free initial configurations were found in the $\RI=2$~\AA\ radius around the first sphere of the tunnel (outside the protein).
@@
 The \RA\ method was launched with the following parameters:
-$\Imax=5,000$, $m=100$, $\gb=0.9$, $\dt=1.5$~\AA, $\rv=2$~\AA\ and $\varepsilon=0.05$~\AA.
+$\Imax=5000$, $m=100$, $\gb=0.9$, $\dt=1.5$~\AA, $\rv=2$~\AA\ and $\varepsilon=0.05$~\AA.
 Both methods utilize the same data structures and low-level routines (e.g., collision-detection) to enable fair comparison.
@@
 For fast motion planning we utilize a subset of the original set $\L$.
-The aim of this experiment was to verify the procedure for selection of the subset.
+The aim of this experiment was to verify the proposed selection of ligand conformations from a large set $\L$ (Sec.~\ref{sec::strat}).
 
@@
 name shows the number of the selected conformations (5 or 50) and the used method (S or P).
-The number after the slash in the ligand's name denotes the number of atoms of the ligand.
+%The number after the slash in the ligand's name denotes the number of atoms of the ligand.
 In case of both tested ligands and 50 selected conformations, the `S' strategy leads to a higher traversability rate than the `P' strategy.
@@
     \small
-    The influence of the conformation selection method on the traversability rates in the first (main) tunnel.
+    The influence of the conformation selection method on the traversability rates in the first (main) tunnel of the 4E46 protein.
     The number after~`$/$' denotes the number of atoms.
-    The rate is shown in [\%].
 }
@@
     \small
-    Traversability rate [\%] over 100 frames for ligands with 50 conformations. 
+    Traversability rate over 100 frames for ligands with 50 conformations in the 4E46 protein.
     The number after '$/$' denotes the number of atoms.
@@
 Examples of elongated/packed conformation for m040 (1,5-Dichloropentane).
-The table shows the average number of conformations (out of 50) used in trajectories reaching the active site.
+The table shows the average number of conformations (out of 50) used in trajectories reaching the active site in the 4E46 protein.
 }
@@
 The tCONCOORD tool~\cite{seeliger2007geometry} was used to generate 1000 conformations of the protein 4L2L.
-The initial PDB structure was minimized by GROMACS 4.0.7~\cite{hess2008gromacs} using the input parameters recommended on the example page for tCONCOORD. 
+The initial PDB structure was minimized by GROMACS 4.0.7
+%\cite{hess2008gromacs} 
+ using the input parameters recommended on the example page tCONCOORD~\cite{tconcoord}.
 The protein conformations were generated using the default settings for tCONCOORD in the PERT mode. % which only perturbs the starting configuration of atoms instead of complete randomization.
@@
 All 1000 conformations contained at least 5 tunnels, more than 900 conformations contained 9 tunnels.
-The ordering of the detected tunnels slightly differ in different conformations.
+The ordering of the detected tunnels slightly differs in different conformations.
 Therefore, the ligand traversability towards the active site was computed for the first three widest tunnels (with average bottlenecks 1.22~\AA, 1.20~\AA\, and 1.19~\AA\ with the average lengths 22.3~\AA, 25.3~\AA\, and 28.4~\AA, respectively).
@@
     \small
-    Traversability rate [\%] in 1000 conformations of the protein 4L2L in the first three tunnels.
+    Traversability rate in 1000 conformations of the protein 4L2L in the first three tunnels.
 }
@@
 As the tested ligand is a known inhibitor of the protein 4L2L~\cite{marques2017enzyme} we assume that the ligand can reach the active site.
-Due to large size of the ligand (33 atoms), low scale-down factors $\smin\in\{0.4,0.5,0.6\}$ were used.
-The traversability detected by the \RA\ method is show in Tab.~\ref{tab::4l2l} and the example
-of a final trajectory to the active site is depicted in Fig.~\ref{fig::inhibitor}.
+Due to large size of the ligand (33 atoms), low scale-down factors $\smin\in\{0.4,0.5,0.6\}$ were used and
+the active site was considered as rached if the distance to the geometric center of the ligand was less than 5~\AA.
+All paramters of the \RA\ were set as in the previous experiment, except $\Imax=100\cdot10^3$.
+The proposed planner was able to find trajectories even for this large ligand inside the long tunnels, which is
+shown by the traversability rates (Tab.~\ref{tab::4l2l}).
+Example of the solution is depicted in Fig.~\ref{fig::inhibitor}.
 

### 2018-02-24 paper.tex
spelling

@@
 To address this problem, we define the computation of the transportation path of non-spherical ligands including their conformation changes (i.e., when the atoms of the ligand move relatively to each other) as a motion planning task in the high-dimensional configuration space.
-We propose a modification of the Rapidly Exploring Random Tree planner for the purpose of trajectory computations for flexible ligands.
+We propose a modification of the Rapidly Exploring Random Tree planner for trajectory computations of flexible ligands.
 To cope with the narrow passage problem, the random samples are generated around a tunnel detected by a Voronoi-based approach.
@@
 
-The RRT-based planners cannot cope with the narrow passage problem simply by increasing the probability of sampling in the difficult regions, as the growth of the configuration tree towards the random samples may be blocked by obstacles.
+In the case of RRT-based planners, increasing the probability of sampling in difficult regions does not need to bring any advantage
+if the growth of the tree is blocked by the obstacles.
+%The RRT-based planners cannot cope with the narrow passage problem by increasing the probability of sampling in the difficult regions, as the growth of the configuration tree towards the random samples may be blocked by obstacles.
 %~\cite{vonasekphd}.
@@
 
-The shrinking technique has been used also in~\cite{cortes2010simulating}, where the exit pathways for a small flexible molecule are computed using a RRT-based method.
+The shrinking technique has also been used in~\cite{cortes2010simulating}, where the exit pathways for a small flexible molecule are computed using a RRT-based method.
 The flexible ligand is modeled as a kinematic chain, which increases the dimension of the configuration space.
@@
 The parameter $\dt$ determines how far can the tree detour from the tunnel centerline.
-In case of tunnels not having other tunnels in the close vicinity (which is very common), we recommend to use the double value of the tunnel average width.
+In case of tunnels not having other tunnels in the close vicinity (which is very common), we recommend using the double value of the tunnel average width.
 The algorithm terminates after a predefined number of planning trials $\Imax$ or if the tree reaches
@@
 $\Imax=5000$, $m=100$, $\gb=0.9$, $\dt=1.5$~\AA, $\rv=2$~\AA\ and $\varepsilon=0.05$~\AA.
-Both methods utilize the same data structures and low-level routines (e.g., collision-detection) to enable fair comparison.
+Both methods utilize the same data structures and low-level routines (e.g., collision-detection) to enable a fair comparison.
 %were implemented in the same software framework and they utilized the same data structures and routines for the collision detection and nearest-neighbor search.
@@
 Due to large size of the ligand (33 atoms), low scale-down factors $\smin\in\{0.4,0.5,0.6\}$ were used and
-the active site was considered as rached if the distance to the geometric center of the ligand was less than 5~\AA.
-All paramters of the \RA\ were set as in the previous experiment, except $\Imax=100\cdot10^3$.
+the active site was considered as reached if the distance to the geometric center of the ligand was less than 5~\AA.
+All parameters of the \RA\ were set as in the previous experiment, except $\Imax=100\cdot10^3$.
 The proposed planner was able to find trajectories even for this large ligand inside the long tunnels, which is
@@
 
-According to the results of the motion planing, the DCP ligand needs to shrink down to approximately 75~\% of its size, but only at one part of the tunnel (Fig.~\ref{fig::dcp}, point B). 
+According to the results of the motion planning, the DCP ligand needs to shrink down to approximately 75~\% of its size, but only at one part of the tunnel (Fig.~\ref{fig::dcp}, point B). 
 Beyond this point, the ligand can move at larger scales again.

### 2018-02-24 paper.tex
spelling 2

@@
 To address this problem, we define the computation of the transportation path of non-spherical ligands including their conformation changes (i.e., when the atoms of the ligand move relatively to each other) as a motion planning task in the high-dimensional configuration space.
-We propose a modification of the Rapidly Exploring Random Tree planner for trajectory computations of flexible ligands.
+We propose a modification of the Rapidly Exploring Random Tree planner for trajectory computation of flexible ligands.
 To cope with the narrow passage problem, the random samples are generated around a tunnel detected by a Voronoi-based approach.
@@
 e.g., to study loop motions~\cite{cortes2004geometric}, to identify binding sites~\cite{bayazit2001ligand}, for exit
-pathway detection~\cite{cortes2010simulating,cortes2005path}, for tunnel detection~\cite{vonasek2017tunnel,vonasek2016application}, and especially for protein folding~\cite{raveh2009rapid,amato2002using}.
-%novinskaya2015improving
+pathway detection~\cite{cortes2010simulating,cortes2005path}, for tunnel detection~\cite{vonasek2017tunnel}, 
+        and especially for protein folding~\cite{raveh2009rapid,amato2002using}. %novinskaya2015improving
 
@@
 In~\cite{kardossRRTKK}, the probability of sampling is increased in several waypoints close to narrow passages, but without specifying how to find these waypoints.
-The generalization of~\cite{kardossRRTKK} is the guided sampling, where the whole path is used to generate samples in the configuration space.
+The generalization of~\cite{kardossRRTKK} is the guided sampling, where a sequence of waypoints (a guiding path) is used to generate samples in the configuration space.
 The guiding path can be computed in the workspace~\cite{vonasek2009rrt}, or iteratively refined in the configuration space based on the solution of a relaxed version of the problem~\cite{bayazitIRC}.
@@
 protein conformation sampling) are rather narrow. 
-However, in protein dynamics with ligand, the tunnels can move, merge and adapt to the shape of the passing ligand and vice versa.
+However, MD simulations of proteins with ligands have shown that the tunnels can move, merge and adapt to the shape of the passing ligand and vice versa.
 %Therefore, it can be expected that the detected tunnels are rather narrow.
@@
 
-MD simulations of ligands passing through proteins further show that the ligands can be highly flexible.
-The possible conformations of a ligand are determined by the torsion and dihedral angles between its atoms. 
-The low-energy conformations are preferred.
+MD simulations of ligands passing through proteins show that the ligands can be highly flexible.
+All possible conformations of a ligand are determined by the torsion and dihedral angles between its atoms, but the low-energy conformations are preferred.
 %the ligand can contain several torsion and dihedral angles between its atoms which determine possible ligand conformations.
@@
 
-The proposed algorithm was verified in two scenarios: on a sequence of frames of MD simulation and on a set of protein conformations provided 
-the conformation sampling approach.
+The proposed algorithm was verified in two scenarios: on a sequence of frames of MD simulation and on a set of protein conformations 
+provided by the conformation sampling approach.
 %by the tCONCOORD tool~\cite{seeliger2007geometry}.
@@
 The \RB\ method can therefore expand the tree by at most 5 new nodes per ligand conformation, while \RA\ extends the tree only by one new node.
-The larger number of nodes in \RB\ method is not problematic, as the KD-trees were used for the nearest-neighbor search and the time to solve these queries is significantly smaller than the computing collision detection.
+The larger number of nodes in \RB\ method is not problematic, as the KD-trees were used for the nearest-neighbor search and the time to solve these queries is significantly smaller than the collision detection.
 The parameters of the methods were selected in such a way that the number of collision detection queries in each expansion step is the same.
@@
 However, their binding was not yet approved by dedicated MD simulations.
-Most of the used 50 ligand conformations where used in the planning, which is shown in Tab.~\ref{tab::m040c}.
+The planning was realized with 50 conformations of the ligands, but not all of them were used, which is shown in Tab.~\ref{tab::m040c}.
+%The selection of 50 conformations was satisfactory.
 
@@
 Examples of elongated/packed conformation for m040 (1,5-Dichloropentane).
-The table shows the average number of conformations (out of 50) used in trajectories reaching the active site in the 4E46 protein.
+The table shows the average number of different ligand conformations (out of 50) used in trajectories reaching the active site in the 4E46 protein.
 }
@@
 %\cite{hess2008gromacs} 
- using the input parameters recommended on the example page tCONCOORD~\cite{tconcoord}.
+ using the input parameters recommended on the example page of tCONCOORD~\cite{tconcoord}.
 The protein conformations were generated using the default settings for tCONCOORD in the PERT mode. % which only perturbs the starting configuration of atoms instead of complete randomization.
@@
 
-As the tested ligand is a known inhibitor of the protein 4L2L~\cite{marques2017enzyme} we assume that the ligand can reach the active site.
+As the tested ligand is a known inhibitor of the protein 4L2L~\cite{marques2017enzyme} we assumed that the ligand can reach the active site.
 Due to large size of the ligand (33 atoms), low scale-down factors $\smin\in\{0.4,0.5,0.6\}$ were used and

### 2018-02-26 paper.tex
image updated
final version for ral

@@
 \usepackage{setspace} % for setstretch in algorithm
+%\usepackage[all]{nowidow}
 \pdfminorversion=4
@@
 
-\title{Assessing the Accessibility of Protein Tunnels using Motion Planning}
+\title{
+%Assessing the Accessibility of Protein Tunnels using Motion Planning % ICRA title
+Motion Planning for Assessing the Accessibility of Protein Tunnels  %iros title, we had to change it
+}
 
@@
 \rotatebox{0}{\hskip 5pt Long} \\ \includegraphics[width=0.07\textwidth]{fig/m040-conf1} \\
-\rotatebox{0}{Packed} \\ \includegraphics[width=0.0375\textwidth]{fig/m040-conf2}
+\rotatebox{0}{Packed} \\ \includegraphics[width=0.055\textwidth]{fig/m040-conf2}
 \end{tabular}
@@
 The tunnels leading to the active site defined by residues 134 (Alanine) and 311 (Phenylalanine) were computed by CAVER 3.0, using the probe 0.9~\AA.
-All 1000 conformations contained at least 5 tunnels, more than 900 conformations contained 9 tunnels.
+All 1000 conformations contained at least 5 tunnels. % more than 900 conformations contained 9 tunnels.
 The ordering of the detected tunnels slightly differs in different conformations.
-Therefore, the ligand traversability towards the active site was computed for the first three widest tunnels (with average bottlenecks 1.22~\AA, 1.20~\AA\, and 1.19~\AA\ with the average lengths 22.3~\AA, 25.3~\AA\, and 28.4~\AA, respectively).
+Therefore, the first three widest tunnels were used (with average bottlenecks 1.22~\AA, 1.20~\AA\, and 1.19~\AA\ with the average lengths 22.3~\AA, 25.3~\AA\, and 28.4~\AA, respectively).
+20 starting configurations were found at the begining of each tunnel, and 10 trajectories were computed for each starting configuration.
 The conformations of the ligand were prepared in the same way as in the previous section.
@@
 \caption{\small
-    \label{fig::inhibitor}Example of a trajectory inside the tunnel No. 1 of the 4L2L protein with the final position of the ligand.
+    \label{fig::inhibitor}Example of a found trajectory inside the tunnel No. 1 of the 4L2L protein with the final position of the ligand.
 }
